### 2019-05-30 peakcharacterization_manuscript.Rmd
started actually writing!

---
title: "Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments"
author: "Robert M Flight"
date: "`r Sys.time()`"
commit: "`r substr(git2r::branch_target(git2r::head(git2r::repository())), 1, 8)`"
output: 
  redoc::redoc
---

## Introduction

## Methods

### Conversion of M/Z to Frequency

The data consists of profile mode M/Z spectra from multiple scans encoded as M/Z and intensity values for each scan. There is no information about the original frequency values available in either the `raw` files or the `mzML` files. However, the frequency values can be approximated by averaging the M/Z of adjacent points, and dividing the M/Z by the difference. A constant difference of adjacent frequency values is expected, and points that have frequency differences within a narrow range can be used to construct a linear model relating M/Z to frequency for individual scans based on the formula:

$$frequency = intercept + slope* \frac{1}{\sqrt{mz}} $$
A model is generated for each scan, and then a single model using the median values of the intercept and slope across scans is chosen, and applied to convert the M/Z values to frequency for all scans. A single model rather than scan specific models is used because this was found to maintain the relative ordering of the peaks in M/Z and frequency space. Integer values (for use with the IRanges R package) are generated by multiplying each frequency by 400 and rounding.

### Frequency Intervals

Two types of frequency intervals are used, sliding and tiled windows. The sliding windows are 10 points wide, offset by one point from each other. The tiled windows are one point wide, offset by one point from each other.

### Interval Range Based Data

To enable interval arithmetic, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package. The sliding and tiled windows are also converted to IRanges objects the same way.

### Peak Containing Intervals

To find intervals that containing data that is not just noise, the number of zero intensity and non-zero intensity frequency points in each sliding window are counted, and any sliding window with a non-zero count greater than or equal to the 99th percentile of non-zero counts is kept, and the remaining sliding windows are reduced, where overlapping sliding windows are merged together to create the initial peak regions.

Within each initial interval region, peaks in each scan are detected (see **Peak Finding**), and their centers binned by the tiled windows. Adjacent tiled windos with non-zero peak counts are merged together, and any zero peak count tiled windows split the initial region into multiple peak interval regions. These interval regions should contain a single **real** peak that was detected in one or more scans.

### Peak Detection and Centroided Values

On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package. These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either M/Z or frequency), and where the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.

$$\ln{intensity} = intercept + a*position+b*position^2$$
From this weighted parabolic fit the center, intensity, integrated area and sum-of-square residuals can be extracted for the peak. These characteristics are equivalent to the centroided peak center and intensity.

### Scan to Scan Normalization

Scans are normalized to a single *median* scan based on the log-intensity ratios of a subset of peaks that are present in the 95th percentile of scans.

### 2019-05-30 peakcharacterization_manuscript.Rmd
lots more writing!

@@
 
+Before further processing, the regions are checked that there is only one peak from each scan.
+If a scan has two or more peaks, then the scan level data in that region is discarded.
+Any regions that subsequently contain zero peaks are removed.
+
 ### Scan to Scan Normalization
 
-Scans are normalized to a single *median* scan based on the log-intensity ratios of a subset of peaks that are present in the 95th percentile of scans.
+Scans are normalized to a single *median* scan based on the log-intensity differences of a subset of peaks that are present in at least the same number of scans as the 95th percentile of scan counts for the peaks.
+In addition, only those peaks that have an intensity greater than 0.7 times the highest intensity peak in the scan are used.
+Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each to all other scans.
+The scan with the lowest summed overall distance is chosen as the "reference" scan, and normalization factors for each scan are calculated as the median log peak intensity differences in scan_i compared to the reference scan.
+This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan number is checked, and those peaks with correlation of greater than 0.5 with scan number are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.
+
+### Full Characterization
+
+Based on the previously detected peaks, the full set of raw data points for each peak in each scan within a region is known.
+The non-zero intensity, normalized raw data points across scans can be combined, and then characterized again using the weighted parabolic fit previously described.
+In addition to the data from the full set of raw points, means and standard deviations of the peak height and location can be extracted from the scan level peak characteristics previously calculated.
+
+In addition to these values, the frequency point-to-point median difference is calculated across all of the raw data points for those points that could be used for modeling frequency to M/Z, and this difference of a single point from the peak center is calculated in frequency space, and converted to M/Z space to provide an "offset" value that is useful to define the search space around the peak for any assignment algorithm.
+
+### Correction of Height and Standard Deviation
+
+Ideally, each peak would be observed in every scan.
+However, in some scans, some peaks are not likely to be observed due to the number of ions falling below the detection threshold.
+This results in a truncated distribution of peak intensities across scans.
+To correct these, either a correction based on a model of the truncated normal distribution can be used, or the differences can be simulated by sampling from data that is present in most of the scans.
+To simulate the effect on the standard deviations, the peaks present in all scans are used.
+For each peak, a sample of the heights across scans are taken (ranging from 5% to 95% of scans), and a new standard deviation calculated for that fraction, and a ratio of the fractional standard deviation to the "true" standard deviation calculated.
+The ratio sd across peaks can then be fitted to a cubic model of the fraction used, and a correction factor predicted for those peaks that are present in fewer scans.
+The corrected sd's can then be used to correct the mean height assuming that it is the result of a truncated normal distribution [wikipedia referenc, https://en.wikipedia.org/wiki/Truncated_normal_distribution].
+
+### Model of M/Z Standard Deviation
+
+**do we use this anywhere**?? Because if not, maybe we shouldn't bother including it.
+
+### Samples and Overall Processing
+
+Samples included two ethyl-chloroformate (ECF) derivatized amino-acid samples [reference] and two samples where lipid extracts were extracted from lung tumors [reference].
+All four samples are positive mode, were converted from raw to mzML using msconvert from ProteoWizard (profile mode) and ThermoRawFileParser (centroided).
+
+For each raw data file in mzML format, they were each processed in these six ways:
+
+  1. No noise removal, no normalization, no frequency SD filtering
+  2. Noise removal, no normalization, no frequency SD filtering
+  3. Noise removal, single pass normalization with all peaks, no frequency SD filtering
+  4. Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering
+  5. Noise removal, two pass normalization, no frequency SD filtering
+  6. Noise removal, two pass normalization, frequency SD filtering
+
+## Results
+
+### Naively Averaged Data Have Bad Relative Intensities
+
+To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur (a process we have observed being used to generate datasets), as well as doing a straight averaging of point intensities using identical scans as was kept by our scan filtering (see Methods) in the xcms R package, and finally averaging peaks that were centroided via the Thermo raw file parsing utilities (wrapped by ThermoRawFileParser).
+In all cases, for a known assigned set of peaks, the intensity ratios between the assigned peaks are off compared to what is expected based on natural abundance probability (NAP), as shown in Figure X, with some solutions giving different amounts of error.
+We do note that when all of the peaks are present in a single scan, their relative intensity ratios are much closer to the theoretical ratios based on NAP.
+
+### M/Z to Frequency
+
+FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data, outside of meta-data, it merely contains the M/Z and intensity values for profile spectra acquired across multiple scans.
+However, it turns out that the frequency can be calculated by dividing the average M/Z of two adjacent points with their difference.
+The differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, shown in Figure Xa and Xb.
+Our instrument seems to have a mode of 0.5, as shown in Figure Xc.
+Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in Figure Xd.
+This model seems to fit the known relationship between frequency and M/Z.
+This is useful, because some of the subsequent steps in our workflow use sliding windows where it is assumed that the sliding windows contain the same number of data points.
+M/Z point-to-point differences are not constant, but can be approximated by a lasso linear model.
+However, it is very difficult to create a lasso model with an intercept of 0, and then vary the width of sliding windows according to the M/Z difference at a particular M/Z.
+Frequency based points suffer none of these drawbacks, and it turns out that the conversion from M/Z can be derived from the raw profile level data itself, which is incredibly useful.
+
+```{r mz_frequency_figure}
+# Figure showing relationship of M/Z to frequency
+# a - mean m/z vs diff m/z
+# b - mean m/z vs diff frequency
+# c - histogram of diff frequency
+# d - mean m/z vs frequency
+
+```
+
+These models are calculated for each scan, and the intercepts from the models compared and examined for outliers.
+Outlier intercepts are determined from the interquartile range, and outliers removed.
+When scan specific models are used for conversion of M/Z to frequency, the peak order observed in M/Z space is not maintained, as shown in Figure 2a-c.
+Therefore, a single model based on the median intercept and slope is used for converting *all* remaining scan level data.
+
+```{r peak_ordering}
+# Figure showing changes in order of peaks
+# a - peaks in M/Z compared to frequency scan specific
+# b - peaks in M/Z compared to frequency single model
+
+```
+
+Although the original model is created from only those points that had frequency point-to-point differences within a narrow range, **all** M/Z points are converted to frequency for subsequent steps in the workflow.
+
+### Sliding Window Density to Remove Noise
+
+In a dataset of this nature, we expect that much of the data is really just noise, and doesn't contribute that much to the analysis.
+It is expected that noise is randomly distributed across the scans.
+Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
+The distribution of counts is shown in Figure X.
+The number of peaks detected without applying a cutoff explodes by ..., with the vast majority of the peaks being present in a single scan.
+By merging those windows that are adjacent **and** have a point density greater than the 99th percentile, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
+
+```{r slidingwindow_count_distribution}
+# Figure showing distribution of nonzero counts
+```
+
+For each region initially created, the peaks in each scan within that region can be characterized using a quadratic fit of log(intensity) to M/Z.
+
+### Peak Characterization Using Quadratic Fit
+
+Although many other types of mass-spectrometry data suffer from a variable, noisy baseline, the scan-level profile data in the direct-injection samples shows a true baseline value of 0, making the determination of the centroided values considerably easier.
+For the peak characterization (centroiding), we use a simple quadratic model of log(intensity) to M/Z.
+We add a small constant to enable using the zero values, and weight the values by their ratio to the most intense value, which is normally the value closest to the center of peak, helping to ensure that the **true** centroid is determined.
+From the fitted model, we can derive the centroided center and the intensity of the peak, as shown in Figure X.
+
+```{r centroided_peaks}
+# figure showing fit of quadratic model to actual data points
+```
+
+### Breaking Up Initial Regions
+
+With the characterized (centroided) peak data from across scans within each region, it is then important to determine if only one or multiple "peaks" are actually present in the region.
+Our solution to this is to define breaks between **actual** peaks as a single frequency bin with zero characterized peaks within it.
+The frequency bins are created from tiled windows that are one frequency point difference wide.
+Adjacent non-zero frequency bins are merged to comprise a single peak region.
+Figure X shows an example where an initial region is broken up into multiple regions based on the characterized peak centers.
+
+```{r breaking_regions}
+# figure showing multiple "peaks" in what is initially a single region
+```
+
+### Normalization of Scans

### 2019-07-26 peakcharacterization_manuscript.Rmd
updating some of the language

@@
 However, the frequency values can be approximated by averaging the M/Z of adjacent points, and dividing the M/Z by the difference.
-A constant difference of adjacent frequency values is expected, and points that have frequency differences within a narrow range can be used to construct a linear model relating M/Z to frequency for individual scans based on the formula:
+The *true* difference can be obtained by examining the mode of the mean frequency differences, and then constraining *useful* points to be within 2% of the mode value.
+These points can be used to construct a linear model relating M/Z to frequency for individual scans based on the formula:
 
@@
 A single model rather than scan specific models is used because this was found to maintain the relative ordering of the peaks in M/Z and frequency space.
-Integer values (for use with the IRanges R package) are generated by multiplying each frequency by 400 and rounding.
+Integer values (for use with the IRanges R package) are generated by multiplying each frequency value by 400 and rounding.
 
@@
 
-To enable interval arithmetic, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package.
+To enable interval arithmetic, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package [citation].
 The sliding and tiled windows are also converted to IRanges objects the same way.
@@
 Within each initial interval region, peaks in each scan are detected (see **Peak Finding**), and their centers binned by the tiled windows.
-Adjacent tiled windos with non-zero peak counts are merged together, and any zero peak count tiled windows split the initial region into multiple peak interval regions.
+Adjacent tiled windows with non-zero peak counts are merged together, and any zero peak count tiled windows split the initial region into multiple peak interval regions.
 These interval regions should contain a single **real** peak that was detected in one or more scans.
@@
 
-On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package.
+On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package [citation].
 These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either M/Z or frequency), and where the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.

### 2019-08-02 data_analysis/R/graph_generation.R
doing some crazy work to generate first set of plots

@@ -77,29 +77,50 @@ plot_frequency_conversion = function(raw_data){
   example_data$pair_frequency = example_data$mean_frequency
   example_data$pair_frequency_offset = example_data$mean_freq_diff
 
-  peak_plot = ggplot(example_data, aes(x = mz, y = log10(intensity+1))) + geom_point() +
-    labs(y = "Log10(Intensity)")
-  peak_plot
-
-  example_data$level1_intensity = 0
+  example_data$level1_intensity = -.1
   example_data$level1_group = 0.1
-  example_data$level1_group_xstart = example_data$mz + 1e-5
-  example_data$level1_group_xend = dplyr::lead(example_data$mz - 1e-5)
-  mz_point_plot = ggplot(example_data, aes(x = mz, y = level1_intensity)) + geom_point() +
-    geom_segment(aes(x = level1_group_xstart, xend = level1_group_xend, y = level1_group, yend = level1_group), color = "red") +
-    geom_point(aes(x = pair_mz, y = level1_group), color = "red") + labs(y = NULL)
+  example_data$mz_group_xstart = example_data$mz + 1e-5
+  example_data$mz_group_xend = dplyr::lead(example_data$mz - 1e-5)
+  example_data$frequency_group_xstart = example_data$mean_frequency - 5e-2
+  example_data$frequency_group_xend = dplyr::lead(example_data$mean_frequency) + 5e-2
+  example_data[1, "pair_mz"] = NA
+  mz_point_plot = ggplot(example_data, aes(x = mz, y = log10(intensity+1))) + geom_point() +
+    geom_segment(aes(x = level1_group_xstart, xend = level1_group_xend, y = level1_intensity, yend = level1_intensity), color = "red") +
+    geom_point(aes(x = pair_mz, y = level1_intensity), color = "red") + labs(y = "Log10(Intensity)", x = "M/Z")
 
   frequency_point_plot = ggplot(example_data, aes(x = pair_frequency, y = log10(intensity+1))) + geom_point() +
-    labs(y = "Log10(Intensity)")
+    labs(y = "Log10(Intensity)", x = "Frequency") + geom_segment(aes(x = frequency_group_xstart, xend = frequency_group_xend, y = level1_intensity, yend = level1_intensity), color = "red")
 
-  point_difference_plot = ggplot(example_data, aes(x = pair_frequency, y = pair_frequency_offset)) + geom_point()
[…]

### 2019-08-02 peakcharacterization_manuscript.Rmd
updated wording

@@
 
-$$frequency = intercept + slope* \frac{1}{\sqrt{mz}} $$
+$$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
+From the physical properties of the orbitrap, only the square root should be necessary, however, in practice we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry and control of the magnetic fields.
 A model is generated for each scan, and then a single model using the median values of the intercept and slope across scans is chosen, and applied to convert the M/Z values to frequency for all scans.
@@
 
-FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data, outside of meta-data, it merely contains the M/Z and intensity values for profile spectra acquired across multiple scans.
-However, it turns out that the frequency can be calculated by dividing the average M/Z of two adjacent points with their difference.
+FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data.
+Outside of the meta-data, it merely contains the M/Z and intensity values for profile spectra acquired across multiple scans.
+However, it turns out that the frequency can be calculated by dividing the average M/Z of two adjacent points with their difference (Figure XA, red points reresenting the average M/Z of two adjacent points, red lines representing the difference between the two adjacent points).
 The differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, shown in Figure Xa and Xb.
-Our instrument seems to have a mode of 0.5, as shown in Figure Xc.
-Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in Figure Xd.
+Our instrument seems to have a mode of 0.5, as shown in Figure XC and XD.
+Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in Figure XE.
 This model seems to fit the known relationship between frequency and M/Z.
@@
 However, it is very difficult to create a lasso model with an intercept of 0, and then vary the width of sliding windows according to the M/Z difference at a particular M/Z.
-Frequency based points suffer none of these drawbacks, and it turns out that the conversion from M/Z can be derived from the raw profile level data itself, which is incredibly useful.
+Frequency based points suffer none of these drawbacks, and the conversion from M/Z can be derived from the raw profile level data itself, which is incredibly useful.
 
 ```{r mz_frequency_figure}
-# Figure showing relationship of M/Z to frequency
-# a - mean m/z vs diff m/z
-# b - mean m/z vs diff frequency
-# c - histogram of diff frequency
-# d - mean m/z vs frequency
-
+drake::loadd(frequency_conversion)
+frequency_conversion
 ```
 
-These models are calculated for each scan, and the intercepts from the models compared and examined for outliers.
-Outlier intercepts are determined from the interquartile range, and outliers removed.
-When scan specific models are used for conversion of M/Z to frequency, the peak order observed in M/Z space is not maintained, as shown in Figure 2a-c.
-Therefore, a single model based on the median intercept and slope is used for converting *all* remaining scan level data.
+These models are calculated for each scan, and the square root term from all scan level models checked for outliers based on the interquartile ranges across all scans in a sample.
+When scan specific models are used for conversion of M/Z to frequency, the peak order observed in M/Z space is not maintained, as shown in Figure XA.
+Therefore, a single model for all scans based on the median of each term in the model across scans is used for converting *all* remaining scan level data.
 

### 2019-08-02 make_plan.R
actually generate the frequency conversion figure

@@ -68,6 +68,9 @@ analysis_plan = drake_plan(
   interesting_peaks = target(
     find_interesting_peaks(raw_assignments),
     transform = map(raw_assignments)
+  ),
+  frequency_conversion = target(
+    plot_frequency_conversion(data)
   )
 )
 

### 2019-08-02 data_analysis/R/processing_functions.R
updated the level of confidence in assignments

@@ -232,7 +232,7 @@ read_assignments = function(in_assign){
 }
 
 find_interesting_peaks = function(assigned_data){
-  evalue_cutoff = 0.5
+  evalue_cutoff = 0.1
   remove_elements = "S"
   tmp_assign = dplyr::filter(assigned_data$assignments, !grepl(remove_elements, complete_EMF))
   emfs = get_sample_emfs(tmp_assign, assigned_data$sample, evalue_cutoff = evalue_cutoff)
@@ -253,6 +253,8 @@ find_interesting_peaks = function(assigned_data){
   max_scan = max(unlist(purrr::map(organized_data$scans, ~ .x)))
   possible_emfs = dplyr::filter(organized_data, clique_size > 3, min_scan < (0.5 * max_scan))
   possible_emfs
+
+  list(emfs = emfs$grouped_emf[possible_emfs$emf], summary_data = possible_emfs)
 
 }
 

### 2019-08-14 peakcharacterization_manuscript.Rmd
changing the default chunk options more appropriate for a manuscript
comparing peak ordering

@@
 ---
+
+```{r setup, echo = FALSE}
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+```
 
@@
 On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package [citation].
-These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either M/Z or frequency), and where the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.
+These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either M/Z or frequency), and the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.
 
@@
 Outside of the meta-data, it merely contains the M/Z and intensity values for profile spectra acquired across multiple scans.
-However, it turns out that the frequency can be calculated by dividing the average M/Z of two adjacent points with their difference (Figure XA, red points reresenting the average M/Z of two adjacent points, red lines representing the difference between the two adjacent points).
-The differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, shown in Figure Xa and Xb.
-Our instrument seems to have a mode of 0.5, as shown in Figure XC and XD.
+However, the frequency can be calculated by dividing the average M/Z of two adjacent points by their difference (Figure XA, red points reresenting the average M/Z of two adjacent points, red lines representing the difference between the two adjacent points).
+The subsequent differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, shown in Figure XB and XC.
+The Thermo Fusion instrument from which most of our collaborators data has been acquired, at resolution of 400K, seems to have a mode of 0.5, as shown in Figure XC and XD.
 Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in Figure XE.
@@
 
-These models are calculated for each scan, and the square root term from all scan level models checked for outliers based on the interquartile ranges across all scans in a sample.
+Frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
 When scan specific models are used for conversion of M/Z to frequency, the peak order observed in M/Z space is not maintained, as shown in Figure XA.
@@
 ```{r peak_ordering}
-# Figure showing changes in order of peaks
-# a - peaks in M/Z compared to frequency scan specific
-# b - peaks in M/Z compared to frequency single model
-
+drake::loadd(plot_peak_ordering)
+plot_peak_ordering
 ```

### 2019-08-14 data_analysis/R/graph_generation.R

@@ -122,3 +122,29 @@ plot_frequency_conversion = function(raw_data){
   frequency_plot
 
 }
+
+plot_peak_ordering(peak_data){
+  scan_level_peaks = as.data.frame(peak_data$zip_ms$peak_finder$peak_regions$scan_peaks[[1]])
+
+  scan_models = peak_data$zip_ms$peak_finder$peak_regions$frequency_point_regions@metadata$frequency_coefficients_all
+  scan_models = dplyr::filter(scan_models, scan %in% scan_level_peaks$scan)
+
+  by_scan = purrr::map_df(seq(1, nrow(scan_level_peaks)), function(in_row){
+    use_mz = scan_level_peaks[in_row, "ObservedMZ"]
+    use_model = dplyr::filter(scan_models, scan %in% scan_level_peaks[in_row, "scan"])
+
+    out_freq = FTMS.peakCharacterization:::predict_exponentials(use_mz, unlist(use_model[1, c("V1", "V2", "V3")]),
+                peak_data$zip_ms$peak_finder$peak_regions$frequency_point_regions@metadata$frequency_fit_description)
+    data.frame(mz = use_mz, single_frequency = scan_level_peaks[in_row, "ObservedFrequency"],
+               scan_frequency = out_freq)
+  })
+
+  by_scan = dplyr::arrange(by_scan, mz) %>% dplyr::mutate(mz_order = seq(1, nrow(.))) %>%
+    dplyr::arrange(single_frequency) %>% dplyr::mutate(single_order = seq(1, nrow(.))) %>%
+    dplyr::arrange(scan_frequency) %>% dplyr::mutate(scan_order = seq(1, nrow(.)))
+
+  by_scan = tidyr::gather(by_scan, key = "type_order", value = "order", single_order, scan_order)
+
+  ggplot(by_scan, aes(x = mz_order, y = order, color = type_order)) + geom_point() +
+    theme(legend.position = c(0.8, 0.9))
+}

### 2019-08-14 make_plan.R

@@ -71,6 +71,9 @@ analysis_plan = drake_plan(
   ),
   frequency_conversion = target(
     plot_frequency_conversion(data)
+  ),
+  peak_ordering = target(
+    plot_peak_ordering(group6_method_data)
   )
 )
 

### 2019-08-14 data_analysis/R/graph_generation.R
forgot to define as a function

@@ -123,7 +123,7 @@ plot_frequency_conversion = function(raw_data){
 
 }
 
-plot_peak_ordering(peak_data){
+plot_peak_ordering = function(peak_data){
   scan_level_peaks = as.data.frame(peak_data$zip_ms$peak_finder$peak_regions$scan_peaks[[1]])
 
   scan_models = peak_data$zip_ms$peak_finder$peak_regions$frequency_point_regions@metadata$frequency_coefficients_all

### 2019-08-14 data_analysis/R/graph_generation.R
add code for sliding window density

@@ -149,3 +149,17 @@ plot_peak_ordering = function(peak_data){
     theme(legend.position = c(0.8, 0.9))
   peak_ordering_plot
 }
+
+plot_sliding_window_density = function(peak_data){
+  sliding_regions = peak_data$char_obj$zip_ms$peak_finder$peak_regions$sliding_regions
+  frequency_points = peak_data$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions
+
+  region_counts = FTMS.peakCharacterization:::count_overlaps(sliding_regions, frequency_points)
+  nonzero_counts = data.frame(nonzero = region_counts@elementMetadata$nonzero_counts)
+
+  data_cutoff = stats::quantile(nonzero_counts$nonzero, 0.99)
+
+  sliding_window_nonzero_count = ggplot(nonzero_counts, aes(x = nonzero)) + geom_histogram(bins = 100) + geom_vline(xintercept = data_cutoff, color = "red") +
+    scale_y_log10(expand = c(0,0), limits = c(1, NA))
+  sliding_window_nonzero_count
+}

### 2019-08-14 make_plan.R

@@ -74,6 +74,9 @@ analysis_plan = drake_plan(
   ),
   peak_ordering = target(
     plot_peak_ordering(group6_method_data)
+  ),
+  sliding_regions = target(
+    plot_sliding_window_density(data)
   )
 )
 

### 2019-08-14 peakcharacterization_manuscript.Rmd
correct variable
figure for sliding window non-zero points

@@
 
+*caption*: **A**: Intensity vs M/Z for a single peak from a single scan.
+Red lines denote the differences between each point, and red dots the average between the pair of points.
+The difference over the point is used to derive the frequency values in **B**.
+**B** plots the intensity vs the converted frequency points derived from **A**.
+The red lines denote pairwise differences, which are shown in **C** for this single peak.
+The differences for all peaks across all scans vs M/Z are shown in **D**, with those differences that lie within 0.49 - 0.51 shown in red.
+**E** shows the plot of derived frequency *vs* M/Z, with fitted values from the linear model in red.
+
 Frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
@@
 ```{r peak_ordering}
-drake::loadd(plot_peak_ordering)
-plot_peak_ordering
+drake::loadd(peak_ordering)
+peak_ordering
 ```
@@
 ```{r slidingwindow_count_distribution}
-# Figure showing distribution of nonzero counts
+drake::loadd(sliding_regions)
+sliding_regions
 ```

### 2019-08-14 peakcharacterization_manuscript.Rmd
captions for new figures

@@
 
+*caption*: Peak ordering in M/Z compared with ordering in frequency space when a single M/Z to frequency model is used or scan specific M/Z to frequency models are used.
+For a single peak, the scan level peak M/Z's were extracted, and then frequency values for those M/Z generated using a single common model of M/Z to frequency (*single_order*), or models derived from each scan (*scan_order*).
+A subset of the peaks end up out of order using scan specific models, implying that a single model should be used across all the scan level data.
+
 Although the original model is created from only those points that had frequency point-to-point differences within a narrow range, **all** M/Z points are converted to frequency for subsequent steps in the workflow.
@@
 ```
+
+*caption*: Histogram of the number of non-zero points across all scans in sliding windows 10 frequency points wide and 1 frequency point apart.
+The red vertical line denotes the lower 99th percentile of the data, only sliding windows with non-zero counts above the red line will be kept for subsequent use.
 

### 2019-08-14 peakcharacterization_manuscript.Rmd
a draft of the abstract

@@
 ```
+
+## Abstract
+
+We introduce a novel method for characterizing peaks from direct injection FT-MS experiments developed on metabolomics datasets that utilizes calculated frequency values derived directly from the spacing of raw M/Z points.
+Our method incorporates several aspects designed to make the derived intensities trustworthy, and suitable for use in our previously published peak assignment algorithm SMIRFE.
+Notably, the inclusion of an intensity independent noise removal, and normalization of scan level data results in a reduction of median relative standard deviation from **X** to **Y**, and much better fit of peak intensities to relative natural abundances of assigned isotopologues.
 

### 2019-08-16 peakcharacterization_manuscript.Rmd
added all the authors and changed output type to PDF until we can get next version of fedora installed

@@
 title: "Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments"
-author: "Robert M Flight"
+author: "Robert M Flight, Joshua M Mitchell, Hunter NB Moseley"
 date: "`r Sys.time()`"
@@
 output: 
-  redoc::redoc
+  pdf_document
 ---

### 2019-08-16 peakcharacterization_manuscript.Rmd
starting the section on changes to RSD, b/c this is *really* the biggest point of the whole paper demonstrating that the characterization as a whole has an effect

@@
 ### Normalization of Scans
+
+### Changes in RSD
+
+Each step in the peak characterization either changes the overall number of peaks coming from each scan (sliding windows and breaking initial regions) or the overall intensity of the points within a scan.
+Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation for the characterized scan level peak intensities, and compare them as each processing step is introduced.

### 2019-08-20 peakcharacterization_manuscript.Rmd
knitting the manuscript as part of the workflow, and using redoc!

@@
 output: 
-  pdf_document
+  redoc::redoc
 ---

### 2019-08-20 make_plan.R

@@ -49,13 +49,18 @@ analysis_plan = drake_plan(
   ),
   sliding_regions = target(
     plot_sliding_window_density(data)
-  )
+  ),
   # # here we want to combine the ways a data file was processed and compare
   # # their rsd's across processing methods.
   # rsd_data = target(
   #   rsd_info(method),
   #   transform = combine(method, .by = data)
   # )
+  manuscript = rmarkdown::render(
+    knitr_in("peakcharacterization_manuscript.Rmd"),
+    output_file = file_out("peakcharacterization.docx"),
+    quiet = TRUE
+  )
 )
 
 make(analysis_plan, lock_envir = FALSE, memory_strategy = "memory")

### 2019-08-20 make_plan.R
making sure the right option is used for new version of the package

@@ -63,4 +63,4 @@ analysis_plan = drake_plan(
   )
 )
 
-make(analysis_plan, lock_envir = FALSE, memory_strategy = "memory")
+make(analysis_plan, lock_envir = FALSE, memory_strategy = "preclean")

### 2019-08-21 data_analysis/R/graph_generation.R
updated the graph and what we are putting for tables

@@ -167,6 +167,8 @@ plot_sliding_window_density = function(peak_data){
 
 plot_rsd_differences = function(rsd_values){
 
-  rsd_comparison_plot = ggplot(rsd_values, aes(x = rsd, y = processed, fill = processed)) + geom_density_ridges()
+  rsd_values$processed = forcats::fct_relevel(rsd_values$processed, "noperc_nonorm", "perc99_nonorm", "singlenorm", "singlenorm_int", "doublenorm",  "filtersd")
+  rsd_comparison_plot = ggplot(rsd_values, aes(x = rsd, y = processed, fill = processed)) + geom_density_ridges() +
+    coord_cartesian(xlim = c(0, 1)) + theme(legend.position = "none")
   rsd_comparison_plot
 }

### 2019-08-21 data_analysis/R/processing_functions.R

@@ -292,5 +292,14 @@ rsd_info = function(...) {
 }
 
 summarize_rsd = function(rsd_df){
- rsd_df
+  get_mode = function(in_values){
+    d_estimate = stats::density(in_values)
+    y_max = which.max(d_estimate$y)
+    d_estimate$x[y_max]
+  }
+  dplyr::group_by(rsd_df, processed) %>%
+    dplyr::summarize(mean = mean(rsd),
+                     median = median(rsd),
+                     mode = get_mode(rsd),
+                     max = max(rsd))
 }

### 2019-08-21 peakcharacterization_manuscript.Rmd
inserting the graph and table

@@
 date: "`r Sys.time()`"
-commit: "`r substr(git2r::branch_target(git2r::head(git2r::repository())), 1, 8)`"
 output: 
@@
 Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation for the characterized scan level peak intensities, and compare them as each processing step is introduced.
+
+```{r rsd_plots}
+drake::loadd(rsd_plot_data)
+rsd_plot_data
+```
+
+```{r rsd_table}
+drake::loadd(rsd_table_data)
+knitr::kable(rsd_table_data, digits = 3)
+```

### 2019-08-21 peakcharacterization_manuscript.Rmd
rsd table and the plot

@@
 ```{r rsd_plots}
-drake::loadd(rsd_plot_data)
-rsd_plot_data
+drake::loadd(rsd_plot_rsd_data)
+rsd_plot_rsd_data
 ```
@@
 ```{r rsd_table}
-drake::loadd(rsd_table_data)
-knitr::kable(rsd_table_data, digits = 3)
+drake::loadd(rsd_table_rsd_data)
+knitr::kable(rsd_table_rsd_data, digits = 3)
 ```

### 2019-10-25 peakcharacterization_manuscript.Rmd
trying a flextable instead of kable
trying flextable again
switch each sentence to it's own line so diffs should become on the sentence, not the paragraph, which should make seeing changes easier

@@
 Two types of frequency intervals are used, sliding and tiled windows. 
-The sliding windows are 10 points wide, offset by one point from each other.
-The tiled windows are one point wide, offset by one point from each other.
+Sliding windows are 10 points wide, offset by one point from each other. 
+Tiled windows are one point wide, offset by one point from each other.
 
@@
 
-To find intervals that containing data that is not just noise, the number of zero intensity and non-zero intensity frequency points in each sliding window are counted, and any sliding window with a non-zero count greater than or equal to the 99th percentile of non-zero counts is kept, and the remaining sliding windows are reduced, where overlapping sliding windows are merged together to create the initial peak regions.
+To find intervals that containing data that is not just noise, the number of non-zero intensity frequency points in each sliding window are counted, and any sliding window with a non-zero count greater than or equal to the 99th percentile of non-zero counts is kept, and the remaining sliding windows are reduced, where overlapping sliding windows are merged together to create the initial peak regions.
 
@@
 
-Before further processing, the regions are checked that there is only one peak from each scan.
-If a scan has two or more peaks, then the scan level data in that region is discarded.
+Before further processing, the regions are checked that there is only one peak from each scan. If a scan has two or more peaks, then the scan level data in that region is discarded. 
 Any regions that subsequently contain zero peaks are removed.
@@
 The ratio sd across peaks can then be fitted to a cubic model of the fraction used, and a correction factor predicted for those peaks that are present in fewer scans. 
-The corrected sd's can then be used to correct the mean height assuming that it is the result of a truncated normal distribution [wikipedia referenc, https://en.wikipedia.org/wiki/Truncated_normal_distribution].
+The corrected sd's can then be used to correct the mean height assuming that it is the result of a truncated normal distribution [wikipedia reference, https://en.wikipedia.org/wiki/Truncated_normal_distribution].
 
@@
 Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in Figure XE.
-This model seems to fit the known relationship between frequency and M/Z.
-This is useful, because some of the subsequent steps in our workflow use sliding windows where it is assumed that the sliding windows contain the same number of data points.
+This model seems to fit the known relationship between frequency and M/Z, where the frequency is related to $1/\sqrt{mz}$. 
+This is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points. 
 M/Z point-to-point differences are not constant, but can be approximated by a lasso linear model. 
@@
 
-```{r rsd_table}
+```{r rsd_table, results = 'asis'}
 drake::loadd(rsd_table_rsd_data)
-knitr::kable(rsd_table_rsd_data, digits = 3)
+knitr::kable(rsd_out, digits = 3)
 ```

### 2019-10-25 peakcharacterization_manuscript.Rmd
discuss normalization results and what we should see

@@
 
-### Model of M/Z Standard Deviation
+### Frequency Standard Deviation Filtering
 
-**do we use this anywhere**?? Because if not, maybe we shouldn't bother including it.
+High-peak-density artifacts [Mitchell et al., 2018] present as singular peaks with higher than expected frequency standard deviations calculated from the scan-to-scan frequency peak locations.
+Outliers are removed by examining the distribution of peak frequency SDs across the entire spectrum.
 
@@
 
+Due to differences in how many ions are captured in the trap, the peak intensities in each scan vary slightly.
+There are two issues with these peak intensities across scans: 1 - Some peak intensities correlate with the scan; 2 - Some peak intensities are correlated with M/Z / frequency.
+The solution to **1** is to do a two pass normalization, detect the correlated peak intensities, remove them, and then calculate normalization factors again.
+Figure XA shows that some peaks will be correlated with scan number.
+The solution to peak intensities correlated with M/Z appears to be tied to their intensities, as shown in Figure XB.
+The highlighted peaks in Figure XB are those with an intensity greater than 0.7, and are **not** correlated with M/Z.
+If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in Figure XC.
+
+```{r normalization_data}
+# figure showing peak intensity correlation with M/Z, intensity varying with M/Z, changes in normalization using all vs most intense
+```
+
 ### Changes in RSD

### 2019-10-25 peakcharacterization_manuscript.Rmd
saying where the 0 baseline really comes from.

@@
 
-Although many other types of mass-spectrometry data suffer from a variable, noisy baseline, the scan-level profile data in the direct-injection samples shows a true baseline value of 0, making the determination of the centroided values considerably easier.
+Although many other types of mass-spectrometry data suffer from a variable, noisy baseline, the scan-level profile data from the Thermo Fusion has a baseline of 0 due to manipulations in the Thermo firmware, making the determination of the centroided values considerably easier.
 For the peak characterization (centroiding), we use a simple quadratic model of log(intensity) to M/Z. 

### 2019-11-13 peakcharacterization_manuscript.Rmd
editing for clarity

@@
 
-We introduce a novel method for characterizing peaks from direct injection FT-MS experiments developed on metabolomics datasets that utilizes calculated frequency values derived directly from the spacing of raw M/Z points.
+We introduce a novel method for characterizing peaks from direct injection FT-MS experiments developed on metabolomics datasets that utilizes frequency values derived directly from the spacing of raw M/Z points.
 Our method incorporates several aspects designed to make the derived intensities trustworthy, and suitable for use in our previously published peak assignment algorithm SMIRFE. 
@@
 However, the frequency values can be approximated by averaging the M/Z of adjacent points, and dividing the M/Z by the difference. 
-The *true* difference can be obtained by examining the mode of the mean frequency differences, and then constraining *useful* points to be within 2% of the mode value. 
-These points can be used to construct a linear model relating M/Z to frequency for individual scans based on the formula:
+Ideally, the difference between subsequent points in frequency space is constant, in practice there is a range of differences in frequency space.
+The *true, constant* difference can be obtained by examining the mode of the mean frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to M/Z) to be within 2% of the mode value.
+These *useful* points can be used to construct a linear model relating M/Z to frequency for individual scans based on the formula:
 
 $$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
-From the physical properties of the orbitrap, only the square root should be necessary, however, in practice we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry and control of the magnetic fields.
-A model is generated for each scan, and then a single model using the median values of the intercept and slope across scans is chosen, and applied to convert the M/Z values to frequency for all scans.
-A single model rather than scan specific models is used because this was found to maintain the relative ordering of the peaks in M/Z and frequency space. 
-Integer values (for use with the IRanges R package) are generated by multiplying each frequency value by 400 and rounding.
+From the known physical properties of the orbitrap, only the square root term should be necessary.
+Practically, we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry and control of the magnetic fields.
+A frequency model is generated for each scan, and then a single model using the median values of each term across scans is generated, and applied to convert the M/Z values to frequency for all scans. 
+A single model rather than scan specific models is used for conversion because this was found to maintain the relative ordering of the peaks in M/Z and frequency space (see Results).
 
@@
 Two types of frequency intervals are used, sliding and tiled windows. 
-Sliding windows are 10 points wide, offset by one point from each other. 
-Tiled windows are one point wide, offset by one point from each other.
+In this work, the sliding windows are 10 points wide and offset by one point.
+The tiled windows are one point wide and offset by one point.
 
@@
 
-To enable interval arithmetic, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package [citation].
+To enable interval algebra, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package [citation].
 The sliding and tiled windows are also converted to IRanges objects the same way.
@@
 
-To find intervals that containing data that is not just noise, the number of non-zero intensity frequency points in each sliding window are counted, and any sliding window with a non-zero count greater than or equal to the 99th percentile of non-zero counts is kept, and the remaining sliding windows are reduced, where overlapping sliding windows are merged together to create the initial peak regions.
+To find intervals that contain points that are not just noise, the number of non-zero intensity frequency points in each sliding window are counted, and any sliding window with a non-zero count greater than or equal to the 99th percentile of non-zero counts is kept, and the remaining sliding windows are reduced, where overlapping sliding windows are merged together to create the initial peak regions.
 
-Within each initial interval region, peaks in each scan are detected (see **Peak Finding**), and their centers binned by the tiled windows.
+Within each initial interval region, peaks in each scan are detected (see **Peak Detection**), and their centers binned by the tiled windows.
 Adjacent tiled windows with non-zero peak counts are merged together, and any zero peak count tiled windows split the initial region into multiple peak interval regions.
@@
 In addition, only those peaks that have an intensity greater than 0.7 times the highest intensity peak in the scan are used. 
-Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each to all other scans.
-The scan with the lowest summed overall distance is chosen as the "reference" scan, and normalization factors for each scan are calculated as the median log peak intensity differences in scan_i compared to the reference scan.
+Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each scan to all other scans.
+The scan with the lowest summed overall distance is chosen as the "reference" scan, and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the reference scan.
 This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan number is checked, and those peaks with correlation of greater than 0.5 with scan number are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.
@@
 The non-zero intensity, normalized raw data points across scans can be combined, and then characterized again using the weighted parabolic fit previously described. 
-In addition to the data from the full set of raw points, means and standard deviations of the peak height and location can be extracted from the scan level peak characteristics previously calculated.
+In addition to the data from the full set of raw points, means and standard deviations of the peak height and location can be derived from the scan level peak characteristics previously calculated.
 
@@
 To correct these, either a correction based on a model of the truncated normal distribution can be used, or the differences can be simulated by sampling from data that is present in most of the scans.
-To simulate the effect on the standard deviations, the peaks present in all scans are used. 
+To simulate the effect of peaks missing from some scans on the standard deviations, the peaks present in all scans are used. 
 For each peak, a sample of the heights across scans are taken (ranging from 5% to 95% of scans), and a new standard deviation calculated for that fraction, and a ratio of the fractional standard deviation to the "true" standard deviation calculated.
@@
 However, the frequency can be calculated by dividing the average M/Z of two adjacent points by their difference (Figure XA, red points reresenting the average M/Z of two adjacent points, red lines representing the difference between the two adjacent points).
-The subsequent differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, shown in Figure XB and XC. 
-The Thermo Fusion instrument from which most of our collaborators data has been acquired, at resolution of 400K, seems to have a mode of 0.5, as shown in Figure XC and XD. 
+The subsequent differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, as shown in Figure XB and XC. 
+The Thermo Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 400K, has a mode of 0.5, as shown in Figure XC and XD. 
 Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in Figure XE.
@@
 Figure XA shows that some peaks will be correlated with scan number.
-The solution to peak intensities correlated with M/Z appears to be tied to their intensities, as shown in Figure XB.
+The solution to **2** (peak intensities correlated with M/Z) appears to be tied to their intensities, as shown in Figure XB.
 The highlighted peaks in Figure XB are those with an intensity greater than 0.7, and are **not** correlated with M/Z.
@@
 
-### Changes in RSD
+### Changes in Relative Standard Deviation (RSD)
 

### 2019-11-13 peakcharacterization_manuscript.Rmd
starting section on relNAP

@@
 Each step in the peak characterization either changes the overall number of peaks coming from each scan (sliding windows and breaking initial regions) or the overall intensity of the points within a scan.
-Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation for the characterized scan level peak intensities, and compare them as each processing step is introduced.
+Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation (RSD) for the characterized scan level peak intensities, and compare them as each processing step is introduced.
 
@@
 ```
+
+### Difference to Relative Natural Abundance
+
+As an alternative to RSD, we can also compare the fit of relative intensities after assignment using SMIRFE [reference] to the theoretical relative natural abundances (relNAP) of the assigned isotopic molecular formula's (IMFs) within the assigned elemental molecular formula's (EMFs).
+Theoretically, we expect that lower quality data will have both lower numbers of assignments, and that for those things that are assigned, the fit between relative intensity and relNAP will be worse.

### 2019-11-14 peakcharacterization_manuscript.Rmd
using the right object
generating comparisons of normalization factors across three different normalization methods

@@
 
+```{r normalization_factors}
+loadd(normalization_plot_normalization_values_data_4)
+normalization_plot_normalization_values_data_4
+```
+
 ### Changes in Relative Standard Deviation (RSD)
@@
 drake::loadd(rsd_table_rsd_data)
-knitr::kable(rsd_out, digits = 3)
+knitr::kable(rsd_table_rsd_data, digits = 3)
 ```

### 2019-11-14 data_analysis/R/graph_generation.R

@@ -186,3 +186,26 @@ proportional_error = function(peak_data){
   })
   peak_sd$rsd = peak_sd$sd / peak_sd$mean
 }
+
+
+normalization_graph = function(normalization_data){
+  normalization_data = dplyr::filter(normalization_data, !(processing %in% "filtersd"))
+
+  hist_plot = ggplot(normalization_data, aes(x = normalization)) + geom_histogram() +
+    facet_wrap(~ processing, ncol = 1)
+
+  split_norm = split(normalization_data, normalization_data$processing)
+  ref_norm = split_norm[["singlenorm"]]
+  diff_norm = purrr::map_df(split_norm, function(in_norm){
+    combine_norm = dplyr::left_join(ref_norm, in_norm, by = "scan", suffix = c(".ref", ".in"))
+    dplyr::mutate(combine_norm, scan = scan, processing = processing.in,
+                  diff = normalization.in - normalization.ref) %>%
+      dplyr::select(scan, diff, processing)
+  })
+
+  diff_norm = dplyr::filter(diff_norm, !(processing %in% "singlenorm"))
+  diff_plot = ggplot(diff_norm, aes(x = scan, y = diff, color = processing)) + geom_point() + geom_line() +
+    theme(legend.position = c(0.8, 0.8))
+
+  (hist_plot | diff_plot) + plot_annotation(tag_levels = "A")
+}

### 2019-11-14 make_plan.R

@@ -61,6 +61,10 @@ analysis_plan = drake_plan(
     normalization_factors(method),
     transform = combine(method, .by = data)
   ),
+  normalization_plot = target(
+    normalization_graph(normalization_values),
+    transform = map(normalization_values)
+  ),
   # here we want to combine the ways a data file was processed and compare
   # their rsd's across processing methods.
   rsd = target(

### 2019-11-14 data_analysis/R/graph_generation.R
trying another image

@@ -205,7 +205,7 @@ normalization_graph = function(normalization_data){
 
   diff_norm = dplyr::filter(diff_norm, !(processing %in% "singlenorm"))
   diff_plot = ggplot(diff_norm, aes(x = scan, y = diff, color = processing)) + geom_point() + geom_line() +
-    theme(legend.position = c(0.5, 0.9))
+    theme(legend.position = c(0.4, 0.9))
 
   (hist_plot | diff_plot) + plot_annotation(tag_levels = "A")
 }

### 2019-12-11 peakcharacterization_manuscript.Rmd
trying another image
some better naming to better represent what is going on, and then actually putting the plots in the manuscript.

@@
 Due to differences in how many ions are captured in the trap, the peak intensities in each scan vary slightly.
-There are two issues with these peak intensities across scans: 1 - Some peak intensities correlate with the scan; 2 - Some peak intensities are correlated with M/Z / frequency.
+There are two issues with these peak intensities across scans: 1 - Some peak intensities correlate with the scan; 2 - Some peak differences across scans are correlated with intensity.
 The solution to **1** is to do a two pass normalization, detect the correlated peak intensities, remove them, and then calculate normalization factors again.
 Figure XA shows that some peaks will be correlated with scan number.
-The solution to **2** (peak intensities correlated with M/Z) appears to be tied to their intensities, as shown in Figure XB.
-The highlighted peaks in Figure XB are those with an intensity greater than 0.7, and are **not** correlated with M/Z.
-If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in Figure XC.
+The solution to **2** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in Figure XB.
+The highlighted peaks in Figure XB are those with an intensity greater than 0.7, and at least visually, their differences are **not** correlated with intensity.
+If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in Figure XA and XB.
 
 ```{r normalization_data}
-# figure showing peak intensity correlation with M/Z, intensity varying with M/Z, changes in normalization using all vs most intense
+# figure showing peak intensity correlation with M/Z, intensity varying with M/Z
+loadd(correlate_scan_height_plot_correlate_scan_height_data)
+correlate_scan_height_plot_correlate_scan_height_data
 ```
@@
 ```{r normalization_factors}
-loadd(normalization_plot_normalization_values_data_4)
-normalization_plot_normalization_values_data_4
+loadd(normalization_plot_normalization_values_data)
+normalization_plot_normalization_values_data
 ```

### 2019-12-11 data_analysis/R/graph_generation.R

@@ -210,7 +210,7 @@ normalization_graph = function(normalization_data){
   (hist_plot | diff_plot) + plot_annotation(tag_levels = "A")
 }
 
-plot_correlate_scan_mz = function(...){
+correlate_scan_height = function(...){
   processed_data = list(...)
   names(processed_data) = purrr::map_chr(processed_data, "processed")
   use_names = grep("perc99", names(processed_data), value = TRUE)
@@ -247,7 +247,7 @@ plot_correlate_scan_mz = function(...){
 
   example_peak = normed_peaks[[high_cor[1, "peak"]]]
 
-  height_2_scan = ggplot(example_peak, aes(x = scan, y = log10(Height))) + geom_point()
+  height_2_scan = ggplot(example_peak, aes(x = scan, y = log10(Height))) + geom_point(size = 3)
 
 
   # do the basic normalization to get M/Z to difference ----
@@ -375,4 +375,9 @@ plot_correlate_scan_mz = function(...){
 
   list(correlation = height_2_scan, diff_df = diff_int_list, diff_plot = diff_int_plots)
 
+}
+
+correlate_scan_height_graph = function(correlation_plots){
+  p = (correlation_plots$correlation | correlation_plots$diff_plot[[1]]) + plot_annotation(tag_levels = "A")
+  p
 }

### 2019-12-11 make_plan.R

@@ -56,9 +56,13 @@ analysis_plan = drake_plan(
     plot_sliding_window_density(data)
   ),
   # look for correlating with scan and correlating normalization values with M/Z
-  correlate_scan_mz = target(
-    plot_correlate_scan_mz(method),
+  correlate_scan_height = target(
+    correlate_scan_height(method),
     transform = combine(method, .by = data)
+  ),
+  correlate_scan_height_plot = target(
+    correlate_scan_height_graph(correlate_scan_height),
+    transform = map(correlate_scan_height)
   ),
   # combine the ways data file was processed and compare
   # normalization factors

### 2019-12-12 peakcharacterization_manuscript.Rmd
trying to use png with cairo output
slight reworking of issues w/ lasso models

@@
 ```{r setup, echo = FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dpi = 600, dev = "png", dev.args = list(type = "cairo"))
 ```
@@
 This is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points. 
-M/Z point-to-point differences are not constant, but can be approximated by a lasso linear model. 
-However, it is very difficult to create a lasso model with an intercept of 0, and then vary the width of sliding windows according to the M/Z difference at a particular M/Z. 
+M/Z point-to-point differences are not constant, but can be approximated by a lasso linear model; however, it is very difficult to create a lasso model with an intercept of 0.
+In addition, we would also need to vary the width of sliding windows according to the M/Z difference at a particular M/Z based on the lasso model.
 Frequency based points suffer none of these drawbacks, and the conversion from M/Z can be derived from the raw profile level data itself, which is incredibly useful.

### 2019-12-12 peakcharacterization_manuscript.Rmd
better description of the m/z to frequency models that are calculated

@@
 
-Frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
-When scan specific models are used for conversion of M/Z to frequency, the peak order observed in M/Z space is not maintained, as shown in Figure XA. 
+M/Z to frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
+While scan specific models **could** be used in the conversion of M/Z to frequency, doing so results in changes to the relative peak ordering compared to M/Z space, as shown in Figure XA. 
 Therefore, a single model for all scans based on the median of each term in the model across scans is used for converting *all* remaining scan level data.

### 2019-12-12 peakcharacterization_manuscript.Rmd
generating figure for peak characterization and inserting it into the manuscript

@@
 Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise. 
-The distribution of counts is shown in Figure X. 
+The distribution of counts in sliding windows ten points wide and offset by one point is shown in Figure X. 
 The number of peaks detected without applying a cutoff explodes by ..., with the vast majority of the peaks being present in a single scan. 
-By merging those windows that are adjacent **and** have a point density greater than the 99th percentile, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
+By discarding those windows that have a non-zero point density below the 99th percentile, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
 
@@
 # figure showing fit of quadratic model to actual data points
+drake::loadd(peak_fit_plot)
+peak_fit_plot
 ```
@@
 # figure showing peak intensity correlation with M/Z, intensity varying with M/Z
-loadd(correlate_scan_height_plot_correlate_scan_height_data)
+drake::loadd(correlate_scan_height_plot_correlate_scan_height_data)
 correlate_scan_height_plot_correlate_scan_height_data
@@
 ```{r normalization_factors}
-loadd(normalization_plot_normalization_values_data)
+drake::loadd(normalization_plot_normalization_values_data)
 normalization_plot_normalization_values_data

### 2019-12-12 data_analysis/R/graph_generation.R

@@ -381,3 +381,35 @@ correlate_scan_height_graph = function(correlation_plots){
   p = (correlation_plots$correlation | correlation_plots$diff_plot[[1]]) + plot_annotation(tag_levels = "A")
   p
 }
+
+
+plot_peak_fitting = function(processed_obj){
+  # to make this something we can test
+  #loadd("method_perc99_nonorm_data")
+  #processed_obj = method_perc99_nonorm_data
+
+  use_peak = processed_obj$char_obj$zip_ms$peak_finder$peak_regions$peak_region_list[[1]]
+  peak_points = as.data.frame(use_peak$points@elementMetadata)
+  use_scan = unique(peak_points$scan)[1]
+  peak_points = dplyr::filter(peak_points, scan %in% use_scan)
+
+  w = peak_points$intensity / max(peak_points$intensity)
+  peak_points$log_int = log(peak_points$intensity + 1e-8)
+  peak_fit = FTMS.peakCharacterization:::parabolic_fit(peak_points$frequency, peak_points$log_int, w = w)
+  peak_points$log_fit = peak_fit$fitted.values
+  fitted_info = FTMS.peakCharacterization:::get_fitted_peak_info(peak_points, "frequency", w)
+
+  log_plot = ggplot(peak_points, aes(x = frequency, y = log_int)) +
+    geom_line(aes(x = frequency, y = log_fit), color = "red", size = 2) +
+    geom_point(size = 3) +
+    geom_point(data = fitted_info, aes(x = ObservedCenter, y = log(Height)), color = "blue", size = 3) +
+    labs(x = "Frequency", y = "Log(Intensity)")
+
+  norm_plot = ggplot(peak_points, aes(x = frequency, y = intensity)) +
+    geom_line(aes(x = frequency, y = exp(log_fit)), color = "red", size = 2) +
+    geom_point(size = 3) +
+    geom_point(data = fitted_info, aes(x = ObservedCenter, y = Height), color = "blue", size = 3) +
+    labs(x = "Frequency", y = "Intensity")
+
+  (log_plot | norm_plot) + plot_annotation(tag_levels = "A")
+}

### 2019-12-12 make_plan.R

@@ -55,6 +55,10 @@ analysis_plan = drake_plan(
   sliding_regions = target(
     plot_sliding_window_density(data)
   ),
+  # peak fitting
+  peak_fit_plot = target(
+    plot_peak_fitting(method_perc99_nonorm_data)
+  ),
   # look for correlating with scan and correlating normalization values with M/Z
   correlate_scan_height = target(
     correlate_scan_height(method),

### 2019-12-17 data_analysis/R/graph_generation.R
add tag levels, and put it in the manuscript

@@ -459,5 +459,5 @@ plot_region_splitting = function(region_list){
     coord_cartesian(xlim = c(xmin, xmax)) +
     labs(x = "frequency", y = "no. of peaks")
 
-  p1 / p2 / p3
+  (p1 / p2 / p3) + plot_annotation(tag_level = "A")
 }

### 2019-12-17 peakcharacterization_manuscript.Rmd
add tag levels, and put it in the manuscript
adding the caption for the region splitting figure

@@
 Adjacent non-zero frequency bins are merged to comprise a single peak region. 
-Figure X shows an example where an initial region is broken up into multiple regions based on the characterized peak centers.
+Figure X shows an example where an initial region is broken up into two regions based on the characterized peak centers.
 
@@
 # figure showing multiple "peaks" in what is initially a single region
+drake::loadd(split_region_plot)
+split_region_plot
 ```
+
+*caption*:Splitting a single region into two regions based on the present peaks.
+**A**: The full set of raw frequency and intensity data across all scans for the region are shown. 
+Clearly the region has two separate peaks within it.
+**B**: The peak centroids (frequency and intensity) for each peak in black.
+The tiled regions (red) are used to quantify the number of peaks.
+**C**: The number of peaks within each tiled region shown as a histogram.
+Each group of non-zero adjacent regions will be merged to form a new peak region.
 

### 2019-12-17 peakcharacterization_manuscript.Rmd
editing normalization and captions

@@
 Due to differences in how many ions are captured in the trap, the peak intensities in each scan vary slightly.
-There are two issues with these peak intensities across scans: 1 - Some peak intensities correlate with the scan; 2 - Some peak differences across scans are correlated with intensity.
-The solution to **1** is to do a two pass normalization, detect the correlated peak intensities, remove them, and then calculate normalization factors again.
-Figure XA shows that some peaks will be correlated with scan number.
+Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
+However, there are two issues with these peak intensities across scans: 1 - Some peak intensities correlate with the scan number; 2 - Some peak differences between scans are correlated with intensity.
+The solution to **1** is to do a two pass normalization.
+In the first pass, the peaks who's intensity is correlated with intensity are detected.
+In the second pass, the correlated peaks are removed, and normalization is carried out again.
+Figure XA contains an example peak whose intensity across scans is correlated with scan number.
 The solution to **2** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in Figure XB.
-The highlighted peaks in Figure XB are those with an intensity greater than 0.7, and at least visually, their differences are **not** correlated with intensity.
+The highlighted peaks in Figure XB are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
 If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in Figure XA and XB.
@@
 
+*caption*: **A**: An example of a peak whose height across scans is correlated with scan number.
+**B**: The peak differences to the same peaks in a reference scan are plotted against peak height.
+Black: Peaks with a height < 0.7 of the maximum. 
+Red: Peaks with a height >= 0.7 of the maximum.
+
 ```{r normalization_factors}
@@
 
+*caption*: **A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity >= 0.7 of the maximum intensity (singlenorm_int), or the double pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
+**B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.
+
 ### Changes in Relative Standard Deviation (RSD)
@@
 Each step in the peak characterization either changes the overall number of peaks coming from each scan (sliding windows and breaking initial regions) or the overall intensity of the points within a scan.
-Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation (RSD) for the characterized scan level peak intensities, and compare them as each processing step is introduced.
+Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation (RSD) for the characterized scan level peak intensities (calculated as the standard deviation of peak heights across scans divided by the mean peak height), and compare them as each processing step is introduced.
 
@@
 ```
+
+*caption*: Density plots of relative standard deviations (RSD) of peak heights across scans for each of the processing methods. 
+A peak had to be present in at least three scans for the RSD value to be reported.
 

### 2019-12-24 peakcharacterization_manuscript.Rmd
discuss hpd sites

@@
 
-For each raw data file in mzML format, they were each processed in these six ways:
+For each raw data file in profile mzML format, they were processed in these six ways:
 
-  1. No noise removal, no normalization, no frequency SD filtering
-  2. Noise removal, no normalization, no frequency SD filtering
-  3. Noise removal, single pass normalization with all peaks, no frequency SD filtering
-  4. Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering
-  5. Noise removal, two pass normalization, no frequency SD filtering
-  6. Noise removal, two pass normalization, frequency SD filtering
+  1. No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
+  2. Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
+  3. Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
+  4. Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
+  5. Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
+  6. Noise removal, two pass normalization, frequency SD filtering (filtersd)
 
@@
 
+### Removal of High Peak Density Artefacts
+
+We previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [Mitchell et al., Metabolomics, 2018]. 
+Ideally, the peak characterization procedure should reduce their presence in the resultant reported peaks. 
+Their presence should be minimized by removing noise peaks, and removing regions that have multiple reported peaks in the same scan. 
+We expect they may present are as characterized peaks that have larger than expected standard deviations in their frequencies when calculated across scans.
+These peaks can be detected by simply examining the distribution of frequency SDs and removing those that are outliers.
+To verify the removal of HPD regions, we converted centroided M/Z's from XCalibur to frequency values using the previously decided values for that sample.
+Peak density was measured using a sliding window 10 points wide and offset by one point.
+
 ### Changes in Relative Standard Deviation (RSD)

### 2019-12-24 data_analysis/R/processing_functions.R
whole bunch of work to get the hpd regions out

@@ -396,3 +396,89 @@ split_regions = function(p99_nonorm_data){
   use_list
 
 }
+
+density_calculate = function(frequency_values, metadata, frequency_range = NULL){
+  freq_points = FTMS.peakCharacterization::frequency_points_to_frequency_regions(as.data.frame(frequency_values), multiplier = metadata$multiplier)
+
+  sliding_regions = FTMS.peakCharacterization::create_frequency_regions(point_spacing = metadata$point_spacing, frequency_range = frequency_range,
+                                             n_point = metadata$n_point, delta_point = metadata$delta_point,
+                                             multiplier = metadata$multiplier)
+
+  sliding_density = IRanges::countOverlaps(sliding_regions, freq_points)
+
+  n_region = 9 * metadata$n_point
+  small_window = c(1 * metadata$n_point, 5 * metadata$n_point)
+  large_window = c((5 * metadata$n_point) + 1, 9 * metadata$n_point)
+
+  start_i = (9 * metadata$n_point) + 1
+  end_i = length(sliding_regions) - (9 * metadata$n_point) - 1
+
+  stats = purrr::map_dbl(seq(start_i, end_i), function(in_i){
+    inner_window = list(sort(in_i - small_window),
+                        sort(in_i + small_window))
+    outer_window = list(sort(in_i - large_window),
+                        sort(in_i + large_window))
+
+    inner_seq = unlist(purrr::map(inner_window, ~ seq(.x[1], .x[2])))
+    outer_seq = unlist(purrr::map(outer_window, ~ seq(.x[1], .x[2])))
+
+    calculate_statistic = function(query_density, reference_window){
+      ref_mean = mean(reference_window)
+      ref_sd = sd(reference_window)
+      if ((ref_mean > 0) & (ref_sd > 0)) {
+        z = ((query_density - ref_mean)^2) / (ref_sd^2)
+        return(z * sign(query_density - ref_mean))
+      } else {
+        return(0)
+      }
+
+    }
+
+    z_inner = calculate_statistic(sliding_density[in_i], sliding_density[inner_seq])
+    z_outer = calculate_statistic(sliding_density[in_i], sliding_density[outer_seq])
+    max(z_inner, z_outer)
+  })
+  all_z = rep(0, length(sliding_density))
+  all_z[seq(start_i, end_i)] = stats
+
[…]

### 2020-01-03 noise_cutoffs/generate_differing_noise.Rmd
checking differences in noise levels

---
title: "Generate Different Noise Levels"
author: "Robert M Flight"
date: "1/3/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(FTMS.peakCharacterization)
disable_logging()
library(furrr)
future::plan(multicore(workers = 8))
FTMS.peakCharacterization::set_internal_map(furrr::future_map)
```

## Noise??

Based on a previous effort, we are pretty sure we are being *very* conservative in our noise cutoff levels (like 4X too conservative). 
So this document is to create a function that can do the new noise calculation, and then do the full characterization while using different noise cutoffs.

## Plan

Create a custom pipeline for doing the noise cutoff that takes a data file and some different multipliers of the noise cutoff, and then apply it.
We will start with 1, and then go all the way up to 4 in 0.5 increments.
The 4X should be equivalent to what we were doing previously.

## Execute!

### Noise Function

```{r noise_function}
reduce_removing_noise <- function(regions, point_regions, multiplier = 1){
  regions = FTMS.peakCharacterization:::count_overlaps(regions, point_regions)
  nz_counts = regions@elementMetadata$nonzero_counts

  chunk_indices = seq(1, length(nz_counts), by = 10000)

  chunk_perc = purrr::map_dbl(chunk_indices, function(in_index){
    use_density = sliding_density[seq(in_index, min(in_index + 9999, length(sliding_density)))]
    if (max(use_density) > 0) {
      return(stats::quantile(use_density, 0.99))
    } else {
      return(0)
    }
  })

  cutoff_value = median(chunk_perc) * multiplier

  regions = regions[nz_counts > cutoff_value]
  IRanges::reduce(regions)
}
```

### Characterization Function

```{r characterization_function}
characterize_varying_noise = function(raw_file, out_file, multiplier = 1){
  in_char <- CharacterizeMSPeaks$new(raw_file, peak_finder = PeakRegionFinder$new())
  in_char$load_file()
  in_char$filter_raw_scans()
  in_char$zip_ms$peak_finder <- in_char$peak_finder
  in_char$zip_ms$peak_finder$add_data(in_char$zip_ms$raw_ms)
  in_char$zip_ms$raw_ms <- NULL # we drop the raw data to save us some memory
  in_char$zip_ms$peak_finder$sample_id <- in_char$zip_ms$id
  in_char$zip_ms$peak_finder$start_time = Sys.time()
  in_char$zip_ms$peak_finder$progress = TRUE

  in_char$zip_ms$peak_finder$add_regions()
  in_char$zip_ms$peak_finder$peak_regions$peak_regions =
    reduce_removing_noise(in_char$peak_finder$peak_regions$sliding_regions,
                         in_char$peak_finder$peak_regions$frequency_point_regions,
                         multiplier = multiplier)
  in_char$zip_ms$peak_finder$split_peak_regions()
  in_char$zip_ms$peak_finder$remove_double_peaks_in_scans()
  in_char$zip_ms$peak_finder$normalize_data()
  in_char$zip_ms$peak_finder$find_peaks_in_regions()
  in_char$zip_ms$peak_finder$add_offset()
  in_char$zip_ms$peak_finder$sort_ascending_mz()

  in_char$out_file = out_file
  in_char$zip_ms$out_file = in_char$out_file

  in_char$zip_ms$peak_finder$stop_time = Sys.time()
  in_char$summarize()
  in_char$save_peaks()
  in_char$write_zip()
  in_char$zip_ms$cleanup()
}
```

### Run It!

```{r run_noise}
in_file = file.path(here::here(), "data_analysis", "data_input", "161212_unlabeledAAs_1_ECF.mzML")

noise_levels = seq(1, 4, 0.5)

purrr::map(noise_levels, function(in_level){
  out_file = paste0(gsub(".mzML", "", basename(in_file)), "_", in_level, ".zip")
  characterize_varying_noise(in_file, out_file, in_level)
})
```

### 2020-01-03 noise_cutoffs/generate_differing_noise.Rmd
missed an instance of sliding density

@@
   chunk_perc = purrr::map_dbl(chunk_indices, function(in_index){
-    use_density = sliding_density[seq(in_index, min(in_index + 9999, length(sliding_density)))]
+    use_counts = nz_counts[seq(in_index, min(in_index + 9999, length(nz_counts)))]
     if (max(use_density) > 0) {

### 2020-01-03 noise_cutoffs/generate_differing_noise.Rmd
and missed use_density, dammit

@@
     use_counts = nz_counts[seq(in_index, min(in_index + 9999, length(nz_counts)))]
-    if (max(use_density) > 0) {
-      return(stats::quantile(use_density, 0.99))
+    if (max(use_counts) > 0) {
+      return(stats::quantile(use_counts, 0.99))
     } else {

### 2020-01-17 peakcharacterization_manuscript.Rmd
adding stuff on peak finding and hpd detection

@@
 Ideally, the difference between subsequent points in frequency space is constant, in practice there is a range of differences in frequency space.
-The *true, constant* difference can be obtained by examining the mode of the mean frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to M/Z) to be within 2% of the mode value.
+The *true, constant* difference can be obtained by examining the median of the mean frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to M/Z) to be within 2% of the mode value.
 These *useful* points can be used to construct a linear model relating M/Z to frequency for individual scans based on the formula:
@@
 
-To find intervals that contain points that are not just noise, the number of non-zero intensity frequency points in each sliding window are counted, and any sliding window with a non-zero count greater than or equal to the 99th percentile of non-zero counts is kept, and the remaining sliding windows are reduced, where overlapping sliding windows are merged together to create the initial peak regions.
+To find intervals that contain points that are not just noise, the number of non-zero intensity frequency points in each sliding window are counted.
+The 99th percentile of non-zero counts across 10000 sliding windows are computed over all sliding windows, and then the median value of these 99th percentiles computed. 
+Any sliding window with a non-zero count less than or equal to 1.5 times the median percentile is removed, and the remaining sliding windows are reduced, where overlapping windows are merged together to create the initial peak regions.
 
@@
 
-High-peak-density artifacts [Mitchell et al., 2018] present as singular peaks with higher than expected frequency standard deviations calculated from the scan-to-scan frequency peak locations.
-Outliers are removed by examining the distribution of peak frequency SDs across the entire spectrum.
+High-peak-density (HPD) artifacts [Mitchell et al., 2018] present as singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
+Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and those FSD that are greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are removed.
+The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work.
+For HPD detection, the peaks in excel output from Xcalibur were used after converting the M/Z peak centers into frequency space.
+Sliding windows 1000 frequency points wide and offset by 100 points were used for the density calculations.
 

### 2020-01-17 peakcharacterization_manuscript.Rmd
describing another way we can evaluate that our peak intensities are better compared to something else

@@
 Theoretically, we expect that lower quality data will have both lower numbers of assignments, and that for those things that are assigned, the fit between relative intensity and relNAP will be worse.
+
+### Changes in Variance Across a Large Dataset
+
+For large datasets, we expect that more correct peak intensities will result in better agreement between sample normalized peak intensities within a sample class. 
+If incorrect peak intensities are being reported, then we would expect between class to within class variances to be very random, resulting in F-statistics close to 1.
+For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using our peak-characterization and those from XCMS.

### 2020-04-10 peakcharacterization_manuscript.Rmd
added conversion back into M/Z into manuscript

@@
 
+To convert M/Z back into frequency, we can use a similar model without the roots.
+
+$$mz = intercept + x* \frac{1}{frequency} + y*\frac{1}{frequency^2} + z*\frac{1}{frequency^3}$$
+
 ### Frequency Intervals
@@
 
-Scans are normalized to a single *median* scan based on the log-intensity differences of a subset of peaks that are present in at least the same number of scans as the 95th percentile of scan counts for the peaks.
+Scans are normalized to a single *reference* scan based on the log-intensity differences of a subset of peaks that are present in at least the same number of scans as the 95th percentile of scan counts for the peaks.
 In addition, only those peaks that have an intensity greater than 0.7 times the highest intensity peak in the scan are used. 
 Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each scan to all other scans.
-The scan with the lowest summed overall distance is chosen as the "reference" scan, and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the reference scan.
+The scan with the lowest summed overall distance is chosen as the *reference* scan, and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the reference scan.
 This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan number is checked, and those peaks with correlation of greater than 0.5 with scan number are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.
@@
 Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and those FSD that are greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are removed.
-The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work.
+The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
 For HPD detection, the peaks in excel output from Xcalibur were used after converting the M/Z peak centers into frequency space.
 Sliding windows 1000 frequency points wide and offset by 100 points were used for the density calculations.
+
+### Multi-Class F-Statistics
+
+To evaluate the F-statistic between multi-class samples, a much larger set of samples was used.
+In this case, 181 matched non-cancer and cancer samples that were previously used in the HPD detection manuscript [ref].
+All 181 samples were peak characterized using the most stringent method below as well as peaks generated from the xcms R package [ref], assigned using SMIRFE, and then peaks matched by shared EMFs across samples (see Supplemental materials for a description of peak matching and deciding the most likely EMF).
+For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using our peak-characterization and those from xcms.
 
@@
 
-We previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [Mitchell et al., Metabolomics, 2018]. 
+We have previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [Mitchell et al., Metabolomics, 2018]. 
 Ideally, the peak characterization procedure should reduce their presence in the resultant reported peaks. 
 Their presence should be minimized by removing noise peaks, and removing regions that have multiple reported peaks in the same scan. 
-We expect they may present are as characterized peaks that have larger than expected standard deviations in their frequencies when calculated across scans.
+We expect they may present as characterized peaks that have larger than expected frequency standard deviations (FSD) when calculated across scans.
 These peaks can be detected by simply examining the distribution of frequency SDs and removing those that are outliers.
-To verify the removal of HPD regions, we converted centroided M/Z's from XCalibur to frequency values using the previously decided values for that sample.
+To verify the removal of HPD regions, we converted centroided M/Z's from XCalibur to frequency values using the previously calculated values for that sample.
 Peak density was measured using a sliding window 10 points wide and offset by one point.
+
+Figure X compares the HPD regions detected using the method from Mitchell et al., and their correspondence with high FSD peaks for one sample (the other comparisons are in Supplemental materials).
+Based on this comparison, the high FSD peaks are indeed HPD sites and should be removed before assignment.
+However, the FSD does not depend on calculating a sliding window based density, but is merely a product of those peaks across multiple scans that span a wider range frequency range than expected based on the peaks in the data.
+We also note that we do expect higher M/Z standard deviations at higher M/Z, but the frequency based method does not suffer the same effect, due to the offsets of frequency points being constant.
 
@@
 Theoretically, we expect that lower quality data will have both lower numbers of assignments, and that for those things that are assigned, the fit between relative intensity and relNAP will be worse.
+To compare relative NAP to relative abundances, we only examined the assignments from the two samples containing ECF derivatized amino-acids, as we can limit the assignments to those that match expected derivatizations of the known amino-acids (see Supplemental for the expected EMFs, and expected relative NAPs for the individual IMFs).
 
@@
 If incorrect peak intensities are being reported, then we would expect between class to within class variances to be very random, resulting in F-statistics close to 1.
-For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using our peak-characterization and those from XCMS.

### 2022-03-02 _targets.R
ready for a first run at least

@@ -9,30 +9,24 @@ excel_files = dir(here::here("data/data_input"), pattern = ".xlsx", full.names =
 
 method_mzml = expand_grid(
   method_function = rlang::syms(c("noperc_nonorm",
-                            "perc99_nonorm",
-                            "singlenorm",
-                            "singlenorm_int",
-                            "doublenorm",
-                            "filtersd")),
+                                  "perc99_nonorm",
+                                  "singlenorm",
+                                  "singlenorm_int",
+                                  "doublenorm",
+                                  "filtersd")),
   mzml = use_files)
 method_mzml = method_mzml %>%
   dplyr::mutate(rest_id = gsub(".mzML", "", basename(mzml)))
 ## tar_plan supports drake-style targets and also tar_target()
-pkg = tar_target(pkg, utils::packageDescription('FTMS.peakCharacterization'))
+pkg_tar = tar_target(pkg, utils::packageDescription('FTMS.peakCharacterization'))
 method_peaks = tar_map(
   unlist = FALSE,
   values = method_mzml,
   names = c("method_function", "rest_id"),
   tar_target(data, reading_scans_tile_windows(mzml, pkg)),
-  tar_target(processing, method_function(data)),
-  tar_target(zip, write_peaks_for_assignment(processing)),
+  tar_target(method, method_function(data)),
+  tar_target(zip, write_peaks_for_assignment(method)),
   tar_target(assign, assign_files(zip))
-)
-
-combine_scan_height = tar_combine(
-  scan_height_cor,
-  method_peaks[["data"]],
-  command = correlate_scan_height(!!!.x)
 )
 
 one_offs = tar_plan(
@@ -44,4 +38,4 @@ one_offs = tar_plan(
   tar_render(manuscript, "doc/peakcharacterization_manuscript.Rmd")
 )
 
-list(pkg, method_peaks, combine_scan_height, one_offs)
+list(pkg_tar, method_peaks, one_offs)

### 2022-03-02 doc/peakcharacterization_manuscript.Rmd
moving things around

--- 2020-04-10 peakcharacterization_manuscript.Rmd
@@
 output: 
-  redoc::redoc
+  word_document:
+    keep_md: true
 ---

### 2022-03-02 R/graph_generation.R

@@ -48,7 +48,9 @@ plot_nap_peaks = function(raw_data, processed_data, assignments, interesting_reg
     geom_point(data = scan_means, aes(x = mean_mz + 0.02, y = log10(mean_height)), color = "brown", size = 4)
 }
 
-plot_frequency_conversion = function(raw_data){
+plot_frequency_conversion = function(...){
+  all_data = as.list(...)
+
   raw_points = as.data.frame(raw_data$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions@elementMetadata)
   raw_scan = dplyr::filter(raw_points, scan %in% unique(scan)[1])
   raw_points2 = convert_mz_frequency(raw_scan)

### 2022-03-02 R/processing_functions.R

@@ -6,6 +6,7 @@ reading_scans_tile_windows <- function(in_file, pkg_desc){
   use_file = in_file
   char_ms <- CharacterizeMSPeaks$new(use_file, peak_finder = PeakRegionFinder$new())
   char_ms$load_file()
+  char_ms$peak_finder$start_time = Sys.time()
   char_ms$filter_raw_scans()
   char_ms$zip_ms$peak_finder <- char_ms$peak_finder
   char_ms$zip_ms$peak_finder$add_data(char_ms$zip_ms$raw_ms)
@@ -14,7 +15,8 @@ reading_scans_tile_windows <- function(in_file, pkg_desc){
   char_ms$zip_ms$cleanup() # don't want to leave stuff in the tmp directory
   char_ms$zip_ms$peak_finder$add_regions()
 
-  list(char_obj = char_ms, pkg = pkg_desc)
+  list(char = char_ms,
+       pkg = pkg_desc)
 }
 
 reduce_removing_zero <- function(regions, point_regions, min_value = 0){
@@ -26,14 +28,14 @@ reduce_removing_zero <- function(regions, point_regions, min_value = 0){
 }
 
 # only remove zero points, no consideration of percentile, and no normalization (zero_normalization)
-noperc_nonorm = function(in_list){
-  in_char = in_list$char_obj$clone(deep = TRUE)
+noperc_nonorm = function(char_obj){
+  in_char = char_obj$clone(deep = TRUE)
   in_char$zip_ms$peak_finder$zero_normalization = TRUE
   in_char$zip_ms$peak_finder$progress = TRUE
 
   in_char$zip_ms$peak_finder$peak_regions$peak_regions =
     reduce_removing_zero(in_char$peak_finder$peak_regions$sliding_regions,
-                         in_char$peak_finder$peak_regions$frequency_point_regions)
+                    in_char$peak_finder$peak_regions$frequency_point_regions)
   in_char$zip_ms$peak_finder$split_peak_regions()
   in_char$zip_ms$peak_finder$remove_double_peaks_in_scans()
   in_char$zip_ms$peak_finder$normalize_data()
@@ -44,8 +46,8 @@ noperc_nonorm = function(in_list){
 }
 
 # remove points based on the 99th percentile, no normalization (see zero_normalization)
-perc99_nonorm = function(in_list){
-  in_char = in_list$char_obj$clone(deep = TRUE)
+perc99_nonorm = function(char_obj){
[…]

### 2022-03-02 data_analysis/run_before_normalization.R

@@ -1,20 +1,0 @@
-library(SIRM.FTMS.peakCharacterization)
-library(rprojroot)
-library(parallel)
-options(mc.cores = 10)
-project_root <- find_root(is_rstudio_project)
-source(file.path(project_root, "data_analysis/R/processing_functions.R"))
-
-peak_pkg_description = utils::packageDescription('SIRM.FTMS.peakCharacterization')
-
-zip_files <- dir(file.path(project_root, "data_analysis", "data_input"), full.names = TRUE)
-
-run_info <- mclapply(zip_files, function(in_file){
-  zip_data <- raw_peakpicking(in_file, peak_pkg_description)
-  zip_data$peak_finder$create_correspondent_peaks(median_corrected = FALSE)
-  zip_data$peak_finder$collapse_correspondent_peaks()
-  out_path <- file.path(project_root, "data_analysis", "data_output")
-  out_file <- gsub(".zip", ".rds", basename(in_file))
-  saveRDS(zip_data, file.path(out_path, out_file))
-  file.path(out_path, out_file)
-})

### 2022-03-02 packages.R

@@ -1,0 +1,24 @@
+## library() calls go here
+library(conflicted)
+library(dotenv)
+options(future.fork.enable = TRUE)
+library(targets)
+library(tarchetypes)
+library(FTMS.peakCharacterization)
+library(methods)
+disable_logging()
+show_progress()
+library(furrr)
+future::plan(multicore)
+FTMS.peakCharacterization::set_internal_map(furrr::future_map)
+library(rprojroot)
+library(ggplot2)
+theme_set(cowplot::theme_cowplot())
+library(patchwork)
+library(ggridges)
+library(dplyr)
+library(glue)
+library(here)
+library(tidyr)
+options(future.globals.maxSize = 2000 * 1024^2)
+library(rmarkdown)

### 2022-03-10 doc/peakcharacterization_manuscript.Rmd
trying to get most of the functionality there!

--- 2020-04-10 peakcharacterization_manuscript.Rmd
@@
 output: 
-  redoc::redoc
+  word_document:
+    keep_md: true
+editor_options: 
+  chunk_output_type: console
 ---
@@
 ```{r setup, echo = FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dpi = 600, dev = "png", dev.args = list(type = "cairo"))
+knitr::opts_chunk$set(echo = FALSE, 
+                      warning = FALSE, 
+                      message = FALSE, 
+                      dpi = 600, 
+                      dev = "png", 
+                      dev.args = list(type = "cairo"),
+                      fig.process = dn_figure_rename)
+```
+
+```{r get_setup}
+source("./packages.R")
+lapply(list.files("./R", full.names = TRUE), source)
+
+figure_count = dn_counter$new("Figure ")
 ```
@@
 To find intervals that contain points that are not just noise, the number of non-zero intensity frequency points in each sliding window are counted.
-The 99th percentile of non-zero counts across 10000 sliding windows are computed over all sliding windows, and then the median value of these 99th percentiles computed. 
-Any sliding window with a non-zero count less than or equal to 1.5 times the median percentile is removed, and the remaining sliding windows are reduced, where overlapping windows are merged together to create the initial peak regions.
+Subsequently, we break these counts into fixed width tiles (default width of 2000) and calculate the 99th percentile of non-zero points for each tile.
+The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions that contain potential signal.
+Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged together to create the initial peak regions.
+
+**Hunter talk about issues around non-detection and why we only use non-zero counts, sliding followed by tiles**.
 
@@
 
-```{r mz_frequency_figure}
-drake::loadd(frequency_conversion)
-frequency_conversion
+```{r increment_mz_frequency}
+figure_count$increment("mz_frequency_conversion")
 ```
 
-*caption*: **A**: Intensity vs M/Z for a single peak from a single scan. 
+```{r mz_frequency_conversion, dn_id = figure_count}
+tar_load(frequency_conversion)
+frequency_conversion$all
+```
+
+`r figure_count$label_text("mz_frequency_conversion")`. **A**: Intensity vs M/Z for a single peak from a single scan. 
 Red lines denote the differences between each point, and red dots the average between the pair of points. 
@@
 
-```{r peak_ordering}
-drake::loadd(peak_ordering)
+```{r increment_peak_ordering}
+figure_count$increment("peak_ordering")
+```
+
+```{r peak_ordering, dn_id = figure_count}
+tar_load(peak_ordering)
 peak_ordering
@@
 
-*caption*: Peak ordering in M/Z compared with ordering in frequency space when a single M/Z to frequency model is used or scan specific M/Z to frequency models are used. 
+`r figure_count$label_text("peak_ordering")`.
+Peak ordering in M/Z compared with ordering in frequency space when a single M/Z to frequency model is used or scan specific M/Z to frequency models are used.
 For a single peak, the scan level peak M/Z's were extracted, and then frequency values for those M/Z generated using a single common model of M/Z to frequency (*single_order*), or models derived from each scan (*scan_order*).
@@
 
-```{r slidingwindow_count_distribution}
-drake::loadd(sliding_regions)
+```{r increment_slidingwindow}
+figure_count$increment("slidingwindow_count")
+```
+
+```{r slidingwindow_count, dn_id = figure_count}
+tar_load(sliding_regions)
 sliding_regions
@@
 
-*caption*: Histogram of the number of non-zero points across all scans in sliding windows 10 frequency points wide and 1 frequency point apart. 
+`r figure_count$label_text("slidingwindow_count")`. Histogram of the number of non-zero points across all scans in sliding windows 10 frequency points wide and 1 frequency point apart. 
 The red vertical line denotes the lower 99th percentile of the data, only sliding windows with non-zero counts above the red line will be kept for subsequent use.
@@
 
-```{r centroided_peaks}
+```{r increment_fit}
+figure_count$increment("centroided_peaks")
+```
+
+```{r centroided_peaks, dn_id = figure_count}
 # figure showing fit of quadratic model to actual data points
-drake::loadd(peak_fit_plot)
+tar_load(peak_fit_plot)
 peak_fit_plot
@@
 
+```{r increment_breaking_regions}
+figure_count$increment("breaking_regions")
+```
+
 With the characterized (centroided) peak data from across scans within each region, it is then important to determine if only one or multiple "peaks" are actually present in the region. 
@@
 Adjacent non-zero frequency bins are merged to comprise a single peak region. 
-Figure X shows an example where an initial region is broken up into two regions based on the characterized peak centers.
+`r figure_count$label_text("breaking_regions")` shows an example where an initial region is broken up into two regions based on the characterized peak centers.
 
-```{r breaking_regions}
+```{r breaking_regions, dn_id = figure_count}
 # figure showing multiple "peaks" in what is initially a single region
-drake::loadd(split_region_plot)
+tar_load(split_region_plot)
 split_region_plot
@@
 
-*caption*:Splitting a single region into two regions based on the present peaks.
+`r figure_count$label_text("breaking_regions")`. 
+Splitting a single region into two regions based on the present peaks.
 **A**: The full set of raw frequency and intensity data across all scans for the region are shown. 
@@
 
-```{r normalization_data}
+```{r normalization_data, eval = FALSE}
 # figure showing peak intensity correlation with M/Z, intensity varying with M/Z
-drake::loadd(correlate_scan_height_plot_correlate_scan_height_data)
+tar_load(correlate_scan_height_plot_correlate_scan_height_data)
 correlate_scan_height_plot_correlate_scan_height_data
@@
 
-```{r normalization_factors}
-drake::loadd(normalization_plot_normalization_values_data)
+```{r normalization_factors, eval = FALSE}
+tar_load(normalization_plot_normalization_values_data)
 normalization_plot_normalization_values_data
@@
 
+```{r increment_rsd}
+figure_count$increment("rsd_method")
+```
+
 Each step in the peak characterization either changes the overall number of peaks coming from each scan (sliding windows and breaking initial regions) or the overall intensity of the points within a scan.
@@
 
-```{r rsd_plots}
-drake::loadd(rsd_plot_rsd_data)
-rsd_plot_rsd_data
+```{r rsd_method, dn_id = figure_count}
+tar_load(rsd_plot)
+rsd_plot
 ```
 
-*caption*: Density plots of relative standard deviations (RSD) of peak heights across scans for each of the processing methods. 
+`r figure_count$label_text("rsd_method")`. 
+Density plots of relative standard deviations (RSD) of peak heights across scans for each of the processing methods. 
 A peak had to be present in at least three scans for the RSD value to be reported.
@@
 ```{r rsd_table, results = 'asis'}
-drake::loadd(rsd_table_rsd_data)
+tar_load(rsd_values)
 knitr::kable(rsd_table_rsd_data, digits = 3)

### 2022-03-10 R/graph_generation.R

@@ -211,7 +211,8 @@ plot_rsd_differences = function(rsd_values){
 
   rsd_values$processed = forcats::fct_relevel(rsd_values$processed, "noperc_nonorm", "perc99_nonorm", "singlenorm", "singlenorm_int", "doublenorm",  "filtersd")
   rsd_comparison_plot = ggplot(rsd_values, aes(x = rsd, y = processed, fill = processed)) + geom_density_ridges() +
-    coord_cartesian(xlim = c(0, 1)) + theme(legend.position = "none")
+    coord_cartesian(xlim = c(0, 1)) + theme(legend.position = "none") +
+    facet_wrap(~ sample)
   rsd_comparison_plot
 }
 
@@ -430,9 +431,8 @@ plot_peak_fitting = function(processed_obj){
   #processed_obj = method_perc99_nonorm_data
 
   use_peak = processed_obj$char_obj$zip_ms$peak_finder$peak_regions$peak_region_list[[1]]
-  peak_points = as.data.frame(use_peak$points@elementMetadata)
+  peak_points = as.data.frame(use_peak$points[[1]]@elementMetadata)
   use_scan = unique(peak_points$scan)[1]
-  peak_points = dplyr::filter(peak_points, scan %in% use_scan)
 
   w = peak_points$intensity / max(peak_points$intensity)
   peak_points$log_int = log(peak_points$intensity + 1e-8)
@@ -456,33 +456,35 @@ plot_peak_fitting = function(processed_obj){
 }
 
 plot_region_splitting = function(region_list){
-  frequency_point_regions = region_list$points
-  tiled_regions = region_list$tiles
-  frequency_point_regions@elementMetadata$log_int <- log(frequency_point_regions@elementMetadata$intensity + 1e-8)
+  use_list = region_list[[1]]
+  frequency_point_regions = use_list$points
+  tiled_regions = use_list$tiles
 
-  scan_runs = rle(frequency_point_regions@elementMetadata$scan)
-  #if (min(scan_runs$lengths) < min_points + 2) {
-  frequency_point_splitscan <- split(frequency_point_regions, frequency_point_regions@elementMetadata$scan)
-  reduced_peaks <- purrr::map_df(names(frequency_point_splitscan), function(in_scan){
-    FTMS.peakCharacterization:::get_reduced_peaks(frequency_point_splitscan[[in_scan]], peak_method = "lm_weighted", min_points = 4)
+  frequency_point_regions = purrr::map(frequency_point_regions, function(in_scan){
[…]

### 2022-03-10 R/msnbase_xcms.R

@@ -1,0 +1,24 @@
+msnbase_only = function(mzml_file){
+  mzml_prof = MSnbase::readMSData(mzml_file, msLevel. = 1, centroided. = FALSE)
+  mzml_info = get_ms_info(mzml_prof)
+
+  all_scans_cent = mzml_prof %>%
+    MSnbase::pickPeaks()
+  all_scans_cent_mz = mz(all_scans_cent)
+  all_scans_cent_intensity = intensity(all_scans_cent)
+
+  all_scans_data = purrr::map_df(seq_len(length(all_scans_cent_mz)), function(in_scan){
+    data.frame(mz = all_scans_cent_mz[[in_scan]],
+               intensity = all_scans_cent_intensity[[in_scan]],
+               scan = in_scan)
+  })
+
+  comb_prof = MSnbase::combineSpectra(mzml_prof, method = meanMzInts, mzd = 0, ppm = 1)
+
+  comb_cent = comb_prof %>%
+    MSnbase::pickPeaks()
+  comb_cent_data = data.frame(mz = mz(comb_cent)[[1]],
+                              intensity = intensity(comb_cent)[[1]])
+
+  list(scanlevel = all_scans_data, comb = comb_cent_data)
+}

### 2022-03-10 R/processing_functions.R

@@ -26,7 +26,6 @@ reduce_removing_zero <- function(regions, point_regions_list, multiplier = 1.5, 
     nz_counts = nz_counts + nz_counts_iregion
   }
 
-
   regions = regions[nz_counts > 0]
   IRanges::reduce(regions)
 }
@@ -280,20 +279,22 @@ normalization_factors = function(...){
 }
 
 calc_rsd = function(in_char, processing){
-  scan_peaks = purrr::map(in_char$zip_ms$peak_finder$peak_regions$peak_region_list, "peaks")
-  n_peak = purrr::map_int(scan_peaks, nrow)
-  keep_peak = n_peak >= 3
-  scan_peaks = scan_peaks[keep_peak]
-  rsd = purrr::map_df(scan_peaks, function(in_peaks){
-    data.frame(mean = mean(in_peaks$Height),
-               sd = sd(in_peaks$Height),
-               n_peak = nrow(in_peaks),
-               stringsAsFactors = FALSE)
-  })
-  rsd$rsd = rsd$sd / rsd$mean
-  rsd$processed = processing
-  rsd$sample = in_char$zip_ms$peak_finder$sample_id
-  rsd
+  scan_peaks = 10^in_char$zip_ms$peak_finder$peak_regions$scan_level_arrays$Log10Height
+  n_scan = ncol(scan_peaks)
+  mean_peaks = rowMeans(scan_peaks, na.rm = TRUE)
+  sd_peaks = apply(scan_peaks, 1, sd, na.rm = TRUE)
+  n_peak = apply(scan_peaks, 1, function(.x){sum(!is.na(.x))})
+
+  rsd_df = data.frame(mean = mean_peaks,
+                      sd = sd_peaks,
+                      rsd = sd_peaks / mean_peaks,
+                      n = n_peak,
+                      n_perc = n_peak / n_scan,
+                      processed = processing,
+                      sample = in_char$zip_ms$peak_finder$sample_id)
+  rsd_df %>%
+    dplyr::filter(n >= 3)
+
 }
 
 single_rsd = function(char_list){
@@ -321,63 +322,99 @@ summarize_rsd = function(rsd_df){
 }
 
 split_regions = function(p99_nonorm_data){
-  # loadd("method_perc99_nonorm_data")
-  # p99_nonorm_data = method_perc99_nonorm_data
+  # loadd("method_perc99_nonorm_49lipid")
[…]

### 2022-03-10 doc/peakcharacterization_manuscript.Rmd
fixing the manuscript

@@
                       dev.args = list(type = "cairo"),
-                      fig.process = dn_figure_rename)
+                      fig.process = dn_modify_path)
 ```

### 2022-03-10 R/graph_generation.R
doing just a single sample for the scan intensity plot

@@ -252,12 +252,9 @@ normalization_graph = function(normalization_data){
   (hist_plot | diff_plot) + plot_annotation(tag_levels = "A")
 }
 
-correlate_scan_height = function(...){
-  processed_data = list(...)
-  names(processed_data) = purrr::map_chr(processed_data, "processed")
-  use_names = grep("perc99", names(processed_data), value = TRUE)
-  message(use_names)
-  processed_obj = processed_data[[use_names]]
+correlate_scan_height = function(char_list){
+  char_obj = char_list$char_obj
+  processed = char_list$processed
   # to make this something we can test
   # loadd("method_perc99_nonorm_data")
   # processed_obj = method_perc99_nonorm_data
@@ -266,13 +263,14 @@ correlate_scan_height = function(...){
   summary_function = median
   normalize_peaks = "both"
 
-  scan_peaks = purrr::map(processed_obj$char_obj$zip_ms$peak_finder$peak_regions$peak_region_list, ~ as.data.frame(.x$peaks))
+  scan_peaks = purrr::map(char_obj$zip_ms$peak_finder$peak_regions$peak_region_list, "peaks")
 
   ## getting the relationship between normalized peak height and scan ----
   normalization_factors <- FTMS.peakCharacterization:::single_pass_normalization(scan_peaks, intensity_measure = intensity_measure, summary_function = summary_function,
                                                                                  min_ratio = 0)
 
   normed_peaks = purrr::map(scan_peaks, function(.x){
+    .x = as.data.frame(.x)
     .x2 = dplyr::inner_join(.x, normalization_factors, by = "scan")
     .x2$Height = exp(log(.x2$RawHeight) - .x2$normalization)
     .x2$normalization = NULL

### 2022-03-10 _targets.R

@@ -60,13 +60,6 @@ rsd_tar = tar_combine(
   iteration = "list"
 )
 
-scan_height_tar = tar_combine(
-  scan_height_correlation,
-  method_tar[[1]],
-  command = correlate_scan_height(!!!.x),
-  iteration = "list"
-)
-
 # rsd_tar = tar_combine(
 #   rsd_data,
 #   values = ends_with("97lipid")
@@ -81,6 +74,8 @@ figures_tar = tar_plan(
 
   tar_target(find_sub_region, split_regions(method_perc99_nonorm_97lipid)
   ),
+  tar_target(intensity_scan_cor, correlate_scan_hight(method_perc99_nonorm_97lipid)),
+  tar_target(intensity_scan_cor_plot, correlate_scan_height_graph(intensity_scan_cor)),
   tar_target(split_region_plot,
              plot_region_splitting(find_sub_region)
   ),

### 2022-03-10 doc/peakcharacterization_manuscript.Rmd
whoops, wrong thing to print
update to the manuscript to have better figures

@@
 
-```{r setup, echo = FALSE}
+```{r setup, include = FALSE}
 knitr::opts_chunk$set(echo = FALSE, 
@@
                       message = FALSE, 
+                      fig.width = 8, 
+                      fig.height = 6, 
+                      fig.process = dn_modify_path,
                       dpi = 600,
-                      dev = "png", 
-                      dev.args = list(type = "cairo"),
-                      fig.process = dn_modify_path)
+                      dev.args = list(png = list(type = "cairo")))
 ```
@@
 tar_load(rsd_values)
-knitr::kable(rsd_table_rsd_data, digits = 3)
+knitr::kable(rsd_values, digits = 3)
 ```

### 2022-03-10 doc/peakcharacterization_manuscript.Rmd
oops

@@
 
-```{r normalization_data, eval = FALSE}
-# figure showing peak intensity correlation with M/Z, intensity varying with M/Z
-tar_load(correlate_scan_height_plot_correlate_scan_height_data)
-correlate_scan_height_plot_correlate_scan_height_data
+```{r increment_intensity_scan}
+figure_count$increment("intensity_scan")
 ```
 
-*caption*: **A**: An example of a peak whose height across scans is correlated with scan number.
+```{r intensity_scan, dn_id = figure_count}
+# figure showing peak intensity correlation with M/Z, intensity varying with M/Z
+tar_load(intensity_scan_cor_plot)
+intensity_scan_cor_plot
+```
+
+`r figure_count$label_text("intensity_scan")`. 
+**A**: An example of a peak whose height across scans is correlated with scan number.
 **B**: The peak differences to the same peaks in a reference scan are plotted against peak height.

### 2022-03-10 _targets.R

@@ -91,5 +91,4 @@ list(pkg_tar,
      method_tar,
      rsd_tar,
      figures_tar,
-     tables_tar,
-     scan_height_tar)
+     tables_tar)

### 2022-03-10 _targets.R
doing horribly at getting right names

@@ -74,7 +74,7 @@ figures_tar = tar_plan(
 
   tar_target(find_sub_region, split_regions(method_perc99_nonorm_97lipid)
   ),
-  tar_target(intensity_scan_cor, correlate_scan_hight(method_perc99_nonorm_97lipid)),
+  tar_target(intensity_scan_cor, correlate_scan_height(method_perc99_nonorm_97lipid)),
   tar_target(intensity_scan_cor_plot, correlate_scan_height_graph(intensity_scan_cor)),
   tar_target(split_region_plot,
              plot_region_splitting(find_sub_region)

### 2022-03-10 doc/peakcharacterization_manuscript.Rmd
new manuscript with better sized figure

@@
 
-```{r get_setup}
+```{r get_setup, include = FALSE}
 source("./packages.R")
@@
 
-```{r mz_frequency_conversion, dn_id = figure_count}
+```{r mz_frequency_conversion, dn_id = figure_count, fig.width = 16, fig.height = 10}
 tar_load(frequency_conversion)

### 2022-03-10 R/msnbase_xcms.R
adding our msnbase stuff to the mix

@@ -1,4 +1,10 @@
-msnbase_only = function(mzml_file){
+msnbase_centroid = function(mzml_file, char_list = NULL){
+
+  if (!is.null(char_list)) {
+    # get the scans from the normalization object,
+    # and then subset the profiled scans using it
+  }
+
   mzml_prof = MSnbase::readMSData(mzml_file, msLevel. = 1, centroided. = FALSE)
   mzml_info = get_ms_info(mzml_prof)
 

### 2022-03-10 _targets.R

@@ -32,7 +32,8 @@ method_data = expand_grid(
                                   "singlenorm",
                                   "intsinglenorm",
                                   "doublenorm",
-                                  "filtersd")),
+                                  "filtersd",
+                       "msnbase_centroid")),
   data_names = paste0("data_", data_mzml$name)) %>%
   dplyr::mutate(name = gsub("data_", "", data_names),
                 data_sym = rlang::syms(data_names),

### 2022-03-15 R/graph_generation.R
handling things a bit better for RSD to include the msnbase peaks

@@ -209,7 +209,7 @@ plot_sliding_window_density = function(peak_data){
 
 plot_rsd_differences = function(rsd_values){
 
-  rsd_values$processed = forcats::fct_relevel(rsd_values$processed, "noperc_nonorm", "perc99_nonorm", "singlenorm", "singlenorm_int", "doublenorm",  "filtersd")
+  rsd_values$processed = forcats::fct_relevel(rsd_values$processed,"msnbase_only", "msnbase_pc", "noperc_nonorm", "perc99_nonorm", "singlenorm", "singlenorm_int", "doublenorm",  "filtersd")
   rsd_comparison_plot = ggplot(rsd_values, aes(x = rsd, y = processed, fill = processed)) + geom_density_ridges() +
     coord_cartesian(xlim = c(0, 1)) + theme(legend.position = "none") +
     facet_wrap(~ sample)
@@ -239,27 +239,40 @@ normalization_graph = function(normalization_combine){
   out_graphs = purrr::map(split_sample, function(norm_sample){
     norm_sample = dplyr::filter(norm_sample, !(processing %in% c("filtersd", "noperc_nonorm", "perc99_nonorm")))
 
-    norm_sample$processing = factor(norm_sample$processing, levels = c("singlenorm", "singlenorm_int", "doublenorm"), ordered = TRUE)
+    norm_sample$`Normalization Type` = factor(norm_sample$processing, levels = c("singlenorm", "singlenorm_int", "doublenorm"), ordered = TRUE)
 
     hist_plot = ggplot(norm_sample, aes(x = normalization)) + geom_histogram() +
-      facet_wrap(~ processing, ncol = 1)
+      facet_wrap(~ `Normalization Type`, ncol = 1) +
+      labs(x = "Normalization Factors", y = "Number of Scans")
 
     split_norm = split(norm_sample, norm_sample$processing)
     ref_norm = split_norm[["singlenorm"]]
     diff_norm = purrr::map_df(split_norm, function(in_norm){
       combine_norm = dplyr::left_join(ref_norm, in_norm, by = "scan", suffix = c(".ref", ".in"))
-      dplyr::mutate(combine_norm, scan = scan, processing = processing.in,
+      dplyr::mutate(combine_norm, scan = scan, `Normalization Type` = processing.in,
                     diff = normalization.in - normalization.ref) %>%
-        dplyr::select(scan, diff, processing)
[…]

### 2022-03-15 R/processing_functions.R

@@ -289,7 +289,7 @@ calc_rsd = function(in_char, processing){
                       n = n_peak,
                       n_perc = n_peak / n_scan,
                       processed = processing,
-                      sample = in_char$zip_ms$peak_finder$sample_id)
+                      sample = rename_samples( in_char$zip_ms$peak_finder$sample_id))
   rsd_df %>%
     dplyr::filter(n >= 3)
 
@@ -298,6 +298,27 @@ calc_rsd = function(in_char, processing){
 single_rsd = function(char_list){
   rsd_df = calc_rsd(char_list$char_obj, char_list$processed)
   rsd_df
+}
+
+calc_rsd_msnbase = function(msnbase_list, processing){
+  scan_peaks = 10^msnbase_list$ScanLevel$Log10Height
+  n_scan = ncol(scan_peaks)
+  mean_peaks = rowMeans(scan_peaks, na.rm = TRUE)
+  sd_peaks = apply(scan_peaks, 1, sd, na.rm = TRUE)
+  n_peak = apply(scan_peaks, 1, function(.x){sum(!is.na(.x))})
+
+  rsd_df = data.frame(mean = mean_peaks,
+                      sd = sd_peaks,
+                      rsd = sd_peaks / mean_peaks,
+                      n = n_peak,
+                      n_perc = n_peak / n_scan,
+                      processed = dplyr::case_when(
+                        grepl("pc", msnbase_list$Sample) ~ "msnbase_pc",
+                        TRUE ~ "msnbase_only"
+                      ),
+                      sample = rename_samples(msnbase_list$Sample))
+  rsd_df %>%
+    dplyr::filter(n >= 3)
 }
 
 merge_rsd = function(...) {
@@ -542,6 +563,10 @@ rename_samples = function(sample_id){
     grepl("97Cpos", sample_id) ~ "97lipid",
     grepl("49Cpos", sample_id) ~ "49lipid",
     grepl("1_ECF", sample_id) ~ "1ecf",
-    grepl("2_ECF", sample_id) ~ "2ecf")
+    grepl("2_ECF", sample_id) ~ "2ecf",
+    grepl("1ecf", sample_id) ~ "1ecf",
+    grepl("2ecf", sample_id) ~ "2ecf",
+    grepl("97lipid", sample_id) ~ "97lipid",
+    grepl("49lipid", sample_id) ~ "49lipid")
   out_id
 }

### 2022-03-15 _targets.R

@@ -5,7 +5,7 @@ source("./packages.R")
 lapply(list.files("./R", full.names = TRUE), source)
 
 ## debugging stuff
-tar_option_set(debug = "scan_height_correlation")
+#tar_option_set(debug = "scan_height_correlation")
 ## targets::tar_make(callr_function = NULL)
 
 use_files = dir("data/data_input", pattern = "mzML", full.names = TRUE)
@@ -53,12 +53,6 @@ method_tar = tar_map(
   tar_target(assign, assign_files(zip))
 )
 
-rsd_tar = tar_combine(
-  rsd_combine,
-  method_tar[[2]],
-  command = merge_rsd(!!!.x),
-  iteration = "list"
-)
 
 normalization_tar = tar_combine(
   normalization_combine,
@@ -111,8 +105,39 @@ msnbase_tar = tar_map(
   names = "name",
   tar_target(msnbase, data_function(mzml)),
   tar_target(msnbase_only, msnbase_match_scan_combined(msnbase)),
+  tar_target(rsd_msnbase_only, calc_rsd_msnbase(msnbase_only)),
   tar_target(msnbase_only_zip, msnbase_zip(msnbase_only))
 )
+
+msnbase_pc_var = tibble(
+  msnbase = rlang::syms(c("msnbase_1ecf",
+              "msnbase_2ecf",
+              "msnbase_49lipid",
+              "msnbase_97lipid")),
+  pc = rlang::syms(c("method_filtersd_1ecf",
+         "method_filtersd_2ecf",
+         "method_filtersd_49lipid",
+         "method_filtersd_97lipid")),
+  name = c("1ecf",
+           "2ecf",
+           "49lipid",
+           "97lipid")
+)
+
+msnbase_pc_tar = tar_map(
+  values = msnbase_pc_var,
+  names = "name",
+  tar_target(msnbase_pc, msnbase_match_pc(msnbase, pc)),
+  tar_target(rsd_msnbase_pc, calc_rsd_msnbase(msnbase_pc))
+)
+
+rsd_tar = tar_combine(
+  rsd_combine,
+  c(method_tar[[2]], msnbase_tar[[3]], msnbase_pc_tar[[2]]),
+  command = merge_rsd(!!!.x),
+  iteration = "list"
+)
+
 
 list(pkg_tar,
      data_tar,
@@ -121,4 +146,5 @@ list(pkg_tar,
      normalization_tar,
      figures_tar,
[…]

### 2022-03-15 doc/peakcharacterization_manuscript.Rmd
general improvements to the manuscript

@@
 title: "Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments"
-author: "Robert M Flight, Joshua M Mitchell, Hunter NB Moseley"
+author:
+  - Robert M Flight:
+      institute: [markey, biochem, rcsirm]
+  - Joshua M Mitchell:
+      institute: lab
+  - Hunter NB Moseley:
+      email: hunter.moseley@uky.edu
+      correspondence: true
+      institute: [markey, biochem, rcsirm, ibi, tox]
+institute:
+  - markey: Markey Cancer Center, University of Kentucky, Lexington, KY 40536, USA
+  - biochem: Department of Molecular & Cellular Biochemistry, University of Kentucky, Lexington, KY 40536, USA
+  - rcsirm: Resource Center for Stable Isotope Resolved Metabolomics, University of Kentucky, Lexington, KY 40536, USA
+  - ibi: Institute for Biomedical Informatics, University of Kentucky, Lexington, KY 40536, USA
+  - tox: Department of Toxicology and Cancer Biology, University of Kentucky, Lexington, KY 40536, USA
+  - lab: Somewhere over the rainbow
 date: "`r Sys.time()`"
@@
     keep_md: true
+    pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
+bibliography: '`r here::here("doc/peakcharacterization.json")`'
+csl: plos-computational-biology.csl
 editor_options: 
@@
 
-To enable interval algebra, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package [citation].
+To enable interval algebra, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package [@lawrenceSoftwareComputingAnnotating2013a].
 The sliding and tiled windows are also converted to IRanges objects the same way.
@@
   1. No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
-  2. Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
-  3. Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
-  4. Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
-  5. Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
-  6. Noise removal, two pass normalization, frequency SD filtering (filtersd)
+  1. Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
+  1. Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
+  1. Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
+  1. Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
+  1. Noise removal, two pass normalization, frequency SD filtering (filtersd)
+  1. Scan level peak centroids generated by MSnbase
+  1. Scans merged and then centroids generated by MSnbase
+
+For the two sets of MSnbase peaks, we also matched scan level peaks to merged peaks via two methods:
+
+  1. All scan level peaks within a 2ppm window are associated with the *merged* peak
+  1. Using the *filtersd* peaks, merged and scan level peaks are associated if they are within the start and stop M/Z of the *filtersd* peaks
 
@@
 The solution to **1** is to do a two pass normalization.
-In the first pass, the peaks who's intensity is correlated with intensity are detected.
+In the first pass, the peaks who's intensity is correlated with scan are detected.
 In the second pass, the correlated peaks are removed, and normalization is carried out again.
@@
 figure_count$increment("intensity_scan")
+figure_count$increment("normalization_factors")
 ```
@@
 
-```{r normalization_factors, eval = FALSE}
-tar_load(normalization_plot_normalization_values_data)
-normalization_plot_normalization_values_data
+```{r normalization_factors, dn_id = figure_count, fig.height = 8, fig.width = 8}
+tar_load(compare_normalization)
+compare_normalization[[1]]
 ```
 
-*caption*: **A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity >= 0.7 of the maximum intensity (singlenorm_int), or the double pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
+`r figure_count$label_text("normalization_factors")`. 
+**A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity >= 0.7 of the maximum intensity (singlenorm_int), or the double pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
 **B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.
@@
 tar_load(rsd_values)
-knitr::kable(rsd_values, digits = 3)
+rsd_ft = flextable(rsd_values) %>%
+  colformat_double(digits = 2) %>%
+  autofit()
+rsd_ft
 ```
@@
 If incorrect peak intensities are being reported, then we would expect between class to within class variances to be very random, resulting in F-statistics close to 1.
+
+## References

### 2022-03-15 doc/peakcharacterization_manuscript.Rmd
updated for latest version of the manuscript

@@
 
+```{r get_setup, include = FALSE}
+source(here::here("packages.R"))
+lapply(list.files(here::here("./R"), full.names = TRUE), source)
+
+figure_count = dn_counter$new("Figure ")
+```
+
 ```{r setup, include = FALSE}
@@
                       dev.args = list(png = list(type = "cairo")))
-```
-
-```{r get_setup, include = FALSE}
-source("./packages.R")
-lapply(list.files("./R", full.names = TRUE), source)
-
-figure_count = dn_counter$new("Figure ")
 ```
@@
 
-To find intervals that contain points that are not just noise, the number of non-zero intensity frequency points in each sliding window are counted.
+To find intervals that contain points that represent actual signal and not just random noise, the number of non-zero intensity points in each sliding window are counted.
 Subsequently, we break these counts into fixed width tiles (default width of 2000) and calculate the 99th percentile of non-zero points for each tile.
@@
 
+```{r increment_mz_frequency}
+figure_count$increment("mz_frequency_conversion")
+```
+
 FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data. 
 Outside of the meta-data, it merely contains the M/Z and intensity values for profile spectra acquired across multiple scans. 
-However, the frequency can be calculated by dividing the average M/Z of two adjacent points by their difference (Figure XA, red points reresenting the average M/Z of two adjacent points, red lines representing the difference between the two adjacent points).
-The subsequent differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, as shown in Figure XB and XC. 
-The Thermo Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 400K, has a mode of 0.5, as shown in Figure XC and XD. 
-Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in Figure XE.
+However, the frequency can be calculated by dividing the average M/Z of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points reresenting the average M/Z of two adjacent points, red lines representing the difference between the two adjacent points).
+The subsequent differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, as shown in `r figure_count$label_text("mz_frequency_conversion")`B and C.
+The Thermo Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 400K, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C and D.
+Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in `r figure_count$label_text("mz_frequency_conversion")`E.
 This model seems to fit the known relationship between frequency and M/Z, where the frequency is related to $1/\sqrt{mz}$. 
@@
 Frequency based points suffer none of these drawbacks, and the conversion from M/Z can be derived from the raw profile level data itself, which is incredibly useful.
-
-```{r increment_mz_frequency}
-figure_count$increment("mz_frequency_conversion")
-```
 
@@
 
-M/Z to frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
-While scan specific models **could** be used in the conversion of M/Z to frequency, doing so results in changes to the relative peak ordering compared to M/Z space, as shown in Figure XA. 
-Therefore, a single model for all scans based on the median of each term in the model across scans is used for converting *all* remaining scan level data.
-
 ```{r increment_peak_ordering}
@@
 ```
+
+M/Z to frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
+While scan specific models **could** be used in the conversion of M/Z to frequency, doing so results in changes to the relative peak ordering compared to M/Z space, as shown in `r figure_count$label_text("peak_ordering")`.
+Therefore, a single model for all scans based on the median of each term in the model across scans is used for converting *all* remaining scan level data.
 
@@
 
+```{r increment_slidingwindow}
+figure_count$increment("slidingwindow_count")
+tar_load(nonoise_vs_noise)
+nn_n_ratio = english::as.english(round(nonoise_vs_noise["noperc"] / nonoise_vs_noise["perc99"]))
+```
+
 In a dataset of this nature, we expect that much of the data is really just noise, and doesn't contribute that much to the analysis. 
@@
 Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
-The distribution of counts in sliding windows ten points wide and offset by one point is shown in Figure X. 
-The number of peaks detected without applying a cutoff explodes by ..., with the vast majority of the peaks being present in a single scan. 
-By discarding those windows that have a non-zero point density below the 99th percentile, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
-
-```{r increment_slidingwindow}
-figure_count$increment("slidingwindow_count")
-```
+Subsequently, we divide the counts into tiled regions of a set size (2000 in this work), and examine the 99th percentile of the sliding window counts within the tiled region.
+A histogram of sliding window counts in a single tiled region is shown in `r figure_count$label_text("slidingwindow_count")`A.
+The distribution of 99th percentiles across all of the tiled regions for a single sample is shown in `r figure_count$label_text("slidingwindow_count")`B.
+We can use the median (50th percentile) value of all the 99th percentiles from all of the tiled regions to determine a cutoff value above which we think any non-zero points are potentially **signal** and not just noise.
+In the example below, the median of the tiled windows is 5, resulting in a cutoff of 7.5.
+The number of peaks detected without applying a cutoff explodes `r nn_n_ratio`-fold (from `r format(nonoise_vs_noise["perc99"], big.mark = ",")` to `r format(nonoise_vs_noise["noperc"], big.mark = ",")` in the 97lipid sample), with a vast majority of the peaks being present in a single scan.
+By discarding those windows that have a non-zero point density below 1.5x the median of 99th percentiles, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
 

### 2022-03-15 _targets.R

@@ -89,6 +89,13 @@ tables_tar = tar_plan(
   tar_target(rsd_values, summarize_rsd(rsd_combine))
 )
 
+other_tar = tar_plan(
+  tar_target(
+    nonoise_vs_noise,
+    compare_noise_cutoff(data_97lipid)
+  )
+)
+
 msnbase_mzml = expand_grid(
   data_function = rlang::syms("msnbase_centroid"),
   mzml = use_files
@@ -147,4 +154,5 @@ list(pkg_tar,
      figures_tar,
      tables_tar,
      msnbase_tar,
-     msnbase_pc_tar)
+     msnbase_pc_tar,
+     other_tar)

### 2022-03-15 packages.R

@@ -26,4 +26,5 @@ library(documentNumbering)
 library(flextable)
 library(ggforce)
 library(MSnbase)
+conflict_prefer("combine", "dplyr")
 tar_option_set(memory = "transient")

### 2022-03-16 R/processing_functions.R
doing hpd stuff

@@ -298,7 +298,7 @@ single_rsd = function(char_list){
   rsd_df
 }
 
-calc_rsd_msnbase = function(msnbase_list, processing){
+calc_rsd_msnbase = function(msnbase_list){
   scan_peaks = 10^msnbase_list$ScanLevel$Log10Height
   n_scan = ncol(scan_peaks)
   mean_peaks = rowMeans(scan_peaks, na.rm = TRUE)
@@ -498,7 +498,7 @@ density_calculate = function(frequency_values, metadata, frequency_range = NULL)
   return(list(hpd_points = hpd_points, freq_points = freq_points, hpd_regions = keep_merged, stats = tmp_df))
 }
 
-hpds_from_excel = function(in_data){
+hpds_from_excel = function(in_data, excel_files){
   sample = in_data$char_obj$zip_ms$id
   match_excel = grep(sample, excel_files, value = TRUE)
 
@@ -506,11 +506,16 @@ hpds_from_excel = function(in_data){
   xl_data = xl_data[, 1:2]
   names(xl_data) = c("mz", "intensity")
   message("got xl data")
-  frequency_coefficients = in_data$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions@metadata$frequency_coefficients
-  frequency_description = in_data$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions@metadata$frequency_fit_description
-  mz_coefficients = in_data$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions@metadata$mz_coefficients
-  mz_description = in_data$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions@metadata$mz_fit_description
-  org_frequency = as.data.frame(in_data$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions@elementMetadata)
+  frequency_coefficients = in_data$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$metadata$frequency_coefficients
+  frequency_description = in_data$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$metadata$frequency_fit_description
+  mz_coefficients = in_data$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$metadata$mz_coefficients
+  mz_description = in_data$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$metadata$mz_fit_description
[…]

### 2022-03-16 _targets.R

@@ -9,7 +9,7 @@ lapply(list.files("./R", full.names = TRUE), source)
 ## targets::tar_make(callr_function = NULL)
 
 use_files = dir("data/data_input", pattern = "mzML", full.names = TRUE)
-excel_files = dir("data/data_input", pattern = ".xlsx", full.names = TRUE)
+excel_files = dir("data/data_input", pattern = "[ECF|pos].xlsx", full.names = TRUE)
 
 pkg_sha = as.character(utils::packageDescription("FTMS.peakCharacterization")$GithubSHA1)
 
@@ -147,6 +147,20 @@ rsd_tar = tar_combine(
   iteration = "list"
 )
 
+hpd_df = tibble(
+  methods = rlang::syms(c("method_filtersd_1ecf",
+              "method_filtersd_2ecf",
+              "method_filtersd_49lipid",
+              "method_filtersd_97lipid")),
+) %>%
+  dplyr::mutate(names = rename_samples(methods))
+
+hpd_tar = tar_map(
+  values = hpd_df,
+  names = "names",
+  tar_target(hpd, hpds_from_excel(methods, excel_files))
+)
+
 
 list(pkg_tar,
      data_tar,
@@ -157,4 +171,5 @@ list(pkg_tar,
      tables_tar,
      msnbase_tar,
      msnbase_pc_tar,
+     hpd_tar,
      other_tar)

### 2022-03-16 doc/peakcharacterization_manuscript.Rmd
incorporating Hunters edits

@@
 ---
-title: "Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments"
+title: "Scan-Centric, Frequency-Based Method for Characterizing Peaks from Direct Injection FT-MS Experiments"
 author:
@@
 
-We introduce a novel method for characterizing peaks from direct injection FT-MS experiments developed on metabolomics datasets that utilizes frequency values derived directly from the spacing of raw M/Z points.
+We introduce a novel method for characterizing peaks from direct injection FT-MS experiments developed on metabolomics datasets that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
 Our method incorporates several aspects designed to make the derived intensities trustworthy, and suitable for use in our previously published peak assignment algorithm SMIRFE. 
@@
 
-### Conversion of M/Z to Frequency
+### Conversion of m/z to Frequency
 
-The data consists of profile mode M/Z spectra from multiple scans encoded as M/Z and intensity values for each scan. 
+The data consists of profile mode m/z spectra from multiple scans encoded as m/z and intensity values for each scan. 
 There is no information about the original frequency values available in either the `raw` files or the `mzML` files. 
-However, the frequency values can be approximated by averaging the M/Z of adjacent points, and dividing the M/Z by the difference. 
+However, the frequency values can be approximated by averaging the m/z of adjacent points, and dividing the m/z by the difference. 
 Ideally, the difference between subsequent points in frequency space is constant, in practice there is a range of differences in frequency space.
-The *true, constant* difference can be obtained by examining the median of the mean frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to M/Z) to be within 2% of the mode value.
-These *useful* points can be used to construct a linear model relating M/Z to frequency for individual scans based on the formula:
+The *true, constant* difference can be obtained by examining the median of the mean frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to m/z) to be within 2% of the mode value.
+These *useful* points can be used to construct a linear model relating m/z to frequency for individual scans based on the formula:
 
 $$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
-From the known physical properties of the orbitrap, only the square root term should be necessary.
-Practically, we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry and control of the magnetic fields.
-A frequency model is generated for each scan, and then a single model using the median values of each term across scans is generated, and applied to convert the M/Z values to frequency for all scans. 
-A single model rather than scan specific models is used for conversion because this was found to maintain the relative ordering of the peaks in M/Z and frequency space (see Results).
+From the known physical properties of the Orbitrap, only the square root term should be necessary.
+Practically, we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
+A frequency model is generated for each scan, and then a single model using the scan with the square-root slope closest to the median of the square-root slopes from all scans.
+The single model is used to convert the m/z values to frequency for all scans.
+A single model rather than scan specific models is used for conversion because this was found to maintain the relative ordering of the peaks in m/z and frequency space (see Results).
 
-To convert M/Z back into frequency, we can use a similar model without the roots.
+To convert m/z back into frequency, we can use a similar model without the roots.
 
@@
 Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged together to create the initial peak regions.
+The presence of zero intensity points are due to flooring implemented by the spectrometer when local spectral quality falls below a certain threshold.
+When this flooring is unstable and incomplete, a partial ringing phenomenom is observed [@mitchellNewMethodsIdentify2018].
+**aside: I'm not sure how this actually relates to the density noise cutoff**.
 
-**Hunter talk about issues around non-detection and why we only use non-zero counts, sliding followed by tiles**.
-
-Within each initial interval region, peaks in each scan are detected (see **Peak Detection**), and their centers binned by the tiled windows.
+Within each initial interval region, peaks in each scan are detected (see **Peak Detection**), and their centers are binned by the tiled windows.
 Adjacent tiled windows with non-zero peak counts are merged together, and any zero peak count tiled windows split the initial region into multiple peak interval regions.
@@
 
-On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package [citation].
-These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either M/Z or frequency), and the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.
+On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package [@borchersPracmaPracticalNumerical2021].
+These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either m/z or frequency), and the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.
 
 $$\ln{intensity} = intercept + a*position+b*position^2$$
-From this weighted parabolic fit the center, intensity, integrated area and sum-of-square residuals can be extracted for the peak. 
+From this weighted parabolic fit, the center, intensity, integrated area and sum-of-square residuals can be extracted for the peak. 
 These characteristics are equivalent to the centroided peak center and intensity.
 
-Before further processing, the regions are checked that there is only one peak from each scan. If a scan has two or more peaks, then the scan level data in that region is discarded. 
+Before further processing, the regions are verified to have only one peak from each scan. 
+If a scan has two or more peaks, then the scan level data in that region is discarded. 
 Any regions that subsequently contain zero peaks are removed.
@@
 The scan with the lowest summed overall distance is chosen as the *reference* scan, and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the reference scan.
-This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan number is checked, and those peaks with correlation of greater than 0.5 with scan number are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.
+This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan order is checked, and those peaks with correlation of greater than 0.5 with scan order are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.
+Peaks correlated to scan order represent an artifact that we speculate results from a gradient in the sample well.
 
@@
 
-In addition to these values, the frequency point-to-point median difference is calculated across all of the raw data points for those points that could be used for modeling frequency to M/Z, and this difference of a single point from the peak center is calculated in frequency space, and converted to M/Z space to provide an "offset" value that is useful to define the search space around the peak for any assignment algorithm.
+In addition to these values, the frequency point-to-point median difference is calculated across all of the raw data points for those points that could be used for modeling frequency to m/z, and this difference of a single point from the peak center is calculated in frequency space, and converted to m/z space to provide an "offset" value that is useful to define the search space around the peak for any assignment algorithm.
 
@@
 However, in some scans, some peaks are not likely to be observed due to the number of ions falling below the detection threshold. 
-This results in a truncated distribution of peak intensities across scans. 
+This results in a left-censored normal distribution of peak intensities across scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
 To correct these, either a correction based on a model of the truncated normal distribution can be used, or the differences can be simulated by sampling from data that is present in most of the scans.
@@
 For each peak, a sample of the heights across scans are taken (ranging from 5% to 95% of scans), and a new standard deviation calculated for that fraction, and a ratio of the fractional standard deviation to the "true" standard deviation calculated.
-The ratio sd across peaks can then be fitted to a cubic model of the fraction used, and a correction factor predicted for those peaks that are present in fewer scans. 
-The corrected sd's can then be used to correct the mean height assuming that it is the result of a truncated normal distribution [wikipedia reference, https://en.wikipedia.org/wiki/Truncated_normal_distribution].
+The ratio standard deviation across peaks can then be fitted to a cubic model of the fraction used, and a correction factor predicted for those peaks that are present in fewer scans. 
+The corrected standard deviations can then be used to correct the mean height assuming that it is the result of a left-censored normal distribution [@TruncatedNormalDistribution2022].
 
@@
 The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
-For HPD detection, the peaks in excel output from Xcalibur were used after converting the M/Z peak centers into frequency space.
-Sliding windows 1000 frequency points wide and offset by 100 points were used for the density calculations.
+For HPD detection, the peaks in excel output from Xcalibur were used after converting the m/z peak centers into frequency space. 
+Sliding windows that are 1000 frequency points wide and offset by 100 points were used for the density calculations.
 
@@
   1. All scan level peaks within a 2ppm window are associated with the *merged* peak
-  1. Using the *filtersd* peaks, merged and scan level peaks are associated if they are within the start and stop M/Z of the *filtersd* peaks
+  1. Using the *filtersd* peaks, merged and scan level peaks are associated if they are within the start and stop m/z of the *filtersd* peaks
 
@@
 
-### M/Z to Frequency
+### m/z to Frequency
 
@@
 FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data. 
-Outside of the meta-data, it merely contains the M/Z and intensity values for profile spectra acquired across multiple scans. 
-However, the frequency can be calculated by dividing the average M/Z of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points reresenting the average M/Z of two adjacent points, red lines representing the difference between the two adjacent points).
-The subsequent differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, as shown in `r figure_count$label_text("mz_frequency_conversion")`B and C.
+Outside of the meta-data, it merely contains the m/z and intensity values for profile spectra acquired across multiple scans. 
+However, the frequency can be calculated by dividing the average m/z of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points representing the average m/z of two adjacent points, red lines representing the difference between the two adjacent points).
+The subsequent differences in frequency should be relatively constant with respect to m/z, in contrast to the differences of adjacent m/z points, as shown in `r figure_count$label_text("mz_frequency_conversion")`B and C.
 The Thermo Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 400K, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C and D.
-Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in `r figure_count$label_text("mz_frequency_conversion")`E.
-This model seems to fit the known relationship between frequency and M/Z, where the frequency is related to $1/\sqrt{mz}$. 
+Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`E.
+This model seems to fit the known relationship between frequency and m/z, where the frequency is related to $1/\sqrt{mz}$. 
 This is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points. 
-M/Z point-to-point differences are not constant, but can be approximated by a lasso linear model; however, it is very difficult to create a lasso model with an intercept of 0.
-In addition, we would also need to vary the width of sliding windows according to the M/Z difference at a particular M/Z based on the lasso model.
-Frequency based points suffer none of these drawbacks, and the conversion from M/Z can be derived from the raw profile level data itself, which is incredibly useful.
+The m/z point-to-point differences are not constant, but can be approximated by a lasso linear model; however, it is very difficult to create a lasso model with an intercept of 0.
+In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the lasso model.
+Frequency based points suffer none of these drawbacks, and the conversion from m/z can be derived from the raw profile level data itself, which is incredibly useful.
 
@@
 
-`r figure_count$label_text("mz_frequency_conversion")`. **A**: Intensity vs M/Z for a single peak from a single scan. 
+`r figure_count$label_text("mz_frequency_conversion")`. **A**: Intensity vs m/z for a single peak from a single scan. 
 Red lines denote the differences between each point, and red dots the average between the pair of points. 
@@
 The red lines denote pairwise differences, which are shown in **C** for this single peak. 
-The differences for all peaks across all scans vs M/Z are shown in **D**, with those differences that lie within 0.49 - 0.51 shown in red. 
-**E** shows the plot of derived frequency *vs* M/Z, with fitted values from the linear model in red.
+The differences for all peaks across all scans vs m/z are shown in **D**, with those differences that lie within 0.49 - 0.51 shown in red. 
+**E** shows the plot of derived frequency *vs* m/z, with fitted values from the linear model in red.
 
@@
 
-M/Z to frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
-While scan specific models **could** be used in the conversion of M/Z to frequency, doing so results in changes to the relative peak ordering compared to M/Z space, as shown in `r figure_count$label_text("peak_ordering")`.
-Therefore, a single model for all scans based on the median of each term in the model across scans is used for converting *all* remaining scan level data.
+The m/z to frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
+While scan specific models **could** be used in the conversion of m/z to frequency, doing so results in changes to the relative peak ordering compared to m/z space, as shown in `r figure_count$label_text("peak_ordering")`.
+Therefore, a single model for all scans based on the scan with the slope closest to the median of slopes across scans is used for converting *all* remaining scan level data.
 
@@
 `r figure_count$label_text("peak_ordering")`.
-Peak ordering in M/Z compared with ordering in frequency space when a single M/Z to frequency model is used or scan specific M/Z to frequency models are used.
-For a single peak, the scan level peak M/Z's were extracted, and then frequency values for those M/Z generated using a single common model of M/Z to frequency (*single_order*), or models derived from each scan (*scan_order*).
+Peak ordering in m/z compared with ordering in frequency space when a single m/z to frequency model is used or scan specific m/z to frequency models are used.
+For a single peak, the scan level peak m/z's were extracted, and then frequency values for those m/z generated using a single common model of m/z to frequency (*single_order*), or models derived from each scan (*scan_order*).
 A subset of the peaks end up out of order using scan specific models, implying that a single model should be used across all the scan level data.
 
-Although the original model is created from only those points that had frequency point-to-point differences within a narrow range, **all** M/Z points are converted to frequency for subsequent steps in the workflow.
+Although the original model is created from only those points that had frequency point-to-point differences within a narrow range, **all** m/z points are converted to frequency for subsequent steps in the workflow.
 
@@
 
-For each region initially created, the peaks in each scan within that region can be characterized using a quadratic fit of log(intensity) to M/Z.
+For each region initially created, the peaks in each scan within that region can be characterized using a quadratic fit of log(intensity) to m/z.
 
 ### Peak Characterization Using Quadratic Fit
-
-Although many other types of mass-spectrometry data suffer from a variable, noisy baseline, the scan-level profile data from the Thermo Fusion has a baseline of 0 due to manipulations in the Thermo firmware, making the determination of the centroided values considerably easier.
-For the peak characterization (centroiding), we use a simple quadratic model of log(intensity) to M/Z. 
-We add a small constant to enable using the zero values, and weight the values by their ratio to the most intense value, which is normally the value closest to the center of peak, helping to ensure that the **true** centroid is determined.
-From the fitted model, we can derive the centroided center and the intensity of the peak, as shown in Figure X.
 
@@
 ```
+
+Although many other types of mass-spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo Fusion has a baseline of 0 due to manipulations in the Thermo firmware, making the determination of the centroided values considerably easier.
+For the peak characterization (centroiding), we use a simple quadratic model of log(intensity) to m/z. 
+We add a small constant to enable using the zero values, and weight the values by their ratio to the most intense value, which is normally the value closest to the center of peak, helping to ensure that the **true** centroid is determined.
+From the fitted model, we can derive the centroided center and the intensity of the peak, as shown in `r figure_count$label_text("centroided_peaks")`.
 
@@
 ```
+
+`r figure_count$label_text("centroided_peaks")`.
+Log-intensity (A) and Intensity (B) of points for a single peak against frequency. Black points are the original data points, and the blue point represents the calculated centroid.
 
@@
 `r figure_count$label_text("breaking_regions")`. 
-Splitting a single region into two regions based on the present peaks.
+Splitting a single region into two regions based on the peaks that are present.
 **A**: The full set of raw frequency and intensity data across all scans for the region are shown. 
@@
 
-Due to differences in how many ions are captured in the trap, the peak intensities in each scan vary slightly.
-Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
-However, there are two issues with these peak intensities across scans: 1 - Some peak intensities correlate with the scan number; 2 - Some peak differences between scans are correlated with intensity.
-The solution to **1** is to do a two pass normalization.
-In the first pass, the peaks who's intensity is correlated with scan are detected.
-In the second pass, the correlated peaks are removed, and normalization is carried out again.
-Figure XA contains an example peak whose intensity across scans is correlated with scan number.
-The solution to **2** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in Figure XB.
-The highlighted peaks in Figure XB are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
-If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in Figure XA and XB.
-
 ```{r increment_intensity_scan}
@@
 
+Due to differences in how many ions are captured in the trap, the peak intensities in each scan vary slightly (see `r figure_count$label_text("breaking_regions")`A and B for example).
+Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
+However, there are two issues with these peak intensities across scans: 1) some peak intensities correlate with the scan order; and 2) some peak differences between scans are correlated with intensity.
+The solution to **1** is to do a two pass normalization.
+After the first pass, the peaks who's intensity is correlated with scan order are detected.
+In the second pass, the correlated peaks are removed, and normalization is carried out again.
+`r figure_count$label_text("intensity_scan")`A contains an example peak whose intensity across scans is correlated with scan number.
+The solution to **2** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
+The highlighted peaks in `r figure_count$label_text("intensity_scan")`B are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
+If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and B.
+
 ```{r intensity_scan, dn_id = figure_count}
-# figure showing peak intensity correlation with M/Z, intensity varying with M/Z
+# figure showing peak intensity correlation with m/z, intensity varying with m/z
 tar_load(intensity_scan_cor_plot)
@@
 
-We have previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [Mitchell et al., Metabolomics, 2018]. 
+We have previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [@mitchellNewMethodsIdentify2018]. 
 Ideally, the peak characterization procedure should reduce their presence in the resultant reported peaks. 
@@
 These peaks can be detected by simply examining the distribution of frequency SDs and removing those that are outliers.
-To verify the removal of HPD regions, we converted centroided M/Z's from XCalibur to frequency values using the previously calculated values for that sample.
+To verify the removal of HPD regions, we converted centroided m/z's from XCalibur to frequency values using the previously calculated values for that sample.
 Peak density was measured using a sliding window 10 points wide and offset by one point.
@@
 However, the FSD does not depend on calculating a sliding window based density, but is merely a product of those peaks across multiple scans that span a wider range frequency range than expected based on the peaks in the data.
-We also note that we do expect higher M/Z standard deviations at higher M/Z, but the frequency based method does not suffer the same effect, due to the offsets of frequency points being constant.
+We also note that we do expect higher m/z standard deviations at higher m/z, but the frequency based method does not suffer the same effect, due to the offsets of frequency points being constant.
 
@@
 tar_load(rsd_values)
+rsd_values = rsd_values %>%
+  dplyr::arrange(sample, processed)
 rsd_ft = flextable(rsd_values) %>%

### 2022-03-16 R/msnbase_xcms.R
fixing a little issue with msnbase stuff

@@ -165,7 +165,7 @@ msnbase_match_pc = function(msnbase_data, char_list){
 
   keep_peaks = n_scan_peak > 1
   scan_mz = scan_mz[keep_peaks, ]
-  scan_intensity = scan_intensity[keep_peaks]
+  scan_intensity = scan_intensity[keep_peaks, ]
 
   cent_data = purrr::map_df(matched_peak_list, ~ .x$peaks)
   cent_data = cent_data[keep_peaks, ]

### 2022-03-18 doc/peakcharacterization_manuscript.Rmd
putting all of the chi-square test results together and then within the manuscript

@@
 figure_count = dn_counter$new("Figure ")
+table_count = dn_counter$new("Table ")
 ```
@@
 
+## Keywords
+
 ## Introduction
-
-## Methods
-
-### Conversion of m/z to Frequency
-
-The data consists of profile mode m/z spectra from multiple scans encoded as m/z and intensity values for each scan. 
-There is no information about the original frequency values available in either the `raw` files or the `mzML` files. 
-However, the frequency values can be approximated by averaging the m/z of adjacent points, and dividing the m/z by the difference. 
-Ideally, the difference between subsequent points in frequency space is constant, in practice there is a range of differences in frequency space.
-The *true, constant* difference can be obtained by examining the median of the mean frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to m/z) to be within 2% of the mode value.
-These *useful* points can be used to construct a linear model relating m/z to frequency for individual scans based on the formula:
-
-$$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
-From the known physical properties of the Orbitrap, only the square root term should be necessary.
-Practically, we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
-A frequency model is generated for each scan, and then a single model using the scan with the square-root slope closest to the median of the square-root slopes from all scans.
-The single model is used to convert the m/z values to frequency for all scans.
-A single model rather than scan specific models is used for conversion because this was found to maintain the relative ordering of the peaks in m/z and frequency space (see Results).
-
-To convert m/z back into frequency, we can use a similar model without the roots.
-
-$$mz = intercept + x* \frac{1}{frequency} + y*\frac{1}{frequency^2} + z*\frac{1}{frequency^3}$$
-
-### Frequency Intervals
-
-Two types of frequency intervals are used, sliding and tiled windows. 
-In this work, the sliding windows are 10 points wide and offset by one point.
-The tiled windows are one point wide and offset by one point.
-
-### Interval Range Based Data
-
-To enable interval algebra, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package [@lawrenceSoftwareComputingAnnotating2013a].
-The sliding and tiled windows are also converted to IRanges objects the same way.
-
-### Peak Containing Intervals
-
-To find intervals that contain points that represent actual signal and not just random noise, the number of non-zero intensity points in each sliding window are counted.
-Subsequently, we break these counts into fixed width tiles (default width of 2000) and calculate the 99th percentile of non-zero points for each tile.
-The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions that contain potential signal.
-Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged together to create the initial peak regions.
-The presence of zero intensity points are due to flooring implemented by the spectrometer when local spectral quality falls below a certain threshold.
-When this flooring is unstable and incomplete, a partial ringing phenomenom is observed [@mitchellNewMethodsIdentify2018].
-**aside: I'm not sure how this actually relates to the density noise cutoff**.
-
-Within each initial interval region, peaks in each scan are detected (see **Peak Detection**), and their centers are binned by the tiled windows.
-Adjacent tiled windows with non-zero peak counts are merged together, and any zero peak count tiled windows split the initial region into multiple peak interval regions.
-These interval regions should contain a single **real** peak that was detected in one or more scans.
-
-### Peak Detection and Centroided Values
-
-On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package [@borchersPracmaPracticalNumerical2021].
-These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either m/z or frequency), and the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.
-
-$$\ln{intensity} = intercept + a*position+b*position^2$$
-From this weighted parabolic fit, the center, intensity, integrated area and sum-of-square residuals can be extracted for the peak. 
-These characteristics are equivalent to the centroided peak center and intensity.
-
-Before further processing, the regions are verified to have only one peak from each scan. 
-If a scan has two or more peaks, then the scan level data in that region is discarded. 
-Any regions that subsequently contain zero peaks are removed.
-
-### Scan to Scan Normalization
-
-Scans are normalized to a single *reference* scan based on the log-intensity differences of a subset of peaks that are present in at least the same number of scans as the 95th percentile of scan counts for the peaks.
-In addition, only those peaks that have an intensity greater than 0.7 times the highest intensity peak in the scan are used. 
-Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each scan to all other scans.
-The scan with the lowest summed overall distance is chosen as the *reference* scan, and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the reference scan.
-This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan order is checked, and those peaks with correlation of greater than 0.5 with scan order are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.
-Peaks correlated to scan order represent an artifact that we speculate results from a gradient in the sample well.
-
-### Full Characterization
-
-Based on the previously detected peaks, the full set of raw data points for each peak in each scan within a region is known. 
-The non-zero intensity, normalized raw data points across scans can be combined, and then characterized again using the weighted parabolic fit previously described. 
-In addition to the data from the full set of raw points, means and standard deviations of the peak height and location can be derived from the scan level peak characteristics previously calculated.
-
-In addition to these values, the frequency point-to-point median difference is calculated across all of the raw data points for those points that could be used for modeling frequency to m/z, and this difference of a single point from the peak center is calculated in frequency space, and converted to m/z space to provide an "offset" value that is useful to define the search space around the peak for any assignment algorithm.
-
-### Correction of Height and Standard Deviation
-
-Ideally, each peak would be observed in every scan. 
-However, in some scans, some peaks are not likely to be observed due to the number of ions falling below the detection threshold. 
-This results in a left-censored normal distribution of peak intensities across scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
-To correct these, either a correction based on a model of the truncated normal distribution can be used, or the differences can be simulated by sampling from data that is present in most of the scans.
-To simulate the effect of peaks missing from some scans on the standard deviations, the peaks present in all scans are used. 
-For each peak, a sample of the heights across scans are taken (ranging from 5% to 95% of scans), and a new standard deviation calculated for that fraction, and a ratio of the fractional standard deviation to the "true" standard deviation calculated.
-The ratio standard deviation across peaks can then be fitted to a cubic model of the fraction used, and a correction factor predicted for those peaks that are present in fewer scans. 
-The corrected standard deviations can then be used to correct the mean height assuming that it is the result of a left-censored normal distribution [@TruncatedNormalDistribution2022].
-
-### Frequency Standard Deviation Filtering
-
-High-peak-density (HPD) artifacts [Mitchell et al., 2018] present as singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
-Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and those FSD that are greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are removed.
-The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
-For HPD detection, the peaks in excel output from Xcalibur were used after converting the m/z peak centers into frequency space. 
-Sliding windows that are 1000 frequency points wide and offset by 100 points were used for the density calculations.
-
-### Multi-Class F-Statistics
-
-To evaluate the F-statistic between multi-class samples, a much larger set of samples was used.
-In this case, 181 matched non-cancer and cancer samples that were previously used in the HPD detection manuscript [ref].
-All 181 samples were peak characterized using the most stringent method below as well as peaks generated from the xcms R package [ref], assigned using SMIRFE, and then peaks matched by shared EMFs across samples (see Supplemental materials for a description of peak matching and deciding the most likely EMF).
-For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using our peak-characterization and those from xcms.
-
-### Samples and Overall Processing
-
-Samples included two ethyl-chloroformate (ECF) derivatized amino-acid samples [reference] and two samples where lipid extracts were extracted from lung tumors [reference]. 
-All four samples are positive mode, were converted from raw to mzML using msconvert from ProteoWizard (profile mode) and ThermoRawFileParser (centroided).
-
-For each raw data file in profile mzML format, they were processed in these six ways:
-
-  1. No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
-  1. Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
-  1. Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
-  1. Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
-  1. Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
-  1. Noise removal, two pass normalization, frequency SD filtering (filtersd)
-  1. Scan level peak centroids generated by MSnbase
-  1. Scans merged and then centroids generated by MSnbase
-
-For the two sets of MSnbase peaks, we also matched scan level peaks to merged peaks via two methods:
-
-  1. All scan level peaks within a 2ppm window are associated with the *merged* peak
-  1. Using the *filtersd* peaks, merged and scan level peaks are associated if they are within the start and stop m/z of the *filtersd* peaks
 
@@
 
+```{r increment_hpd}
+figure_count$increment("hpd_fsd")
+```
+
 We have previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [@mitchellNewMethodsIdentify2018]. 
@@
 
-Figure X compares the HPD regions detected using the method from Mitchell et al., and their correspondence with high FSD peaks for one sample (the other comparisons are in Supplemental materials).
+`r figure_count$label_text("hpd_fsd")` compares the HPD regions detected using the method from Mitchell et al., and their correspondence with high FSD peaks for one sample (the other comparisons are in Supplemental materials).
+`r table_count$label_text("hpd_fsd")` summarizes the results of Chi-square tests using contingency tables of HPD and high FSD peaks.
 Based on this comparison, the high FSD peaks are indeed HPD sites and should be removed before assignment.
@@
 We also note that we do expect higher m/z standard deviations at higher m/z, but the frequency based method does not suffer the same effect, due to the offsets of frequency points being constant.
+
+```{r hpd_fsd, dn_id = figure_count}
+tar_load(hpd_plots_49lipid)
+hpd_plots_49lipid
+```
+
+`r figure_count$label_text("hpd_fsn")`.
+Comparison of HPD and high FSD sites in the lipid 49 sample.
+Black are the original fully characterized peaks.
+Red indicates peaks within HPD sites detected using the XCalibur peak sets.
+Blue indicates peaks that had abnormally high FSD values.
+
+`r table_count$label_text("hpd_fsn")`.
+Chi-square statistics from contingency tables of shared presence and absence of peaks in HPD sites and labeled as high FSD.
+
+```{r hpd_fsd_table}
+tar_load(hpd_chisq_all)
+sci_formatter = 
+hpd_chisq_table = hpd_chisq_all %>%
+  dplyr::select(sample, p.value, statistic) %>%
+  flextable() %>%
+  set_formatter(p.value = function(x){
+    formatC(x, format = "e", digits = 2)
+  }) %>%
+  colformat_double(digits = 1, j = 3) %>%
+  autofit()
+```
 
@@
 
+## Discussion
+
+## Materials and Methods
+
+### Conversion of m/z to Frequency
+
+The data consists of profile mode m/z spectra from multiple scans encoded as m/z and intensity values for each scan. 
+There is no information about the original frequency values available in either the `raw` files or the `mzML` files. 
+However, the frequency values can be approximated by averaging the m/z of adjacent points, and dividing the m/z by the difference. 
+Ideally, the difference between subsequent points in frequency space is constant, in practice there is a range of differences in frequency space.
+The *true, constant* difference can be obtained by examining the median of the mean frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to m/z) to be within 2% of the mode value.
+These *useful* points can be used to construct a linear model relating m/z to frequency for individual scans based on the formula:
+
+$$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
+From the known physical properties of the Orbitrap, only the square root term should be necessary.
+Practically, we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
+A frequency model is generated for each scan, and then a single model using the scan with the square-root slope closest to the median of the square-root slopes from all scans.
+The single model is used to convert the m/z values to frequency for all scans.
+A single model rather than scan specific models is used for conversion because this was found to maintain the relative ordering of the peaks in m/z and frequency space (see Results).
+
+To convert m/z back into frequency, we can use a similar model without the roots.
+
+$$mz = intercept + x* \frac{1}{frequency} + y*\frac{1}{frequency^2} + z*\frac{1}{frequency^3}$$
+
+### Frequency Intervals
+
+Two types of frequency intervals are used, sliding and tiled windows. 
+In this work, the sliding windows are 10 points wide and offset by one point.
+The tiled windows are one point wide and offset by one point.
+
+### Interval Range Based Data
+
+To enable interval algebra, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package [@lawrenceSoftwareComputingAnnotating2013a].
+The sliding and tiled windows are also converted to IRanges objects the same way.
+
+### Peak Containing Intervals
+
+To find intervals that contain points that represent actual signal and not just random noise, the number of non-zero intensity points in each sliding window are counted.
+Subsequently, we break these counts into fixed width tiles (default width of 2000) and calculate the 99th percentile of non-zero points for each tile.
+The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions that contain potential signal.
+Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged together to create the initial peak regions.
+The presence of zero intensity points are due to flooring implemented by the spectrometer when local spectral quality falls below a certain threshold.
+When this flooring is unstable and incomplete, a partial ringing phenomenom is observed [@mitchellNewMethodsIdentify2018].
+**aside: I'm not sure how this actually relates to the density noise cutoff**.
+
+Within each initial interval region, peaks in each scan are detected (see **Peak Detection**), and their centers are binned by the tiled windows.
+Adjacent tiled windows with non-zero peak counts are merged together, and any zero peak count tiled windows split the initial region into multiple peak interval regions.
+These interval regions should contain a single **real** peak that was detected in one or more scans.
+
+### Peak Detection and Centroided Values
+
+On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package [@borchersPracmaPracticalNumerical2021].
+These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either m/z or frequency), and the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.
+
+$$\ln{intensity} = intercept + a*position+b*position^2$$
+From this weighted parabolic fit, the center, intensity, integrated area and sum-of-square residuals can be extracted for the peak. 
+These characteristics are equivalent to the centroided peak center and intensity.
+
+Before further processing, the regions are verified to have only one peak from each scan. 
+If a scan has two or more peaks, then the scan level data in that region is discarded. 
+Any regions that subsequently contain zero peaks are removed.
+
+### Scan to Scan Normalization
+
+Scans are normalized to a single *reference* scan based on the log-intensity differences of a subset of peaks that are present in at least the same number of scans as the 95th percentile of scan counts for the peaks.
+In addition, only those peaks that have an intensity greater than 0.7 times the highest intensity peak in the scan are used. 
+Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each scan to all other scans.
+The scan with the lowest summed overall distance is chosen as the *reference* scan, and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the reference scan.
+This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan order is checked, and those peaks with correlation of greater than 0.5 with scan order are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.
+Peaks correlated to scan order represent an artifact that we speculate results from a gradient in the sample well.
+
+### Full Characterization
+
+Based on the previously detected peaks, the full set of raw data points for each peak in each scan within a region is known. 
+The non-zero intensity, normalized raw data points across scans can be combined, and then characterized again using the weighted parabolic fit previously described. 
+In addition to the data from the full set of raw points, means and standard deviations of the peak height and location can be derived from the scan level peak characteristics previously calculated.
+
+In addition to these values, the frequency point-to-point median difference is calculated across all of the raw data points for those points that could be used for modeling frequency to m/z, and this difference of a single point from the peak center is calculated in frequency space, and converted to m/z space to provide an "offset" value that is useful to define the search space around the peak for any assignment algorithm.
+
+### Correction of Height and Standard Deviation
+
+Ideally, each peak would be observed in every scan. 
+However, in some scans, some peaks are not likely to be observed due to the number of ions falling below the detection threshold. 
+This results in a left-censored normal distribution of peak intensities across scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
+To correct these, either a correction based on a model of the truncated normal distribution can be used, or the differences can be simulated by sampling from data that is present in most of the scans.
+To simulate the effect of peaks missing from some scans on the standard deviations, the peaks present in all scans are used. 
+For each peak, a sample of the heights across scans are taken (ranging from 5% to 95% of scans), and a new standard deviation calculated for that fraction, and a ratio of the fractional standard deviation to the "true" standard deviation calculated.
+The ratio standard deviation across peaks can then be fitted to a cubic model of the fraction used, and a correction factor predicted for those peaks that are present in fewer scans. 
+The corrected standard deviations can then be used to correct the mean height assuming that it is the result of a left-censored normal distribution [@TruncatedNormalDistribution2022].
+
+### Frequency Standard Deviation Filtering
+
+High-peak-density (HPD) artifacts [Mitchell et al., 2018] present as singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
+Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and those FSD that are greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are removed.
+The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
+For HPD detection, the peaks in excel output from Xcalibur were used after converting the m/z peak centers into frequency space. 
+Sliding windows that are 1000 frequency points wide and offset by 100 points were used for the density calculations.
+
+### Multi-Class F-Statistics
+
+To evaluate the F-statistic between multi-class samples, a much larger set of samples was used.
+In this case, 181 matched non-cancer and cancer samples that were previously used in the HPD detection manuscript [ref].
+All 181 samples were peak characterized using the most stringent method below as well as peaks generated from the xcms R package [ref], assigned using SMIRFE, and then peaks matched by shared EMFs across samples (see Supplemental materials for a description of peak matching and deciding the most likely EMF).
+For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using our peak-characterization and those from xcms.
+
+### Samples and Overall Processing
+
+Samples included two ethyl-chloroformate (ECF) derivatized amino-acid samples [reference] and two samples where lipid extracts were extracted from lung tumors [reference]. 
+All four samples are positive mode, were converted from raw to mzML using msconvert from ProteoWizard (profile mode) and ThermoRawFileParser (centroided).
+
+For each raw data file in profile mzML format, they were processed in these six ways:
+
+  1. No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
+  1. Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
+  1. Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
+  1. Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
+  1. Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
+  1. Noise removal, two pass normalization, frequency SD filtering (filtersd)
+  1. Scan level peak centroids generated by MSnbase
+  1. Scans merged and then centroids generated by MSnbase
+
+For the two sets of MSnbase peaks, we also matched scan level peaks to merged peaks via two methods:
+
+  1. All scan level peaks within a 2ppm window are associated with the *merged* peak
+  1. Using the *filtersd* peaks, merged and scan level peaks are associated if they are within the start and stop m/z of the *filtersd* peaks
+
+## Conclusions
+
 ## References

### 2022-03-18 _targets.R

@@ -147,6 +147,12 @@ hpd_width_tar = tar_combine(
   command = combine_hpd_width(!!!.x)
 )
 
+hpd_chisq_tar = tar_combine(
+  hpd_chisq_all,
+  hpd_tar[[2]],
+  command = dplyr::bind_rows(!!!.x)
+)
+
 
 list(pkg_tar,
      data_tar,
@@ -158,4 +164,5 @@ list(pkg_tar,
      msnbase_tar,
      hpd_tar,
      hpd_width_tar,
+     hpd_chisq_tar,
      other_tar)

### 2022-03-18 doc/peakcharacterization_manuscript.Rmd
some updates for the hpd stuff

@@
   - Robert M Flight:
+      email: robert.flight@uky.edu
       institute: [markey, biochem, rcsirm]
   - Joshua M Mitchell:
-      institute: lab
+      email: jmmi243@uky.edu
+      institute: [markey, biochem, rcsirm, ibi]
   - Hunter NB Moseley:
@@
   - tox: Department of Toxicology and Cancer Biology, University of Kentucky, Lexington, KY 40536, USA
-  - lab: Somewhere over the rainbow
 date: "`r Sys.time()`"
@@
 
+`r table_count$label_text("rsd_method")`.
+Relative standard deviation means, medians, modes and maximum observed values for each sample with different overall processing.
+
 ```{r rsd_table, results = 'asis'}

### 2022-03-18 R/examine_hpd.R

@@ -151,10 +151,55 @@ plot_hpds = function(hpd_res){
     dplyr::filter(HPD) %>%
     dplyr::mutate(Height = Height + 100000)
 
-  ggplot(peak_data, aes(x = ObservedMZ, xend = ObservedMZ, y = 0, yend = log10(Height))) +
+  plot_all = ggplot(peak_data, aes(x = ObservedMZ, xend = ObservedMZ, y = 0, yend = log10(Height))) +
     geom_segment() +
     geom_segment(data = highsd, aes(y = 6), color = "red") +
     geom_segment(data = hpd, aes(y = 7), color = "blue")
+
+  n_both = peak_data %>%
+    dplyr::filter(!is.na(HPDID)) %>%
+    dplyr::filter(HighSD | HPD) %>%
+    dplyr::group_by(HPDID) %>%
+    dplyr::summarise(hisd = sum(HighSD), scanlevel = sum(HPD)) %>%
+    dplyr::arrange(dplyr::desc(hisd))
+
+  n_xcal = purrr::imap_dfr(hpd_res$hpd_ranges, function(in_range, range_id){
+    xcal_in = hpd_res$xl_data %>%
+      dplyr::filter(dplyr::between(frequency, in_range[1], in_range[2]))
+    data.frame(xcal = nrow(xcal_in), HPDID = range_id)
+  })
+
+  peak_compare = dplyr::left_join(n_both, n_xcal, by = "HPDID")
+
+  use_group = n_both %>%
+    dplyr::slice(1) %>%
+    dplyr::pull(HPDID)
+
+  pc_in_group = peak_data %>%
+    dplyr::filter(HPDID %in% use_group)
+  xl_in_group = hpd_res$xl_data %>%
+    dplyr::filter(dplyr::between(frequency, hpd_res$hpd_ranges[[use_group]][1], hpd_res$hpd_ranges[[use_group]][2]))
+
+  use_range = range(c(pc_in_group$ObservedFrequency, xl_in_group$frequency))
+  pc_mode = metabolomicsUtilities::calculate_mode(pc_in_group$Height)
+  xl_mode = metabolomicsUtilities::calculate_mode(xl_in_group$intensity)
+  pc_plot = pc_in_group %>%
+    ggplot(aes(x = ObservedFrequency, xend = ObservedFrequency,
+               y = 0, yend = Height)) +
+    geom_segment() +
[…]

### 2022-03-18 R/processing_functions.R

@@ -30,6 +30,32 @@ reduce_removing_zero <- function(regions, point_regions_list, multiplier = 1.5, 
   IRanges::reduce(regions)
 }
 
+find_signal_98 <- function(regions, point_regions_list, multiplier = 1.5, n_point_region = 2000){
+  nz_counts = FTMS.peakCharacterization:::count_overlaps(regions, point_regions_list[[1]])
+  n_region = seq(2, length(point_regions_list))
+  for (iregion in n_region) {
+    nz_counts_iregion = FTMS.peakCharacterization:::count_overlaps(regions, point_regions_list[[iregion]])
+    nz_counts = nz_counts + nz_counts_iregion
+  }
+
+  chunk_indices = seq(1, length(nz_counts), by = n_point_region)
+
+  chunk_perc = purrr::map_dbl(chunk_indices, function(in_index){
+    use_counts = nz_counts[seq(in_index, min(in_index + (n_point_region - 1), length(nz_counts)))]
+    if (max(use_counts) > 0) {
+      return(stats::quantile(use_counts, 0.98))
+    } else {
+      return(0)
+    }
+  })
+
+  cutoff_value = ceiling(median(chunk_perc) * multiplier)
+
+  regions = regions[nz_counts > cutoff_value]
+  IRanges::reduce(regions)
+}
+
+
 single_adjustable_normalization = function(peak_regions, min_ratio = 0){
   intensity_measure = c("RawHeight", "Height")
   summary_function = median
@@ -174,6 +200,26 @@ filtersd = function(use_char){
   in_char$zip_ms$peak_finder$sort_ascending_mz()
   in_char$zip_ms$peak_finder$model_mzsd()
   list(char_obj = in_char, processed = "filtersd")
+}
+
+filtersd98 = function(use_char){
+  in_char = use_char$clone(deep = TRUE)
+  in_char$zip_ms$peak_finder$zero_normalization = FALSE
+  in_char$zip_ms$peak_finder$peak_regions$peak_regions =
+    find_signal_98(in_char$peak_finder$peak_regions$sliding_regions,
+                         in_char$peak_finder$peak_regions$frequency_point_regions$frequency,
+                         in_char$peak_finder$quantile_multiplier,
+                         in_char$peak_finder$n_point_region)
+  in_char$zip_ms$peak_finder$reduce_sliding_regions()
[…]

### 2022-03-18 _targets.R

@@ -32,7 +32,8 @@ method_data = expand_grid(
                                   "singlenorm",
                                   "intsinglenorm",
                                   "doublenorm",
-                                  "filtersd")),
+                                  "filtersd",
+                                  "filtersd98")),
   data_names = paste0("data_", data_mzml$name)) %>%
   dplyr::mutate(name = gsub("data_", "", data_names),
                 data_sym = rlang::syms(data_names),

### 2022-03-21 doc/peakcharacterization_manuscript.Rmd
new version of HPD graphs.

@@
 ```{r increment_hpd}
-figure_count$increment("hpd_fsd")
+figure_count$increment("hpd_compare_fsd")
 ```
@@
 These peaks can be detected by simply examining the distribution of frequency SDs and removing those that are outliers.
-To verify the removal of HPD regions, we converted centroided m/z's from XCalibur to frequency values using the previously calculated values for that sample.
+To verify the removal of HPD peaks, we converted centroided m/z's from XCalibur to frequency values using the previously calculated values for that sample.
 Peak density was measured using a sliding window 10 points wide and offset by one point.
 
-`r figure_count$label_text("hpd_fsd")` compares the HPD regions detected using the method from Mitchell et al., and their correspondence with high FSD peaks for one sample (the other comparisons are in Supplemental materials).
-`r table_count$label_text("hpd_fsd")` summarizes the results of Chi-square tests using contingency tables of HPD and high FSD peaks.
-Based on this comparison, the high FSD peaks are indeed HPD sites and should be removed before assignment.
-However, the FSD does not depend on calculating a sliding window based density, but is merely a product of those peaks across multiple scans that span a wider range frequency range than expected based on the peaks in the data.
-We also note that we do expect higher m/z standard deviations at higher m/z, but the frequency based method does not suffer the same effect, due to the offsets of frequency points being constant.
+`r figure_count$label_text("hpd_compare_fsd")` shows a single HPD site detected in the ECF 2 sample, with the peaks from Xcalibur, as well as various scan-level processing and the peaks from centroiding using MSnbase.
+From this figure, we can see that the point density based noise filtering removes a large number of the peaks in the HPD site, while the frequency standard deviation removes further peaks that may be suspect.
+After removing the high FSD peaks, the number of peaks left in the HPD site are the same or less than those from MSnbase, and at least in this example, look more likely to be real peaks compared to those from MSnbase.
+Therefore, scan-level characterization allows us to keep what are likely **real** peaks in HPD sites, without consideration of peak intensity.
 
-```{r hpd_fsd, dn_id = figure_count}
-tar_load(hpd_plots_49lipid)
-hpd_plots_49lipid
+```{r hpd_compare_fsd, dn_id = figure_count, fig.width = 16, fig.height = 8}
+tar_load(hpd_plots_2ecf)
+
+hpd_left_panel = wrap_plots(hpd_plots_2ecf$plots, ncol = 1)
+hpd_right_panel = hpd_plots_2ecf$n %>%
+  ggplot(aes(x = xcalibur,
+             y = count,
+             color = method)) +
+  geom_point(size = 3) +
+  labs(x = "Xcalibur peak count",
+       y = "peak count") +
+  theme(legend.position = c(0.1, 0.8))
+hpd_left_panel | hpd_right_panel
 ```
 
-`r figure_count$label_text("hpd_fsn")`.
-Comparison of HPD and high FSD sites in the lipid 49 sample.
-Black are the original fully characterized peaks.
-Red indicates peaks within HPD sites detected using the XCalibur peak sets.
-Blue indicates peaks that had abnormally high FSD values.
-
-`r table_count$label_text("hpd_fsn")`.
-Chi-square statistics from contingency tables of shared presence and absence of peaks in HPD sites and labeled as high FSD.
-
-```{r hpd_fsd_table}
-tar_load(hpd_chisq_all)
-sci_formatter = 
-hpd_chisq_table = hpd_chisq_all %>%
-  dplyr::select(sample, p.value, statistic) %>%
-  flextable() %>%
-  set_formatter(p.value = function(x){
-    formatC(x, format = "e", digits = 2)
-  }) %>%
-  colformat_double(digits = 1, j = 3) %>%
-  autofit()
-```
+`r figure_count$label_text("hpd_compare_fsd")`.
+Comparison of HPD and high FSD sites in the ECF 2 sample.
+Left: Peak plots for various peak processing modalities in a single HPD site. 
+*xcalibur*: peaks exported from Thermo Xcalibur after averaging scans.
+*scalevel_00*: Scan-level peak characterization without any density based filtering (see noperc_nonorm in Methods).
+*scanlevel_99*: Scan-level peak characterization using the default point density filtering (see filtersd in Methods).
+*scanlevel_99_lowsd*: Same as *scanlevel_99*, but removing any peaks marked as having a high frequency standard deviation.
+*msnbase*: peak centroids generated from *MSnbase*.
+Right: Scatterplot of peak counts across all of the HPDs detected in the ECF 2 sample against the Xcalibur peak counts.
 

### 2022-03-21 R/examine_hpd.R

@@ -167,18 +167,18 @@ width_sd_hpds = function(hpd_res){
 plot_hpds = function(hpd_res){
   peak_data = hpd_res$hpd_data
 
+
   n_any = peak_data %>%
     dplyr::group_by(source, HPDID) %>%
     dplyr::summarise(n = n(),
                      n_lowsd = n - sum(HighSD))
   n_wide = n_any %>%
-    tidyr::pivot_wider(id_cols = HPDID, names_from = source, values_from = n) %>%
+    tidyr::pivot_wider(id_cols = HPDID, names_from = source, values_from = n, values_fill = 0) %>%
     dplyr::arrange(dplyr::desc(scanlevel_99))
   n_wide_lowsd = n_any %>%
-    tidyr::pivot_wider(id_cols = HPDID, names_from = source, values_from = n_lowsd) %>%
+    tidyr::pivot_wider(id_cols = HPDID, names_from = source, values_from = n_lowsd, values_fill = 0) %>%
     dplyr::arrange(dplyr::desc(scanlevel_99)) %>%
     dplyr::transmute(HPDID = HPDID,
-                     scanlevel_00_lowsd = scanlevel_00,
                      scanlevel_99_lowsd = scanlevel_99)
 
   all_n_wide = dplyr::left_join(n_wide, n_wide_lowsd, by = "HPDID")
@@ -196,6 +196,9 @@ plot_hpds = function(hpd_res){
     dplyr::filter(HPDID %in% use_hpdid)
 
   peaks_type = split(peaks_use, peaks_use$source)
+  peaks_type$scanlevel_99_lowsd = peaks_type$scanlevel_99 %>%
+    dplyr::filter(!HighSD) %>%
+    dplyr::mutate(source = "scanlevel_99_lowsd")
   plot_range = range(peaks_use$ObservedFrequency)
 
   plots_type = purrr::map(peaks_type, function(in_type){
@@ -206,26 +209,16 @@ plot_hpds = function(hpd_res){
                  xend = ObservedFrequency,
                  y = 0,
                  yend = Height)) +
-      geom_segment()
-    if (grepl("scanlevel", in_type$source[1])) {
-      hi_sd = in_type %>%
-        dplyr::filter(HighSD)
-      new_plot = base_plot +
-        geom_segment(data = hi_sd, color = "red")
-    } else {
-      new_plot = base_plot
-    }
-    new_plot = new_plot +
+      geom_segment() +
       coord_cartesian(xlim = plot_range,
[…]

### 2022-03-21 doc/peakcharacterization_manuscript.Rmd
just a few more tweaks to the figure, and I think we are good.

@@
   theme(legend.position = c(0.1, 0.8))
-hpd_left_panel | hpd_right_panel
+(hpd_left_panel | hpd_right_panel) + plot_annotation(tag_levels = "A")
 ```
@@
 Comparison of HPD and high FSD sites in the ECF 2 sample.
-Left: Peak plots for various peak processing modalities in a single HPD site. 
-*xcalibur*: peaks exported from Thermo Xcalibur after averaging scans.
-*scalevel_00*: Scan-level peak characterization without any density based filtering (see noperc_nonorm in Methods).
-*scanlevel_99*: Scan-level peak characterization using the default point density filtering (see filtersd in Methods).
-*scanlevel_99_lowsd*: Same as *scanlevel_99*, but removing any peaks marked as having a high frequency standard deviation.
-*msnbase*: peak centroids generated from *MSnbase*.
-Right: Scatterplot of peak counts across all of the HPDs detected in the ECF 2 sample against the Xcalibur peak counts.
+**A** - **E**:Peak plots for various peak processing modalities in a single HPD site. 
+**A**:*xcalibur* peaks exported from Thermo Xcalibur after averaging scans.
+**B**:*scalevel_00*: Scan-level peak characterization without any density based filtering (see noperc_nonorm in Methods).
+**C**:*scanlevel_99*: Scan-level peak characterization using the default point density filtering (see filtersd in Methods).
+**D**:*scanlevel_99_lowsd*: Same as *scanlevel_99*, but removing any peaks marked as having a high frequency standard deviation.
+**E**:*msnbase*: peak centroids generated from *MSnbase*.
+**F**: Scatterplot of peak counts across all of the HPDs detected in the ECF 2 sample against the Xcalibur peak counts.
 
@@
 figure_count$increment("rsd_method")
+table_count$increment("rsd_method")
 ```

### 2022-03-22 doc/peakcharacterization_manuscript.Rmd
updated figures.
Closes #10

@@
 figure_count$increment("slidingwindow_count")
-tar_load(nonoise_vs_noise)
-nn_n_ratio = english::as.english(round(nonoise_vs_noise["noperc"] / nonoise_vs_noise["perc99"]))
+tar_load(mn_ratios)
+tar_load(sliding_regions)
+tar_load(noise_plot)
 ```
@@
 In the example below, the median of the tiled windows is 5, resulting in a cutoff of 7.5.
-The number of peaks detected without applying a cutoff explodes `r nn_n_ratio`-fold (from `r format(nonoise_vs_noise["perc99"], big.mark = ",")` to `r format(nonoise_vs_noise["noperc"], big.mark = ",")` in the 97lipid sample), with a vast majority of the peaks being present in a single scan.
+The number of peaks detected without applying a cutoff explodes anywhere from `r mn_ratios[["97lipid"]]$ratio`-fold (from `r format(mn_ratios[["97lipid"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["97lipid"]]$count_0, big.mark = ",")` in the 97lipid sample), to `r mn_ratios[["2ecf"]]$ratio`-fold (from `r format(mn_ratios[["2ecf"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["2ecf"]]$count_0, big.mark = ",")` in the 2ecf sample) with a vast majority of the peaks being present in a single scan.
 By discarding those windows that have a non-zero point density below 1.5x the median of 99th percentiles, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
 
-```{r slidingwindow_count, dn_id = figure_count}
-tar_load(sliding_regions)
-sliding_regions
+```{r slidingwindow_count, dn_id = figure_count, fig.width = 8, fig.height = 8}
+(sliding_regions / noise_plot) + plot_annotation(tag_levels = "A")
 ```
 
-`r figure_count$label_text("slidingwindow_count")`. Histogram of the number of non-zero points across all scans in sliding windows 10 frequency points wide and 1 frequency point apart. 
-The red vertical line denotes the lower 99th percentile of the data, only sliding windows with non-zero counts above the red line will be kept for subsequent use.
+`r figure_count$label_text("slidingwindow_count")`. 
+**A**: Histogram of the number of non-zero points from sliding windows 10 frequency points wide and 1 frequency point apart, from a single tiled window 2000 points wide.
+Blue line is the 99th percentile.
+**B**: Histogram of the 99th percentile cutoffs from all of the tiled windows.
+Red line is the median x 1.5.
+**C** and **D** show the number of initial regions as a function of the percentile cutoff used for either the AA ECF (C) or lipid (D) samples.
 

### 2022-03-22 doc/peakcharacterization_manuscript.Rmd
incorporating Joshs edits

@@
 
-To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur (a process we have observed being used to generate datasets), as well as doing a straight averaging of point intensities using identical scans as was kept by our scan filtering (see Methods) in the xcms R package, and finally averaging peaks that were centroided via the Thermo raw file parsing utilities (wrapped by ThermoRawFileParser).
-In all cases, for a known assigned set of peaks, the intensity ratios between the assigned peaks are off compared to what is expected based on natural abundance probability (NAP), as shown in Figure X, with some solutions giving different amounts of error.
+To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur (a process we have observed being used to generate datasets), as well as merging scans (see Methods) in the MSnbase R package.
+In all cases, for a known assigned set of peaks, the intensity ratios between the assigned peaks deviate from the intensity ratio predicted by their natural abundance probabilities (NAPs), as shown in Figure X, with some solutions giving different amounts of error.
 We do note that when all of the peaks are present in a single scan, their relative intensity ratios are much closer to the theoretical ratios based on NAP.
@@
 
-FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data. 
+FT-MS data from the Thermo-Fisher Orbitrap instruments used to acquire the data does not provide any information about the raw frequency data. 
 Outside of the meta-data, it merely contains the m/z and intensity values for profile spectra acquired across multiple scans. 
@@
 In a dataset of this nature, we expect that much of the data is really just noise, and doesn't contribute that much to the analysis. 
-It is expected that noise is randomly distributed across the scans. 
+Furthermore, it is expected that noise is randomly distributed across the scans. 
 Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
@@
 The distribution of 99th percentiles across all of the tiled regions for a single sample is shown in `r figure_count$label_text("slidingwindow_count")`B.
-We can use the median (50th percentile) value of all the 99th percentiles from all of the tiled regions to determine a cutoff value above which we think any non-zero points are potentially **signal** and not just noise.
+We can use the median (50th percentile) value of all the 99th percentiles from all of the tiled regions to determine a cutoff value above which any non-zero points are potentially **signal** and not just noise.
 In the example below, the median of the tiled windows is 5, resulting in a cutoff of 7.5.
 The number of peaks detected without applying a cutoff explodes anywhere from `r mn_ratios[["97lipid"]]$ratio`-fold (from `r format(mn_ratios[["97lipid"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["97lipid"]]$count_0, big.mark = ",")` in the 97lipid sample), to `r mn_ratios[["2ecf"]]$ratio`-fold (from `r format(mn_ratios[["2ecf"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["2ecf"]]$count_0, big.mark = ",")` in the 2ecf sample) with a vast majority of the peaks being present in a single scan.
-By discarding those windows that have a non-zero point density below 1.5x the median of 99th percentiles, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
+By discarding those windows with a non-zero point density below 1.5x the median of 99th percentiles, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
 
@@
 `r figure_count$label_text("slidingwindow_count")`. 
-**A**: Histogram of the number of non-zero points from sliding windows 10 frequency points wide and 1 frequency point apart, from a single tiled window 2000 points wide.
+**A**: Histogram of the number of non-zero points from sliding windows ten frequency points wide and a stride of one point, from a single tiled window 2000 points wide.
 Blue line is the 99th percentile.
@@
 
-Due to differences in how many ions are captured in the trap, the peak intensities in each scan vary slightly (see `r figure_count$label_text("breaking_regions")`A and B for example).
+Due to differences in how many ions are captured in the trap and the limited dynamic range of the detector, the observed peak intensities for the same analyte vary slightly between scans (see `r figure_count$label_text("breaking_regions")`A and B for example).
 Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
-However, there are two issues with these peak intensities across scans: 1) some peak intensities correlate with the scan order; and 2) some peak differences between scans are correlated with intensity.
+However, there are two issues with these peak intensities across scans: 1) some peak intensities correlate with the scan number (i.e.
+scan acquisition order); and 2) some peak differences between scans are correlated with intensity.
 The solution to **1** is to do a two pass normalization.
-After the first pass, the peaks who's intensity is correlated with scan order are detected.
+After the first pass, the peaks whose intensity is correlated with scan order are detected.
 In the second pass, the correlated peaks are removed, and normalization is carried out again.
@@
 Ideally, the peak characterization procedure should reduce their presence in the resultant reported peaks. 
-Their presence should be minimized by removing noise peaks, and removing regions that have multiple reported peaks in the same scan. 
-We expect they may present as characterized peaks that have larger than expected frequency standard deviations (FSD) when calculated across scans.
-These peaks can be detected by simply examining the distribution of frequency SDs and removing those that are outliers.
+Their presence should be minimized by removing noise peaks, and removing peak regions that have multiple reported peaks from the same scan. 
+However, we expect they may also present as characterized peaks that have larger than expected frequency level standard deviations (FSD) when calculated across scans.
+These peaks can be detected by simply examining the distribution of FSDs and removing those that are outliers.
 To verify the removal of HPD peaks, we converted centroided m/z's from XCalibur to frequency values using the previously calculated values for that sample.
-Peak density was measured using a sliding window 10 points wide and offset by one point.
+Peak density was measured using a sliding window ten points wide and a stride of one point.
 
@@
 The data consists of profile mode m/z spectra from multiple scans encoded as m/z and intensity values for each scan. 
-There is no information about the original frequency values available in either the `raw` files or the `mzML` files. 
-However, the frequency values can be approximated by averaging the m/z of adjacent points, and dividing the m/z by the difference. 
-Ideally, the difference between subsequent points in frequency space is constant, in practice there is a range of differences in frequency space.
-The *true, constant* difference can be obtained by examining the median of the mean frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to m/z) to be within 2% of the mode value.
+No information about the original observed frequency values is available in either the `raw` files or the `mzML` files. 
+However, proxy frequency values can be generated by averaging the m/z of adjacent points and dividing them by the difference. 
+Ideally, the difference between subsequent points in this proxy frequency space is constant, in practice there is a range of differences in frequency space.
+The *actual, constant* difference can be obtained by examining the median of the frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to m/z) to be within 2% of the mode value.
 These *useful* points can be used to construct a linear model relating m/z to frequency for individual scans based on the formula:
@@
 $$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
-From the known physical properties of the Orbitrap, only the square root term should be necessary.
-Practically, we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
-A frequency model is generated for each scan, and then a single model using the scan with the square-root slope closest to the median of the square-root slopes from all scans.
-The single model is used to convert the m/z values to frequency for all scans.
+From the known physical properties of the Orbitrap, only the square root term should be necessary [@ledfordSpaceChargeEffects1984].
+Practically, we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry, contributions from space charge effects and magnetronic motion, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
+A frequency model is generated for each scan, and then a single model using the scan with the square-root term closest to the median of the square-root terms from all scans.
+We observed that this single model better preserved the relative ordering of the peaks in both m/z and frequency-space compared to the scan specific models (see Results).
 A single model rather than scan specific models is used for conversion because this was found to maintain the relative ordering of the peaks in m/z and frequency space (see Results).
@@
 Two types of frequency intervals are used, sliding and tiled windows. 
-In this work, the sliding windows are 10 points wide and offset by one point.
-The tiled windows are one point wide and offset by one point.
+In this work, the sliding windows are 10 points wide with a stride of one point.
+The tiled windows are one point wide with a stride of one point.
 
@@
 
-To enable interval algebra, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package [@lawrenceSoftwareComputingAnnotating2013a].
-The sliding and tiled windows are also converted to IRanges objects the same way.
+To enable interval algebra, the frequency points are converted to single width intervals by multiplying (by a multiplier of 400 in this work), rounding to the nearest integer, and storing them as IRanges objects from the IRanges Bioconductor package [@lawrenceSoftwareComputingAnnotating2013a].
+The sliding and tiled windows are also converted to IRanges objects using this process.
 
@@
 Subsequently, we break these counts into fixed width tiles (default width of 2000) and calculate the 99th percentile of non-zero points for each tile.
-The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions that contain potential signal.
-Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged together to create the initial peak regions.
+The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions containing potential signal.
+Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged to create the initial peak regions.
 The presence of zero intensity points are due to flooring implemented by the spectrometer when local spectral quality falls below a certain threshold.
 When this flooring is unstable and incomplete, a partial ringing phenomenom is observed [@mitchellNewMethodsIdentify2018].
-**aside: I'm not sure how this actually relates to the density noise cutoff**.
 
@@
 
-Scans are normalized to a single *reference* scan based on the log-intensity differences of a subset of peaks that are present in at least the same number of scans as the 95th percentile of scan counts for the peaks.
-In addition, only those peaks that have an intensity greater than 0.7 times the highest intensity peak in the scan are used. 
+Scans are normalized to a single *reference* scan based on the log-intensity differences of a subset of peaks present in at least the same number of scans as the 95th percentile of scan counts for the peaks.
+In addition, only those peaks with an intensity greater than 0.7 times the highest intensity peak in the scan are used. 
 Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each scan to all other scans.
@@
 
-Based on the previously detected peaks, the full set of raw data points for each peak in each scan within a region is known. 
-The non-zero intensity, normalized raw data points across scans can be combined, and then characterized again using the weighted parabolic fit previously described. 
-In addition to the data from the full set of raw points, means and standard deviations of the peak height and location can be derived from the scan level peak characteristics previously calculated.
+The full set of raw data points for each peak in each scan within a region is known based on the previously detected peaks. 
+Therefore, the non-zero intensity, normalized raw data points across scans can be combined, and then characterized again using the weighted parabolic fit previously described. 
+In addition to the data from the full set of raw points, the means and the standard deviations of the peak height and location can be derived from the scan-level peak characteristics previously calculated.
 
@@
 Ideally, each peak would be observed in every scan. 
-However, in some scans, some peaks are not likely to be observed due to the number of ions falling below the detection threshold. 
+However, some peaks are not observed in some scans due to the number of ions falling below the detection threshold. 
 This results in a left-censored normal distribution of peak intensities across scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
@@
 The ratio standard deviation across peaks can then be fitted to a cubic model of the fraction used, and a correction factor predicted for those peaks that are present in fewer scans. 
-The corrected standard deviations can then be used to correct the mean height assuming that it is the result of a left-censored normal distribution [@TruncatedNormalDistribution2022].
+The corrected standard deviations can then be used to correct the mean height assuming that it is the result of a left-censored normal distribution [@TruncatedNormalDistribution2022; @burkardtTruncatedNormalDistribution].
 
@@
 
-High-peak-density (HPD) artifacts [Mitchell et al., 2018] present as singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
-Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and those FSD that are greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are removed.
+High-peak-density (HPD) artifacts [@mitchellNewMethodsIdentify2018] present as groups of singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
+Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and peaks with FSDs greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are removed.
 The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
 For HPD detection, the peaks in excel output from Xcalibur were used after converting the m/z peak centers into frequency space.
-Sliding windows that are 1000 frequency points wide and offset by 100 points were used for the density calculations.
+Sliding windows that are 1000 frequency points wide with a stride of 100 points were used for the density calculations.
 
@@
 To evaluate the F-statistic between multi-class samples, a much larger set of samples was used.
-In this case, 181 matched non-cancer and cancer samples that were previously used in the HPD detection manuscript [ref].
-All 181 samples were peak characterized using the most stringent method below as well as peaks generated from the xcms R package [ref], assigned using SMIRFE, and then peaks matched by shared EMFs across samples (see Supplemental materials for a description of peak matching and deciding the most likely EMF).
-For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using our peak-characterization and those from xcms.
+In this case, 181 matched non-cancer and cancer samples previously used in the HPD detection manuscript [@mitchellNewMethodsIdentify2018].
+All 181 samples were peak characterized using the most stringent method below as well as peaks generated from the MSnbase R package [ref], assigned using SMIRFE, and then peaks matched by shared EMFs across samples (see Supplemental materials for a description of peak matching and deciding the most likely EMF).
+For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using scan-level peak-characterization and those from MSnbase.
 
@@
 Samples included two ethyl-chloroformate (ECF) derivatized amino-acid samples [reference] and two samples where lipid extracts were extracted from lung tumors [reference]. 
-All four samples are positive mode, were converted from raw to mzML using msconvert from ProteoWizard (profile mode) and ThermoRawFileParser (centroided).
+All four samples are positive mode, were converted from raw to mzML using msconvert from ProteoWizard (profile mode).
 
@@
   1. Noise removal, two pass normalization, frequency SD filtering (filtersd)
-  1. Scan level peak centroids generated by MSnbase
-  1. Scans merged and then centroids generated by MSnbase
+  1. Scan level peak centroids generated by MSnbase (using `pickPeaks` on the individual scans)
+  1. Scans merged and then centroids generated by MSnbase (using `combineSpectra` and `pickPeaks`)
 

### 2022-03-22 R/get_assignments.R
bunch more of the pieces, and getting the assignments read in

@@ -1,0 +1,5 @@
+get_assignments = function(assignments_file){
+  #assignments_file = tar_read(assign_filtersd_1ecf)
+  assign_data = smirfeTools::read_smirfe_assignment(assignments_file$file)
+  assign_data
+}

### 2022-03-22 _targets.R

@@ -50,7 +50,9 @@ method_tar = tar_map(
   tar_target(method, method_sym(data_sym)),
   tar_target(rsd, single_rsd(method)),
   tar_target(zip, write_peaks_for_assignment(method)),
-  tar_target(assign, assign_files(zip))
+  tar_target(assign, assign_files(zip)),
+  tar_target(n_peak, n_final_peak(method)),
+  tar_target(emf, get_assignments(assign))
 )
 
 
@@ -59,6 +61,37 @@ normalization_tar = tar_combine(
   method_tar[[1]],
   command = normalization_factors(!!!.x),
   iteration = "list"
+)
+
+n_peak_tar = tar_combine(
+  n_peak_combine,
+  method_tar[[5]],
+  command = dplyr::bind_rows(!!!.x)
+)
+
+noise_tar = tibble(
+  data_name = (c(
+    "data_97lipid",
+    "data_49lipid",
+    "data_1ecf",
+    "data_2ecf"
+  ))
+) %>%
+  dplyr::mutate(
+    data_syms = rlang::syms(data_name),
+    data_id = rename_samples(data_name)
+  )
+
+compare_noise_tar = tar_map(
+  values = noise_tar,
+  names = "data_id",
+  tar_target(noise, compare_noise_cutoffs(data_syms))
+)
+
+combine_noise_tar = tar_combine(
+  noise_combine,
+  compare_noise_tar,
+  command = dplyr::bind_rows(!!!.x)
 )
 
 # rsd_tar = tar_combine(
@@ -82,6 +115,10 @@ figures_tar = tar_plan(
   ),
   tar_target(compare_normalization,
              normalization_graph(normalization_combine)),
+  tar_target(noise_plot,
+             create_noise_plot(noise_combine)),
+  tar_target(mn_ratios,
+             calculate_m_n_ratio(noise_combine)),
   tar_render(manuscript, "doc/peakcharacterization_manuscript.Rmd")
 )
 
@@ -89,12 +126,12 @@ tables_tar = tar_plan(
   tar_target(rsd_values, summarize_rsd(rsd_combine))
 )
 
-other_tar = tar_plan(
-  tar_target(
-    nonoise_vs_noise,
-    compare_noise_cutoff(data_97lipid)
-  )
-)
+# other_tar = tar_plan(
+#   tar_target(
+#     nonoise_vs_noise,
+#     compare_noise_cutoff(data_97lipid)
+#   )
+# )
 
[…]

### 2022-03-29 doc/peakcharacterization_manuscript.Rmd
MDPI style

@@
   word_document:
+    reference_docx: metabolites-template-v3.docx
     keep_md: true
@@
   chunk_output_type: console
+  markdown: 
+    wrap: sentence
 ---
@@
 
-## Abstract
-
-We introduce a novel method for characterizing peaks from direct injection FT-MS experiments developed on metabolomics datasets that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
+::: {custom-style="MDPI_1.7_abstract"}
+**Abstract:** We introduce a novel method for characterizing peaks from direct injection FT-MS experiments developed on metabolomics datasets that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
 Our method incorporates several aspects designed to make the derived intensities trustworthy, and suitable for use in our previously published peak assignment algorithm SMIRFE.
 Notably, the inclusion of an intensity independent noise removal, and normalization of scan level data results in a reduction of median relative standard deviation from **X** to **Y**, and much better fit of peak intensities to relative natural abundances of assigned isotopologues.
+:::
 
-## Keywords
+::: {custom-style="MDPI_1.8_keywords"}
+**Keywords**: Fourier-transform mass-spectrometry
+:::
 
+::: {custom-style="MDPI_2.1_heading1"}
 ## Introduction
+:::
 
+::: {custom-style="MDPI_3.1_text"}
+What can we say about FT-MS data?
+There is more and more of it, and more of it applied to metabolomics.
+Profile mode, multi-scan direct injection FT-MS is an interesting approach because it allows us to handle cases where something might go wrong in an individual scan, while still extracting useful data.
+:::
+
+::: {custom-style="MDPI_2.1_heading1"}
 ## Results
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Naively Averaged Data Have Bad Relative Intensities
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur (a process we have observed being used to generate datasets), as well as merging scans (see Methods) in the MSnbase R package.
@@
 We do note that when all of the peaks are present in a single scan, their relative intensity ratios are much closer to the theoretical ratios based on NAP.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### m/z to Frequency
+:::
 
@@
 
+::: {custom-style="MDPI_3.1_text"}
 FT-MS data from the Thermo-Fisher Orbitrap instruments used to acquire the data does not provide any information about the raw frequency data.
@@
 Frequency based points suffer none of these drawbacks, and the conversion from m/z can be derived from the raw profile level data itself, which is incredibly useful.
+:::
 
+::: {custom-style="MDPI_5.2_figure"}
 ```{r mz_frequency_conversion, dn_id = figure_count, fig.width = 16, fig.height = 10}
@@
 ```
+:::
 
-`r figure_count$label_text("mz_frequency_conversion")`. **A**: Intensity vs m/z for a single peak from a single scan. 
+::: {custom-style="MDPI_5.1_figure_caption"}
+**`r figure_count$label_text("mz_frequency_conversion")`.** **A**: Intensity vs m/z for a single peak from a single scan.
 Red lines denote the differences between each point, and red dots the average between the pair of points.
@@
 **E** shows the plot of derived frequency *vs* m/z, with fitted values from the linear model in red.
+:::
 
@@
 
+::: {custom-style="MDPI_3.1_text"}
 The m/z to frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
@@
 Therefore, a single model for all scans based on the scan with the slope closest to the median of slopes across scans is used for converting *all* remaining scan level data.
+:::
 
+::: {custom-style="MDPI_5.2_figure"}
 ```{r peak_ordering, dn_id = figure_count}
@@
 ```
+:::
 
-`r figure_count$label_text("peak_ordering")`.
-Peak ordering in m/z compared with ordering in frequency space when a single m/z to frequency model is used or scan specific m/z to frequency models are used.
+::: {custom-style="MDPI_5.1_figure_caption"}
+**`r figure_count$label_text("peak_ordering")`.** Peak ordering in m/z compared with ordering in frequency space when a single m/z to frequency model is used or scan specific m/z to frequency models are used.
 For a single peak, the scan level peak m/z's were extracted, and then frequency values for those m/z generated using a single common model of m/z to frequency (*single_order*), or models derived from each scan (*scan_order*).
 A subset of the peaks end up out of order using scan specific models, implying that a single model should be used across all the scan level data.
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 Although the original model is created from only those points that had frequency point-to-point differences within a narrow range, **all** m/z points are converted to frequency for subsequent steps in the workflow.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Sliding Window Density to Remove Noise
+:::
 
@@
 
+::: {custom-style="MDPI_3.1_text"}
 In a dataset of this nature, we expect that much of the data is really just noise, and doesn't contribute that much to the analysis.
@@
 By discarding those windows with a non-zero point density below 1.5x the median of 99th percentiles, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
+:::
+
+```{=html}
+<div custom-style="MDPI_5.2_figure>
 
@@
 
-`r figure_count$label_text("slidingwindow_count")`. 
+</div>
+
+<div custom-style="MDPI_5.1_figure_caption">
+**`r figure_count$label_text("slidingwindow_count")`.**
 **A**: Histogram of the number of non-zero points from sliding windows ten frequency points wide and a stride of one point, from a single tiled window 2000 points wide.
@@
 **C** and **D** show the number of initial regions as a function of the percentile cutoff used for either the AA ECF (C) or lipid (D) samples.
+</div>
+```
+::: {custom-style="MDPI_3.1_text"}
+For each region initially created, the peaks in each scan within that region can be characterized using a quadratic fit of log(intensity) to m/z.
+:::
 
-For each region initially created, the peaks in each scan within that region can be characterized using a quadratic fit of log(intensity) to m/z.
-
+```{=html}
+<div custom-style="MDPI_2.2_heading2>
 ### Peak Characterization Using Quadratic Fit
+</div>
 
@@
 
+<div custom-style="MDPI_3.1_text">
 Although many other types of mass-spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo Fusion has a baseline of 0 due to manipulations in the Thermo firmware, making the determination of the centroided values considerably easier.
@@
 From the fitted model, we can derive the centroided center and the intensity of the peak, as shown in `r figure_count$label_text("centroided_peaks")`.
-
+</div>
+```
+```{=html}
+<div custom-style="MDPI_5.2_figure>
 ```{r centroided_peaks, dn_id = figure_count}
@@
 ```
+</div>
 
-`r figure_count$label_text("centroided_peaks")`.
+<div custom-style="MDPI_5.1_figure_caption">
+**`r figure_count$label_text("centroided_peaks")`.**
 Log-intensity (A) and Intensity (B) of points for a single peak against frequency. Black points are the original data points, and the blue point represents the calculated centroid.
-
+</div>
+```
+::: {custom-style="MDPI_2.2_heading2"}
 ### Breaking Up Initial Regions
+:::
 
@@
 
+::: {custom-style="MDPI_3.1_text"}
 With the characterized (centroided) peak data from across scans within each region, it is then important to determine if only one or multiple "peaks" are actually present in the region.
@@
 `r figure_count$label_text("breaking_regions")` shows an example where an initial region is broken up into two regions based on the characterized peak centers.
+:::
 
+::: {custom-style="MDPI_5.2_figure"}
 ```{r breaking_regions, dn_id = figure_count}
@@
 ```
+:::
 
-`r figure_count$label_text("breaking_regions")`. 
-Splitting a single region into two regions based on the peaks that are present.
+::: {custom-style="MDPI_5.1_figure_caption"}
+**`r figure_count$label_text("breaking_regions")`.** Splitting a single region into two regions based on the peaks that are present.
 **A**: The full set of raw frequency and intensity data across all scans for the region are shown.
@@
 Each group of non-zero adjacent regions will be merged to form a new peak region.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Normalization of Scans
+:::
 
@@
 
+::: {custom-style="MDPI_3.1_text"}
 Due to differences in how many ions are captured in the trap and the limited dynamic range of the detector, the observed peak intensities for the same analyte vary slightly between scans (see `r figure_count$label_text("breaking_regions")`A and B for example).
@@
 If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and B.
+:::
 
+::: {custom-style="MDPI_5.2_figure"}
 ```{r intensity_scan, dn_id = figure_count}
@@
 ```
+:::
 
-`r figure_count$label_text("intensity_scan")`. 
-**A**: An example of a peak whose height across scans is correlated with scan number.
+::: {custom-style="MDPI_5.1_figure_caption"}
+**`r figure_count$label_text("intensity_scan")`.** **A**: An example of a peak whose height across scans is correlated with scan number.
 **B**: The peak differences to the same peaks in a reference scan are plotted against peak height.
-Black: Peaks with a height < 0.7 of the maximum. 
-Red: Peaks with a height >= 0.7 of the maximum.
+Black: Peaks with a height \< 0.7 of the maximum.
+Red: Peaks with a height \>= 0.7 of the maximum.
+:::
 
+::: {custom-style="MDPI_5.2_figure"}
 ```{r normalization_factors, dn_id = figure_count, fig.height = 8, fig.width = 8}
@@
 ```
+:::
 
-`r figure_count$label_text("normalization_factors")`. 
-**A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity >= 0.7 of the maximum intensity (singlenorm_int), or the double pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
+::: {custom-style="MDPI_5.1_figure_caption"}
+**`r figure_count$label_text("normalization_factors")`.** **A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity \>= 0.7 of the maximum intensity (singlenorm_int), or the double pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
 **B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Removal of High Peak Density Artefacts
+:::
 
@@
 
+::: {custom-style="MDPI_3.1_text"}
 We have previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [@mitchellNewMethodsIdentify2018].
@@
 Therefore, scan-level characterization allows us to keep what are likely **real** peaks in HPD sites, without consideration of peak intensity.
+:::
 
+::: {custom-style="MDPI_5.2_figure"}
 ```{r hpd_compare_fsd, dn_id = figure_count, fig.width = 16, fig.height = 8}
@@
 ```
+:::
 
-`r figure_count$label_text("hpd_compare_fsd")`.
-Comparison of HPD and high FSD sites in the ECF 2 sample.
+::: {custom-style="MDPI_5.1_figure_caption"}
+**`r figure_count$label_text("hpd_compare_fsd")`.** Comparison of HPD and high FSD sites in the ECF 2 sample.
 **A** - **E**:Peak plots for various peak processing modalities in a single HPD site.
@@
 **F**: Scatterplot of peak counts across all of the HPDs detected in the ECF 2 sample against the Xcalibur peak counts.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Changes in Relative Standard Deviation (RSD)
+:::
 
@@
 
+::: {custom-style="MDPI_3.1_text"}
 Each step in the peak characterization either changes the overall number of peaks coming from each scan (sliding windows and breaking initial regions) or the overall intensity of the points within a scan.
 Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation (RSD) for the characterized scan level peak intensities (calculated as the standard deviation of peak heights across scans divided by the mean peak height), and compare them as each processing step is introduced.
+:::
 
@@
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Difference to Relative Natural Abundance
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 As an alternative to RSD, we can also compare the fit of relative intensities after assignment using SMIRFE [reference] to the theoretical relative natural abundances (relNAP) of the assigned isotopic molecular formula's (IMFs) within the assigned elemental molecular formula's (EMFs).
@@
 To compare relative NAP to relative abundances, we only examined the assignments from the two samples containing ECF derivatized amino-acids, as we can limit the assignments to those that match expected derivatizations of the known amino-acids (see Supplemental for the expected EMFs, and expected relative NAPs for the individual IMFs).
+:::
 
+```{r nap_intensity}
+tar_load(aa_diffs_filtersd_1ecf)
+tar_load(aa_diffs_filtersd_2ecf)
+
+all_diffs = rbind(aa_diffs_filtersd_1ecf$all_diffs %>%
+                    dplyr::mutate(sample = "1ecf"),
+                  aa_diffs_filtersd_2ecf$all_diffs %>%
+                    dplyr::mutate(sample = "2ecf"))
+
+threonine_ratios = aa_diffs_filtersd_1ecf$specific_diffs$ratios %>%
+  dplyr::filter(source %in% c("scan-level", "xcalibur"))
+threonine_diffs = aa_diffs_filtersd_1ecf$specific_diffs$diffs %>%
+  dplyr::filter(source %in% "xcal_raw_char")
+
+threonine_ratioplot = ggplot(threonine_ratios, aes(x = ratio, y = diff, color = source, size = n_missing)) +
+    geom_point() +
+    theme(legend.position = c(0.7, 0.8)) +
+    labs(x = "Pairwise Comparison",
+         y = "Peak-Peak NAP - Height Difference")
+
+threonine_diffplot = ggplot(threonine_diffs, aes(x = n_missing, y = diff)) +
+    geom_point(size = 2) +
+    labs(x = "Number of Missing Scans",
+         y = "Xcalibur - Scan-Level NAP - Height Difference")
+
+aa_diffplot = all_diffs %>%
+  dplyr::filter(source %in% "xcal_raw_char") %>%
+  ggplot(aes(x = n_missing, y = diff)) +
+  geom_point() +
+  labs(x = "Number of Missing Scans",
+         y = "Xcalibur - Scan-Level NAP - Height Difference")
+
+```
+
+::: {custom-style="MDPI_2.2_heading2"}
 ### Changes in Variance Across a Large Dataset
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 For large datasets, we expect that more correct peak intensities will result in better agreement between sample normalized peak intensities within a sample class.
 If incorrect peak intensities are being reported, then we would expect between class to within class variances to be very random, resulting in F-statistics close to 1.
+:::
 
@@
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Conversion of m/z to Frequency
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 The data consists of profile mode m/z spectra from multiple scans encoded as m/z and intensity values for each scan.
@@
 These *useful* points can be used to construct a linear model relating m/z to frequency for individual scans based on the formula:
+:::
 
 $$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
+
+::: {custom-style="MDPI_3.1_text"}
 From the known physical properties of the Orbitrap, only the square root term should be necessary [@ledfordSpaceChargeEffects1984].
@@
 To convert m/z back into frequency, we can use a similar model without the roots.
+:::
 
@@
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Frequency Intervals
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 Two types of frequency intervals are used, sliding and tiled windows.
@@
 The tiled windows are one point wide with a stride of one point.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Interval Range Based Data
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 To enable interval algebra, the frequency points are converted to single width intervals by multiplying (by a multiplier of 400 in this work), rounding to the nearest integer, and storing them as IRanges objects from the IRanges Bioconductor package [@lawrenceSoftwareComputingAnnotating2013a].
 The sliding and tiled windows are also converted to IRanges objects using this process.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Peak Containing Intervals
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 To find intervals that contain points that represent actual signal and not just random noise, the number of non-zero intensity points in each sliding window are counted.
@@
 These interval regions should contain a single **real** peak that was detected in one or more scans.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Peak Detection and Centroided Values
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package [@borchersPracmaPracticalNumerical2021].
 These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either m/z or frequency), and the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.
+:::
 
 $$\ln{intensity} = intercept + a*position+b*position^2$$
+
+::: {custom-style="MDPI_3.1_text"}
 From this weighted parabolic fit, the center, intensity, integrated area and sum-of-square residuals can be extracted for the peak.
@@
 Any regions that subsequently contain zero peaks are removed.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Scan to Scan Normalization
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 Scans are normalized to a single *reference* scan based on the log-intensity differences of a subset of peaks present in at least the same number of scans as the 95th percentile of scan counts for the peaks.
@@
 Peaks correlated to scan order represent an artifact that we speculate results from a gradient in the sample well.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Full Characterization
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 The full set of raw data points for each peak in each scan within a region is known based on the previously detected peaks.
@@
 In addition to these values, the frequency point-to-point median difference is calculated across all of the raw data points for those points that could be used for modeling frequency to m/z, and this difference of a single point from the peak center is calculated in frequency space, and converted to m/z space to provide an "offset" value that is useful to define the search space around the peak for any assignment algorithm.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Correction of Height and Standard Deviation
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 Ideally, each peak would be observed in every scan.
@@
 The corrected standard deviations can then be used to correct the mean height assuming that it is the result of a left-censored normal distribution [@TruncatedNormalDistribution2022; @burkardtTruncatedNormalDistribution].
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Frequency Standard Deviation Filtering
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 High-peak-density (HPD) artifacts [@mitchellNewMethodsIdentify2018] present as groups of singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
@@
 Sliding windows that are 1000 frequency points wide with a stride of 100 points were used for the density calculations.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Multi-Class F-Statistics
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 To evaluate the F-statistic between multi-class samples, a much larger set of samples was used.
@@
 For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using scan-level peak-characterization and those from MSnbase.
+:::
 
+::: {custom-style="MDPI_2.2_heading2"}
 ### Samples and Overall Processing
+:::
 
+::: {custom-style="MDPI_3.1_text"}
 Samples included two ethyl-chloroformate (ECF) derivatized amino-acid samples [reference] and two samples where lipid extracts were extracted from lung tumors [reference].
 All four samples are positive mode, were converted from raw to mzML using msconvert from ProteoWizard (profile mode).
+:::
 
@@
 1.  No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
-  1. Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
-  1. Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
-  1. Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
-  1. Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
-  1. Noise removal, two pass normalization, frequency SD filtering (filtersd)
-  1. Scan level peak centroids generated by MSnbase (using `pickPeaks` on the individual scans)
-  1. Scans merged and then centroids generated by MSnbase (using `combineSpectra` and `pickPeaks`)
+2.  Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
+3.  Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
+4.  Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
+5.  Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
+6.  Noise removal, two pass normalization, frequency SD filtering (filtersd)
+7.  Scan level peak centroids generated by MSnbase (using `pickPeaks` on the individual scans)
+8.  Scans merged and then centroids generated by MSnbase (using `combineSpectra` and `pickPeaks`)
 
@@
 1.  All scan level peaks within a 2ppm window are associated with the *merged* peak
-  1. Using the *filtersd* peaks, merged and scan level peaks are associated if they are within the start and stop m/z of the *filtersd* peaks
+2.  Using the *filtersd* peaks, merged and scan level peaks are associated if they are within the start and stop m/z of the *filtersd* peaks
 

### 2022-03-29 R/get_assignments.R
getting all of the amino acid Height / NAP ratio differences across both samples.
more to address #12

@@ -443,48 +443,40 @@ find_large_diffs = function(aa_diffs){
   check_largish
 }
 
-large_figure = function(difference_data, nap_height_data, xcalibur_data, aa_id = "Threonine", formula_id = "C9H17N1Na1O5.15N.0", use_data = "raw"){
+ratio_diffs_extracted = function(difference_data, aa_id = "Threonine", formula_id = "C9H17N1Na1O5.15N.0"){
   # difference_data = tar_read(aa_compared_filtersd_1ecf)
   # nap_height_data = tar_read(nap_height_1ecf)
   # aa_id = "Threonine"
   # formula_id = "C9H17N1Na1O5.15N.0"
   diff_formula = difference_data[[aa_id]][[formula_id]]
-  nap_height_formula = nap_height_data[[aa_id]][[formula_id]]
 
-  scan_level = nap_height_formula$scanlevel$ratios$Log10Height$nap_intensity
-  n_na = data.frame(ratio = seq(1, ncol(scan_level)),
-                    n_missing = apply(scan_level, 2, function(.x){
-                      sum(is.na(.x))
-                    })) %>%
-    dplyr::arrange(n_missing) %>%
-    dplyr::mutate(miss_order = seq(1, nrow(.)))
+  raw_ratios = diff_formula$data_ratios %>%
+    dplyr::mutate(source2 = source,
+                  source = dplyr::case_when(
+      grepl("char_height", source2) ~ "scan-level",
+      TRUE ~ source2
+                  ))
 
-  raw_ratios_char = nap_height_formula$characterized$ratios$Log10Height$nap_intensity_diff %>%
-      diff_vector_2_df(source = "Log10Height")
-  raw_ratios_xcal = nap_height_formula$xcalibur$ratios$nap_intensity_diff %>%
-      diff_vector_2_df(source = "xcalibur")
-  raw_ratios_diff = dplyr::left_join(raw_ratios_char,
-                                     raw_ratios_xcal, by = "ratio", suffix = c(".char", ".xcal")) %>%
-    dplyr::mutate(xcal_char = diff.xcal - diff.char)
+  raw_ratios_diff = diff_formula$method_differences
 
-  raw_ratios_diff = dplyr::left_join(raw_ratios_diff, n_na, by = "ratio")
+  all_diffs = purrr::imap_dfr(difference_data,
+                             function(in_diff, aa_id){
+                               message(aa_id)
+    diff_df = purrr::imap_dfr(in_diff,
[…]

### 2022-03-29 _targets.R

@@ -236,7 +236,11 @@ height_nap_tar = list(
   tar_target(nap_height_1ecf, aa_height_nap_all(aa_filtersd_1ecf, xcalibur_1ecf, msnbase_1ecf)),
   tar_target(nap_height_2ecf, aa_height_nap_all(aa_filtersd_2ecf, xcalibur_2ecf, msnbase_2ecf)),
   tar_target(aa_compared_filtersd_1ecf, compare_peak_ratios(nap_height_1ecf)),
-  tar_target(aa_compared_filtersd_2ecf, compare_peak_ratios(nap_height_2ecf))
+  tar_target(aa_compared_filtersd_2ecf, compare_peak_ratios(nap_height_2ecf)),
+  tar_target(aa_large_filtersd_1ecf, find_large_diffs(aa_compared_filtersd_1ecf)),
+  tar_target(aa_large_filtersd_2ecf, find_large_diffs(aa_compared_filtersd_2ecf)),
+  tar_target(aa_diffs_filtersd_1ecf, ratio_diffs_extracted(aa_compared_filtersd_1ecf)),
+  tar_target(aa_diffs_filtersd_2ecf, ratio_diffs_extracted(aa_compared_filtersd_2ecf))
 )
 
 list(pkg_tar,

### 2022-03-29 doc/peakcharacterization_manuscript.Rmd
updated styles and then actually getting the figure to render and the caption out

@@
 
-```{=html}
-<div custom-style="MDPI_5.2_figure>
+::: {custom-style="MDPI_5.2_figure"}
 
@@
 
-</div>
+:::
 
-<div custom-style="MDPI_5.1_figure_caption">
+::: {custom-style="MDPI_5.1_figure_caption"}
 **`r figure_count$label_text("slidingwindow_count")`.**
@@
 **C** and **D** show the number of initial regions as a function of the percentile cutoff used for either the AA ECF (C) or lipid (D) samples.
-</div>
-```
+:::
+
 ::: {custom-style="MDPI_3.1_text"}
@@
 
-```{=html}
-<div custom-style="MDPI_2.2_heading2>
+::: {custom-style="MDPI_2.2_heading2"}
 ### Peak Characterization Using Quadratic Fit
-</div>
+:::
 
@@
 
-<div custom-style="MDPI_3.1_text">
+::: {custom-style="MDPI_3.1_text"}
 Although many other types of mass-spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo Fusion has a baseline of 0 due to manipulations in the Thermo firmware, making the determination of the centroided values considerably easier.
@@
 From the fitted model, we can derive the centroided center and the intensity of the peak, as shown in `r figure_count$label_text("centroided_peaks")`.
-</div>
-```
-```{=html}
-<div custom-style="MDPI_5.2_figure>
+:::
+
+::: {custom-style="MDPI_5.2_figure"}
 ```{r centroided_peaks, dn_id = figure_count}
@@
 ```
-</div>
+:::
 
-<div custom-style="MDPI_5.1_figure_caption">
+::: {custom-style="MDPI_5.1_figure_caption"}
 **`r figure_count$label_text("centroided_peaks")`.**
 Log-intensity (A) and Intensity (B) of points for a single peak against frequency. Black points are the original data points, and the blue point represents the calculated centroid.
-</div>
-```
+:::
+
 ::: {custom-style="MDPI_2.2_heading2"}
@@
 
+::: {custom-style="MDPI_5.2_figure"}
 ```{r rsd_method, dn_id = figure_count}
@@
 ```
+:::
 
+::: {custom-style="MDPI_5.1_figure_caption"}
 `r figure_count$label_text("rsd_method")`.
@@
 A peak had to be present in at least three scans for the RSD value to be reported.
+:::
 
+::: {custom-style="MDPI_4.1_table_caption"}
 `r table_count$label_text("rsd_method")`.
 Relative standard deviation means, medians, modes and maximum observed values for each sample with different overall processing.
+:::
 
+::: {custom-style="MDPI_4.2_table_body"}
 ```{r rsd_table, results = 'asis'}
@@
 ```
+:::
 
@@
 
+```{r nap_intensity_increment}
+figure_count$increment("nap_intensity")
+```
+
 ::: {custom-style="MDPI_3.1_text"}
@@
 To compare relative NAP to relative abundances, we only examined the assignments from the two samples containing ECF derivatized amino-acids, as we can limit the assignments to those that match expected derivatizations of the known amino-acids (see Supplemental for the expected EMFs, and expected relative NAPs for the individual IMFs).
+
+`r figure_count$label_text("nap_intensity")` compares the peak - peak natural abundance and height log-ratio differences generated using heights from Xcalibur and from our scan-level peak characterization.
 :::
 
-```{r nap_intensity}
+::: {custom-style="MDPI_5.2_figure"}
+```{r nap_intensity, dn_id = figure_count, fig.width = 16, fig.height = 8}
+nap_fontsize = 12
 tar_load(aa_diffs_filtersd_1ecf)
@@
     labs(x = "Pairwise Comparison",
-         y = "Peak-Peak NAP - Height Difference")
+         y = "Peak-Peak NAP - Height Differences") +
+  theme(axis.title.y = element_text(size = nap_fontsize))
 
@@
     labs(x = "Number of Missing Scans",
-         y = "Xcalibur - Scan-Level NAP - Height Difference")
+         y = "Xcalibur - Scan-Level NAP - Height Differences")+
+  theme(axis.title.y = element_text(size = nap_fontsize))
 
@@
   labs(x = "Number of Missing Scans",
-         y = "Xcalibur - Scan-Level NAP - Height Difference")
+         y = "Xcalibur - Scan-Level NAP - Height Differences") +
+  theme(axis.title.y = element_text(size = nap_fontsize))
 
+out_plot = (threonine_ratioplot | threonine_diffplot) / aa_diffplot + plot_annotation(tag_levels = "A")
+out_plot
 ```
+:::
+
+::: {custom-style="MDPI_5.1_figure_caption"}
+`r figure_count$label_text("nap_intensity")`.
+**A**.
+Peak - peak NAP - intensity differences from scan-level peak heights (red) and Xcalibur peak heights (blue) from the threonine amino acid assignments with Na adduct, with point size reflecting how many peaks were missing across scans.
+**B**. The difference of Xcalibur to scan-level ratios plotted directly as a function of the number of scans the peak was not found in.
+**C**. The differences of Xcalibur to scan-level ratios for all of the amino acid assignments in EMFs with more than a single peak in both ECF samples.
+:::
 

### 2022-03-30 doc/peakcharacterization_manuscript.Rmd
introduction, heavily cribbed from the HPD paper

@@
 ::: {custom-style="MDPI_1.7_abstract"}
-**Abstract:** We introduce a novel method for characterizing peaks from direct injection FT-MS experiments developed on metabolomics datasets that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
+**Abstract:** We introduce a novel, scan-centric method for characterizing peaks from direct injection multi-scan FT-MS experiments developed on metabolomics datasets that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
 Our method incorporates several aspects designed to make the derived intensities trustworthy, and suitable for use in our previously published peak assignment algorithm SMIRFE.
-Notably, the inclusion of an intensity independent noise removal, and normalization of scan level data results in a reduction of median relative standard deviation from **X** to **Y**, and much better fit of peak intensities to relative natural abundances of assigned isotopologues.
+Notably, the use of intensity independent noise removal, and normalization of scan level data results in a much better fit of relative intensity to natural abundance probability for low abundance isotopologues that are not present in all of the acquired scans.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-What can we say about FT-MS data?
-There is more and more of it, and more of it applied to metabolomics.
-Profile mode, multi-scan direct injection FT-MS is an interesting approach because it allows us to handle cases where something might go wrong in an individual scan, while still extracting useful data.
+Fourier transform mass spectrometry (FT-MS) provides high performance in terms of sensitivity, resolution and mass accuracy all in one analytical instrumentation.
+The combination of these capabilities provide several analytical and interpretive improvements including: (i) the ability to resolve distinct isotopologues with identical unit masses but different accurate masses [@higashiStableIsotopeLabeledTracers2014], enabling multi-element isotopic natural abundance correction [@moseleyCorrectingEffectsNatural2010; @carreerComputationalFrameworkHighThroughput2013]; (ii) improved assignment accuracy [@kindMetabolomicDatabaseAnnotations2006]; and (iii) the detection of metabolites in the sub-femtomolar range [@eylesMethodsStudyProtein2004; @dettmerMassSpectrometrybasedMetabolomics2007].
+In the metabolomics field, these improvements permit more complicated, but more informative experimental designs such as the use of multiple isotope-labeled precursors in stable isotope-resolved metabolomics (SIRM) experiments [@yangChloroformateDerivatizationTracing2017.
+Multiple isotope-labeling experiments can provide the information needed to elucidate unknown metabolic pathways [@creekStableIsotopeAssistedMetabolomics2012; @higashiStableIsotopeLabeledTracers2014], quantify relative fluxes through multiple metabolic pathways [@hillerNontargetedElucidationMetabolic2010], identify metabolite pools in different celullar compartments [@fanStableIsotoperesolvedMetabolomics2012], and identify active metabolic pathways under various cellular conditions [@moseleyNovelDeconvolutionMethod2011; @sellersPyruvateCarboxylaseCritical2015; @verdegemMAIMSSoftwareTool2017].
+These gains in information enable more complete modeling of cellular metabolism and better understanding of physiological and pathological processes at a mechanistic level, facilitating the identification of potential therapeutic targets [@fanAlteredRegulationMetabolic2009] and the quantification of differential drug response [@harrisDiverseMechanismsGrowth2012].
+
+While the advantages of FT-MS are significant, when deployed in a high-throughput environment, the volume of data produced requires automated tools for data reduction, quality control, feature assignment, and downstream analyses.
+These requirements are amplified when FT-MS is used with direct infusion where existing and well-validated assignment methods that rely on orthogonal sources of information (such as chromatography) are either not applicable or would further increase the amount of data to be processed.
+Therefore, without automation, rigorous assignment of non-polymeric biomolecules in untargeted MS analyses remains difficult even with FT-MS’s high resolution.
+
+Within this high-throughput environment we previously developed methods to remove high-peak-density artifacts [@mitchellNewMethodsIdentify2018], annotate peaks with assignments using SMIRFE [@mitchellSmallMoleculeIsotope2019], and generate lipid classifications for those peak assignments [@mitchellDerivingLipidClassification2020].
+All of these methods together culminated in detecting differential lipids in non-small-cell lung cancer (NSCLC) [@mitchellUntargetedLipidomicsNonSmall2021].
+
+The input peak-data in these studies are derived from multi-scan direct-injection FT-MS data.
+SMIRFE in particular uses expected relationships of relative peak height or intensity to natural abundance probability (NAP) across multiple peaks to determine the likelihood of an assignment for the peaks in question.
+In the Thermo-Fisher Orbitrap FT-MS instruments, data is acquired in micro-scans and scans, where each scan is an aggregate of multiple micro-scans.
+Depending on the sample source, it can be advantageous to keep the data at the scan level, so that any given "bad" scan can be removed prior to aggregating the data across scans and generating a centroided peak m/z and intensity.
+If the scan level data is not processed and aggregated correctly, the final peak intensities and / or m/z's will be incorrect.
+In particular, if peaks are missing in some scans due to low abundance, the final aggregate peak intensities will not fit the expected relative intensity to NAP relationships.
+Here we describe novel scan-centric FT-MS metabolomics data processing methods that better preserve the expected relative intensity to NAP.
+In addition, due to the tendency of increasing point spacing with increasing m/z in Orbitrap type instruments, we derive a method for transforming the m/z data to axial frequency, which has the desirable property of being equally spaced across the full spectrum.
+Finally, we show that in addition to better preserving the expected intensity - NAP relationships, our scan-centric method also results in improved relative standard deviations, automatic identification of high-peak-density artifact peaks, and separation of samples in a lipidomics data analysis.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur (a process we have observed being used to generate datasets), as well as merging scans (see Methods) in the MSnbase R package.
+To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur (a process we have observed being used to generate datasets), as well as merging scans (see Methods) with the MSnbase R package.
 In all cases, for a known assigned set of peaks, the intensity ratios between the assigned peaks deviate from the intensity ratio predicted by their natural abundance probabilities (NAPs), as shown in Figure X, with some solutions giving different amounts of error.

### 2022-03-30 R/strip_mdpi.R
ability to keep or strip out the MDPI formatting shit

@@ -1,0 +1,24 @@
+strip_mdpi = function(in_rmd = "doc/peakcharacterization_manuscript.Rmd", out_rmd = "doc/peakcharacterization_nostyle.Rmd"){
+  rmd_doc = readLines(in_rmd)
+
+  template_line = grepl("metabolites-template", rmd_doc)
+  rmd_notemplate = rmd_doc[!template_line]
+
+  has_style = grepl(":::", rmd_notemplate)
+  rmd_nostyle = rmd_notemplate[!has_style]
+
+  cat(rmd_nostyle, file = out_rmd, sep = "\n")
+  out_rmd
+}
+
+strip_headers = function(in_rmd = "doc/peakcharacterization_manuscript.Rmd",
+                         out_rmd = "doc/peakcharacterization_mdpi.Rmd"){
+
+  rmd_doc = readLines(in_rmd)
+
+  has_header = grepl("^##+ \\w+", rmd_doc)
+  doc_out = rmd_doc
+  doc_out[has_header] = gsub("^##+ ", "", rmd_doc[has_header])
+  cat(doc_out, file = out_rmd, sep = "\n")
+  out_rmd
+}

### 2022-03-30 _targets.R

@@ -119,7 +119,11 @@ figures_tar = tar_plan(
              create_noise_plot(noise_combine)),
   tar_target(mn_ratios,
              calculate_m_n_ratio(noise_combine)),
-  tar_render(manuscript, "doc/peakcharacterization_manuscript.Rmd")
+  tar_render(manuscript, "doc/peakcharacterization_manuscript.Rmd"),
+  tar_target(rmd_no_mdpi, strip_mdpi("doc/peakcharacterization_manuscript.Rmd")),
+  tar_target(rmd_mdpi, strip_headers("doc/peakcharacterization_manuscript.Rmd")),
+  tar_render(manuscript_mdpi, "doc/peakcharacterization_mdpi.Rmd"),
+  tar_render(manuscript_nostyle, "doc/peakcharacterization_nostyle.Rmd")
 )
 
 tables_tar = tar_plan(

### 2022-03-30 R/get_assignments.R
making these do rendering b/c otherwise things are weird trying to force it.

@@ -480,3 +480,37 @@ ratio_diffs_extracted = function(difference_data, aa_id = "Threonine", formula_i
                               diffs = raw_ratios_diff),
        all_diffs = all_diffs)
 }
+
+motivating_plot = function(nap_height_1ecf, aa_id = "Threonine", formula_id = "C9H17N1Na1O5.15N.0"){
+  #tar_load(nap_height_1ecf)
+  use_nap_height = nap_height_1ecf[[aa_id]][[formula_id]]
+
+  peak_data = use_nap_height$xcalibur$peaks %>%
+    dplyr::arrange(ObservedMZ)
+  peak_data$ObservedMZ[4] = peak_data$ObservedMZ[4] + 0.01
+  peak_data = peak_data %>%
+    dplyr::mutate(NAP = 10^NAP,
+                  intensity = 10^intensity,
+                  relNAP = NAP / max(NAP),
+                  intNAP = relNAP * max(intensity),
+                  log10intNAP = log10(intNAP),
+                  log10intensity = log10(intensity),
+                  mzNAP = ObservedMZ + 0.05)
+
+  peak_data = dplyr::left_join(peak_data, nap_height_1ecf[[aa_id]][[formula_id]]$characterized$peaks[, c("Sample_Peak", "complete_IMF")], by = "Sample_Peak")
+
+  out_plot = peak_data %>%
+    ggplot(aes(x = ObservedMZ, xend = ObservedMZ,
+               y = 0, yend = log10intensity)) +
+    geom_segment() +
+    geom_segment(aes(x = mzNAP, xend = mzNAP,
+                     y = 0, yend = log10intNAP),
+                 color = "red") +
+    annotate("text", x = 242.45, y = 7, label = peak_data$complete_IMF[1]) +
+    annotate("text", x = 243.3, y = 6.1, label = peak_data$complete_IMF[2]) +
+    annotate("text", x = 243.7, y = 5, label = peak_data$complete_IMF[3]) +
+    annotate("text", x = 243.75, y = 4.2, label = peak_data$complete_IMF[3]) +
+    labs(x = "m/z", y = "Log10(Intensity)") +
+    scale_y_continuous(expand = expansion(mult = c(0, 0.05)))
+  out_plot
+}

### 2022-03-30 R/strip_mdpi.R

@@ -1,4 +1,4 @@
-strip_mdpi = function(in_rmd = "doc/peakcharacterization_manuscript.Rmd", out_rmd = "doc/peakcharacterization_nostyle.Rmd"){
+strip_mdpi_render = function(in_rmd = "doc/peakcharacterization_manuscript.Rmd", out_rmd = "doc/peakcharacterization_nostyle.Rmd"){
   rmd_doc = readLines(in_rmd)
 
   template_line = grepl("metabolites-template", rmd_doc)
@@ -8,17 +8,21 @@ strip_mdpi = function(in_rmd = "doc/peakcharacterization_manuscript.Rmd", out_rm
   rmd_nostyle = rmd_notemplate[!has_style]
 
   cat(rmd_nostyle, file = out_rmd, sep = "\n")
+  rmarkdown::render(out_rmd, knit_root_dir = getwd(), quiet = TRUE)
   out_rmd
 }
 
-strip_headers = function(in_rmd = "doc/peakcharacterization_manuscript.Rmd",
+strip_headers_render = function(in_rmd = "doc/peakcharacterization_manuscript.Rmd",
                          out_rmd = "doc/peakcharacterization_mdpi.Rmd"){
 
   rmd_doc = readLines(in_rmd)
 
   has_header = grepl("^##+ \\w+", rmd_doc)
+  has_bullet = grepl("^\\* ", rmd_doc)
   doc_out = rmd_doc
   doc_out[has_header] = gsub("^##+ ", "", rmd_doc[has_header])
+  doc_out[has_bullet] = gsub("^\\* ", "", rmd_doc[has_bullet])
   cat(doc_out, file = out_rmd, sep = "\n")
+  rmarkdown::render(out_rmd, knit_root_dir = getwd(), quiet = TRUE)
   out_rmd
 }

### 2022-03-30 doc/peakcharacterization_manuscript.Rmd
finally some motivation

@@
 The input peak-data in these studies are derived from multi-scan direct-injection FT-MS data.
-SMIRFE in particular uses expected relationships of relative peak height or intensity to natural abundance probability (NAP) across multiple peaks to determine the likelihood of an assignment for the peaks in question.
+SMIRFE uses expected relationships of relative peak height or intensity to natural abundance probability (NAP) across multiple peak pairs to determine the likelihood of an assignment for the peaks in question.
 In the Thermo-Fisher Orbitrap FT-MS instruments, data is acquired in micro-scans and scans, where each scan is an aggregate of multiple micro-scans.
@@
 If the scan level data is not processed and aggregated correctly, the final peak intensities and / or m/z's will be incorrect.
+
 In particular, if peaks are missing in some scans due to low abundance, the final aggregate peak intensities will not fit the expected relative intensity to NAP relationships.
@@
 
+```{r increment_motivation}
+figure_count$increment("motivation")
+```
+
 ::: {custom-style="MDPI_3.1_text"}
-To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur (a process we have observed being used to generate datasets), as well as merging scans (see Methods) with the MSnbase R package.
-In all cases, for a known assigned set of peaks, the intensity ratios between the assigned peaks deviate from the intensity ratio predicted by their natural abundance probabilities (NAPs), as shown in Figure X, with some solutions giving different amounts of error.
-We do note that when all of the peaks are present in a single scan, their relative intensity ratios are much closer to the theoretical ratios based on NAP.
+To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur as well as our scan-centric peak characterization, and found matching Xcalibur peaks to our assignments.
+As an example, in `r figure_count$label_text("motivation")` we show the Xcalibur calculated intensities and theoretical peak intensities based on NAP for four peaks matching threonine.
+As the NAP gets lower, the deviations of the Xcalibur calculated intensities from the NAP become larger.
+:::
+
+::: {custom-style="MDPI_5.2_figure"}
+```{r motivation, dn_id = figure_count, fig.width = 10, fig.height = 6}
+tar_load(motivation_plot)
+motivation_plot
+```
+:::
+
+::: {custom-style="MDPI_5.1_figure_caption"}
+`r figure_count$label_text("motivation")`.
+Xcalibur intensities (black) and theoretical intensities based on relative NAP (red) for ECF derived threonine.
+The 18O isotopologue is shifted by 0.01 m/z and the NAP peaks by 0.05 m/z to aid visualization.
 :::

### 2022-03-30 doc/peakcharacterization_manuscript.Rmd
more styling

@@
 
+::: {custom-style="MDPI_2.1_heading1"}
 ## Discussion
+:::
 
+::: {custom-style="MDPI_3.1_text"}
+we should have some discussion here about the results
+:::
+
+::: {custom-style="MDPI_2.1_heading1"}
 ## Materials and Methods
+:::
 
@@
 
+::: {custom-style="MDPI_3.9_equation"}
 $$mz = intercept + x* \frac{1}{frequency} + y*\frac{1}{frequency^2} + z*\frac{1}{frequency^3}$$
+:::
 
@@
 ::: {custom-style="MDPI_2.2_heading2"}
+### Assigning Scan-Centric Peaks
+:::
+
+We created potential EMF databases for SMIRFE assignments for ECF and 15N labeled amino acids or for lipids (see supplemental materials for the actual limits).
+We assigned ECF derivatized amino acid samples with 15N labeling, a single positive charge, and H and Na as adducts.
+We assigned lipid samples with no labeling, a single positive charge, and H, Na, K, and NH4 as possible adducts.
+We attempted to create peaks that could be assigned by SMIRFE from the MSnbase and Xcalibur peak lists, but SMIRFE would not assign them.
+
+::: {custom-style="MDPI_2.2_heading2"}
+### Matching Peaks
+:::
+
+We matched the scan-centric peaks to the other methods by looking for the closest m/z peak that was within 2 ppm of the scan-centric reported m/z.
+
+::: {custom-style="MDPI_2.2_heading2"}
+### Peak - Peak NAP Height Ratios
+:::
+
+::: {custom-style="MDPI_2.2_heading2"}
 ### Multi-Class F-Statistics
@@
 
+::: {custom-style="MDPI_3.5_text_before_list"}
 For each raw data file in profile mzML format, they were processed in these six ways:
+:::
 
-1.  No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
-2.  Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
-3.  Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
-4.  Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
-5.  Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
-6.  Noise removal, two pass normalization, frequency SD filtering (filtersd)
-7.  Scan level peak centroids generated by MSnbase (using `pickPeaks` on the individual scans)
-8.  Scans merged and then centroids generated by MSnbase (using `combineSpectra` and `pickPeaks`)
+::: {custom-style="MDPI_3.8_bullet"}
+* No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
+:::
+::: {custom-style="MDPI_3.8_bullet"}
+* Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
+:::
+::: {custom-style="MDPI_3.8_bullet"}
+* Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
+:::
+::: {custom-style="MDPI_3.8_bullet"}
+* Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
+:::
+::: {custom-style="MDPI_3.8_bullet"}
+* Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
+:::
+::: {custom-style="MDPI_3.8_bullet"}
+* Noise removal, two pass normalization, frequency SD filtering (filtersd)
+:::
+::: {custom-style="MDPI_3.8_bullet"}
+* Scan level peak centroids generated by MSnbase (using `pickPeaks` on the individual scans)
+:::
+::: {custom-style="MDPI_3.8_bullet"}
+* Scans merged and then centroids generated by MSnbase (using `combineSpectra` and `pickPeaks`)
+:::
 
+::: {custom-style="MDPI_3.5_text_before_list"}
 For the two sets of MSnbase peaks, we also matched scan level peaks to merged peaks via two methods:
+:::
 
-1.  All scan level peaks within a 2ppm window are associated with the *merged* peak
-2.  Using the *filtersd* peaks, merged and scan level peaks are associated if they are within the start and stop m/z of the *filtersd* peaks
+::: {custom-style="MDPI_3.8_bullet"}
+* All scan level peaks within a 2ppm window are associated with the *merged* peak
+:::
 
+::: {custom-style="MDPI_3.8_bullet"}
+* Using the *filtersd* peaks, merged and scan level peaks are associated if they are within the start and stop m/z of the *filtersd* peaks
+:::
+
+::: {custom-style="MDPI_2.1_heading1"}
 ## Conclusions
+:::
 
+::: {custom-style="MDPI_2.1_heading1"}
 ## References
+:::

### 2022-03-31 doc/peakcharacterization_manuscript.Rmd
whole lot of methods
Pulling in methods for the samples from previous publications, including SMIRFE and NSCLC

@@
 
+::: {custom-style="MDPI_3.9_equation"}
 $$\ln{intensity} = intercept + a*position+b*position^2$$
+:::
 
@@
 
-We created potential EMF databases for SMIRFE assignments for ECF and 15N labeled amino acids or for lipids (see supplemental materials for the actual limits).
-We assigned ECF derivatized amino acid samples with 15N labeling, a single positive charge, and H and Na as adducts.
-We assigned lipid samples with no labeling, a single positive charge, and H, Na, K, and NH4 as possible adducts.
+::: {custom-style="MDPI_3.1_text"}
+Our previously described SMIRFE algorithm [@mitchellSmallMoleculeIsotope2019] was used to assign molecular formulas to scan-centric characterized peaks in an untargeted manner. 
+For the lipid samples, an initial EMF database was generated using an M/Z limit of 1605 M/Z, and maximum numbers for each element were set to C: 130, N: 7, O: 28, P: 3, H: 230. 
+Assigned formulas were allowed to have K, Na, H, and NH4 adducts (only positive mode samples were assigned). 
+Assigned molecular formulas were then classified into one or more lipid categories using our lipid classifier tool [@mitchellDerivingLipidClassification2020]. 
+Lipid classes that were previously observed to be overclassified by our models (neutral and acidic glycosphingolipids) were excluded.
+For the ECF derivatized amino-acid samples, the EMF database was generated using an m/z limit of 1005 m/z, and maximum numbers for each element were set to C: 100, N: 7, O: 40, H: 230, P: 3, S: 3.
 We attempted to create peaks that could be assigned by SMIRFE from the MSnbase and Xcalibur peak lists, but SMIRFE would not assign them.
+:::
+
+::: {custom-style="MDPI_2.2_heading2"}
+Consistently Assigned Lipid Spectral Feature (Corresponded Peak) Generation and Peak Intensity Normalization
+:::
+
+Our SMIRFE assignment method assigns isotope-resolved molecular formulas (IMFs) to characterized peaks in each spectrum. 
+Each IMF represents an isotopologue of a given elemental molecular formula (EMF) (e.g., 13C112C51H1216O6 is an IMF representing the m+13C1 isotopologue of EMF C6H12O6).
+Consistently assigned (i.e., corresponded peaks) are identified using an in-house method we have named EMF voting which is described in Appendix A.
+EMF voting identified 3485 total corresponded peaks across all 157 spectra with 855 corresponded peaks present in 25% or more of either the disease or non-disease group of samples. 
+To minimize false assignment and their effects, non_lipid and multiple lipids were not considered in downstream analyses. 
+All lipid isotopologue intensities were normalized by dividing the isotopologue intensity by the median intensity of all the peaks in the sample. 
+For differential abundance analysis, missing values were replaced with a threshold value that is ½ of the lower confidence interval of the distribution of all log-transformed intensities from that tissue.
 
@@
 :::
+
+Each pair of peaks in a specifically labeled and adducted elemental molecular formula are related by their natural abundance probability (NAP).
+Theoretically, the log-ratios of two peaks NAPs are approximately equal to the log-ratios of two peaks intensities.
+
+::: {custom-style="MDPI_3.9_equation"}
+$$log(NAP_{P1} / NAP_{P2}) - log(Int_{P1} / Int_{P2}) \approx 0$$
+:::
+
+SMIRFE assignments include the NAP for each of the isotopologue molecular formulas (IMFs) in the particular labelled and adducted EMF.
+For each peak pair in the peak group for a particular EMF, we can calculate the log-ratio for the NAPs and the log-ratio for the intensities, and their absolute differences.
+This calculation is done for the final characterized intensity, as well as the intensities at the scan level, and for both the *raw* intensity and *corrected* intensity, and for matched Xcalibur peak intensities, and matched MSnbase peak intensities if there were two or more matching peaks in the labeled, adducted EMF.
 
@@
 ::: {custom-style="MDPI_3.1_text"}
-Samples included two ethyl-chloroformate (ECF) derivatized amino-acid samples [reference] and two samples where lipid extracts were extracted from lung tumors [reference].
-All four samples are positive mode, were converted from raw to mzML using msconvert from ProteoWizard (profile mode).
+The method for generating the amino-acid dataset was adapted from a previously published method for performing ethylchloroformate (ECF) amino acid derivatization [@yangChloroformateDerivatizationTracing2017].
+Two replicate samples were prepared and spectra were obtained for both samples using a Tribrid Fusion Orbitrap at 500k resolution and from 150 to 1000m/z.
+
+The collection, preparation, and mass spectrometry analysis of the paired cancer and non-cancer samples has been previously described [@sellersPyruvateCarboxylaseCritical2015]. 
+In summary, cancer and non-cancer samples were acquired from eight-six non-diabetic patients with suspected resectable stage I or IIA non-small cell lung cancer. 
+Written informed consent was collected from all subjects prior to inclusion and all samples were collected under a University of Louisville or University of Kentucky IRB protocol. 
+Lipid extracts were prepared using a modified Folch extraction and reconstituted for direct infusion ultra-high resolution mass spectrometry on a pair of Thermo Tribrid Fusion Orbitrap instruments (FSN10115 and FSN10352, referred to as Fusion 1 and Fusion 2 respectively) coupled to an Advion nanoelectrospray system.
+53 patients (102 spectra total) were acquired from Fusion 1 and 40 patients (77 spectra total) were acquired from Fusion 2.
+Fusion 1 samples were exclusively from the University of Louisville while Fusion 2 were a mix.
+Three of the patients had only a cancer or non-cancer sample acquired.
+Two of the cancer lipid samples (49Cpos and 97Cpos) were used for examples in this work.
+The rest were used for the examination of F-statistics.
 :::

### 2022-03-31 doc/peakcharacterization_manuscript.Rmd
got unassigned peak matching and description of results done
related to #11

@@
 ::: {custom-style="MDPI_2.2_heading2"}
+### Method Specific Peaks
+:::
+
+```{r increment_peak_comparisons}
+table_count$increment("unassigned_1ecf")
+table_count$increment("unassigned_97lipid")
+
+figure_count$increment("unassigned_upset")
+
+tar_load(um_peaks_1ecf)
+tar_load(um_peaks_97lipid)
+```
+
+::: {custom-style="MDPI_3.1_text"}
+Each set of peaks generated may or may not be specific to the particular method used to generate centroided peak m/z and intensities.
+This is true for the different scan-centric combinations, as well as the peaks from Xcalibur and MSnbase.
+Here we examine the overlap of the unassigned (`r figure_count$label_text("unassigned_upset")`) and assigned peaks from the full scan-centric processing with Xcalibur and MSnbase.
+These same counts are also summarized in `r table_count$label_text("unassigned_1ecf")` and `r table_count$label_text("unassigned_97lipid")`.
+In these two examples, there are some striking differences.
+The 1ecf sample has all of the scan-centric peaks shared with either Xcalibur or MSnbase peaks, whereas the 97lipid sample has 2/3 of the peaks specific to scan-centric characterization and not matched to either of the other methods.
+Notably, for both samples, the scan-centric characterization produces similar numbers of peaks, even though the upper mass limit in 1ecf is 1000 m/z compared to 1600 m/z for the 97lipid sample.
+Whereas the number of peaks from MSnbase and Xcalibur are three-fold and 40-fold higher in the 1ecf sample compared to the 97lipid sample.
+:::
+
+```{r unassigned_upset, dn_id = figure_count, fig.width = 16, fig.height = 8}
+set_order = c("scancentric", "msnbase", "xcalibur")
+u1 = grid::grid.grabExpr(draw(UpSet(um_peaks_1ecf$comb_matrix, 
+                           set_order = set_order)))
+u2 = grid::grid.grabExpr(draw(UpSet(um_peaks_97lipid$comb_matrix, 
+                           set_order = set_order)))
+
+up_plot = wrap_elements(u1, clip = FALSE) + wrap_elements(u2, clip = FALSE)
+up_plot + plot_annotation(tag_levels = "A")
+```
+
+::: {custom-style="MDPI_5.1_figure_caption"}
+`r figure_count$label_text("unassigned_upset")`.
+UpSet plot with the counts of common and specific peaks for each of scancentric, MSnbase and XCalibur generated peaks for the 1ecf (**A**) and 97lipid (**B**) samples.
+:::
+
+::: {custom-style="MDPI_4.1_table_caption"}
+`r table_count$label_text("unassigned_1ecf")`.
+Number of matched peaks between peak processing methods for 1ecf.
+:::
+::: {custom-style="MDPI_4.2_table_body"}
+```{r unassigned_1ecf}
+um_peaks_1ecf$ft_table %>%
+  autofit()
+```
+:::
+
+::: {custom-style="MDPI_4.1_table_caption"}
+`r table_count$label_text("unassigned_97lipid")`.
+Number of matched peaks between peak processing methods for 97lipid.
+:::
+::: {custom-style="MDPI_4.2_table_body"}
+```{r unassigned_97lipid}
+um_peaks_97lipid$ft_table %>%
+  autofit()
+```
+:::
+
+::: {custom-style="MDPI_2.2_heading2"}
 ### Changes in Variance Across a Large Dataset

### 2022-04-01 doc/peakcharacterization_manuscript.Rmd
added set intersection and mz differences
I think this closes #11

@@
     geom_point() +
-    theme(legend.position = c(0.7, 0.8)) +
+    theme(legend.position = c(0.7, 0.75)) +
     labs(x = "Pairwise Comparison",
@@
 figure_count$increment("unassigned_upset")
+figure_count$increment("assigned_mz_diff")
 
@@
 tar_load(um_peaks_97lipid)
+tar_load(ass_peaks_1ecf)
+tar_load(ass_peaks_97lipid)
 ```
@@
 ```
+:::
+
+::: {custom-style="MDPI_5.2_figure"}
+```{r assigned_mz_diff, dn_id = figure_count}
+diffs_1ecf = ass_peaks_1ecf$peaks %>%
+  dplyr::mutate(source = factor(source, levels = c("scanlevel", "xcalibur", "msnbase"), ordered = TRUE))
+diffs_97lipid = ass_peaks_97lipid$peaks %>%
+  dplyr::mutate(source = factor(source, levels = c("scanlevel", "xcalibur", "msnbase"), ordered = TRUE))
+
+diffs_plot = (diffs_1ecf %>%
+                ggplot(aes(x = mz_diff)) +
+  geom_histogram(bins = 30) +
+  facet_wrap(~ source, ncol = 1) +
+  labs(x = "m/z Difference")) |
+  (diffs_97lipid %>%
+     ggplot(aes(x = mz_diff)) +
+     geom_histogram(bins = 30) +
+     facet_wrap(~ source, ncol = 1) +
+     labs(x = "m/z Difference")) +
+  plot_annotation(tag_levels = "A")
+diffs_plot
+```
+:::
+
+::: {custom-style="MDPI_5.1_figure_caption"}
+`r figure_count$label_text("assigned_mz_diff")`.
+Histograms of the m/z differences to the assigned and matched peak for 1ecf (**A**) and 97lipid (**B**).
 :::

### 2022-04-01 R/get_assignments.R

@@ -14,7 +14,12 @@ check_nap_intensity = function(nap_df){
   nap_int_ratios = calculate_log_ratio_differences(nap_df)
 }
 
-find_aa_assignments = function(assign_data, aa_formulas, e_cutoff = NULL){
+find_assignments = function(assign_data, formulas_df = NULL, e_cutoff = NULL){
+  # assign_data = tar_read(emf_filtersd_97lipid)
+  # assign_data = tar_read(emf_filtersd_1ecf)
+  # formulas_df = tar_read(aa_formula)
+  # e_cutoff = 0.1
+  # e_cutoff = NULL
   if (is.null(e_cutoff)) {
     scored_assignments = smirfeTools::score_filter_assignments(assign_data)
   } else {
@@ -23,21 +28,32 @@ find_aa_assignments = function(assign_data, aa_formulas, e_cutoff = NULL){
 
   within_emfs = smirfeTools:::get_sample_emfs(scored_assignments$assignments, assign_data$sample, evalue_cutoff = 0.98, use_corroborating = TRUE)
 
-  possible_aa = purrr::map_lgl(within_emfs, function(in_emf){
-    if (any(aa_formulas$complete_EMF %in% in_emf$complete_EMF)) {
-      TRUE
-    } else {
-      FALSE
-    }
-  })
-  within_aa = within_emfs[possible_aa]
-  all_aa = purrr::map_df(within_aa, ~ .x)
-  all_aa = dplyr::left_join(all_aa, dplyr::select(aa_formulas, -isotopologue_EMF), by = "complete_EMF")
-  all_aa = dplyr::left_join(dplyr::select(all_aa, -PeakID, -Sample, -ObservedMZ), scored_assignments$data, by = "Sample_Peak")
+  if (!is.null(formulas_df)) {
+    possible_formula = purrr::map_lgl(within_emfs, function(in_emf){
+      if (any(formulas_df$complete_EMF %in% in_emf$complete_EMF)) {
+        TRUE
+      } else {
+        FALSE
+      }
+    })
+    within_list = within_emfs[possible_formula]
+    all_list = purrr::map_df(within_list, ~ .x)
+    all_list = dplyr::left_join(all_list, dplyr::select(formulas_df, -isotopologue_EMF), by = "complete_EMF")
+  } else {
+    all_list = purrr::map_df(within_emfs, ~ .x)
+  }
 
[…]

### 2022-04-01 R/match_peaks.R

@@ -1,11 +1,53 @@
-match_assigned_peaks = function(filtersd_assign, xcalibur_peaks, msnbase_peaks){
-  NULL
+match_assigned_peaks = function(filtersd_assign, xcalibur_data, msnbase_data){
+  # filtersd_assign = tar_read(aa_filtersd_1ecf)
+  # xcalibur_data = tar_read(xcalibur_1ecf)
+  # msnbase_data = tar_read(msnbase_1ecf)
+  xcalibur_peaks = xcalibur_data$comb
+  xcalibur_peaks$PeakID = paste0("xcalibur_", seq(1, nrow(xcalibur_peaks)))
+  xcalibur_peaks$rowid = seq(1, nrow(xcalibur_peaks))
+  msnbase_peaks = msnbase_data$comb
+  msnbase_peaks$PeakID = paste0("msnbase_", seq(1, nrow(msnbase_peaks)))
+  msnbase_peaks$rowid = seq(1, nrow(msnbase_peaks))
+  assigned_peaks = purrr::map_df(filtersd_assign$emfs, ~ .x)
+  theoretical_peaks = assigned_peaks %>%
+    dplyr::transmute(theoreticalmz = ObservedMZ - mass_error,
+                     ppm = theoreticalmz * 2e-6,
+                     lowmz = theoreticalmz - ppm,
+                     highmz = theoreticalmz + ppm)
+  theoretical_peaks$PeakID = paste0("scanlevel_", seq(1, nrow(theoretical_peaks)))
+  scanlevel_peaks = assigned_peaks %>%
+    dplyr::transmute(mz = ObservedMZ,
+                     intensity = Height)
+  scanlevel_peaks$PeakID = theoretical_peaks$PeakID
+  scanlevel_peaks$rowid = seq(1, nrow(scanlevel_peaks))
+  scanlevel_peaks$mz_diff = scanlevel_peaks$mz - theoretical_peaks$theoreticalmz
+  scanlevel_peaks$source = "scanlevel"
+
+  xcalibur_matched = sub_match(theoretical_peaks, xcalibur_peaks) %>%
+    dplyr::filter(grepl("^scanlevel", PeakID)) %>%
+    dplyr::mutate(source = "xcalibur")
+  msnbase_matched = sub_match(theoretical_peaks, msnbase_peaks) %>%
+    dplyr::filter(grepl("^scanlevel", PeakID)) %>%
+    dplyr::mutate(source = "msnbase")
+
[…]

### 2022-04-01 _targets.R

@@ -227,8 +227,29 @@ formula_tar = tar_target(aa_formula,
 aa_tar = tar_map(
   values = aa_methods,
   names = "name",
-  tar_target(aa_high, find_aa_assignments(in_assign, aa_formula)),
-  tar_target(aa, find_aa_assignments(in_assign, aa_formula, 0.1))
+  tar_target(aa_high, find_assignments(in_assign, aa_formula)),
+  tar_target(aa, find_assignments(in_assign, aa_formula, 0.1))
+)
+
+lipid_methods = expand_grid(
+  assign = "emf",
+  method = c("noperc_nonorm",
+             "perc99_nonorm",
+             "singlenorm",
+             "intsinglenorm",
+             "doublenorm",
+             "filtersd"),
+  sample = c("97lipid",
+             "49lipid")
+) %>%
+  dplyr::mutate(in_assign = rlang::syms(paste0(assign, "_", method, "_", sample)),
+                name = paste0(method, "_", sample))
+
+lipid_tar = tar_map(
+  values = lipid_methods,
+  names = "name",
+  tar_target(lipid_high, find_assignments(in_assign)),
+  tar_target(lipid, find_assignments(in_assign, e_cutoff = 0.1))
 )
 
 xcalibur_df = tibble(
@@ -270,9 +291,9 @@ assigned_match_tar = list(
   tar_target(ass_peaks_2ecf,
              match_assigned_peaks(aa_filtersd_2ecf, xcalibur_2ecf, msnbase_2ecf)),
   tar_target(ass_peaks_97lipid,
-             match_assigned_peaks(aa_filtersd_97lipid, xcalibur_97lipid, msnbase_97lipid)),
+             match_assigned_peaks(lipid_filtersd_97lipid, xcalibur_97lipid, msnbase_97lipid)),
   tar_target(ass_peaks_49lipid,
-             match_assigned_peaks(aa_filtersd_49lipid, xcalibur_49lipid, msnbase_49lipid))
+             match_assigned_peaks(lipid_filtersd_49lipid, xcalibur_49lipid, msnbase_49lipid))
 )
 
 list(pkg_tar,
@@ -292,4 +313,5 @@ list(pkg_tar,
      height_nap_tar,
      unassigned_match_tar,
      assigned_match_tar,
-     aa_tar)
+     aa_tar,
+     lipid_tar)

### 2022-04-04 doc/peakcharacterization_manuscript.Rmd
missed some formatting issues

@@
 ::: {custom-style="MDPI_2.2_heading2"}
-Consistently Assigned Lipid Spectral Feature (Corresponded Peak) Generation and Peak Intensity Normalization
+### Consistently Assigned Lipid Spectral Feature (Corresponded Peak) Generation and Peak Intensity Normalization
 :::
 
+::: {custom-style="MDPI_3.1_text"}
 Our SMIRFE assignment method assigns isotope-resolved molecular formulas (IMFs) to characterized peaks in each spectrum. 
@@
 For differential abundance analysis, missing values were replaced with a threshold value that is ½ of the lower confidence interval of the distribution of all log-transformed intensities from that tissue.
+:::
 
@@
 
+::: {custom-style="MDPI_3.1_text"}
 We matched the scan-centric peaks to the other methods by looking for the closest m/z peak that was within 2 ppm of the scan-centric reported m/z.
+:::
 
@@
 
+::: {custom-style="MDPI_3.1_text"}
 Each pair of peaks in a specifically labeled and adducted elemental molecular formula are related by their natural abundance probability (NAP).
 Theoretically, the log-ratios of two peaks NAPs are approximately equal to the log-ratios of two peaks intensities.
+:::
 

### 2022-04-04 doc/peakcharacterization_manuscript.Rmd
incorporating @hunter-moseley latest revisions

@@
 ---
-title: "Scan-Centric, Frequency-Based Method for Characterizing Peaks from Direct Injection FT-MS Experiments"
+title: "Scan-Centric, Frequency-Based Method for Characterizing Peaks from Direct Injection Fourier transform Mass Spectrometry Experiments"
 author:
@@
 ::: {custom-style="MDPI_1.7_abstract"}
-**Abstract:** We introduce a novel, scan-centric method for characterizing peaks from direct injection multi-scan FT-MS experiments developed on metabolomics datasets that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
-Our method incorporates several aspects designed to make the derived intensities trustworthy, and suitable for use in our previously published peak assignment algorithm SMIRFE.
-Notably, the use of intensity independent noise removal, and normalization of scan level data results in a much better fit of relative intensity to natural abundance probability for low abundance isotopologues that are not present in all of the acquired scans.
+**Abstract:** We introduce a novel, scan-centric method for characterizing peaks from direct injection multi-scan Fourier transform mass spectra of complex samples that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
+Our peak characterization method utilizes intensity independent noise removal and normalization of scan level data to provide a much better fit of relative intensity to natural abundance probabilities for low abundance isotopologues that are not present in all of the acquired scans.
+Moreover, our method calculates both peak- and scan-specific statistics incorporated within a series of quality control steps that are designed to robustly derive peak centers, intensities and intensity ratios with their scan-level variances.
+These cross-scan characterized peaks are suitable for use in our previously published peak assignment methodology, Small Molecule Isotope Resolved Formula Enumeration (SMIRFE).
 :::
@@
 Fourier transform mass spectrometry (FT-MS) provides high performance in terms of sensitivity, resolution and mass accuracy all in one analytical instrumentation.
-The combination of these capabilities provide several analytical and interpretive improvements including: (i) the ability to resolve distinct isotopologues with identical unit masses but different accurate masses [@higashiStableIsotopeLabeledTracers2014], enabling multi-element isotopic natural abundance correction [@moseleyCorrectingEffectsNatural2010; @carreerComputationalFrameworkHighThroughput2013]; (ii) improved assignment accuracy [@kindMetabolomicDatabaseAnnotations2006]; and (iii) the detection of metabolites in the sub-femtomolar range [@eylesMethodsStudyProtein2004; @dettmerMassSpectrometrybasedMetabolomics2007].
-In the metabolomics field, these improvements permit more complicated, but more informative experimental designs such as the use of multiple isotope-labeled precursors in stable isotope-resolved metabolomics (SIRM) experiments [@yangChloroformateDerivatizationTracing2017.
-Multiple isotope-labeling experiments can provide the information needed to elucidate unknown metabolic pathways [@creekStableIsotopeAssistedMetabolomics2012; @higashiStableIsotopeLabeledTracers2014], quantify relative fluxes through multiple metabolic pathways [@hillerNontargetedElucidationMetabolic2010], identify metabolite pools in different celullar compartments [@fanStableIsotoperesolvedMetabolomics2012], and identify active metabolic pathways under various cellular conditions [@moseleyNovelDeconvolutionMethod2011; @sellersPyruvateCarboxylaseCritical2015; @verdegemMAIMSSoftwareTool2017].
-These gains in information enable more complete modeling of cellular metabolism and better understanding of physiological and pathological processes at a mechanistic level, facilitating the identification of potential therapeutic targets [@fanAlteredRegulationMetabolic2009] and the quantification of differential drug response [@harrisDiverseMechanismsGrowth2012].
+The combination of these capabilities provide several analytical and interpretive improvements including: (i) the ability to resolve distinct isotopologues with identical unit masses but different accurate masses [@higashiStableIsotopeLabeledTracers2014]; (ii)  enable multi-element isotopic natural abundance correction for at least the lower portion of the detected mass range  [@moseleyCorrectingEffectsNatural2010; @carreerComputationalFrameworkHighThroughput2013; @wangAccuCor2IsotopeNatural2021]; (iii) improved assignment accuracy [@kindMetabolomicDatabaseAnnotations2006; @mitchellSmallMoleculeIsotope2019]; and (iv) the detection of metabolites in the sub-femtomolar range, when combined with chromatographic separation  [@eylesMethodsStudyProtein2004; @dettmerMassSpectrometrybasedMetabolomics2007].
+In the metabolomics field, these improvements permit more complicated, but more informative experimental designs such as the use of single and multiple isotope-labeled precursors in stable isotope-resolved metabolomics (SIRM) experiments [@yangChloroformateDerivatizationTracing2017].
+These stable isotop tracing experiments provide a wealth of isotope flux data that is interpretable in terms of metabolite flux information that is specific to a metabolic model, pathway, and subcellular location [@@higashiStableIsotopeLabeledTracers2014; @creekStableIsotopeAssistedMetabolomics2012; @hillerNontargetedElucidationMetabolic2010; @fanStableIsotoperesolvedMetabolomics2012; @moseleyNovelDeconvolutionMethod2011; @sellersPyruvateCarboxylaseCritical2015; @verdegemMAIMSSoftwareTool2017; @jinMoietyModelingFramework2019; @jinRobustMoietyModel2020].
 
-While the advantages of FT-MS are significant, when deployed in a high-throughput environment, the volume of data produced requires automated tools for data reduction, quality control, feature assignment, and downstream analyses.
-These requirements are amplified when FT-MS is used with direct infusion where existing and well-validated assignment methods that rely on orthogonal sources of information (such as chromatography) are either not applicable or would further increase the amount of data to be processed.
-Therefore, without automation, rigorous assignment of non-polymeric biomolecules in untargeted MS analyses remains difficult even with FT-MS’s high resolution.
+While these advantages of FT-MS are significant, when deployed in a high-throughput environment, the volume of data produced requires automated tools for data reduction, quality control, feature assignment, and downstream analyses.
+Furthermore within the context of direct infusion, assignment of FT-MS spectral features lack orthogonal sources of information such as chromatographic retention times, reducing the reliability of assignment with most MS assignment software tools.
 
-Within this high-throughput environment we previously developed methods to remove high-peak-density artifacts [@mitchellNewMethodsIdentify2018], annotate peaks with assignments using SMIRFE [@mitchellSmallMoleculeIsotope2019], and generate lipid classifications for those peak assignments [@mitchellDerivingLipidClassification2020].
-All of these methods together culminated in detecting differential lipids in non-small-cell lung cancer (NSCLC) [@mitchellUntargetedLipidomicsNonSmall2021].
+Within this combined high-throughput with direct injection context, we previously developed methods to remove high-peak-density artifacts [@mitchellNewMethodsIdentify2018], annotate peaks with assignments using SMIRFE [@mitchellSmallMoleculeIsotope2019], and generate lipid classifications for spectral peak assignments [@mitchellDerivingLipidClassification2020].
+We applied all of these methods together in detecting and assigning differential lipids in non-small-cell lung cancer (NSCLC) [@mitchellUntargetedLipidomicsNonSmall2021].
 
@@
 In the Thermo-Fisher Orbitrap FT-MS instruments, data is acquired in micro-scans and scans, where each scan is an aggregate of multiple micro-scans.
-Depending on the sample source, it can be advantageous to keep the data at the scan level, so that any given "bad" scan can be removed prior to aggregating the data across scans and generating a centroided peak m/z and intensity.
-If the scan level data is not processed and aggregated correctly, the final peak intensities and / or m/z's will be incorrect.
+While these micro-scan and subsequent scans are expected to be analytical replicates, spray instability and temporal delay in automatic gain control pragmatically break this expectation.
+Therefore, it is advantageous to keep the data at the scan level, so that any given poor quality (i.e.
+"bad") scan can be removed prior to aggregating the data across scans and generating a centroided peak m/z and intensity.
+If the scan level data is not processed and aggregated correctly, the final peak intensities in arbitrary units and centers in m/z will have high uncertainty (total variance) [@moseleyErrorAnalysisPropagation2013].
 
@@
 To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur as well as our scan-centric peak characterization, and found matching Xcalibur peaks to our assignments.
-As an example, in `r figure_count$label_text("motivation")` we show the Xcalibur calculated intensities and theoretical peak intensities based on NAP for four peaks matching threonine.
+As an example, in `r figure_count$label_text("motivation")`, we show the Xcalibur calculated intensities and theoretical peak intensities based on NAP for four peaks matching threonine.
 As the NAP gets lower, the deviations of the Xcalibur calculated intensities from the NAP become larger.
@@
 The subsequent differences in frequency should be relatively constant with respect to m/z, in contrast to the differences of adjacent m/z points, as shown in `r figure_count$label_text("mz_frequency_conversion")`B and C.
-The Thermo Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 400K, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C and D.
+The Thermo Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 450K or 500K depending on the sample, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C and D.
 Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`E.
@@
 This is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points.
-The m/z point-to-point differences are not constant, but can be approximated by a lasso linear model; however, it is very difficult to create a lasso model with an intercept of 0.
-In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the lasso model.
+The m/z point-to-point differences are not constant, but can be approximated by a loess linear model [@Cleveland1992loess]; however, it is very difficult to create a loess model with an intercept of 0.
+In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the loess model.
 Frequency based points suffer none of these drawbacks, and the conversion from m/z can be derived from the raw profile level data itself, which is incredibly useful.
@@
 ::: {custom-style="MDPI_3.1_text"}
-Due to differences in how many ions are captured in the trap and the limited dynamic range of the detector, the observed peak intensities for the same analyte vary slightly between scans (see `r figure_count$label_text("breaking_regions")`A and B for example).
+Due to differences in how many ions are captured in the trap and the limited dynamic range of the detector, the observed peak intensities for the same analyte vary between scans (see `r figure_count$label_text("breaking_regions")`A and B for example).
 Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
@@
 If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and B.
+Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors"`B and C), removing the scan-correlated peaks does change some of the normalization factors.
 :::
@@
 **B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.
+**C**: **B**, but with a zoomed y-axis to illustrate the differences in the normalization factors obtained with the different methods.
 :::
@@
 **A** - **E**:Peak plots for various peak processing modalities in a single HPD site.
-**A**:*xcalibur* peaks exported from Thermo Xcalibur after averaging scans.
-**B**:*scalevel_00*: Scan-level peak characterization without any density based filtering (see noperc_nonorm in Methods).
-**C**:*scanlevel_99*: Scan-level peak characterization using the default point density filtering (see filtersd in Methods).
-**D**:*scanlevel_99_lowsd*: Same as *scanlevel_99*, but removing any peaks marked as having a high frequency standard deviation.
-**E**:*msnbase*: peak centroids generated from *MSnbase*.
+**A**: *xcalibur* peaks exported from Thermo Xcalibur after averaging scans.
+**B**: *scalevel_00*: Scan-level peak characterization without any density based filtering (see noperc_nonorm in Methods).
+**C**: *scanlevel_99*: Scan-level peak characterization using the default point density filtering (see filtersd in Methods).
+**D**: *scanlevel_99_lowsd*: Same as *scanlevel_99*, but removing any peaks marked as having a high frequency standard deviation.
+**E**: *msnbase*: peak centroids generated from *MSnbase*.
 **F**: Scatterplot of peak counts across all of the HPDs detected in the ECF 2 sample against the Xcalibur peak counts.
@@
 Each step in the peak characterization either changes the overall number of peaks coming from each scan (sliding windows and breaking initial regions) or the overall intensity of the points within a scan.
-Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation (RSD) for the characterized scan level peak intensities (calculated as the standard deviation of peak heights across scans divided by the mean peak height), and compare them as each processing step is introduced.
+Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation (RSD) for the characterized scan level peak intensities (calculated as the standard deviation of peak heights across scans divided by the mean peak height) and compare them as each processing step is introduced.
 :::
@@
 `r table_count$label_text("rsd_method")`.
-Relative standard deviation means, medians, modes and maximum observed values for each sample with different overall processing.
+Relative standard deviation means, medians, modes, and maximum observed values for each sample with different overall processing.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-As an alternative to RSD, we can also compare the fit of relative intensities after assignment using SMIRFE [reference] to the theoretical relative natural abundances (relNAP) of the assigned isotopic molecular formula's (IMFs) within the assigned elemental molecular formula's (EMFs).
-Theoretically, we expect that lower quality data will have both lower numbers of assignments, and that for those things that are assigned, the fit between relative intensity and relNAP will be worse.
+As an alternative to RSD, we can also compare the fit of relative intensities after assignment using SMIRFE [@mitchellSmallMoleculeIsotope2019] to the theoretical relative natural abundances (relNAP) of the assigned isotopic molecular formula's (IMFs) within the assigned elemental molecular formula's (EMFs).
+Theoretically, we expect lower quality data to have both lower numbers of assignments, and for those things that are assigned, the fit between relative intensity and relNAP to be worse.
 To compare relative NAP to relative abundances, we only examined the assignments from the two samples containing ECF derivatized amino-acids, as we can limit the assignments to those that match expected derivatizations of the known amino-acids (see Supplemental for the expected EMFs, and expected relative NAPs for the individual IMFs).
 
-`r figure_count$label_text("nap_intensity")` compares the peak - peak natural abundance and height log-ratio differences generated using heights from Xcalibur and from our scan-level peak characterization.
+`r figure_count$label_text("nap_intensity")` compares the peak-peak isotopic natural abundance probability and height log-ratio differences (Equation X in Methods) generated using heights from Xcalibur and from our scan-level peak characterization.
 :::
@@
 **A**.
-Peak - peak NAP - intensity differences from scan-level peak heights (red) and Xcalibur peak heights (blue) from the threonine amino acid assignments with Na adduct, with point size reflecting how many peaks were missing across scans.
+The peak-peak NAP - intensity log differences from scan-level peak heights (red) and Xcalibur peak heights (blue) from the threonine amino acid assignments with Na adduct, with point size reflecting how many peaks were missing across scans.
 **B**. The difference of Xcalibur to scan-level ratios plotted directly as a function of the number of scans the peak was not found in.
@@
 Number of matched peaks between peak processing methods for 1ecf.
+There are no overlapping peaks only between scancentric and msnbase, as well as no peaks found only by the scancentric method.
 :::
@@
 Two types of frequency intervals are used, sliding and tiled windows.
-In this work, the sliding windows are 10 points wide with a stride of one point.
+In this work, the sliding windows are the equivalent of 10 frequency points wide with a stride of one point.
 The tiled windows are one point wide with a stride of one point.
+Each point here is the equivalent of the difference between data points in frequency space, which is a spacing of 0.5.
 :::
@@
 For the ECF derivatized amino-acid samples, the EMF database was generated using an m/z limit of 1005 m/z, and maximum numbers for each element were set to C: 100, N: 7, O: 40, H: 230, P: 3, S: 3.
-We attempted to create peaks that could be assigned by SMIRFE from the MSnbase and Xcalibur peak lists, but SMIRFE would not assign them.
+We attempted to create peaks that could be assigned by SMIRFE from the MSnbase and Xcalibur peak lists, but SMIRFE would not assign them given its requirements for scan-level peak characterization data.
 :::
@@
 EMF voting identified 3485 total corresponded peaks across all 157 spectra with 855 corresponded peaks present in 25% or more of either the disease or non-disease group of samples. 
-To minimize false assignment and their effects, non_lipid and multiple lipids were not considered in downstream analyses. 
 All lipid isotopologue intensities were normalized by dividing the isotopologue intensity by the median intensity of all the peaks in the sample. 
-For differential abundance analysis, missing values were replaced with a threshold value that is ½ of the lower confidence interval of the distribution of all log-transformed intensities from that tissue.
+For differential abundance analysis, missing values were removed. 
 :::
@@
 SMIRFE assignments include the NAP for each of the isotopologue molecular formulas (IMFs) in the particular labelled and adducted EMF.
-For each peak pair in the peak group for a particular EMF, we can calculate the log-ratio for the NAPs and the log-ratio for the intensities, and their absolute differences.
+For each peak pair in the peak group for a particular EMF, we calculate the log-ratio for the NAPs and the log-ratio for the intensities, and their absolute differences.
 This calculation is done for the final characterized intensity, as well as the intensities at the scan level, and for both the *raw* intensity and *corrected* intensity, and for matched Xcalibur peak intensities, and matched MSnbase peak intensities if there were two or more matching peaks in the labeled, adducted EMF.

### 2022-04-05 doc/peakcharacterization_manuscript.Rmd
basic numbers added to a first equation.
closes #17

@@
 table_count = dn_counter$new("Table ")
+equation_count = dn_counter$new("")
 ```
@@
                       dev.args = list(png = list(type = "cairo")))
+
+format_equation = function(equation_text, label){
+  number_out = equation_count$label_text(label)
+  paste0('$$\n\\begin{align}\n', equation_text, '&& \\text{(', number_out, ')}', '\n\\end{align}$$')
+}
 ```
@@
 In the metabolomics field, these improvements permit more complicated, but more informative experimental designs such as the use of single and multiple isotope-labeled precursors in stable isotope-resolved metabolomics (SIRM) experiments [@yangChloroformateDerivatizationTracing2017].
-These stable isotop tracing experiments provide a wealth of isotope flux data that is interpretable in terms of metabolite flux information that is specific to a metabolic model, pathway, and subcellular location [@@higashiStableIsotopeLabeledTracers2014; @creekStableIsotopeAssistedMetabolomics2012; @hillerNontargetedElucidationMetabolic2010; @fanStableIsotoperesolvedMetabolomics2012; @moseleyNovelDeconvolutionMethod2011; @sellersPyruvateCarboxylaseCritical2015; @verdegemMAIMSSoftwareTool2017; @jinMoietyModelingFramework2019; @jinRobustMoietyModel2020].
+These stable isotop tracing experiments provide a wealth of isotope flux data that is interpretable in terms of metabolite flux information that is specific to a metabolic model, pathway, and subcellular location [@higashiStableIsotopeLabeledTracers2014; @creekStableIsotopeAssistedMetabolomics2012; @hillerNontargetedElucidationMetabolic2010; @fanStableIsotoperesolvedMetabolomics2012; @moseleyNovelDeconvolutionMethod2011; @sellersPyruvateCarboxylaseCritical2015; @verdegemMAIMSSoftwareTool2017; @jinMoietyModelingFramework2019; @jinRobustMoietyModel2020].
 
@@
 If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and B.
-Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors"`B and C), removing the scan-correlated peaks does change some of the normalization factors.
+Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B and C), removing the scan-correlated peaks does change some of the normalization factors.
 :::
@@
 
+```{r increment_frequency}
+equation_count$increment("mzfrequency")
+```
+
 ::: {custom-style="MDPI_3.1_text"}
@@
 
+::: {custom-style="MDPI_3.9_equation"}
+`r format_equation('frequency = intercept + x* \\frac{1}{\\sqrt{mz}} + y * \\frac{1}{\\sqrt[3]{mz}}', 'mzfrequency')`
+:::
+
+::: {custom-style="MDPI_3.9_equation"}
 $$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
+:::
 

### 2022-04-05 doc/test_equations.Rmd
basic numbers added to a first equation.
closes #17

--- 2022-04-05 doc/peakcharacterization_manuscript.Rmd
@@
 
-format_equation = function(equation_text, label){
+format_eqn = function(eqn_text, label){
   number_out = equation_count$label_text(label)
-  paste0('$$\n\\begin{align}\n', equation_text, '&& \\text{(', number_out, ')}', '\n\\end{align}$$')
+  paste0('$$\n\\begin{align}\n', eqn_text, '&& \\text{(', number_out, ')}', '\n\\end{align}$$')
 }
@@
 
-::: {custom-style="MDPI_2.1_heading1"}
-## Introduction
-:::
+### Deleteme
 
-::: {custom-style="MDPI_3.1_text"}
-Fourier transform mass spectrometry (FT-MS) provides high performance in terms of sensitivity, resolution and mass accuracy all in one analytical instrumentation.
-The combination of these capabilities provide several analytical and interpretive improvements including: (i) the ability to resolve distinct isotopologues with identical unit masses but different accurate masses [@higashiStableIsotopeLabeledTracers2014]; (ii)  enable multi-element isotopic natural abundance correction for at least the lower portion of the detected mass range  [@moseleyCorrectingEffectsNatural2010; @carreerComputationalFrameworkHighThroughput2013; @wangAccuCor2IsotopeNatural2021]; (iii) improved assignment accuracy [@kindMetabolomicDatabaseAnnotations2006; @mitchellSmallMoleculeIsotope2019]; and (iv) the detection of metabolites in the sub-femtomolar range, when combined with chromatographic separation  [@eylesMethodsStudyProtein2004; @dettmerMassSpectrometrybasedMetabolomics2007].
-In the metabolomics field, these improvements permit more complicated, but more informative experimental designs such as the use of single and multiple isotope-labeled precursors in stable isotope-resolved metabolomics (SIRM) experiments [@yangChloroformateDerivatizationTracing2017].
-These stable isotop tracing experiments provide a wealth of isotope flux data that is interpretable in terms of metabolite flux information that is specific to a metabolic model, pathway, and subcellular location [@higashiStableIsotopeLabeledTracers2014; @creekStableIsotopeAssistedMetabolomics2012; @hillerNontargetedElucidationMetabolic2010; @fanStableIsotoperesolvedMetabolomics2012; @moseleyNovelDeconvolutionMethod2011; @sellersPyruvateCarboxylaseCritical2015; @verdegemMAIMSSoftwareTool2017; @jinMoietyModelingFramework2019; @jinRobustMoietyModel2020].
-
-While these advantages of FT-MS are significant, when deployed in a high-throughput environment, the volume of data produced requires automated tools for data reduction, quality control, feature assignment, and downstream analyses.
-Furthermore within the context of direct infusion, assignment of FT-MS spectral features lack orthogonal sources of information such as chromatographic retention times, reducing the reliability of assignment with most MS assignment software tools.
-
-Within this combined high-throughput with direct injection context, we previously developed methods to remove high-peak-density artifacts [@mitchellNewMethodsIdentify2018], annotate peaks with assignments using SMIRFE [@mitchellSmallMoleculeIsotope2019], and generate lipid classifications for spectral peak assignments [@mitchellDerivingLipidClassification2020].
-We applied all of these methods together in detecting and assigning differential lipids in non-small-cell lung cancer (NSCLC) [@mitchellUntargetedLipidomicsNonSmall2021].
-
-The input peak-data in these studies are derived from multi-scan direct-injection FT-MS data.
-SMIRFE uses expected relationships of relative peak height or intensity to natural abundance probability (NAP) across multiple peak pairs to determine the likelihood of an assignment for the peaks in question.
-In the Thermo-Fisher Orbitrap FT-MS instruments, data is acquired in micro-scans and scans, where each scan is an aggregate of multiple micro-scans.
-While these micro-scan and subsequent scans are expected to be analytical replicates, spray instability and temporal delay in automatic gain control pragmatically break this expectation.
-Therefore, it is advantageous to keep the data at the scan level, so that any given poor quality (i.e.
-"bad") scan can be removed prior to aggregating the data across scans and generating a centroided peak m/z and intensity.
-If the scan level data is not processed and aggregated correctly, the final peak intensities in arbitrary units and centers in m/z will have high uncertainty (total variance) [@moseleyErrorAnalysisPropagation2013].
-
-In particular, if peaks are missing in some scans due to low abundance, the final aggregate peak intensities will not fit the expected relative intensity to NAP relationships.
-Here we describe novel scan-centric FT-MS metabolomics data processing methods that better preserve the expected relative intensity to NAP.
-In addition, due to the tendency of increasing point spacing with increasing m/z in Orbitrap type instruments, we derive a method for transforming the m/z data to axial frequency, which has the desirable property of being equally spaced across the full spectrum.
-Finally, we show that in addition to better preserving the expected intensity - NAP relationships, our scan-centric method also results in improved relative standard deviations, automatic identification of high-peak-density artifact peaks, and separation of samples in a lipidomics data analysis.
-:::
-
-::: {custom-style="MDPI_2.1_heading1"}
-## Results
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Naively Averaged Data Have Bad Relative Intensities
-:::
-
-```{r increment_motivation}
-figure_count$increment("motivation")
+```{r test_equation}
+eqn_count$increment("deleteme")
 ```
 
-::: {custom-style="MDPI_3.1_text"}
-To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur as well as our scan-centric peak characterization, and found matching Xcalibur peaks to our assignments.
-As an example, in `r figure_count$label_text("motivation")`, we show the Xcalibur calculated intensities and theoretical peak intensities based on NAP for four peaks matching threonine.
-As the NAP gets lower, the deviations of the Xcalibur calculated intensities from the NAP become larger.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-```{r motivation, dn_id = figure_count, fig.width = 10, fig.height = 6}
-tar_load(motivation_plot)
-motivation_plot
-```
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("motivation")`.
-Xcalibur intensities (black) and theoretical intensities based on relative NAP (red) for ECF derived threonine.
-The 18O isotopologue is shifted by 0.01 m/z and the NAP peaks by 0.05 m/z to aid visualization.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### m/z to Frequency
-:::
-
-```{r increment_mz_frequency}
-figure_count$increment("mz_frequency_conversion")
-```
-
-::: {custom-style="MDPI_3.1_text"}
-FT-MS data from the Thermo-Fisher Orbitrap instruments used to acquire the data does not provide any information about the raw frequency data.
-Outside of the meta-data, it merely contains the m/z and intensity values for profile spectra acquired across multiple scans.
-However, the frequency can be calculated by dividing the average m/z of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points representing the average m/z of two adjacent points, red lines representing the difference between the two adjacent points).
-The subsequent differences in frequency should be relatively constant with respect to m/z, in contrast to the differences of adjacent m/z points, as shown in `r figure_count$label_text("mz_frequency_conversion")`B and C.
-The Thermo Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 450K or 500K depending on the sample, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C and D.
-Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`E.
-This model seems to fit the known relationship between frequency and m/z, where the frequency is related to $1/\sqrt{mz}$.
-This is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points.
-The m/z point-to-point differences are not constant, but can be approximated by a loess linear model [@Cleveland1992loess]; however, it is very difficult to create a loess model with an intercept of 0.
-In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the loess model.
-Frequency based points suffer none of these drawbacks, and the conversion from m/z can be derived from the raw profile level data itself, which is incredibly useful.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-```{r mz_frequency_conversion, dn_id = figure_count, fig.width = 16, fig.height = 10}
-tar_load(frequency_conversion)
-frequency_conversion$all
-```
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("mz_frequency_conversion")`.** **A**: Intensity vs m/z for a single peak from a single scan.
-Red lines denote the differences between each point, and red dots the average between the pair of points.
-The difference over the point is used to derive the frequency values in **B**.
-**B** plots the intensity vs the converted frequency points derived from **A**.
-The red lines denote pairwise differences, which are shown in **C** for this single peak.
-The differences for all peaks across all scans vs m/z are shown in **D**, with those differences that lie within 0.49 - 0.51 shown in red.
-**E** shows the plot of derived frequency *vs* m/z, with fitted values from the linear model in red.
-:::
-
-```{r increment_peak_ordering}
-figure_count$increment("peak_ordering")
-```
-
-::: {custom-style="MDPI_3.1_text"}
-The m/z to frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
-While scan specific models **could** be used in the conversion of m/z to frequency, doing so results in changes to the relative peak ordering compared to m/z space, as shown in `r figure_count$label_text("peak_ordering")`.
-Therefore, a single model for all scans based on the scan with the slope closest to the median of slopes across scans is used for converting *all* remaining scan level data.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-```{r peak_ordering, dn_id = figure_count}
-tar_load(peak_ordering)
-peak_ordering
-```
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("peak_ordering")`.** Peak ordering in m/z compared with ordering in frequency space when a single m/z to frequency model is used or scan specific m/z to frequency models are used.
-For a single peak, the scan level peak m/z's were extracted, and then frequency values for those m/z generated using a single common model of m/z to frequency (*single_order*), or models derived from each scan (*scan_order*).
-A subset of the peaks end up out of order using scan specific models, implying that a single model should be used across all the scan level data.
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-Although the original model is created from only those points that had frequency point-to-point differences within a narrow range, **all** m/z points are converted to frequency for subsequent steps in the workflow.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Sliding Window Density to Remove Noise
-:::
-
-```{r increment_slidingwindow}
-figure_count$increment("slidingwindow_count")
-tar_load(mn_ratios)
-tar_load(sliding_regions)
-tar_load(noise_plot)
-```
-
-::: {custom-style="MDPI_3.1_text"}
-In a dataset of this nature, we expect that much of the data is really just noise, and doesn't contribute that much to the analysis.
-Furthermore, it is expected that noise is randomly distributed across the scans.
-Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
-Subsequently, we divide the counts into tiled regions of a set size (2000 in this work), and examine the 99th percentile of the sliding window counts within the tiled region.
-A histogram of sliding window counts in a single tiled region is shown in `r figure_count$label_text("slidingwindow_count")`A.
-The distribution of 99th percentiles across all of the tiled regions for a single sample is shown in `r figure_count$label_text("slidingwindow_count")`B.
-We can use the median (50th percentile) value of all the 99th percentiles from all of the tiled regions to determine a cutoff value above which any non-zero points are potentially **signal** and not just noise.
-In the example below, the median of the tiled windows is 5, resulting in a cutoff of 7.5.
-The number of peaks detected without applying a cutoff explodes anywhere from `r mn_ratios[["97lipid"]]$ratio`-fold (from `r format(mn_ratios[["97lipid"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["97lipid"]]$count_0, big.mark = ",")` in the 97lipid sample), to `r mn_ratios[["2ecf"]]$ratio`-fold (from `r format(mn_ratios[["2ecf"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["2ecf"]]$count_0, big.mark = ",")` in the 2ecf sample) with a vast majority of the peaks being present in a single scan.
-By discarding those windows with a non-zero point density below 1.5x the median of 99th percentiles, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-
-```{r slidingwindow_count, dn_id = figure_count, fig.width = 8, fig.height = 8}
-(sliding_regions / noise_plot) + plot_annotation(tag_levels = "A")
-```
-
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("slidingwindow_count")`.**
-**A**: Histogram of the number of non-zero points from sliding windows ten frequency points wide and a stride of one point, from a single tiled window 2000 points wide.
-Blue line is the 99th percentile.
-**B**: Histogram of the 99th percentile cutoffs from all of the tiled windows.
-Red line is the median x 1.5.
-**C** and **D** show the number of initial regions as a function of the percentile cutoff used for either the AA ECF (C) or lipid (D) samples.
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-For each region initially created, the peaks in each scan within that region can be characterized using a quadratic fit of log(intensity) to m/z.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Peak Characterization Using Quadratic Fit
-:::
-
-```{r increment_fit}
-figure_count$increment("centroided_peaks")
-```
-
-::: {custom-style="MDPI_3.1_text"}
-Although many other types of mass-spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo Fusion has a baseline of 0 due to manipulations in the Thermo firmware, making the determination of the centroided values considerably easier.
-For the peak characterization (centroiding), we use a simple quadratic model of log(intensity) to m/z. 
-We add a small constant to enable using the zero values, and weight the values by their ratio to the most intense value, which is normally the value closest to the center of peak, helping to ensure that the **true** centroid is determined.
-From the fitted model, we can derive the centroided center and the intensity of the peak, as shown in `r figure_count$label_text("centroided_peaks")`.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-```{r centroided_peaks, dn_id = figure_count}
-# figure showing fit of quadratic model to actual data points
-tar_load(peak_fit_plot)
-peak_fit_plot
-```
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("centroided_peaks")`.**
-Log-intensity (A) and Intensity (B) of points for a single peak against frequency. Black points are the original data points, and the blue point represents the calculated centroid.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Breaking Up Initial Regions
-:::
-
-```{r increment_breaking_regions}
-figure_count$increment("breaking_regions")
-```
-
-::: {custom-style="MDPI_3.1_text"}
-With the characterized (centroided) peak data from across scans within each region, it is then important to determine if only one or multiple "peaks" are actually present in the region.
-Our solution to this is to define breaks between **actual** peaks as a single frequency bin with zero characterized peaks within it.
-The frequency bins are created from tiled windows that are one frequency point difference wide.
-Adjacent non-zero frequency bins are merged to comprise a single peak region.
-`r figure_count$label_text("breaking_regions")` shows an example where an initial region is broken up into two regions based on the characterized peak centers.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-```{r breaking_regions, dn_id = figure_count}
-# figure showing multiple "peaks" in what is initially a single region
-tar_load(split_region_plot)
-split_region_plot
-```
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("breaking_regions")`.** Splitting a single region into two regions based on the peaks that are present.
-**A**: The full set of raw frequency and intensity data across all scans for the region are shown.
-Clearly the region has two separate peaks within it.
-**B**: The peak centroids (frequency and intensity) for each peak in black.
-The tiled regions (red) are used to quantify the number of peaks.
-**C**: The number of peaks within each tiled region shown as a histogram.
-Each group of non-zero adjacent regions will be merged to form a new peak region.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Normalization of Scans
-:::
-
-```{r increment_intensity_scan}
-figure_count$increment("intensity_scan")
-figure_count$increment("normalization_factors")
-```
-
-::: {custom-style="MDPI_3.1_text"}
-Due to differences in how many ions are captured in the trap and the limited dynamic range of the detector, the observed peak intensities for the same analyte vary between scans (see `r figure_count$label_text("breaking_regions")`A and B for example).
-Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
-However, there are two issues with these peak intensities across scans: 1) some peak intensities correlate with the scan number (i.e.
-scan acquisition order); and 2) some peak differences between scans are correlated with intensity.
-The solution to **1** is to do a two pass normalization.
-After the first pass, the peaks whose intensity is correlated with scan order are detected.
-In the second pass, the correlated peaks are removed, and normalization is carried out again.
-`r figure_count$label_text("intensity_scan")`A contains an example peak whose intensity across scans is correlated with scan number.
-The solution to **2** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
-The highlighted peaks in `r figure_count$label_text("intensity_scan")`B are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
-If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and B.
-Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B and C), removing the scan-correlated peaks does change some of the normalization factors.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-```{r intensity_scan, dn_id = figure_count}
-# figure showing peak intensity correlation with m/z, intensity varying with m/z
-tar_load(intensity_scan_cor_plot)
-intensity_scan_cor_plot
-```
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("intensity_scan")`.** **A**: An example of a peak whose height across scans is correlated with scan number.
-**B**: The peak differences to the same peaks in a reference scan are plotted against peak height.
-Black: Peaks with a height \< 0.7 of the maximum.
-Red: Peaks with a height \>= 0.7 of the maximum.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-```{r normalization_factors, dn_id = figure_count, fig.height = 8, fig.width = 8}
-tar_load(compare_normalization)
-compare_normalization[[1]]
-```
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("normalization_factors")`.** **A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity \>= 0.7 of the maximum intensity (singlenorm_int), or the double pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
-**B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.
-**C**: **B**, but with a zoomed y-axis to illustrate the differences in the normalization factors obtained with the different methods.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Removal of High Peak Density Artefacts
-:::
-
-```{r increment_hpd}
-figure_count$increment("hpd_compare_fsd")
-```
-
-::: {custom-style="MDPI_3.1_text"}
-We have previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [@mitchellNewMethodsIdentify2018].
-Ideally, the peak characterization procedure should reduce their presence in the resultant reported peaks.
-Their presence should be minimized by removing noise peaks, and removing peak regions that have multiple reported peaks from the same scan.
-However, we expect they may also present as characterized peaks that have larger than expected frequency level standard deviations (FSD) when calculated across scans.
-These peaks can be detected by simply examining the distribution of FSDs and removing those that are outliers.
-To verify the removal of HPD peaks, we converted centroided m/z's from XCalibur to frequency values using the previously calculated values for that sample.
-Peak density was measured using a sliding window ten points wide and a stride of one point.
-
-`r figure_count$label_text("hpd_compare_fsd")` shows a single HPD site detected in the ECF 2 sample, with the peaks from Xcalibur, as well as various scan-level processing and the peaks from centroiding using MSnbase.
-From this figure, we can see that the point density based noise filtering removes a large number of the peaks in the HPD site, while the frequency standard deviation removes further peaks that may be suspect.
-After removing the high FSD peaks, the number of peaks left in the HPD site are the same or less than those from MSnbase, and at least in this example, look more likely to be real peaks compared to those from MSnbase.
-Therefore, scan-level characterization allows us to keep what are likely **real** peaks in HPD sites, without consideration of peak intensity.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-```{r hpd_compare_fsd, dn_id = figure_count, fig.width = 16, fig.height = 8}
-tar_load(hpd_plots_2ecf)
-
-hpd_left_panel = wrap_plots(hpd_plots_2ecf$plots, ncol = 1)
-hpd_right_panel = hpd_plots_2ecf$n %>%
-  ggplot(aes(x = xcalibur,
-             y = count,
-             color = method)) +
-  geom_point(size = 3) +
-  labs(x = "Xcalibur peak count",
-       y = "peak count") +
-  theme(legend.position = c(0.1, 0.8))
-(hpd_left_panel | hpd_right_panel) + plot_annotation(tag_levels = "A")
-```
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("hpd_compare_fsd")`.** Comparison of HPD and high FSD sites in the ECF 2 sample.
-**A** - **E**:Peak plots for various peak processing modalities in a single HPD site.
-**A**: *xcalibur* peaks exported from Thermo Xcalibur after averaging scans.
-**B**: *scalevel_00*: Scan-level peak characterization without any density based filtering (see noperc_nonorm in Methods).
-**C**: *scanlevel_99*: Scan-level peak characterization using the default point density filtering (see filtersd in Methods).
-**D**: *scanlevel_99_lowsd*: Same as *scanlevel_99*, but removing any peaks marked as having a high frequency standard deviation.
-**E**: *msnbase*: peak centroids generated from *MSnbase*.
-**F**: Scatterplot of peak counts across all of the HPDs detected in the ECF 2 sample against the Xcalibur peak counts.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Changes in Relative Standard Deviation (RSD)
-:::
-
-```{r increment_rsd}
-figure_count$increment("rsd_method")
-table_count$increment("rsd_method")
-```
-
-::: {custom-style="MDPI_3.1_text"}
-Each step in the peak characterization either changes the overall number of peaks coming from each scan (sliding windows and breaking initial regions) or the overall intensity of the points within a scan.
-Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation (RSD) for the characterized scan level peak intensities (calculated as the standard deviation of peak heights across scans divided by the mean peak height) and compare them as each processing step is introduced.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-```{r rsd_method, dn_id = figure_count}
-tar_load(rsd_plot)
-rsd_plot
-```
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("rsd_method")`.
-Density plots of relative standard deviations (RSD) of peak heights across scans for each of the processing methods.
-A peak had to be present in at least three scans for the RSD value to be reported.
-:::
-
-::: {custom-style="MDPI_4.1_table_caption"}
-`r table_count$label_text("rsd_method")`.
-Relative standard deviation means, medians, modes, and maximum observed values for each sample with different overall processing.
-:::
-
-::: {custom-style="MDPI_4.2_table_body"}
-```{r rsd_table, results = 'asis'}
-tar_load(rsd_values)
-rsd_values = rsd_values %>%
-  dplyr::arrange(sample, processed)
-rsd_ft = flextable(rsd_values) %>%
-  colformat_double(digits = 2) %>%
-  autofit()
-rsd_ft
-```
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Difference to Relative Natural Abundance
-:::
-
-```{r nap_intensity_increment}
-figure_count$increment("nap_intensity")
-```
-
-::: {custom-style="MDPI_3.1_text"}
-As an alternative to RSD, we can also compare the fit of relative intensities after assignment using SMIRFE [@mitchellSmallMoleculeIsotope2019] to the theoretical relative natural abundances (relNAP) of the assigned isotopic molecular formula's (IMFs) within the assigned elemental molecular formula's (EMFs).
-Theoretically, we expect lower quality data to have both lower numbers of assignments, and for those things that are assigned, the fit between relative intensity and relNAP to be worse.
-To compare relative NAP to relative abundances, we only examined the assignments from the two samples containing ECF derivatized amino-acids, as we can limit the assignments to those that match expected derivatizations of the known amino-acids (see Supplemental for the expected EMFs, and expected relative NAPs for the individual IMFs).
-
-`r figure_count$label_text("nap_intensity")` compares the peak-peak isotopic natural abundance probability and height log-ratio differences (Equation X in Methods) generated using heights from Xcalibur and from our scan-level peak characterization.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-```{r nap_intensity, dn_id = figure_count, fig.width = 16, fig.height = 8}
-nap_fontsize = 12
-tar_load(aa_diffs_filtersd_1ecf)
-tar_load(aa_diffs_filtersd_2ecf)
-
-all_diffs = rbind(aa_diffs_filtersd_1ecf$all_diffs %>%
-                    dplyr::mutate(sample = "1ecf"),
-                  aa_diffs_filtersd_2ecf$all_diffs %>%
-                    dplyr::mutate(sample = "2ecf"))
-
-threonine_ratios = aa_diffs_filtersd_1ecf$specific_diffs$ratios %>%
-  dplyr::filter(source %in% c("scan-level", "xcalibur"))
-threonine_diffs = aa_diffs_filtersd_1ecf$specific_diffs$diffs %>%
-  dplyr::filter(source %in% "xcal_raw_char")
-
-threonine_ratioplot = ggplot(threonine_ratios, aes(x = ratio, y = diff, color = source, size = n_missing)) +
-    geom_point() +
-    theme(legend.position = c(0.7, 0.75)) +
-    labs(x = "Pairwise Comparison",
-         y = "Peak-Peak NAP - Height Differences") +
-  theme(axis.title.y = element_text(size = nap_fontsize))
-
-threonine_diffplot = ggplot(threonine_diffs, aes(x = n_missing, y = diff)) +
-    geom_point(size = 2) +
-    labs(x = "Number of Missing Scans",
-         y = "Xcalibur - Scan-Level NAP - Height Differences")+
-  theme(axis.title.y = element_text(size = nap_fontsize))
-
-aa_diffplot = all_diffs %>%
-  dplyr::filter(source %in% "xcal_raw_char") %>%
-  ggplot(aes(x = n_missing, y = diff)) +
-  geom_point() +
-  labs(x = "Number of Missing Scans",
-         y = "Xcalibur - Scan-Level NAP - Height Differences") +
-  theme(axis.title.y = element_text(size = nap_fontsize))
-
-out_plot = (threonine_ratioplot | threonine_diffplot) / aa_diffplot + plot_annotation(tag_levels = "A")
-out_plot
-```
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("nap_intensity")`.
-**A**.
-The peak-peak NAP - intensity log differences from scan-level peak heights (red) and Xcalibur peak heights (blue) from the threonine amino acid assignments with Na adduct, with point size reflecting how many peaks were missing across scans.
-**B**. The difference of Xcalibur to scan-level ratios plotted directly as a function of the number of scans the peak was not found in.
-**C**. The differences of Xcalibur to scan-level ratios for all of the amino acid assignments in EMFs with more than a single peak in both ECF samples.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Method Specific Peaks
-:::
-
-```{r increment_peak_comparisons}
-table_count$increment("unassigned_1ecf")
-table_count$increment("unassigned_97lipid")
-
-figure_count$increment("unassigned_upset")
-figure_count$increment("assigned_mz_diff")
-
-tar_load(um_peaks_1ecf)
-tar_load(um_peaks_97lipid)
-tar_load(ass_peaks_1ecf)
-tar_load(ass_peaks_97lipid)
-```
-
-::: {custom-style="MDPI_3.1_text"}
-Each set of peaks generated may or may not be specific to the particular method used to generate centroided peak m/z and intensities.
-This is true for the different scan-centric combinations, as well as the peaks from Xcalibur and MSnbase.
-Here we examine the overlap of the unassigned (`r figure_count$label_text("unassigned_upset")`) and assigned peaks from the full scan-centric processing with Xcalibur and MSnbase.
-These same counts are also summarized in `r table_count$label_text("unassigned_1ecf")` and `r table_count$label_text("unassigned_97lipid")`.
-In these two examples, there are some striking differences.
-The 1ecf sample has all of the scan-centric peaks shared with either Xcalibur or MSnbase peaks, whereas the 97lipid sample has 2/3 of the peaks specific to scan-centric characterization and not matched to either of the other methods.
-Notably, for both samples, the scan-centric characterization produces similar numbers of peaks, even though the upper mass limit in 1ecf is 1000 m/z compared to 1600 m/z for the 97lipid sample.
-Whereas the number of peaks from MSnbase and Xcalibur are three-fold and 40-fold higher in the 1ecf sample compared to the 97lipid sample.
-:::
-
-```{r unassigned_upset, dn_id = figure_count, fig.width = 16, fig.height = 8}
-set_order = c("scancentric", "msnbase", "xcalibur")
-u1 = grid::grid.grabExpr(draw(UpSet(um_peaks_1ecf$comb_matrix, 
-                           set_order = set_order)))
-u2 = grid::grid.grabExpr(draw(UpSet(um_peaks_97lipid$comb_matrix, 
-                           set_order = set_order)))
-
-up_plot = wrap_elements(u1, clip = FALSE) + wrap_elements(u2, clip = FALSE)
-up_plot + plot_annotation(tag_levels = "A")
-```
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("unassigned_upset")`.
-UpSet plot with the counts of common and specific peaks for each of scancentric, MSnbase and XCalibur generated peaks for the 1ecf (**A**) and 97lipid (**B**) samples.
-:::
-
-::: {custom-style="MDPI_4.1_table_caption"}
-`r table_count$label_text("unassigned_1ecf")`.
-Number of matched peaks between peak processing methods for 1ecf.
-There are no overlapping peaks only between scancentric and msnbase, as well as no peaks found only by the scancentric method.
-:::
-::: {custom-style="MDPI_4.2_table_body"}
-```{r unassigned_1ecf}
-um_peaks_1ecf$ft_table %>%
-  autofit()
-```
-:::
-
-::: {custom-style="MDPI_4.1_table_caption"}
-`r table_count$label_text("unassigned_97lipid")`.
-Number of matched peaks between peak processing methods for 97lipid.
-:::
-::: {custom-style="MDPI_4.2_table_body"}
-```{r unassigned_97lipid}
-um_peaks_97lipid$ft_table %>%
-  autofit()
-```
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-```{r assigned_mz_diff, dn_id = figure_count}
-diffs_1ecf = ass_peaks_1ecf$peaks %>%
-  dplyr::mutate(source = factor(source, levels = c("scanlevel", "xcalibur", "msnbase"), ordered = TRUE))
-diffs_97lipid = ass_peaks_97lipid$peaks %>%
-  dplyr::mutate(source = factor(source, levels = c("scanlevel", "xcalibur", "msnbase"), ordered = TRUE))
-
-diffs_plot = (diffs_1ecf %>%
-                ggplot(aes(x = mz_diff)) +
-  geom_histogram(bins = 30) +
-  facet_wrap(~ source, ncol = 1) +
-  labs(x = "m/z Difference")) |
-  (diffs_97lipid %>%
-     ggplot(aes(x = mz_diff)) +
-     geom_histogram(bins = 30) +
-     facet_wrap(~ source, ncol = 1) +
-     labs(x = "m/z Difference")) +
-  plot_annotation(tag_levels = "A")
-diffs_plot
-```
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("assigned_mz_diff")`.
-Histograms of the m/z differences to the assigned and matched peak for 1ecf (**A**) and 97lipid (**B**).
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Changes in Variance Across a Large Dataset
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-For large datasets, we expect that more correct peak intensities will result in better agreement between sample normalized peak intensities within a sample class.
-If incorrect peak intensities are being reported, then we would expect between class to within class variances to be very random, resulting in F-statistics close to 1.
-:::
-
-::: {custom-style="MDPI_2.1_heading1"}
-## Discussion
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-we should have some discussion here about the results
-:::
-
-::: {custom-style="MDPI_2.1_heading1"}
-## Materials and Methods
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Conversion of m/z to Frequency
-:::
-
-```{r increment_frequency}
-equation_count$increment("mzfrequency")
-```
-
-::: {custom-style="MDPI_3.1_text"}
-The data consists of profile mode m/z spectra from multiple scans encoded as m/z and intensity values for each scan.
-No information about the original observed frequency values is available in either the `raw` files or the `mzML` files.
-However, proxy frequency values can be generated by averaging the m/z of adjacent points and dividing them by the difference.
-Ideally, the difference between subsequent points in this proxy frequency space is constant, in practice there is a range of differences in frequency space.
-The *actual, constant* difference can be obtained by examining the median of the frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to m/z) to be within 2% of the mode value.
-These *useful* points can be used to construct a linear model relating m/z to frequency for individual scans based on the formula:
+::: {custom-style="MDPI_3.9_equation"}
+`r format_eqn(eqn_count, 'frequency = intercept + x* \\frac{1}{\\sqrt{mz}} + y * \\frac{1}{\\sqrt[3]{mz}}', 'deleteme')`
 :::
@@
 ::: {custom-style="MDPI_3.9_equation"}
-`r format_equation('frequency = intercept + x* \\frac{1}{\\sqrt{mz}} + y * \\frac{1}{\\sqrt[3]{mz}}', 'mzfrequency')`
+$$
+\begin{align}
+frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}} && \text{(1)}
+\end{align}
+$$
 :::
@@
 ::: {custom-style="MDPI_3.9_equation"}
-$$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
+*frequency* = a + x * *mz*^-1/2^ + y * *mz*^-1/3^
+:::
+::: {custom-style="MDPI_3.a_equation_number"}
+(1)
 :::
 
-::: {custom-style="MDPI_3.1_text"}
-From the known physical properties of the Orbitrap, only the square root term should be necessary [@ledfordSpaceChargeEffects1984].
-Practically, we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry, contributions from space charge effects and magnetronic motion, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
-A frequency model is generated for each scan, and then a single model using the scan with the square-root term closest to the median of the square-root terms from all scans.
-We observed that this single model better preserved the relative ordering of the peaks in both m/z and frequency-space compared to the scan specific models (see Results).
-A single model rather than scan specific models is used for conversion because this was found to maintain the relative ordering of the peaks in m/z and frequency space (see Results).
+mz = a + x * *frequency*^-1^ + y * *frequency*^-2^ + z * *frequency*^-3^
 
-To convert m/z back into frequency, we can use a similar model without the roots.
-:::
+ln(*intensity*) = a + x * *position* + y * *position*^2^
 
-::: {custom-style="MDPI_3.9_equation"}
-$$mz = intercept + x* \frac{1}{frequency} + y*\frac{1}{frequency^2} + z*\frac{1}{frequency^3}$$
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Frequency Intervals
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-Two types of frequency intervals are used, sliding and tiled windows.
-In this work, the sliding windows are the equivalent of 10 frequency points wide with a stride of one point.
-The tiled windows are one point wide with a stride of one point.
-Each point here is the equivalent of the difference between data points in frequency space, which is a spacing of 0.5.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Interval Range Based Data
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-To enable interval algebra, the frequency points are converted to single width intervals by multiplying (by a multiplier of 400 in this work), rounding to the nearest integer, and storing them as IRanges objects from the IRanges Bioconductor package [@lawrenceSoftwareComputingAnnotating2013a].
-The sliding and tiled windows are also converted to IRanges objects using this process.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Peak Containing Intervals
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-To find intervals that contain points that represent actual signal and not just random noise, the number of non-zero intensity points in each sliding window are counted.
-Subsequently, we break these counts into fixed width tiles (default width of 2000) and calculate the 99th percentile of non-zero points for each tile.
-The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions containing potential signal.
-Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged to create the initial peak regions.
-The presence of zero intensity points are due to flooring implemented by the spectrometer when local spectral quality falls below a certain threshold.
-When this flooring is unstable and incomplete, a partial ringing phenomenom is observed [@mitchellNewMethodsIdentify2018].
-
-Within each initial interval region, peaks in each scan are detected (see **Peak Detection**), and their centers are binned by the tiled windows.
-Adjacent tiled windows with non-zero peak counts are merged together, and any zero peak count tiled windows split the initial region into multiple peak interval regions.
-These interval regions should contain a single **real** peak that was detected in one or more scans.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Peak Detection and Centroided Values
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package [@borchersPracmaPracticalNumerical2021].
-These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either m/z or frequency), and the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.
-:::
-
-::: {custom-style="MDPI_3.9_equation"}
-$$\ln{intensity} = intercept + a*position+b*position^2$$
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-From this weighted parabolic fit, the center, intensity, integrated area and sum-of-square residuals can be extracted for the peak.
-These characteristics are equivalent to the centroided peak center and intensity.
-
-Before further processing, the regions are verified to have only one peak from each scan.
-If a scan has two or more peaks, then the scan level data in that region is discarded.
-Any regions that subsequently contain zero peaks are removed.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Scan to Scan Normalization
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-Scans are normalized to a single *reference* scan based on the log-intensity differences of a subset of peaks present in at least the same number of scans as the 95th percentile of scan counts for the peaks.
-In addition, only those peaks with an intensity greater than 0.7 times the highest intensity peak in the scan are used.
-Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each scan to all other scans.
-The scan with the lowest summed overall distance is chosen as the *reference* scan, and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the reference scan.
-This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan order is checked, and those peaks with correlation of greater than 0.5 with scan order are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.
-Peaks correlated to scan order represent an artifact that we speculate results from a gradient in the sample well.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Full Characterization
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-The full set of raw data points for each peak in each scan within a region is known based on the previously detected peaks.
-Therefore, the non-zero intensity, normalized raw data points across scans can be combined, and then characterized again using the weighted parabolic fit previously described.
-In addition to the data from the full set of raw points, the means and the standard deviations of the peak height and location can be derived from the scan-level peak characteristics previously calculated.
-
-In addition to these values, the frequency point-to-point median difference is calculated across all of the raw data points for those points that could be used for modeling frequency to m/z, and this difference of a single point from the peak center is calculated in frequency space, and converted to m/z space to provide an "offset" value that is useful to define the search space around the peak for any assignment algorithm.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Correction of Height and Standard Deviation
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-Ideally, each peak would be observed in every scan.
-However, some peaks are not observed in some scans due to the number of ions falling below the detection threshold.
-This results in a left-censored normal distribution of peak intensities across scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
-To correct these, either a correction based on a model of the truncated normal distribution can be used, or the differences can be simulated by sampling from data that is present in most of the scans.
-To simulate the effect of peaks missing from some scans on the standard deviations, the peaks present in all scans are used.
-For each peak, a sample of the heights across scans are taken (ranging from 5% to 95% of scans), and a new standard deviation calculated for that fraction, and a ratio of the fractional standard deviation to the "true" standard deviation calculated.
-The ratio standard deviation across peaks can then be fitted to a cubic model of the fraction used, and a correction factor predicted for those peaks that are present in fewer scans.
-The corrected standard deviations can then be used to correct the mean height assuming that it is the result of a left-censored normal distribution [@TruncatedNormalDistribution2022; @burkardtTruncatedNormalDistribution].
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Frequency Standard Deviation Filtering
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-High-peak-density (HPD) artifacts [@mitchellNewMethodsIdentify2018] present as groups of singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
-Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and peaks with FSDs greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are removed.
-The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
-For HPD detection, the peaks in excel output from Xcalibur were used after converting the m/z peak centers into frequency space.
-Sliding windows that are 1000 frequency points wide with a stride of 100 points were used for the density calculations.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Assigning Scan-Centric Peaks
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-Our previously described SMIRFE algorithm [@mitchellSmallMoleculeIsotope2019] was used to assign molecular formulas to scan-centric characterized peaks in an untargeted manner. 
-For the lipid samples, an initial EMF database was generated using an M/Z limit of 1605 M/Z, and maximum numbers for each element were set to C: 130, N: 7, O: 28, P: 3, H: 230. 
-Assigned formulas were allowed to have K, Na, H, and NH4 adducts (only positive mode samples were assigned). 
-Assigned molecular formulas were then classified into one or more lipid categories using our lipid classifier tool [@mitchellDerivingLipidClassification2020]. 
-Lipid classes that were previously observed to be overclassified by our models (neutral and acidic glycosphingolipids) were excluded.
-For the ECF derivatized amino-acid samples, the EMF database was generated using an m/z limit of 1005 m/z, and maximum numbers for each element were set to C: 100, N: 7, O: 40, H: 230, P: 3, S: 3.
-We attempted to create peaks that could be assigned by SMIRFE from the MSnbase and Xcalibur peak lists, but SMIRFE would not assign them given its requirements for scan-level peak characterization data.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Consistently Assigned Lipid Spectral Feature (Corresponded Peak) Generation and Peak Intensity Normalization
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-Our SMIRFE assignment method assigns isotope-resolved molecular formulas (IMFs) to characterized peaks in each spectrum. 
-Each IMF represents an isotopologue of a given elemental molecular formula (EMF) (e.g., 13C112C51H1216O6 is an IMF representing the m+13C1 isotopologue of EMF C6H12O6).
-Consistently assigned (i.e., corresponded peaks) are identified using an in-house method we have named EMF voting which is described in Appendix A.
-EMF voting identified 3485 total corresponded peaks across all 157 spectra with 855 corresponded peaks present in 25% or more of either the disease or non-disease group of samples. 
-All lipid isotopologue intensities were normalized by dividing the isotopologue intensity by the median intensity of all the peaks in the sample. 
-For differential abundance analysis, missing values were removed. 
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Matching Peaks
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-We matched the scan-centric peaks to the other methods by looking for the closest m/z peak that was within 2 ppm of the scan-centric reported m/z.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Peak - Peak NAP Height Ratios
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-Each pair of peaks in a specifically labeled and adducted elemental molecular formula are related by their natural abundance probability (NAP).
-Theoretically, the log-ratios of two peaks NAPs are approximately equal to the log-ratios of two peaks intensities.
-:::
-
-::: {custom-style="MDPI_3.9_equation"}
-$$log(NAP_{P1} / NAP_{P2}) - log(Int_{P1} / Int_{P2}) \approx 0$$
-:::
-
-SMIRFE assignments include the NAP for each of the isotopologue molecular formulas (IMFs) in the particular labelled and adducted EMF.
-For each peak pair in the peak group for a particular EMF, we calculate the log-ratio for the NAPs and the log-ratio for the intensities, and their absolute differences.
-This calculation is done for the final characterized intensity, as well as the intensities at the scan level, and for both the *raw* intensity and *corrected* intensity, and for matched Xcalibur peak intensities, and matched MSnbase peak intensities if there were two or more matching peaks in the labeled, adducted EMF.
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Multi-Class F-Statistics
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-To evaluate the F-statistic between multi-class samples, a much larger set of samples was used.
-In this case, 181 matched non-cancer and cancer samples previously used in the HPD detection manuscript [@mitchellNewMethodsIdentify2018].
-All 181 samples were peak characterized using the most stringent method below as well as peaks generated from the MSnbase R package [ref], assigned using SMIRFE, and then peaks matched by shared EMFs across samples (see Supplemental materials for a description of peak matching and deciding the most likely EMF).
-For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using scan-level peak-characterization and those from MSnbase.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Samples and Overall Processing
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-The method for generating the amino-acid dataset was adapted from a previously published method for performing ethylchloroformate (ECF) amino acid derivatization [@yangChloroformateDerivatizationTracing2017].
-Two replicate samples were prepared and spectra were obtained for both samples using a Tribrid Fusion Orbitrap at 500k resolution and from 150 to 1000m/z.
-
-The collection, preparation, and mass spectrometry analysis of the paired cancer and non-cancer samples has been previously described [@sellersPyruvateCarboxylaseCritical2015]. 
-In summary, cancer and non-cancer samples were acquired from eight-six non-diabetic patients with suspected resectable stage I or IIA non-small cell lung cancer. 
-Written informed consent was collected from all subjects prior to inclusion and all samples were collected under a University of Louisville or University of Kentucky IRB protocol. 
-Lipid extracts were prepared using a modified Folch extraction and reconstituted for direct infusion ultra-high resolution mass spectrometry on a pair of Thermo Tribrid Fusion Orbitrap instruments (FSN10115 and FSN10352, referred to as Fusion 1 and Fusion 2 respectively) coupled to an Advion nanoelectrospray system.
-53 patients (102 spectra total) were acquired from Fusion 1 and 40 patients (77 spectra total) were acquired from Fusion 2.
-Fusion 1 samples were exclusively from the University of Louisville while Fusion 2 were a mix.
-Three of the patients had only a cancer or non-cancer sample acquired.
-Two of the cancer lipid samples (49Cpos and 97Cpos) were used for examples in this work.
-The rest were used for the examination of F-statistics.
-:::
-
-::: {custom-style="MDPI_3.5_text_before_list"}
-For each raw data file in profile mzML format, they were processed in these six ways:
-:::
-
-::: {custom-style="MDPI_3.8_bullet"}
-* No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, two pass normalization, frequency SD filtering (filtersd)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Scan level peak centroids generated by MSnbase (using `pickPeaks` on the individual scans)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Scans merged and then centroids generated by MSnbase (using `combineSpectra` and `pickPeaks`)
-:::
-
-::: {custom-style="MDPI_3.5_text_before_list"}
-For the two sets of MSnbase peaks, we also matched scan level peaks to merged peaks via two methods:
-:::
-
-::: {custom-style="MDPI_3.8_bullet"}
-* All scan level peaks within a 2ppm window are associated with the *merged* peak
-:::
-
-::: {custom-style="MDPI_3.8_bullet"}
-* Using the *filtersd* peaks, merged and scan level peaks are associated if they are within the start and stop m/z of the *filtersd* peaks
-:::
-
-::: {custom-style="MDPI_2.1_heading1"}
-## Conclusions
-:::
-
-::: {custom-style="MDPI_2.1_heading1"}
-## References
-:::
+ln(*NAP*~P1~ / *NAP*~P2~) - ln(*Int*~P1~ / *Int*~P2~) \U2248 0

### 2022-04-05 _targets.R

@@ -347,5 +347,5 @@ list(pkg_tar,
      unassigned_match_tar,
      assigned_match_tar,
      aa_tar,
-     lipid_tar,
-     lungcancer_tar)
+     lipid_tar)
+     #lungcancer_tar)

### 2022-04-05 lungcancer_all/lungcancer_smirfe_assignmentprocessing.R

@@ -1,6 +1,6 @@
-# python3 -m AssignMany.py 12 "python3 ./Main.py /mlab/scratch/cesb_data/smirfe_dbs/none_nosulfur_1600.db FILE '_assigned.json' '[]' '[\"H\", \"NH4\", \"Na\", \"K\"]' '[1]'" /mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2020-01-27/*.zip
-# python3 -m AssignMany.py 12 "python3 ./Main.py /mlab/scratch/cesb_data/smirfe_dbs/none_1600.db FILE '_assigned_sulfur.json' '[]' '[\"H\", \"NH4\", \"Na\", \"K\"]' '[1]'" /mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2020-01-27/*.zip
-# Example Workflow Across 3 Groups
+# cd smirfe_related/smirfe_code/SMIRFE_assigner
+# python3 -m pipenv shell
+# python3 -m AssignMany.py 12 "../../smirfe_dbs/none_nosulfur_1600.db FILE '_assigned.json' '[]' '[\"H\", \"Na\", \"K\", \"NH4\"]' '[1]'" /mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-04-02/*.zip
 # Database limits: C: 130, N: 7, O: 28, S: 0, P: 3, H: 230, max M/Z 1605, adducts: K, Na, NH4, NAP cutoff: 0.4
 library(smirfeTools)
 library(ggplot2)
@@ -14,7 +14,7 @@ plan(multicore)
 set_internal_map(furrr::future_map)
 options(future.globals.maxSize = 800 * 1024 ^ 2)
 
-assigned_files <- dir("/mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-03-08",
+assigned_files <- dir("/mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-04-02",
                       full.names = TRUE, pattern = "assigned.json")
 pb <- progress_estimated(length(assigned_files))
 
@@ -29,11 +29,11 @@ names(assigned_data) = purrr::map_chr(assigned_data, ~ .x$sample)
 
 all_emfs = unique(unlist(purrr::map(assigned_data, ~ .x$assignments$isotopologue_EMF)))
 
[…]

### 2022-04-05 lungcancer_all/msnbase_peakset.R

@@ -4,7 +4,7 @@ library(knitrProgressBar)
 all_mzml_files <- dir("/mlab/data/archives/FTMS_raw/CESB/mzml_data", recursive = TRUE, full.names = TRUE, pattern = "mzML$")
 names(all_mzml_files) = gsub(".mzML$", "", basename(all_mzml_files))
 
-all_zip_files = dir("/mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-03-08", pattern = ".zip$", full.names = TRUE)
+all_zip_files = dir("/mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-04-02", pattern = ".zip$", full.names = TRUE)
 names(all_zip_files) = gsub(".zip$", "", basename(all_zip_files))
 
 zip_mzml = intersect(names(all_mzml_files), names(all_zip_files))

### 2022-04-07 doc/peakcharacterization_manuscript.Rmd
p-value comparisons among different intensities
closes #3

@@
 }
+```
+
+```{r increment_equations}
+equation_count$increment("mzfrequency")
+equation_count$increment("frequencymz")
+equation_count$increment("modelintensity")
+equation_count$increment("napintensity")
 ```
@@
 
-`r figure_count$label_text("nap_intensity")` compares the peak-peak isotopic natural abundance probability and height log-ratio differences (Equation X in Methods) generated using heights from Xcalibur and from our scan-level peak characterization.
+`r figure_count$label_text("nap_intensity")` compares the peak-peak isotopic natural abundance probability and height log-ratio differences (Equation `r equation_count$label_text("napintensity")` in Methods) generated using heights from Xcalibur and from our scan-centric peak characterization.
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Changes in Variance Across a Large Dataset
+### Changes in P-Values On a Large Dataset
 :::
+
+```{r increment_p_value}
+figure_count$increment("p_value_comparison")
+table_count$increment("p_value_comparison")
+```
 
@@
 For large datasets, we expect that more correct peak intensities will result in better agreement between sample normalized peak intensities within a sample class.
-If incorrect peak intensities are being reported, then we would expect between class to within class variances to be very random, resulting in F-statistics close to 1.
+One way to test this is to compare p-values generated from the different intensities.
+`r figure_count$label_text("p_value_comparison")` compares the -1 * Log10(p-values) from XX or XX isotope resolved molecular formulas (IMFs) assigned by SMIRFE for the scan-centric peaks and then peaks matched and intensities extracted from the other methods.
+This shows that the p-values generated by the raw scan-centric intensities are generally the same if we use the intensities corrected for missing scans (on the line), and the p-values using raw scan-centric are better than either the XCalibur or MSnbase generated peak intensities (below the line due to being worse in XCalibur and MSnbase).
+We can test whether these differences are random or not using a binomial test of the actual difference in log-p-values.
+The binomial test p-values of the differences are reported in `r table_count$label_text("p_value_comparison")`.
+:::
+
+::: {custom-style="MDPI_5.2_figure"}
+```{r p_value_comparison, dn_id = figure_count, fig.width = 4, fig.height = 9}
+tar_load(lung_compare)
+
+lung_compare$id.test = factor(lung_compare$id.test, c("corrected", "xcalibur", "msnbase"))
+
+lung_compare %>%
+  ggplot(aes(x = log_p.ref,
+             y = log_p.test)) +
+  geom_point() +
+  geom_abline(slope = 1, color = "red") +
+  facet_wrap(~ id.test, ncol = 1) +
+  labs(x = "-1 * Log10(P-Value) Raw Scan-Centric",
+       y = "-1 * Log10(P-Value) Other")
+```
+:::
+
+::: {custom-style="MDPI_5.1_figure_caption"}
+`r figure_count$label_text("p_value_comparison")`.
+Log-p-values generated by comparing non-cancer and cancer sample IMFs using peak intensities from different methods.
+:::
+
+::: {custom-style="MDPI_4.1_table_caption"}
+`r table_count$label_text("p_value_comparison")`.
+Binomial test results on log-p-value differences to p-values generated by raw scan-centric IMF intensities.
+P-values were adjusted using the Bonferroni method.
+:::
+
+::: {custom-style="MDPI_4.2_table_body"}
+```{r p_value_comparison_table}
+tar_load(lung_binomial)
+binomial_order = data.frame("comparison" = c("corrected", "xcalibur", "msnbase"))
+lung_binomial = dplyr::left_join(binomial_order,
+                                 lung_binomial,
+                                 by = "comparison")
+lung_binomial %>%
+  dplyr::transmute(Comparison = comparison,
+                `N-Positive` = n_positive,
+                `N-Negative` = n_negative,
+                `P-Value` = sprintf("%.02e", p.value),
+                `Adj-P-Value` = sprintf("%.02e", p.adjust)) %>%
+  flextable() %>%
+  align(j = c(4, 5), align = "right", part = "all") %>%
+  autofit()
+```
 :::
@@
 
-```{r increment_frequency}
-equation_count$increment("mzfrequency")
-```
-
 ::: {custom-style="MDPI_3.1_text"}
@@
 ::: {custom-style="MDPI_3.9_equation"}
-`r format_equation('frequency = intercept + x* \\frac{1}{\\sqrt{mz}} + y * \\frac{1}{\\sqrt[3]{mz}}', 'mzfrequency')`
-:::
-
-::: {custom-style="MDPI_3.9_equation"}
-$$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
+`r format_equation('frequency = a + x* \\frac{1}{\\sqrt{mz}} + y * \\frac{1}{\\sqrt[3]{mz}}', 'mzfrequency')`
 :::
@@
 ::: {custom-style="MDPI_3.9_equation"}
-$$mz = intercept + x* \frac{1}{frequency} + y*\frac{1}{frequency^2} + z*\frac{1}{frequency^3}$$
+`r format_equation('mz = a + x * \\frac{1}{frequency} + y * \\frac{1}{frequency^2} + z *\\frac{1}{frequency^3}', 'frequencymz')`
 :::
@@
 ::: {custom-style="MDPI_3.9_equation"}
-$$\ln{intensity} = intercept + a*position+b*position^2$$
+`r format_equation('\\ln{intensity} = a + x * position + y * position^2', 'modelintensity')`
 :::
@@
 Our previously described SMIRFE algorithm [@mitchellSmallMoleculeIsotope2019] was used to assign molecular formulas to scan-centric characterized peaks in an untargeted manner. 
-For the lipid samples, an initial EMF database was generated using an M/Z limit of 1605 M/Z, and maximum numbers for each element were set to C: 130, N: 7, O: 28, P: 3, H: 230. 
+For the lipid samples, an initial EMF database was generated using an m/z limit of 1605 m/z, and maximum numbers for each element were set to C: 130, N: 7, O: 28, P: 3, H: 230. 
 Assigned formulas were allowed to have K, Na, H, and NH4 adducts (only positive mode samples were assigned). 
@@
 ::: {custom-style="MDPI_3.9_equation"}
-$$log(NAP_{P1} / NAP_{P2}) - log(Int_{P1} / Int_{P2}) \approx 0$$
+`r format_equation('\\ln(NAP_{P1} / NAP_{P2}) - \\ln(Int_{P1} / Int_{P2}) \\approx 0', 'napintensity')`
 :::
 
+::: {custom-style="MDPI_3.1_text"}
 SMIRFE assignments include the NAP for each of the isotopologue molecular formulas (IMFs) in the particular labelled and adducted EMF.
@@
 This calculation is done for the final characterized intensity, as well as the intensities at the scan level, and for both the *raw* intensity and *corrected* intensity, and for matched Xcalibur peak intensities, and matched MSnbase peak intensities if there were two or more matching peaks in the labeled, adducted EMF.
+:::
 

### 2022-04-07 R/lungcancer_statistics.R

@@ -1,17 +1,46 @@
-create_sample_info_df = function(sample_list_file){
-  sample_info = read.table(sample_list_file, header = TRUE, sep = "\t")
+create_sample_info_df = function(sample_file,
+                                 bad_samples,
+                                 patient_file){
+  # tar_load(sample_file)
+  # tar_load(bad_samples)
+  # tar_load(patient_file)
+  sample_info = read.table(sample_file, header = TRUE, sep = "\t")
+
+  patient_subtype = read.table(patient_file, sep = ",", header = TRUE,
+                               stringsAsFactors = FALSE)
+  patient_subtype$Patient = as.character(patient_subtype$Patient)
+
+  bad_str = paste0(bad_samples, collapse = "|")
+  sample_info = sample_info %>%
+    dplyr::filter(!grepl(bad_str, sample))
+  sample_info = dplyr::mutate(sample_info, patient = stringr::str_extract(sample, "(uk)?[[:digit:]]*"))
+  dup_sample = sample_info$sample[duplicated(sample_info$sample)]
+
+  sample_info = sample_info %>%
+    dplyr::filter(!(sample %in% dup_sample))
+
+  sample_info = dplyr::left_join(sample_info, patient_subtype, by = c("patient" = "Patient"))
+
+  sample_info[sample_info$disease == "non-cancer", "Cancer.Type"] = "non-cancer"
+
+  granuloma_patients = dplyr::filter(sample_info, (Cancer.Type %in% "Granuloma")) %>%
+    dplyr::pull(patient)
+
+  sample_info = dplyr::filter(sample_info, !(patient %in% granuloma_patients))
+
   sample_info
 }
 
-extract_scancentric_imfs = function(emf_file){
+extract_scancentric_imfs = function(emf_file, which = "raw"){
   emf_data = readRDS(emf_file)
-  imf_height = smirfeTools::extract_imf_emf_data(emf_data, by = "IMF", intensity = Height)
-  imf_corrected_height = smirfeTools::extract_imf_emf_data(emf_data, by = "IMF", intensity = CorrectedLog10Height)
 
-  raw_intensity = imf_height$intensity
-  cor_intensity = 10^imf_corrected_height$intensity
+  if (which %in% "raw") {
+    imf_height = smirfeTools::extract_imf_emf_data(emf_data, by = "IMF", intensity = Height)
+  } else {
+    imf_height = smirfeTools::extract_imf_emf_data(emf_data, by = "IMF", intensity = CorrectedLog10Height)
[…]

### 2022-04-07 packages.R

@@ -25,6 +25,7 @@ library(rmarkdown)
 library(documentNumbering)
 library(flextable)
 library(ggforce)
+library(visualizationQualityControl)
 suppressPackageStartupMessages(library(MSnbase))
 suppressPackageStartupMessages(library(ComplexHeatmap))
 conflict_prefer("combine", "dplyr")

### 2022-04-08 doc/peakcharacterization_manuscript.Rmd
methods around the p-value comparison

@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Multi-Class F-Statistics
+### Differential Analysis of Large Dataset
 :::
 
+```{r load_qcqa}
+tar_load(qcqa)
+n_out = qcqa$outlier_data %>%
+  dplyr::group_by(sample_class) %>%
+  dplyr::summarise(n_out = sum(outlier))
+n_out_vector = n_out$n_out
+names(n_out_vector) = n_out$sample_class
+```
+
 ::: {custom-style="MDPI_3.1_text"}
-To evaluate the F-statistic between multi-class samples, a much larger set of samples was used.
-In this case, 181 matched non-cancer and cancer samples previously used in the HPD detection manuscript [@mitchellNewMethodsIdentify2018].
-All 181 samples were peak characterized using the most stringent method below as well as peaks generated from the MSnbase R package [ref], assigned using SMIRFE, and then peaks matched by shared EMFs across samples (see Supplemental materials for a description of peak matching and deciding the most likely EMF).
-For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using scan-level peak-characterization and those from MSnbase.
+To compare p-value changes in a large multi-class sample dataset, we used a much larger set of samples.
+We started with the 181 matched non-cancer and cancer samples previously used for HPD detection and lipidomics of non-small-cell-lung-cancer (NSCLC) [@mitchellNewMethodsIdentify2018; @mitchellUntargetedLipidomicsNonSmall2021].
+All samples were characterized using the full scan-centric workflow, as well as peaks generated from the MSnbase R package [@gattoMSnbaseBioconductorPackage2012; @gattoMSnbaseEfficientElegant2020], assigned using SMIRFE, and then peaks matched by shared EMFs across samples (see other methods for a description of peak matching and deciding the most likely EMF).
+After extraction of the scan-centric IMF peak locations across all samples, Xcalibur and MSnbase peaks were matched within each sample, and locations and intensities extracted.
+
+Of the starting 182 samples, five did not finish during scan-centric characterization due to not having any scans left during normalization.
+In addition, some of the samples were acquired multiple times, creating duplicate samples.
+Not knowing which sample run was most appropriate to use, we removed all of the duplicated samples.
+Using information-content-informed Kendall-tau correlation, we compared the median correlations of each sample to all others in the same disease class for outliers, and removed `r n_out_vector["cancer"]` cancer and `r n_out_vector["non-cancer"]` samples.
+
+Each samples intensities were normalized by dividing by the median intensities from that sample.
+Only those peaks present in 50% of both the non-cancer and cancer samples were kept for differential analysis.
+Differential analysis used the logged intensity values, with p-values calculated using rowttests function from the genefilter Bioconductor package (v1.76.0) [@gentlemanGenefilterMethodsFiltering2021], removing any missing values before calculation.
+P-values were adjusted using the Benjamini-Hochberg method in the p.adjust function from the base R stats package (v4.1.0) [@rcoreteamLanguageEnvironmentStatistical2021a].
+
+For comparisons across peak sources, we converted both the raw p-values to log-p-values by calculating -1 * Log10(p-value).
+The reference peak source is the raw scan-centric p-values.
 :::

### 2022-04-09 doc/peakcharacterization_manuscript.Rmd
updated some text around HPD and marking FSD instead of removing

@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Naively Averaged Data Have Bad Relative Intensities
+### Naïvely Averaged Data Have Bad Relative Intensities
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Removal of High Peak Density Artefacts
+### Mitigation of High Peak Density Artefacts
 :::
@@
 We have previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [@mitchellNewMethodsIdentify2018].
-Ideally, the peak characterization procedure should reduce their presence in the resultant reported peaks.
+Ideally, the peak characterization procedure should reduce and mitigate their presence and effect in the resultant reported peaks.
 Their presence should be minimized by removing noise peaks, and removing peak regions that have multiple reported peaks from the same scan.
 However, we expect they may also present as characterized peaks that have larger than expected frequency level standard deviations (FSD) when calculated across scans.
-These peaks can be detected by simply examining the distribution of FSDs and removing those that are outliers.
-To verify the removal of HPD peaks, we converted centroided m/z's from XCalibur to frequency values using the previously calculated values for that sample.
-Peak density was measured using a sliding window ten points wide and a stride of one point.
+These peaks can be detected by simply examining the distribution of FSDs and marking outlier peaks.
+To verify the mitigation of HPD regions, we converted centroided m/z's from XCalibur to frequency values using the previously calculated model for that sample, and measured peak density to detect HPD regions, and compare them with the scan-centric peaks and FSD outliers.
+XCalibur peak density was measured using a sliding window ten points wide and a stride of one point.
 
-`r figure_count$label_text("hpd_compare_fsd")` shows a single HPD site detected in the ECF 2 sample, with the peaks from Xcalibur, as well as various scan-level processing and the peaks from centroiding using MSnbase.
+`r figure_count$label_text("hpd_compare_fsd")` shows a single HPD site detected in the 2ecf sample, with the peaks from Xcalibur, as well as various scan-centric processing and the peaks from centroiding using MSnbase.
 From this figure, we can see that the point density based noise filtering removes a large number of the peaks in the HPD site, while the frequency standard deviation removes further peaks that may be suspect.
@@
 ```{r increment_p_value}
+tar_load(lung_compare)
+lung_compare = lung_compare %>%
+  dplyr::mutate(compare_id = dplyr::case_when(
+    id.test %in% "corrected" ~ "Corrected",
+    id.test %in% "xcalibur" ~ "XCalibur",
+    id.test %in% "msnbase" ~ "MSnbase"
+  ))
+
+compare_split = split(lung_compare, lung_compare$compare_id)
 figure_count$increment("p_value_comparison")
-table_count$increment("p_value_comparison")
 ```
@@
 One way to test this is to compare p-values generated from the different intensities.
-`r figure_count$label_text("p_value_comparison")` compares the -1 * Log10(p-values) from XX or XX isotope resolved molecular formulas (IMFs) assigned by SMIRFE for the scan-centric peaks and then peaks matched and intensities extracted from the other methods.
-This shows that the p-values generated by the raw scan-centric intensities are generally the same if we use the intensities corrected for missing scans (on the line), and the p-values using raw scan-centric are better than either the XCalibur or MSnbase generated peak intensities (below the line due to being worse in XCalibur and MSnbase).
-We can test whether these differences are random or not using a binomial test of the actual difference in log-p-values.
-The binomial test p-values of the differences are reported in `r table_count$label_text("p_value_comparison")`.
 :::
@@
 ::: {custom-style="MDPI_5.2_figure"}
-```{r p_value_comparison, dn_id = figure_count, fig.width = 4, fig.height = 9}
-tar_load(lung_compare)
-
-lung_compare$id.test = factor(lung_compare$id.test, c("corrected", "xcalibur", "msnbase"))
 
@@
 Two of the cancer lipid samples (49Cpos and 97Cpos) were used for examples in this work.
-The rest were used for the examination of F-statistics.
+The rest were used for the examination of changes in differential analysis p-values using different peak intensities (see **Differential Analysis of Large Dataset**)..
 :::

### 2022-04-09 _targets.R

@@ -309,8 +309,13 @@ lungcancer_tar = list(
                patient_file
              )),
   tar_target(scancentric_json,
-             "data/data_output/lung_json_data_2022-04-02.rds",
+             "data/data_output/lung_data/lung_json_data_2022-04-02.rds",
              format = "file"),
+  tar_target(scancentric_med_file_corrected,
+             "data/data_output/lung_data/corrected_medians_2022-04-02.rds",
+             format = "file"),
+  tar_target(scancentric_medians_corrected,
+             readRDS(scancentric_med_file_corrected)),
   tar_target(scancentric_medians,
              get_scancentric_medians(scancentric_json)),
   tar_target(emf_file,
@@ -360,7 +365,7 @@ lungcancer_tar = list(
                sample_info,
                id = "raw"
              )),
-  tar_target(lung_corrected,
+  tar_target(lung_corrected_wrong,
              run_imf_models(
                scancentric_imfs_corrected,
                scancentric_medians,
@@ -384,6 +389,24 @@ lungcancer_tar = list(
                sample_info,
                id = "xcalibur"
              )),
+  tar_target(lung_compare_wrong,
+             compare_ttests(
+               list(corrected = lung_corrected_wrong,
+                    raw = lung_raw,
+                    msnbase = lung_msnbase,
+                    xcalibur = lung_xcalibur),
+               reference = "raw")
+             ),
+  tar_target(lung_binomial,
+             binomial_compare(lung_compare)),
+  tar_target(lung_corrected,
+             run_imf_models(
+               scancentric_imfs_corrected,
+               scancentric_medians_corrected,
+               qcqa,
+               sample_info,
+               id = "corrected"
+             )),
   tar_target(lung_compare,
              compare_ttests(
                list(corrected = lung_corrected,
@@ -391,9 +414,11 @@ lungcancer_tar = list(
                     msnbase = lung_msnbase,
                     xcalibur = lung_xcalibur),
                reference = "raw")
-             ),
-  tar_target(lung_binomial,
-             binomial_compare(lung_compare))
+  ),
+  tar_target(lung_compare_ttest,
+             ttest_compare_diffs(
+               lung_compare
+             ))
 )
 
 list(pkg_tar,

### 2022-04-09 doc/peakcharacterization_manuscript.Rmd
new version of p-value comparison by peak source
using a t-test for examination of the differences we see, and putting the summary values directly on the sina-plot so we don't actually need a table

@@
 One way to test this is to compare p-values generated from the different intensities.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-
-lung_compare %>%
-  ggplot(aes(x = log_p.ref,
-             y = log_p.test)) +
-  geom_point() +
+`r figure_count$label_text("p_value_comparison")` compares the -1 * Log10(p-values) from `r nrow(compare_split$Corrected)` (corrected and XCalibur) or `r nrow(compare_split$MSnbase)` (MSnbase) isotope resolved molecular formulas (IMFs) assigned by SMIRFE for the scan-centric peaks and then peaks matched and intensities extracted from the other methods.
+Although all of the p-vaues are somewhat (and statistically) different from those reported by the raw intensities, there are some interesting patterns of differences.
+MSnbase generated p-values show the widest distribution of differences, as well as the smallest number of peaks that are present in 50% of both the cancer and non-cancer samples.
+This echoes the patterns of low overlapping assigned peaks observed in both ECF samples, where many of the scan-centric amino-acid peaks only found one match from the MSnbase peaks.
+Surprisingly, the corrected intensities generated very different p-values compared to the raw intensities, and XCalibur showed the most agreement with the raw p-values.
+We tested the statistical significance of these differences using a t-test of the actual difference in log-p-values (shown in `r figure_count$label_text("p_value_comparison")`D).
+:::
+
+::: {custom-style="MDPI_5.2_figure"}
+```{r p_value_comparison, dn_id = figure_count, fig.width = 8, fig.height = 8}
+tar_load(lung_compare_ttest)
+
+compare_plots = purrr::imap(compare_split, function(in_compare, id){
+  out_plot = in_compare %>%
+    ggplot(aes(x = log_p.ref, y = log_p.test)) +
     geom_abline(slope = 1, color = "red") +
-  facet_wrap(~ id.test, ncol = 1) +
-  labs(x = "-1 * Log10(P-Value) Raw Scan-Centric",
-       y = "-1 * Log10(P-Value) Other")
+    geom_point(alpha = 0.5) +
+    coord_equal() +
+    labs(x = "Log-P-Value Raw",
+         y = paste0("Log-P-Value ", id))
+  out_plot
+})
+
+long_ttest = lung_compare_ttest %>%
+  dplyr::select(p.adjust, comparison, estimate, conf.low, conf.high) %>%
+  tidyr::pivot_longer(cols = !comparison, names_to = "measure", values_to = "value") %>%
+  dplyr::mutate(compare_id = dplyr::case_when(
+    comparison %in% "corrected" ~ "Corrected",
+    comparison %in% "xcalibur" ~ "XCalibur",
+    comparison %in% "msnbase" ~ "MSnbase"
+  ))
+
+long_pvalue = long_ttest %>%
+  dplyr::filter(measure %in% "p.adjust")
+
+long_measure = long_ttest %>%
+  dplyr::filter(!(measure %in% "p.adjust"))
+
+compare_sina = lung_compare %>%
+  ggplot(aes(x = compare_id, y = p_diff)) +
+  geom_hline(yintercept = 0, color = "red") +
+  geom_sina(alpha = 0.5) +
+  geom_point(data = long_measure, aes(x = compare_id, y = value), color = "green") +
+  geom_text(data = long_pvalue, aes(x = compare_id, y = -5, label = format(value, digits = 2))) +
+  labs(x = "", y = "Log-P-Value Difference")
+
+lung_plot = (compare_plots$Corrected | compare_plots$MSnbase) / (compare_plots$XCalibur | compare_sina) +
+  plot_annotation(tag_levels = "A")
+
+lung_plot
 ```
@@
 `r figure_count$label_text("p_value_comparison")`.
-Log-p-values generated by comparing non-cancer and cancer sample IMFs using peak intensities from different methods.
-:::
-
-::: {custom-style="MDPI_4.1_table_caption"}
-`r table_count$label_text("p_value_comparison")`.
-Binomial test results on log-p-value differences to p-values generated by raw scan-centric IMF intensities.
-P-values were adjusted using the Bonferroni method.
-:::
-
-::: {custom-style="MDPI_4.2_table_body"}
-```{r p_value_comparison_table}
-tar_load(lung_binomial)
-binomial_order = data.frame("comparison" = c("corrected", "xcalibur", "msnbase"))
-lung_binomial = dplyr::left_join(binomial_order,
-                                 lung_binomial,
-                                 by = "comparison")
-lung_binomial %>%
-  dplyr::transmute(Comparison = comparison,
-                `N-Positive` = n_positive,
-                `N-Negative` = n_negative,
-                `P-Value` = sprintf("%.02e", p.value),
-                `Adj-P-Value` = sprintf("%.02e", p.adjust)) %>%
-  flextable() %>%
-  align(j = c(4, 5), align = "right", part = "all") %>%
-  autofit()
-```
+**A - C**: Log-p-values generated by comparing non-cancer and cancer sample IMFs using peak intensities from different methods.
+Red line denotes perfect agreement.
+**D**: Sina plot of differences in the log-p-values generated by different methods compared to the raw scan-centric log-p-values.
+Also shown are the Bonferroni adjusted p-values from a t-test of the log-p-value differences for each method.
+Green points denote the high, mean, and low-confidence limits reported from the t-test.
+:::
+
 :::

### 2022-04-09 lungcancer_all/lungcancer_smirfe_assignmentprocessing.R
getting the medians for corrected peaks

@@ -91,3 +91,11 @@ all_vote = extract_assigned_data(assigned_data, difference_cutoff = mz_cutoff,
 saveRDS(all_vote, file = "data/data_output/lung_data/lung_voted_all_2022-04-02.rds")
 
 #textme::textme("Lung is all done!")
+all_json_files = dir("/mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-04-02", pattern = "json$", full.names=TRUE)
+all_json_files = grep("assigned", all_json_files, value = TRUE, invert = TRUE)
+
+json_data = purrr::map(all_json_files, function(in_json){
+  tmp_json = jsonlite::fromJSON(in_json)
+})
+names(json_data) = gsub(".json$", "", basename(all_json_files))
+saveRDS(json_data, file = "data/data_output/lung_json_data_2022-04-02.rds")

### 2022-04-09 lungcancer_all/lungcancer_smirfe_correctedmedians.R

@@ -1,0 +1,23 @@
+# cd smirfe_related/smirfe_code/SMIRFE_assigner
+# python3 -m pipenv shell
+# python3 -m AssignMany.py 12 "../../smirfe_dbs/none_nosulfur_1600.db FILE '_assigned.json' '[]' '[\"H\", \"Na\", \"K\", \"NH4\"]' '[1]'" /mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-04-02/*.zip
+# Database limits: C: 130, N: 7, O: 28, S: 0, P: 3, H: 230, max M/Z 1605, adducts: K, Na, NH4, NAP cutoff: 0.4
+library(FTMS.peakCharacterization)
+library(knitrProgressBar)
+
+zip_files <- dir("/mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-04-02",
+                      full.names = TRUE, pattern = ".zip$")
+pb <- progress_estimated(length(zip_files))
+
+corrected_medians <- purrr::map_df(zip_files, function(in_file){
+  zip_ms = ZipMS$new(in_file)
+  zip_ms$load_peak_finder()
+
+  out_df = data.frame(sample = zip_ms$id,
+             median = median(10^zip_ms$peak_finder$peak_regions$peak_data$CorrectedLog10Height))
+  zip_ms$cleanup()
+  update_progress(pb)
+  out_df
+})
+
+saveRDS(corrected_medians, file = "data/data_output/lung_data/corrected_medians_2022-04-02.rds")

### 2022-04-09 R/get_sample_medians.R
code for t-test of p-value differences

@@ -1,0 +1,14 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param lung_json
+#' @return
+#' @author rmflight
+#' @export
+get_sample_medians <- function(lung_json) {
+
+  NULL
+
+}

### 2022-04-09 R/lungcancer_statistics.R

@@ -284,3 +284,15 @@ binomial_compare = function(lung_compare, p_use = "p_diff"){
   binom_res$p.adjust = p.adjust(binom_res$p.value, method = "bonferroni")
   binom_res
 }
+
+ttest_compare_diffs = function(lung_compare, p_use = "p_diff"){
+  split_compare = split(lung_compare[[p_use]], lung_compare$id.test)
+
+  ttest_res = purrr::map_df(split_compare, function(in_compare){
+    test_res = broom::tidy(t.test(in_compare, alternative = "two.sided"))
+
+  })
+  ttest_res$comparison = names(split_compare)
+  ttest_res$p.adjust = p.adjust(ttest_res$p.value, method = "bonferroni")
+  ttest_res
+}

### 2022-04-09 doc/peakcharacterization_manuscript.Rmd
updated methods to include EMF voting

@@
 `r figure_count$label_text("hpd_compare_fsd")` shows a single HPD site detected in the 2ecf sample, with the peaks from Xcalibur, as well as various scan-centric processing and the peaks from centroiding using MSnbase.
-From this figure, we can see that the point density based noise filtering removes a large number of the peaks in the HPD site, while the frequency standard deviation removes further peaks that may be suspect.
+From this figure, we can see that the point density based noise filtering removes a large number of the peaks in the HPD site, while the FSD outliers removes further peaks that may be suspect.
 After removing the high FSD peaks, the number of peaks left in the HPD site are the same or less than those from MSnbase, and at least in this example, look more likely to be real peaks compared to those from MSnbase.
-Therefore, scan-level characterization allows us to keep what are likely **real** peaks in HPD sites, without consideration of peak intensity.
+Therefore, scan-level characterization allows us to keep what are likely **real** peaks in HPD sites, without consideration of peak intensity, and mark peaks that may be artifactual from HPD regions.
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
+### Samples and Overall Processing
+:::
+
+::: {custom-style="MDPI_3.1_text"}
+Two different sets of samples were used to evaluate the various methods, ethylchloroformate (ECF) amino acid samples and non-small-cell-lung-cancer (NSCLC) lipid samples.
+
+The method for generating the amino-acid samples was adapted from a previously published method for performing ECF amino acid derivatization [@yangChloroformateDerivatizationTracing2017]. 
+Two replicate samples were prepared and spectra were obtained for both samples using a Tribrid Fusion Orbitrap at 500k resolution and 150 to 1000m/z.
+
+The collection, preparation, and mass spectrometry analysis of the paired cancer and non-cancer samples has been previously described [@sellersPyruvateCarboxylaseCritical2015]. 
+In summary, cancer and non-cancer samples were acquired from eight-six non-diabetic patients with suspected resectable stage I or IIA non-small cell lung cancer. 
+Written informed consent was collected from all subjects prior to inclusion and all samples were collected under a University of Louisville or University of Kentucky IRB protocol. 
+Lipid extracts were prepared using a modified Folch extraction and reconstituted for direct infusion ultra-high resolution mass spectrometry on a pair of Thermo Tribrid Fusion Orbitrap instruments (FSN10115 and FSN10352, referred to as Fusion 1 and Fusion 2 respectively) coupled to an Advion nanoelectrospray system.
+53 patients (102 spectra total) were acquired from Fusion 1 (serial # FSN10115) and 40 patients (77 spectra total) were acquired from Fusion 2 (serial # FSN10352).
+Fusion 1 samples were exclusively from the University of Louisville while Fusion 2 were a mix.
+Three of the patients had only a cancer or non-cancer sample acquired.
+Two of the cancer lipid samples (49Cpos and 97Cpos) were used for the majority of examples in this manuscript.
+The rest were included for the examination of changes in differential analysis p-values using different peak source intensities (see **Differential Analysis of Large Dataset**).
+:::
+
+::: {custom-style="MDPI_3.5_text_before_list"}
+For each of the two ECF and two lipid samples, the raw data file was converted to profile mzML format.
+Only the MS1 scans were used, and the scan-scan time difference had to be >= 4 seconds, and for the lipid samples only scans acquired before 450 seconds (7.5 minutes) were kept.
+The scan-level data was then processed in the following eight ways:
+:::
+
+::: {custom-style="MDPI_3.8_bullet"}
+* No noise removal, no normalization (noperc_nonorm)
+:::
+::: {custom-style="MDPI_3.8_bullet"}
+* Noise removal, no normalization (perc99_nonorm)
+:::
+::: {custom-style="MDPI_3.8_bullet"}
+* Noise removal, single pass normalization with all peaks (singlenorm)
+:::
+::: {custom-style="MDPI_3.8_bullet"}
+* Noise removal, single pass normalization with high ratio peaks (singlenorm_int)
+:::
+::: {custom-style="MDPI_3.8_bullet"}
+* Noise removal, two pass normalization (doublenorm)
+:::
+::: {custom-style="MDPI_3.8_bullet"}
+* Noise removal, two pass normalization (filtersd)
+:::
+
+::: {custom-style="MDPI_3.8_bullet"}
+* Scans merged and then centroids generated by MSnbase (using `combineSpectra` and `pickPeaks`)
+:::
+
+::: {custom-style="MDPI_3.8_bullet"}
+* Scans merged and peak-list exported by XCalibur
+:::
+
+::: {custom-style="MDPI_2.2_heading2"}
+### Matching Peaks
+:::
+
+::: {custom-style="MDPI_3.9_text"}
+To associate the XCalibur and MSnbase peaks with the scan-centric peaks, a 4 ppm (2ppm low and high) window is calculated for the scan-centric centroid, and if any peaks are found within the window, the one with the smallest m/z difference is kept as the matching peak to the scan-centric one under consideration.
+:::
+
+::: {custom-style="MDPI_2.2_heading2"}
 ### Conversion of m/z to Frequency
@@
 ::: {custom-style="MDPI_3.1_text"}
-The data consists of profile mode m/z spectra from multiple scans encoded as m/z and intensity values for each scan.
+The input data consists of profile mode m/z spectra from multiple scans encoded as m/z and intensity values for each scan.
 No information about the original observed frequency values is available in either the `raw` files or the `mzML` files.
-However, proxy frequency values can be generated by averaging the m/z of adjacent points and dividing them by the difference.
+However, proxy frequency values can be generated by averaging the m/z of adjacent points and dividing them by the m/z difference.
 Ideally, the difference between subsequent points in this proxy frequency space is constant, in practice there is a range of differences in frequency space.
-The *actual, constant* difference can be obtained by examining the median of the frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to m/z) to be within 2% of the mode value.
+The *actual, constant* difference can be obtained by examining the median of the calculated frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to m/z) to be within 2% of the mode value.
 These *useful* points can be used to construct a linear model relating m/z to frequency for individual scans based on the formula:
@@
 ::: {custom-style="MDPI_3.9_equation"}
-`r format_equation('frequency = a + x* \\frac{1}{\\sqrt{mz}} + y * \\frac{1}{\\sqrt[3]{mz}}', 'mzfrequency')`
+`r format_equation('frequency = a + y * \\frac{1}{\\sqrt{mz}} + z * \\frac{1}{\\sqrt[3]{mz}}', 'mzfrequency')`
 :::
@@
 We observed that this single model better preserved the relative ordering of the peaks in both m/z and frequency-space compared to the scan specific models (see Results).
-A single model rather than scan specific models is used for conversion because this was found to maintain the relative ordering of the peaks in m/z and frequency space (see Results).
-
-To convert m/z back into frequency, we can use a similar model without the roots.
+
+To convert m/z back into frequency, we can use a similar model without the roots, as well as an extra simple linear term that doesn't have an equivalent in the frequency model.
 :::
@@
 The tiled windows are one point wide with a stride of one point.
-Each point here is the equivalent of the difference between data points in frequency space, which is a spacing of 0.5.
+Each **point** above is the equivalent of the difference between data points in frequency space, which in these samples have a spacing of 0.5.
 :::
@@
 The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions containing potential signal.
-Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged to create the initial peak regions.
-The presence of zero intensity points are due to flooring implemented by the spectrometer when local spectral quality falls below a certain threshold.
+Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are kept and overlapping sliding windows are merged to create the initial peak regions.
+The presence of zero intensity points in these Thermo-Fisher Orbitrap spectra are primarily due to flooring implemented by the spectrometer when local spectral quality falls below a certain threshold.
 When this flooring is unstable and incomplete, a partial ringing phenomenom is observed [@mitchellNewMethodsIdentify2018].
 
-Within each initial interval region, peaks in each scan are detected (see **Peak Detection**), and their centers are binned by the tiled windows.
+Within each initial interval region, peaks in each scan are detected (see **Peak Detection and Centroided Values**), and their centers are binned by the tiled windows.
 Adjacent tiled windows with non-zero peak counts are merged together, and any zero peak count tiled windows split the initial region into multiple peak interval regions.
@@
 From this weighted parabolic fit, the center, intensity, integrated area and sum-of-square residuals can be extracted for the peak.
-These characteristics are equivalent to the centroided peak center and intensity.
-
-Before further processing, the regions are verified to have only one peak from each scan.
-If a scan has two or more peaks, then the scan level data in that region is discarded.
+The center and intensity from this model are equivalent to the centroided peak center and intensity.
+
+Before further processing, each region is verified to have only one peak from each scan.
+If a scan has two or more peaks in a region, then the scan level data in that region is discarded.
 Any regions that subsequently contain zero peaks are removed.
@@
 Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each scan to all other scans.
-The scan with the lowest summed overall distance is chosen as the *reference* scan, and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the reference scan.
+The scan with the lowest summed overall distance is chosen as the *reference* scan ($scan_{ref}$), and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the $scan_{ref}$.
 This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan order is checked, and those peaks with correlation of greater than 0.5 with scan order are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.
-Peaks correlated to scan order represent an artifact that we speculate results from a gradient in the sample well.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Full Characterization
+Peaks correlated to scan order represent an artifact that we speculate results from a gradient in the sample well and are marked in the final output.
+:::
+
+::: {custom-style="MDPI_2.2_heading2"}
+### Full Scan-Centric Characterization
 :::
@@
 The full set of raw data points for each peak in each scan within a region is known based on the previously detected peaks.
-Therefore, the non-zero intensity, normalized raw data points across scans can be combined, and then characterized again using the weighted parabolic fit previously described.
+Therefore, the non-zero intensity, normalized raw data points across scans can be combined, and then characterized again using the weighted parabolic fit method previously described.
 In addition to the data from the full set of raw points, the means and the standard deviations of the peak height and location can be derived from the scan-level peak characteristics previously calculated.
 
-In addition to these values, the frequency point-to-point median difference is calculated across all of the raw data points for those points that could be used for modeling frequency to m/z, and this difference of a single point from the peak center is calculated in frequency space, and converted to m/z space to provide an "offset" value that is useful to define the search space around the peak for any assignment algorithm.
+In addition to these values, the frequency point-to-point median difference is calculated across all of the raw data points for those points that could be used for modeling frequency to m/z, and this difference of a single point from the peak center is calculated in frequency space, and converted to m/z space to provide an "offset" value that is potentially useful to define the search space around the peak for any assignment algorithm.
 :::
@@
 Ideally, each peak would be observed in every scan.
-However, some peaks are not observed in some scans due to the number of ions falling below the detection threshold.
+However, some peaks are not observed in some scans due to the number of ions falling below the detection threshold, or being excluded after filling the trap.
 This results in a left-censored normal distribution of peak intensities across scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Frequency Standard Deviation Filtering
+### Marking High Frequency Standard Deviation Peaks
 :::
@@
 High-peak-density (HPD) artifacts [@mitchellNewMethodsIdentify2018] present as groups of singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
-Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and peaks with FSDs greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are removed.
+Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and peaks with FSDs greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are marked.
 The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Assigning Scan-Centric Peaks
+### Scan-Centric Peak Assignment
 :::
@@
 For the ECF derivatized amino-acid samples, the EMF database was generated using an m/z limit of 1005 m/z, and maximum numbers for each element were set to C: 100, N: 7, O: 40, H: 230, P: 3, S: 3.
+
 We attempted to create peaks that could be assigned by SMIRFE from the MSnbase and Xcalibur peak lists, but SMIRFE would not assign them given its requirements for scan-level peak characterization data.
@@
 Each IMF represents an isotopologue of a given elemental molecular formula (EMF) (e.g., 13C112C51H1216O6 is an IMF representing the m+13C1 isotopologue of EMF C6H12O6).
-Consistently assigned (i.e., corresponded peaks) are identified using an in-house method we have named EMF voting which is described in Appendix A.
-EMF voting identified 3485 total corresponded peaks across all 157 spectra with 855 corresponded peaks present in 25% or more of either the disease or non-disease group of samples. 
+Consistently assigned (i.e., corresponded peaks) are identified using an in-house method we have named EMF voting.
+
+Elemental molecular formula (EMF) voting was used to match peaks across samples and determine the most likely assignments from SMIRFE. 
+First, for each sample, the assignments are extracted, filtered, and scored.
+Any assignments that contain only peaks that were marked as being questionable (high FSD or correlated with scan number) removed.
+Scores are calculated as 1 – E-value, and for the lipids, EMFs that were classified as lipids had their score multiplied by 2.
+
+For each sample, peaks were grouped to a sample specific EMF by determining the list of shared EMFs across a group of peaks (grouped_EMF). 
+Scores for each formula in the grouped_EMF in the sample were taken as the best score for that formula from available scores in the group. 
+After all sample grouped_EMFs are generated, additional scores for a formula are considered in actual voting by looking for the same unadducted base EMF with different adducts and adding these scores into the final total score.
+
+The pseudo_EMFs are collections of grouped_EMFs across samples. 
+They are generated across samples by iteratively merging grouped_EMFs with shared formulas, creating a new list of formulas in an EMF, and merging any pseudo_EMFs that have shared formulas again.
+
+For each pseudo_EMFs, the most likely formulas are determined by voting. 
+Voting uses the sum of EMF scores across samples, including those from the same formula with a different adduct. 
+Those formulas with a total score in the top 90% of all total scores were considered “winning” formulas and kept as the "voted formula" set. 
+Any peaks that did not originally have a "voted formula" were then checked to see if the peak location was within previously defined tolerance, and if the ordered peak intensities for the set of peaks from a sample are in the same order as the natural abundance probabilities of IMFs for the voted EMF.
+If so, then the peaks are kept as having the "voted formula" set.
+
+After voting, all of the peaks are checked to make sure that the peak locations are within previously defined m/z specific cutoffs. 
+Any peak outside of its specific cutoff is removed.
+
+Finally, those pseudo_EMFs that share greater than 50% of their peaks are merged together, and voting on the EMFs is performed again.
+
+Peak locations from our custom data processing pipeline are reported both in m/z and frequency, which are derived from the m/z values. 
+The frequency values are more reliable (i.e., more consistent with less variance), but differ between instruments. 
+Therefore, for each set of samples from a particular instrument, we use high confidence assignments (e-value <= 0.1, m/z <= 600) to derive a frequency cutoff using the mode of the distribution of frequency standard deviations across both groups of samples.
+This frequency cutoff is used to do EMF voting within an instrument. 
+The m/z standard deviations of the voted peaks are then fit to m/z using a generalized additive model, and the standard deviation of the predicted m/z values are multiplied by 2 to derive an m/z cutoff so that voting can be performed across the instruments.
+
+```{r load_lipid_full_data}
+tar_load(scancentric_imfs_raw)
+```
+
+EMF voting identified `r nrow(scancentric_imfs_raw$intensity)` total corresponded peaks across all `r ncol(scancentric_imfs_raw$intensity)` spectra.
 All lipid isotopologue intensities were normalized by dividing the isotopologue intensity by the median intensity of all the peaks in the sample. 
-For differential abundance analysis, missing values were removed. 
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Matching Peaks
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-We matched the scan-centric peaks to the other methods by looking for the closest m/z peak that was within 2 ppm of the scan-centric reported m/z.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-To compare p-value changes in a large multi-class sample dataset, we used a much larger set of samples.
+To compare p-value changes in a large multi-class sample dataset, we used the full set of NSCLC lipid samples described earlier.
 We started with the 181 matched non-cancer and cancer samples previously used for HPD detection and lipidomics of non-small-cell-lung-cancer (NSCLC) [@mitchellNewMethodsIdentify2018; @mitchellUntargetedLipidomicsNonSmall2021].
-All samples were characterized using the full scan-centric workflow, as well as peaks generated from the MSnbase R package [@gattoMSnbaseBioconductorPackage2012; @gattoMSnbaseEfficientElegant2020], assigned using SMIRFE, and then peaks matched by shared EMFs across samples (see other methods for a description of peak matching and deciding the most likely EMF).
+All samples were characterized using the full scan-centric workflow, assigned using SMIRFE (see **Scan-Centric Peak Assignments**), and then peaks matched by shared EMFs across samples (see **Consistently Assigned Lipid Spectral Feature (Corresponded Peak) Generation and Peak Intensity Normalization**).
 After extraction of the scan-centric IMF peak locations across all samples, Xcalibur and MSnbase peaks were matched within each sample, and locations and intensities extracted.
 
-Of the starting 182 samples, five did not finish during scan-centric characterization due to not having any scans left during normalization.
+Of the starting 181 samples, five did not finish during scan-centric characterization due to not having any scans left during normalization.
 In addition, some of the samples were acquired multiple times, creating duplicate samples.
 Not knowing which sample run was most appropriate to use, we removed all of the duplicated samples.
-Using information-content-informed Kendall-tau correlation, we compared the median correlations of each sample to all others in the same disease class for outliers, and removed `r n_out_vector["cancer"]` cancer and `r n_out_vector["non-cancer"]` samples.
-
-Each samples intensities were normalized by dividing by the median intensities from that sample.
+Using information-content-informed Kendall-tau correlation [@flightInformationContentInformedKendalltauCorrelation2022], we compared the median correlations of each sample to all others in the same disease class for outliers, and removed `r n_out_vector["cancer"]` cancer and `r n_out_vector["non-cancer"]` samples.
+
+Each samples intensities were normalized by dividing by the median intensities from that sample and intensity method.
 Only those peaks present in 50% of both the non-cancer and cancer samples were kept for differential analysis.
-Differential analysis used the logged intensity values, with p-values calculated using rowttests function from the genefilter Bioconductor package (v1.76.0) [@gentlemanGenefilterMethodsFiltering2021], removing any missing values before calculation.
+This resulted in `r nrow(compare_split$Corrected)` IMF peaks for differential analysis.
+Differential analysis used the logged intensity values, with p-values calculated using `rowttests` function from the `genefilter` Bioconductor package (v1.76.0) [@gentlemanGenefilterMethodsFiltering2021], removing any missing values before calculation.
 P-values were adjusted using the Benjamini-Hochberg method in the p.adjust function from the base R stats package (v4.1.0) [@rcoreteamLanguageEnvironmentStatistical2021a].
@@
 For comparisons across peak sources, we converted both the raw p-values to log-p-values by calculating -1 * Log10(p-value).
-The reference peak source is the raw scan-centric p-values.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Samples and Overall Processing
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-The method for generating the amino-acid dataset was adapted from a previously published method for performing ethylchloroformate (ECF) amino acid derivatization [@yangChloroformateDerivatizationTracing2017].
-Two replicate samples were prepared and spectra were obtained for both samples using a Tribrid Fusion Orbitrap at 500k resolution and from 150 to 1000m/z.
-
-The collection, preparation, and mass spectrometry analysis of the paired cancer and non-cancer samples has been previously described [@sellersPyruvateCarboxylaseCritical2015]. 
-In summary, cancer and non-cancer samples were acquired from eight-six non-diabetic patients with suspected resectable stage I or IIA non-small cell lung cancer. 
-Written informed consent was collected from all subjects prior to inclusion and all samples were collected under a University of Louisville or University of Kentucky IRB protocol. 
-Lipid extracts were prepared using a modified Folch extraction and reconstituted for direct infusion ultra-high resolution mass spectrometry on a pair of Thermo Tribrid Fusion Orbitrap instruments (FSN10115 and FSN10352, referred to as Fusion 1 and Fusion 2 respectively) coupled to an Advion nanoelectrospray system.
-53 patients (102 spectra total) were acquired from Fusion 1 and 40 patients (77 spectra total) were acquired from Fusion 2.
-Fusion 1 samples were exclusively from the University of Louisville while Fusion 2 were a mix.
-Three of the patients had only a cancer or non-cancer sample acquired.
-Two of the cancer lipid samples (49Cpos and 97Cpos) were used for examples in this work.
-The rest were used for the examination of changes in differential analysis p-values using different peak intensities (see **Differential Analysis of Large Dataset**)..
-:::
-
-::: {custom-style="MDPI_3.5_text_before_list"}
-For each raw data file in profile mzML format, they were processed in these six ways:
-:::
-
-::: {custom-style="MDPI_3.8_bullet"}
-* No noise removal, no normalization, no frequency SD filtering (noperc_nonorm)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, no normalization, no frequency SD filtering (perc99_nonorm)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, single pass normalization with all peaks, no frequency SD filtering (singlenorm)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, single pass normalization with high ratio peaks, no frequency SD filtering (singlenorm_int)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, two pass normalization, no frequency SD filtering (doublenorm)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, two pass normalization, frequency SD filtering (filtersd)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Scan level peak centroids generated by MSnbase (using `pickPeaks` on the individual scans)
-:::
-::: {custom-style="MDPI_3.8_bullet"}
-* Scans merged and then centroids generated by MSnbase (using `combineSpectra` and `pickPeaks`)
-:::
-
-::: {custom-style="MDPI_3.5_text_before_list"}
-For the two sets of MSnbase peaks, we also matched scan level peaks to merged peaks via two methods:
-:::
-
-::: {custom-style="MDPI_3.8_bullet"}
-* All scan level peaks within a 2ppm window are associated with the *merged* peak
-:::
-
-::: {custom-style="MDPI_3.8_bullet"}
-* Using the *filtersd* peaks, merged and scan level peaks are associated if they are within the start and stop m/z of the *filtersd* peaks
+The reference peak p-value is the raw scan-centric p-values.
+:::
+
+::: {custom-style="MDPI_2.1_heading2"}
+### Software Used
+:::
+
 :::

### 2022-04-09 doc/peakcharacterization_manuscript.Rmd
citing all the software packages we used

@@
 
-:::
+Thermo-Fisher XCalibur was used to export peak lists from all samples used.
+SMIRFE v 1.0 [@mitchellSmallMoleculeIsotope2019] running under Python 3.8 [@vanrossumPython3] was used for assignments.
+Lipid classifications were generated by LipidClassifier v 1.0 [@mitchellDerivingLipidClassification2020].
+All other calculations were performed in R v 4.1.0 [@rcoreteam_rlanguage_2020].
+The targets package v 0.10.0 was used to control processing and aggregation of results [@landauTargetsPackageDynamic2021], and renv v 0.15.3 [@usheyRenvProjectEnvironments2022] to create a reproducible R package environment.
+Plots were generated using ggplot2 v 3.3.5 [@wickhamGgplot2ElegantGraphics2016], patchwork v 1.1.1 [@pedersenPatchworkComposerPlots2020], ComplexHeatmap v 2.10.0 [@guComplexHeatmapsReveal2016a], ggridges v 0.5.3 [@wilkeGgridgesRidgelinePlots2021], and ggforce v 0.3.3 [@pedersenGgforceAcceleratingGgplot22021].
+MSnbase v 2.20.4 [@gattoMSnbaseEfficientElegant2020; @gattoMSnbaseBioconductorPackage2012] provided facilities for reading in scan-level data, merging scans and calculating centroided peaks.
+ICI-Kt correlation values among lipid samples were calculated using ICIKendallTau v 0.1.16 [@flightInformationContentInformedKendalltauCorrelation2022].
+Outlier lipid samples were determined using visualizationQualityControl v 0.4.7 [@flightVisualizationQualityControlDevelopmentVisualization2021].
+Specific data manipulation facilities were provided by dplyr v 1.0.8 [@wickhamDplyrGrammarData2022], tidyr v 1.2.0 [@wickhamTidyrTidyMessy2022], furrr v 0.2.3 [@vaughanFurrrApplyMapping2021a], 
+This manuscript was generated from rmarkdown v 2.11 [@allaireRmarkdownDynamicDocuments2021; @xieMarkdownDefinitiveGuide2018; xieMarkdownCookbook2020].
+
+::: {custom-style="MDPI_2.1_heading2"}
+### Availability
+:::
+
+The code used in the reporting of results is available on GitHub at https://github.com/MoseleyBioinformaticsLab/manuscript.peakCharacterization, and archived on Zenodo [ref].
+The scan-centric peak-characterization is packaged as an R package at https://github.com/MoseleyBioinformaticsLab/FTMS.peakCharacterization, and the specific version used in this work is archived on Zenodo [ref].
 

### 2022-04-09 doc/peakcharacterization_manuscript.Rmd
fixing styles

@@
 
-::: {custom-style="MDPI_3.9_text"}
+::: {custom-style="MDPI_3.1_text"}
 To associate the XCalibur and MSnbase peaks with the scan-centric peaks, a 4 ppm (2ppm low and high) window is calculated for the scan-centric centroid, and if any peaks are found within the window, the one with the smallest m/z difference is kept as the matching peak to the scan-centric one under consideration.
@@
 
-::: {custom-style="MDPI_2.1_heading2"}
+::: {custom-style="MDPI_2.2_heading2"}
 ### Software Used
@@
 
+::: {custom-style="MDPI_3.1_text"}
 Thermo-Fisher XCalibur was used to export peak lists from all samples used.
@@
 This manuscript was generated from rmarkdown v 2.11 [@allaireRmarkdownDynamicDocuments2021; @xieMarkdownDefinitiveGuide2018; xieMarkdownCookbook2020].
-
-::: {custom-style="MDPI_2.1_heading2"}
+:::
+
+::: {custom-style="MDPI_2.2_heading2"}
 ### Availability
@@
 
+::: {custom-style="MDPI_3.1_text"}
 The code used in the reporting of results is available on GitHub at https://github.com/MoseleyBioinformaticsLab/manuscript.peakCharacterization, and archived on Zenodo [ref].
 The scan-centric peak-characterization is packaged as an R package at https://github.com/MoseleyBioinformaticsLab/FTMS.peakCharacterization, and the specific version used in this work is archived on Zenodo [ref].
+:::
 

### 2022-04-11 R/qcqa_analysis.R
getting the coefficients qcqa example
related to #19

@@ -1,0 +1,30 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param coefficients_data
+#' @return
+#' @author rmflight
+#' @export
+coefficients_analysis <- function(coefficients_data) {
+
+  n_freq = purrr::map_dbl(coefficients_data, function(.x){
+    length(.x$frequency$bad)
+  })
+
+  example_freq = coefficients_data[[(which(n_freq == 7)[1])]]
+
+  df_freq = data.frame(coefficient = example_freq$frequency$coefficients,
+                       sample = gsub(".zip$", "", basename(example_freq$zip)))
+  df_freq = df_freq %>%
+    dplyr::mutate(outlier = coefficient %in% example_freq$frequency$bad)
+
+  n_norm = purrr::map_dbl(coefficients_data, function(.x){
+    length(.x$normalization)
+  })
+
+  list(n_coef_out = n_freq,
+       example = df_freq)
+
+}

### 2022-04-11 _targets.R

@@ -421,6 +421,16 @@ lungcancer_tar = list(
              ))
 )
 
+qcqa_tar = list(
+  tar_target(coefficients_file,
+             "data/data_output/lung_data/lung_scandiffs_2022-04-02.rds",
+             format = "file"),
+  tar_target(coefficients_data,
+             readRDS(coefficients_file)),
+  tar_target(coefficients_qcqa,
+             coefficients_analysis(coefficients_data))
+)
+
 list(pkg_tar,
      data_tar,
      method_tar,
@@ -440,4 +450,5 @@ list(pkg_tar,
      assigned_match_tar,
      aa_tar,
      lipid_tar,
-     lungcancer_tar)
+     lungcancer_tar,
+     qcqa_tar)

### 2022-04-11 doc/peakcharacterization_manuscript.Rmd
adding results for coefficients qcqa

@@
 
-:::
+::: {custom-style="MDPI_2.2_heading2"}
+### Quality Control and Quality Analysis
+:::
+
+```{r load_increment_qcqa}
+figure_count$increment("coefficients_qcqa")
+figure_count$increment("frequency_mode_qcqa")
+tar_load(coefficients_qcqa)
+```
+
+::: {custom-style="MDPI_3.1_text"}
+Having a scan-centric workflow for generating the peak centroids means that we have opportunities to evaluate the scans as a whole, as well as the peaks across scans.
+For example, we can mark peaks with unusually high frequency standard deviations (FSD) and mitigate the presence of high-peak-density artefacts (see **Mitigation of High Peak Density Artefacts**).
+We also sometimes find that from initial peak region that has been marked as a single region, still results in two peaks being characterized in a single scan.
+This results in that peak being removed from that scan entirely.
+
+The conversion of m/z to frequency also provides opportunities to evaluate each scan and whether it should be kept for further processing.
+One way is to examine the coefficients of the square-root term for each scan (see term **y** in Equation `r equation_count$label_text("mzfrequency")`) and look for outliers there.
+Across the `r length(coefficients_qcqa$n_coef_out)` raw files that completed scan-centric processing, `r sum(coefficients_qcqa$n_coef_out > 0)` had at least one scan removed based on the square-root coefficient.
+`r figure_count$label_text("coefficients_qcqa")`A shows a histogram of the number of scans removed based on being outliers of the overall distribution.
+`r figure_count$label_text("coefficients_qcqa")`B shows what outlier scan coefficients look like in a single sample.
+:::
+
+::: {custom-style="MDPI_5.2_figure"}
+```{r coefficients_qcqa}
+coef_df1 = data.frame(N_out = coefficients_qcqa$n_coef_out)
+coef_g1 = coef_df1 %>%
+  ggplot(aes(x = N_out)) + 
+  geom_histogram() +
+  labs(x = "# of Scan Outliers", y = "# of Samples")
+
+coef_g2 = coefficients_qcqa$example %>%
+  ggplot(aes(x = sample, y = coefficient, color = outlier, group = sample)) +
+  geom_sina(size = 2) +
+  labs(x = "", y = "Square-Root Term") +
+  theme(legend.position = c(0.1, 0.8))
+
+(coef_g1 | coef_g2) + plot_annotation(tag_levels = "A")
+```
+:::
+
+::: {custom-style="MDPI_5.1_figure_caption"}
+`r figure_count$label_text("coefficients_qcqa")`.
+**A**: Number of scan outliers for each sample in the NSCLC dataset.
+**B**: Single example for sample 199Cpos of the scan square-root terms marked as outliers and removed from further consideration.
+:::
+
+One other check we implemented in the conversion to frequency is checking the range of frequency point-point differences allowed for creating the m/z to frequency model, and making sure that they are the same across scans.
+The initial set of scans we allowed to use included **all** of the MS1 scans, including the precursor scans for the MS2 scans. 
+The samples were acquired as a mix of MS1 only scans for the first 7.5 minutes, and then a mix of MS1 precursor scans followed by MS2 scans for the second 7.5 minutes.
+In 20 samples, the MS1 precursor scans have a different mode in the frequency point-point difference than the primary MS1 scans, as shown in `r figure_count$label_text("frequency_mode_qcqa")`.
 
@@
 Specific data manipulation facilities were provided by dplyr v 1.0.8 [@wickhamDplyrGrammarData2022], tidyr v 1.2.0 [@wickhamTidyrTidyMessy2022], furrr v 0.2.3 [@vaughanFurrrApplyMapping2021a], 
-This manuscript was generated from rmarkdown v 2.11 [@allaireRmarkdownDynamicDocuments2021; @xieMarkdownDefinitiveGuide2018; xieMarkdownCookbook2020].
+This manuscript was generated from rmarkdown v 2.11 [@allaireRmarkdownDynamicDocuments2021; @xieMarkdownDefinitiveGuide2018; @xieMarkdownCookbook2020].
 :::

### 2022-04-12 doc/peakcharacterization_manuscript.Rmd
all of qcqa in, I think. Closes #19

@@
 figure_count$increment("coefficients_qcqa")
-figure_count$increment("frequency_mode_qcqa")
+figure_count$increment("resolution_qcqa")
 tar_load(coefficients_qcqa)
@@
 The conversion of m/z to frequency also provides opportunities to evaluate each scan and whether it should be kept for further processing.
-One way is to examine the coefficients of the square-root term for each scan (see term **y** in Equation `r equation_count$label_text("mzfrequency")`) and look for outliers there.
-Across the `r length(coefficients_qcqa$n_coef_out)` raw files that completed scan-centric processing, `r sum(coefficients_qcqa$n_coef_out > 0)` had at least one scan removed based on the square-root coefficient.
+Checks that have helped us in the development of the scan-centric peak characterization have included: (i) verifying that the ordering of data points remains identical in m/z and frequency space after conversion to frequency; and (ii) examining the R^2^ fit of the predicted frequency points to the original frequency points across scans for outliers.
+A third quality control step is to examine the coefficients of the square-root term for each scan (see term **y** in Equation `r equation_count$label_text("mzfrequency")`) and look for any outliers, and remove them before continuing the sample processing.
+Across the `r length(coefficients_qcqa$n_coef_out)` NSCLC raw files that completed scan-centric processing, `r sum(coefficients_qcqa$n_coef_out > 0)` had at least one scan removed based on the square-root coefficient.
 `r figure_count$label_text("coefficients_qcqa")`A shows a histogram of the number of scans removed based on being outliers of the overall distribution.
@@
 **A**: Number of scan outliers for each sample in the NSCLC dataset.
-**B**: Single example for sample 199Cpos of the scan square-root terms marked as outliers and removed from further consideration.
-:::
-
-One other check we implemented in the conversion to frequency is checking the range of frequency point-point differences allowed for creating the m/z to frequency model, and making sure that they are the same across scans.
+**B**: Single example for sample 199Cpos of the scan square-root terms, with outliers marked and removed from further consideration.
+:::
+
+::: {custom-style="MDPI_3.1_text"}
+A final check we implemented in the conversion to frequency was verifying that the range of frequency point-point differences allowed for creating the m/z to frequency model are the same across scans.
 The initial set of scans we allowed to use included **all** of the MS1 scans, including the precursor scans for the MS2 scans. 
-The samples were acquired as a mix of MS1 only scans for the first 7.5 minutes, and then a mix of MS1 precursor scans followed by MS2 scans for the second 7.5 minutes.
-In 20 samples, the MS1 precursor scans have a different mode in the frequency point-point difference than the primary MS1 scans, as shown in `r figure_count$label_text("frequency_mode_qcqa")`.
+However, the samples were acquired as a mix of MS1 only scans for the first 7.5 minutes, and then a mix of MS1 precursor scans followed by MS2 scans for the second 7.5 minutes.
+In 20 samples, the MS1 precursor scans have a different resolution than the MS1 main scans, as shown in `r figure_count$label_text("resolution_qcqa")`.
+This is actually easy to detect by differences in the square-root coefficients, but due to using a single frequency model for converting all scans actually **manifested** as different modes in the point-point frequency differences.
+Oddly enough, the check for outlier scans discussed previously, does not flag any of the scans as outliers in this case.
+Ultimately, we were able to resolve this anomaly by only using those scans with an injection time before 7.5 minutes.
+The package will be updated to include checks that the resolution used is the same across the scans, and produce an informative error for the end user before proceeding.
+:::
+
+::: {custom-style="MDPI_5.2_figure"}
+```{r resolution_qcqa}
+tar_load(resolution_qcqa)
+
+resolution_qcqa %>%
+  ggplot(aes(x = rtime, xend = rtime, y = 0, yend = tic, color = sqrt)) +
+  geom_segment() +
+  theme(legend.position = c(0.8, 0.8)) +
+  labs(x = "Injection Time (s)", y = "TIC", color = "Coefficient") +
+  coord_cartesian(ylim = c(NA, 2e7)) +
+  scale_y_continuous(expand = expansion(mult = c(0, .1))) +
+  scale_x_continuous(expand = expansion(mult = c(0, .1)))
+```
+:::
+
+::: {custom-style="MDPI_5.1_figure_caption"}
+`r figure_count$label_text("resolution_qcqa")`.
+Scan TIC plot noting the square-root coefficient for each scan, where the coefficient is reflective of the resolution used for each scan.
+The y-axis is cropped to 2x10^7^ so that the MS1 precursor scans are visible, as their TIC's are much smaller than the others.
+:::
+
+::: {custom-style="MDPI_3.1_text"}
+Finally, in the scan-scan normalization step, we require that there are at least 25 peaks in each of the scans that can be used to calculate normalization factors between the scans and the reference scan.
+If there are no scans with at least 25 peaks, then the processing of that sample will error and stop.
+:::
 

### 2022-04-12 R/qcqa_analysis.R

@@ -28,3 +28,78 @@ coefficients_analysis <- function(coefficients_data) {
        example = df_freq)
 
 }
+
+resolution_analysis = function(zip_sample){
+  # zip_sample = "data/data_output/149Cpos.zip"
+  zip_ms = ZipMS$new(zip_sample)
+  zip_ms$load_raw()
+  zip_ms$load_peak_finder()
+
+  raw_info = zip_ms$raw_ms$ms_info
+  raw_info$scan = as.character(raw_info$scan)
+
+  zip_ms$raw_ms$set_scans()
+
+  mz_df_list = zip_ms$raw_ms$extract_raw_data()
+  frequency_fit_description = zip_ms$peak_finder$peak_regions$frequency_fit_description
+  mz_fit_description = zip_ms$peak_finder$peak_regions$mz_fit_description
+  ... = NULL
+
+  if (is.null(names(mz_df_list))) {
+    names(mz_df_list) = purrr::map_chr(mz_df_list, ~ .x$scan[1])
+  }
+
+  mz_frequency = FTMS.peakCharacterization:::internal_map$map_function(mz_df_list, function(in_scan){
+    #message(scan_name)
+    out_scan = FTMS.peakCharacterization:::convert_mz_frequency(in_scan, ...)
+    out_scan
+  })
+  #str(mz_frequency[[1]])
+  #log_message("converted frequencies")
+  frequency_fits = FTMS.peakCharacterization:::internal_map$map_function(mz_frequency, function(in_freq){
+    use_peaks = in_freq$convertable
+    tmp_fit = FTMS.peakCharacterization:::fit_exponentials(in_freq$mean_mz[use_peaks], in_freq$mean_frequency[use_peaks], frequency_fit_description)
+    tmp_fit$scan = in_freq[1, "scan"]
+    tmp_fit
+  })
+  #str(frequency_fits[[1]])
+  #log_message("fit frequencies")
+
+  mz_fits = FTMS.peakCharacterization:::internal_map$map_function(mz_frequency, function(in_freq){
+    use_peaks = in_freq$convertable
+    tmp_fit = FTMS.peakCharacterization:::fit_exponentials(in_freq$mean_frequency[use_peaks], in_freq$mean_mz[use_peaks], mz_fit_description)
+    tmp_fit$scan = in_freq[1, "scan"]
+    tmp_fit
+  })
+  #str(mz_fits[[1]])
+  #log_message("fit mz")
+
[…]

### 2022-04-12 _targets.R

@@ -428,7 +428,12 @@ qcqa_tar = list(
   tar_target(coefficients_data,
              readRDS(coefficients_file)),
   tar_target(coefficients_qcqa,
-             coefficients_analysis(coefficients_data))
+             coefficients_analysis(coefficients_data)),
+  tar_target(bad_file,
+             "data/data_output/149Cpos.zip",
+             format = "file"),
+  tar_target(resolution_qcqa,
+             resolution_analysis(bad_file))
 )
 
 list(pkg_tar,

### 2022-04-12 doc/peakcharacterization_manuscript.Rmd
all of Hunters edits

@@
 
-::: {custom-style="MDPI_3.1_text"}
-For each region initially created, the peaks in each scan within that region can be characterized using a quadratic fit of log(intensity) to m/z.
-:::
-
 ::: {custom-style="MDPI_2.2_heading2"}
@@
 Although many other types of mass-spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo Fusion has a baseline of 0 due to manipulations in the Thermo firmware, making the determination of the centroided values considerably easier.
-For the peak characterization (centroiding), we use a simple quadratic model of log(intensity) to m/z. 
+For each region initially created, the peaks in each scan within that region can be characterized (i.e. centroiding).
+For centroiding, we use a simple weighted quadratic model of log(intensity) to m/z. 
 We add a small constant to enable using the zero values, and weight the values by their ratio to the most intense value, which is normally the value closest to the center of peak, helping to ensure that the **true** centroid is determined.
@@
 Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
-However, there are two issues with these peak intensities across scans: 1) some peak intensities correlate with the scan number (i.e.
-scan acquisition order); and 2) some peak differences between scans are correlated with intensity.
-The solution to **1** is to do a two pass normalization.
+However, there are two issues with these peak intensities across scans: (i) some peak intensities correlate with the scan number (i.e.
+scan acquisition order); and (ii) some peak differences between scans are correlated with intensity.
+The solution to **i** is to do a two pass normalization.
 After the first pass, the peaks whose intensity is correlated with scan order are detected.
@@
 `r figure_count$label_text("intensity_scan")`A contains an example peak whose intensity across scans is correlated with scan number.
-The solution to **2** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
+The solution to **ii** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
 The highlighted peaks in `r figure_count$label_text("intensity_scan")`B are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("normalization_factors")`.** **A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity \>= 0.7 of the maximum intensity (singlenorm_int), or the double pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
+**`r figure_count$label_text("normalization_factors")`.** **A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity \>= 0.7 of the maximum intensity (singlenorm_int), or the two pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
 **B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.
@@
 Therefore, one way to quantify any potential *improvements* in the characterized peaks is to look at the relative standard deviation (RSD) for the characterized scan level peak intensities (calculated as the standard deviation of peak heights across scans divided by the mean peak height) and compare them as each processing step is introduced.
+`r figure_count$label_text("rsd_method")` illustrates the peak height RSD distributions for four different samples.
+For samples 1ecf and 2ecf, there is a general shift in the RSD to the left, going from the bottom processing mehtods (msnbase only) to the top processing method (filtersd), representing a visually clear improvement.
+For samples 49lipid and 97lipid, the improvements are visually more subtle, having a slight shift to the left as well as a narrowing and smoothing of the RSD distribution.
+Part of this sublety is likely due to the bimodality of these two distributions.
+This is not surprising, since these two samples are non-polar extractions from tissue and are biochemically more complex.
+`r table_count$label_text("rsd_method")` provides more quantitative metrics.
+At two decimal places, the filtersd, doublenorm, and singlenorm processing methods give superior and nearly identical results for three of the four samples, especially in terms of mean and median RSD.
+However, for sample 1ecf, the max RSD is much higher for singlenorm, highlighting its instability.
 :::
@@
 `r figure_count$label_text("nap_intensity")` compares the peak-peak isotopic natural abundance probability and height log-ratio differences (Equation `r equation_count$label_text("napintensity")` in Methods) generated using heights from Xcalibur and from our scan-centric peak characterization.
+From this figure, it is clear that inconsistency in peak presence across scans leads to larger deviations between measured peak heights from aggregate spectra and expected relNAP.
 :::
@@
 This echoes the patterns of low overlapping assigned peaks observed in both ECF samples, where many of the scan-centric amino-acid peaks only found one match from the MSnbase peaks.
-Surprisingly, the corrected intensities generated very different p-values compared to the raw intensities, and XCalibur showed the most agreement with the raw p-values.
+Surprisingly, the truncated normal distribution corrected intensities generated very different p-values compared to the raw intensities, and XCalibur showed the most agreement with the raw p-values.
 We tested the statistical significance of these differences using a t-test of the actual difference in log-p-values (shown in `r figure_count$label_text("p_value_comparison")`D).
+It is clear that the raw scan-level intensities are superior.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-we should have some discussion here about the results
+FT-MS provides spectral scans; however, unlike nuclear magnetic resonance, these spectral scans are not identical, due to several instrumentation issues.
+Therefore, simple scan summation into an aggregate spectrum followed by peak characterization does not provide optimal data.
+Also, this approach prevents the calculation of useful peak-specific statistics, especially variance.
+In the results presented, we demonstrate superior performance by scan-level peak characterization in terms of: i) improved QC/QA and noise reduction, ii) improved peak height relative standard deviations, iii) improved peak intensity ratios that better match natural abundance probabilities, and iv) better separation between biological groups.
+These improvements are necessary for downstream derivation of molecular formula based on peak ratio matching to NAP in SMIRFE.
+However, scan-level peak characterization requires a sophisticated pipeline of QC/QA and processing steps to derive a superior characterization with informative statistics and QC/QA metrics.
+Years of methods development, testing, and optimization have gone into the methods presented here.
+But once implemented, large datasets can be processed in a straight-forward manner and in a manner that supports full computational reproducibility. 
+Not all of the methods presented here provided an improvement.
+We included our negative results with correcting mean and standard deviations for the intensity of peaks not detected in all scans, so that others do not waste effort pursuing this.
+We initially expected the data to follow a truncated normal distribution; however, upon further contemplation of the results presented here, we believe that the differences between scans still present after normalization imparts a dispersive variance component.
+Thus, we hypothesize that a truncated negative binomial distribution better represents the intensity of peaks not detected across all scans.
+Fortunately, this distribution has been previously studied [@sampfordTruncatedNegativeBinomial1955a] and we plan to explore this possibility for a future improvement to our methods. 
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-Two different sets of samples were used to evaluate the various methods, ethylchloroformate (ECF) amino acid samples and non-small-cell-lung-cancer (NSCLC) lipid samples.
+Two different sets of samples were used to evaluate the various methods, ethylchloroformate (ECF) amino acid samples and non-small-cell-lung-cancer (NSCLC) lipid-extracted tissue samples.
 
@@
 The collection, preparation, and mass spectrometry analysis of the paired cancer and non-cancer samples has been previously described [@sellersPyruvateCarboxylaseCritical2015]. 
-In summary, cancer and non-cancer samples were acquired from eight-six non-diabetic patients with suspected resectable stage I or IIA non-small cell lung cancer. 
+In summary, cancer and nearby non-cancer tissue samples were acquired from eight-six non-diabetic patients with suspected resectable stage I or IIA non-small cell lung cancer. 
 Written informed consent was collected from all subjects prior to inclusion and all samples were collected under a University of Louisville or University of Kentucky IRB protocol. 
@@
 From the known physical properties of the Orbitrap, only the square root term should be necessary [@ledfordSpaceChargeEffects1984].
-Practically, we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry, contributions from space charge effects and magnetronic motion, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
+Practically, we have found the combination of square and cube-roots to provide a better fit when processing data from the Thermo-Fisher Fusion instrument, likely due to issues with slight imperfections in the orbitrap geometry, contributions from space charge effects and magnetronic motion, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
+We do note here that when working with Bruker SolariX ICR data, the equation is slightly different, and does not require the cubic square root term (unpublished results, see Supplemental Materials).
 A frequency model is generated for each scan, and then a single model using the scan with the square-root term closest to the median of the square-root terms from all scans.

### 2022-04-12 doc/peakcharacterization_supplemental.Rmd
all of Hunters edits

---
title: "Supplemental Materials for: Scan-Centric, Frequency-Based Method for Characterizing Peaks from Direct Injection Fourier transform Mass Spectrometry Experiments"
author:
  - Robert M Flight:
      email: robert.flight@uky.edu
      institute: [markey, biochem, rcsirm]
  - Joshua M Mitchell:
      email: jmmi243@uky.edu
      institute: [markey, biochem, rcsirm, ibi]
  - Hunter NB Moseley:
      email: hunter.moseley@uky.edu
      correspondence: true
      institute: [markey, biochem, rcsirm, ibi, tox]
institute:
  - markey: Markey Cancer Center, University of Kentucky, Lexington, KY 40536, USA
  - biochem: Department of Molecular & Cellular Biochemistry, University of Kentucky, Lexington, KY 40536, USA
  - rcsirm: Resource Center for Stable Isotope Resolved Metabolomics, University of Kentucky, Lexington, KY 40536, USA
  - ibi: Institute for Biomedical Informatics, University of Kentucky, Lexington, KY 40536, USA
  - tox: Department of Toxicology and Cancer Biology, University of Kentucky, Lexington, KY 40536, USA
date: "`r Sys.time()`"
output: 
  word_document:
    keep_md: true
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: '`r here::here("doc/peakcharacterization.json")`'
csl: plos-computational-biology.csl
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

## Bruker SolariX ICR Frequency Conversion

In contrast to the m/z to frequency conversion equations used for Thermo-Fisher Fusion and other Orbitrap instruments, the one we have used for Bruker SolariX ICR instruments is simpler:

$$frequency = a + x * \frac{1}{mz} + y * \frac{1}{\sqrt{mz}}$$

### 2022-04-12 _targets.R

@@ -126,6 +126,8 @@ figures_tar = tar_plan(
              format = "file"),
   tar_render(manuscript_both,
              "doc/peakcharacterization_manuscript.Rmd"),
+  tar_render(supplemental_materials,
+             "doc/peakcharacterization_supplemental.Rmd"),
   tar_target(manuscript_nostyle,
              strip_mdpi_render(base_manuscript)),
   tar_target(manuscript_mdpi,

### 2022-04-12 doc/peakcharacterization_manuscript.Rmd
closes #23

@@
 ```{r nap_intensity, dn_id = figure_count, fig.width = 16, fig.height = 8}
-nap_fontsize = 12
+nap_fontsize = 10
 tar_load(aa_diffs_filtersd_1ecf)

### 2022-04-12 doc/peakcharacterization_manuscript.Rmd
making scan-centric where appropriate
closes #24

@@
 ::: {custom-style="MDPI_3.1_text"}
-Although many other types of mass-spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo Fusion has a baseline of 0 due to manipulations in the Thermo firmware, making the determination of the centroided values considerably easier.
+Although many other types of mass-spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo-Fisher Fusion has a baseline of 0 due to manipulations in the Thermo-Fisher firmware, making the determination of the centroided values considerably easier.
 For each region initially created, the peaks in each scan within that region can be characterized (i.e. centroiding).
@@
 After removing the high FSD peaks, the number of peaks left in the HPD site are the same or less than those from MSnbase, and at least in this example, look more likely to be real peaks compared to those from MSnbase.
-Therefore, scan-level characterization allows us to keep what are likely **real** peaks in HPD sites, without consideration of peak intensity, and mark peaks that may be artifactual from HPD regions.
+Therefore, scan-centric characterization allows us to keep what are likely **real** peaks in HPD sites, without consideration of peak intensity, and mark peaks that may be artifactual from HPD regions.
 :::
@@
 **A**: *xcalibur* peaks exported from Thermo Xcalibur after averaging scans.
-**B**: *scalevel_00*: Scan-level peak characterization without any density based filtering (see noperc_nonorm in Methods).
-**C**: *scanlevel_99*: Scan-level peak characterization using the default point density filtering (see filtersd in Methods).
-**D**: *scanlevel_99_lowsd*: Same as *scanlevel_99*, but removing any peaks marked as having a high frequency standard deviation.
+**B**: *scalevel_00*: Scan-centric peak characterization without any density based filtering (see noperc_nonorm in Methods).
+**C**: *scancentric_99*: Scan-centric peak characterization using the default point density filtering (see filtersd in Methods).
+**D**: *scancentric_99_lowsd*: Same as *scanlevel_99*, but removing any peaks marked as having a high frequency standard deviation.
 **E**: *msnbase*: peak centroids generated from *MSnbase*.
@@
 threonine_ratios = aa_diffs_filtersd_1ecf$specific_diffs$ratios %>%
-  dplyr::filter(source %in% c("scan-level", "xcalibur"))
+  dplyr::filter(source %in% c("scan-centric", "xcalibur"))
 threonine_diffs = aa_diffs_filtersd_1ecf$specific_diffs$diffs %>%
@@
     labs(x = "Number of Missing Scans",
-         y = "Xcalibur - Scan-Level NAP - Height Differences")+
+         y = "Xcalibur - Scan-Centric NAP - Height Differences")+
   theme(axis.title.y = element_text(size = nap_fontsize))
@@
   labs(x = "Number of Missing Scans",
-         y = "Xcalibur - Scan-Level NAP - Height Differences") +
+         y = "Xcalibur - Scan-Centric NAP - Height Differences") +
   theme(axis.title.y = element_text(size = nap_fontsize))
@@
 **A**.
-The peak-peak NAP - intensity log differences from scan-level peak heights (red) and Xcalibur peak heights (blue) from the threonine amino acid assignments with Na adduct, with point size reflecting how many peaks were missing across scans.
-**B**. The difference of Xcalibur to scan-level ratios plotted directly as a function of the number of scans the peak was not found in.
-**C**. The differences of Xcalibur to scan-level ratios for all of the amino acid assignments in EMFs with more than a single peak in both ECF samples.
+The peak-peak NAP - intensity log differences from scan-centric peak heights (red) and Xcalibur peak heights (blue) from the threonine amino acid assignments with Na adduct, with point size reflecting how many peaks were missing across scans.
+**B**. The difference of Xcalibur to scan-centric ratios plotted directly as a function of the number of scans the peak was not found in.
+**C**. The differences of Xcalibur to scan-centric ratios for all of the amino acid assignments in EMFs with more than a single peak in both ECF samples.
 :::
@@
 We tested the statistical significance of these differences using a t-test of the actual difference in log-p-values (shown in `r figure_count$label_text("p_value_comparison")`D).
-It is clear that the raw scan-level intensities are superior.
+It is clear that the raw scan-centric intensities are superior.
 :::
@@
 Also, this approach prevents the calculation of useful peak-specific statistics, especially variance.
-In the results presented, we demonstrate superior performance by scan-level peak characterization in terms of: i) improved QC/QA and noise reduction, ii) improved peak height relative standard deviations, iii) improved peak intensity ratios that better match natural abundance probabilities, and iv) better separation between biological groups.
+In the results presented, we demonstrate superior performance by scan-centric peak characterization in terms of: i) improved QC/QA and noise reduction, ii) improved peak height relative standard deviations, iii) improved peak intensity ratios that better match natural abundance probabilities, and iv) better separation between biological groups.
 These improvements are necessary for downstream derivation of molecular formula based on peak ratio matching to NAP in SMIRFE.
-However, scan-level peak characterization requires a sophisticated pipeline of QC/QA and processing steps to derive a superior characterization with informative statistics and QC/QA metrics.
+However, scan-centric peak characterization requires a sophisticated pipeline of QC/QA and processing steps to derive a superior characterization with informative statistics and QC/QA metrics.
 Years of methods development, testing, and optimization have gone into the methods presented here.
@@
 Only the MS1 scans were used, and the scan-scan time difference had to be >= 4 seconds, and for the lipid samples only scans acquired before 450 seconds (7.5 minutes) were kept.
-The scan-level data was then processed in the following eight ways:
+The scan-centric data was then processed in the following eight ways:
 :::
@@
 
-We attempted to create peaks that could be assigned by SMIRFE from the MSnbase and Xcalibur peak lists, but SMIRFE would not assign them given its requirements for scan-level peak characterization data.
+We attempted to create peaks that could be assigned by SMIRFE from the MSnbase and Xcalibur peak lists, but SMIRFE would not assign them given its requirements for scan-centric peak characterization data.
 :::

### 2022-04-12 R/examine_hpd.R

@@ -60,27 +60,27 @@ density_calculate = function(frequency_values, metadata, frequency_range = NULL)
 }
 
 
-hpds_from_excel = function(scanlevel_99, scanlevel_00, msnbase, excel_files){
-  sample = scanlevel_99$char_obj$zip_ms$id
+hpds_from_excel = function(scancentric_99, scancentric_00, msnbase, excel_files){
+  sample = scancentric_99$char_obj$zip_ms$id
   match_excel = grep(sample, excel_files, value = TRUE)
 
   xl_data = readxl::read_excel(match_excel, skip = 8, col_names = FALSE)
   xl_data = xl_data[, 1:2]
   names(xl_data) = c("mz", "intensity")
   message("got xl data")
-  frequency_coefficients = scanlevel_99$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$metadata$frequency_coefficients
-  frequency_description = scanlevel_99$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$metadata$frequency_fit_description
-  mz_coefficients = scanlevel_99$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$metadata$mz_coefficients
-  mz_description = scanlevel_99$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$metadata$mz_fit_description
-  org_frequency_peaks = purrr::map_df(seq_len(length(scanlevel_99$char_obj$zip_ms$peak_finder$peak_regions$peak_region_list)),
+  frequency_coefficients = scancentric_99$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$metadata$frequency_coefficients
+  frequency_description = scancentric_99$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$metadata$frequency_fit_description
+  mz_coefficients = scancentric_99$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$metadata$mz_coefficients
+  mz_description = scancentric_99$char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$metadata$mz_fit_description
+  org_frequency_peaks = purrr::map_df(seq_len(length(scancentric_99$char_obj$zip_ms$peak_finder$peak_regions$peak_region_list)),
                                       function(.x){
[…]

### 2022-04-12 R/get_assignments.R

@@ -469,7 +469,7 @@ ratio_diffs_extracted = function(difference_data, aa_id = "Threonine", formula_i
   raw_ratios = diff_formula$data_ratios %>%
     dplyr::mutate(source2 = source,
                   source = dplyr::case_when(
-      grepl("char_height", source2) ~ "scan-level",
+      grepl("char_height", source2) ~ "scan-centric",
       TRUE ~ source2
                   ))
 

### 2022-04-12 doc/peakcharacterization_manuscript.Rmd
more edits

@@
 The subsequent differences in frequency should be relatively constant with respect to m/z, in contrast to the differences of adjacent m/z points, as shown in `r figure_count$label_text("mz_frequency_conversion")`B and C.
-The Thermo Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 450K or 500K depending on the sample, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C and D.
+The Thermo-Fisher Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 450K or 500K depending on the sample, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C and D.
 Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`E.
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("hpd_compare_fsd")`.** Comparison of HPD and high FSD sites in the ECF 2 sample.
+`r figure_count$label_text("hpd_compare_fsd")`.
+Comparison of HPD and high FSD sites in the 2ecf sample.
 **A** - **E**:Peak plots for various peak processing modalities in a single HPD site.
-**A**: *xcalibur* peaks exported from Thermo Xcalibur after averaging scans.
-**B**: *scalevel_00*: Scan-centric peak characterization without any density based filtering (see noperc_nonorm in Methods).
+**A**: *xcalibur* peaks exported from Thermo-Fisher Xcalibur after averaging scans.
+**B**: *scancentric_00*: Scan-centric peak characterization without any density based filtering (see noperc_nonorm in Methods).
 **C**: *scancentric_99*: Scan-centric peak characterization using the default point density filtering (see filtersd in Methods).
-**D**: *scancentric_99_lowsd*: Same as *scanlevel_99*, but removing any peaks marked as having a high frequency standard deviation.
+**D**: *scancentric_99_lowsd*: Same as *scancentric_99*, but removing any peaks marked as having a high frequency standard deviation.
 **E**: *msnbase*: peak centroids generated from *MSnbase*.
-**F**: Scatterplot of peak counts across all of the HPDs detected in the ECF 2 sample against the Xcalibur peak counts.
+**F**: Scatterplot of peak counts across all of the HPDs detected in the 2ecf sample against the Xcalibur peak counts.
 :::
@@
 ```{r rsd_table, results = 'asis'}
-tar_load(rsd_values)
-rsd_values = rsd_values %>%
-  dplyr::arrange(sample, processed)
-rsd_ft = flextable(rsd_values) %>%
-  colformat_double(digits = 2) %>%
+tar_load(rsd_best)
+rsd_best %>%
   autofit()
-rsd_ft
 ```
@@
 ```{r nap_intensity, dn_id = figure_count, fig.width = 16, fig.height = 8}
-nap_fontsize = 10
+nap_fontsize = 9
 tar_load(aa_diffs_filtersd_1ecf)
@@
 ::: {custom-style="MDPI_3.1_text"}
-FT-MS provides spectral scans; however, unlike nuclear magnetic resonance, these spectral scans are not identical, due to several instrumentation issues.
-Therefore, simple scan summation into an aggregate spectrum followed by peak characterization does not provide optimal data.
+[**Is it just me, or does this feel like the conclusions? If so, do we need a Discussion? I think for MDPI they are optional ...**]
+FT-MS direct-injection data provides spectral scans; however, unlike nuclear magnetic resonance, these spectral scans are not identical, due to several instrumentation issues.
+Therefore, simple scan summation to an aggregate spectrum followed by peak characterization does not provide optimal data.
 Also, this approach prevents the calculation of useful peak-specific statistics, especially variance.
-In the results presented, we demonstrate superior performance by scan-centric peak characterization in terms of: i) improved QC/QA and noise reduction, ii) improved peak height relative standard deviations, iii) improved peak intensity ratios that better match natural abundance probabilities, and iv) better separation between biological groups.
+In the results presented, we demonstrate superior performance by scan-centric peak characterization in terms of: i) improved QC/QA and noise reduction; ii) improved peak height relative standard deviations; iii) improved peak intensity ratios that better match natural abundance probabilities; and iv) better separation between biological groups.
 These improvements are necessary for downstream derivation of molecular formula based on peak ratio matching to NAP in SMIRFE.
@@
 Years of methods development, testing, and optimization have gone into the methods presented here.
-But once implemented, large datasets can be processed in a straight-forward manner and in a manner that supports full computational reproducibility. 
+But once implemented, large datasets can be processed in a straight-forward manner that supports full computational reproducibility. 
 Not all of the methods presented here provided an improvement.
@@
 Written informed consent was collected from all subjects prior to inclusion and all samples were collected under a University of Louisville or University of Kentucky IRB protocol. 
-Lipid extracts were prepared using a modified Folch extraction and reconstituted for direct infusion ultra-high resolution mass spectrometry on a pair of Thermo Tribrid Fusion Orbitrap instruments (FSN10115 and FSN10352, referred to as Fusion 1 and Fusion 2 respectively) coupled to an Advion nanoelectrospray system.
+Lipid extracts were prepared using a modified Folch extraction and reconstituted for direct infusion ultra-high resolution mass spectrometry on a pair of Thermo-Fisher Tribrid Fusion Orbitrap instruments (FSN10115 and FSN10352, referred to as Fusion 1 and Fusion 2 respectively) coupled to an Advion nanoelectrospray system.
 53 patients (102 spectra total) were acquired from Fusion 1 (serial # FSN10115) and 40 patients (77 spectra total) were acquired from Fusion 2 (serial # FSN10352).
@@
 Practically, we have found the combination of square and cube-roots to provide a better fit when processing data from the Thermo-Fisher Fusion instrument, likely due to issues with slight imperfections in the orbitrap geometry, contributions from space charge effects and magnetronic motion, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
-We do note here that when working with Bruker SolariX ICR data, the equation is slightly different, and does not require the cubic square root term (unpublished results, see Supplemental Materials).
+We do note that when working with Bruker SolariX ICR data, the equation is slightly different, and does not require the cubic square root term (unpublished results, see Supplemental Materials).
 A frequency model is generated for each scan, and then a single model using the scan with the square-root term closest to the median of the square-root terms from all scans.
@@
 
-To convert m/z back into frequency, we can use a similar model without the roots, as well as an extra simple linear term that doesn't have an equivalent in the frequency model.
+To convert m/z back into frequency, we can use a similar model without the roots, as well as an extra simple linear term that doesn't have an equivalent in the above frequency model.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-To enable interval algebra, the frequency points are converted to single width intervals by multiplying (by a multiplier of 400 in this work), rounding to the nearest integer, and storing them as IRanges objects from the IRanges Bioconductor package [@lawrenceSoftwareComputingAnnotating2013a].
+To enable interval algebra, the frequency points are converted to single width intervals by multiplying by a constant factor to maintain the differences in individual points (a multiplier of 400 in this work), rounding to the nearest integer, and storing them as IRanges objects from the IRanges Bioconductor package [@lawrenceSoftwareComputingAnnotating2013a].
 The sliding and tiled windows are also converted to IRanges objects using this process.
@@
 ::: {custom-style="MDPI_3.1_text"}
-The code used in the reporting of results is available on GitHub at https://github.com/MoseleyBioinformaticsLab/manuscript.peakCharacterization, and archived on Zenodo [ref].
-The scan-centric peak-characterization is packaged as an R package at https://github.com/MoseleyBioinformaticsLab/FTMS.peakCharacterization, and the specific version used in this work is archived on Zenodo [ref].
+The code used in the reporting of results is available on GitHub at https://github.com/MoseleyBioinformaticsLab/manuscript.peakCharacterization, and archived on Zenodo [@flightMoseleyBioinformaticsLabManuscriptPeakCharacterization2022].
+The scan-centric peak-characterization is packaged as an R package at https://github.com/MoseleyBioinformaticsLab/FTMS.peakCharacterization, and the specific version used in this work is archived on Zenodo [@flightMoseleyBioinformaticsLabFTMSPeakCharacterization2022].
 :::
@@
 
+::: {custom-style="MDPI_3.1_text"}
+
+:::
+
 ::: {custom-style="MDPI_2.1_heading1"}

### 2022-04-12 R/best_rsd.R
closes #21

@@ -1,0 +1,30 @@
+find_best_rsd = function(rsd_values){
+  other_frame = readr::read_delim(
+    'processed   | sample
+     singlenorm  | 2ecf
+     singlenorm  | 97lipid',
+    delim = '|', trim_ws = TRUE)
+
+  order_frame = tidyr::expand_grid(
+    sample = c("1ecf", "2ecf", "49lipid", "97lipid"),
+    processed = c("filtersd",
+                   "doublenorm",
+                   "singlenorm_int",
+                   "singlenorm",
+                   "perc99_nonorm",
+                   "noperc_nonorm",
+                   "msnbase_only")
+  )
+
+  rsd_values2 = dplyr::left_join(order_frame, rsd_values, by = c("sample", "processed"))
+  filtersd_rsd = which(rsd_values2$processed %in% "filtersd")
+  other_loc = c(11, 25)
+
+  rsd_ft = rsd_values2 %>%
+    flextable() %>%
+    hline(i = filtersd_rsd[2:(length(filtersd_rsd))] - 1) %>%
+    colformat_double(digits = 2) %>%
+    bold(i = c(other_loc, filtersd_rsd), j = seq(3, 6), part = "body")
+  rsd_ft
+
+}

### 2022-04-12 _targets.R

@@ -135,7 +135,8 @@ figures_tar = tar_plan(
 )
 
 tables_tar = tar_plan(
-  tar_target(rsd_values, summarize_rsd(rsd_combine))
+  tar_target(rsd_values, summarize_rsd(rsd_combine)),
+  tar_target(rsd_best, find_best_rsd(rsd_values))
 )
 
 # other_tar = tar_plan(

### 2022-04-13 doc/peakcharacterization_manuscript.Rmd
supplemental stuff and referring to it
closes #22 and #15

@@
 
+::: {custom-style="MDPI_3.1_text"}
+Although we expected that the corrected scan-centric intensities would behave similarly or better compared to the raw intensities, Figure S1 shows that while they have smaller differences than XCalibur, there is still a trend of increasing difference with the raw scan-centric intensity - NAP ratios as the number of missing scans increases.
+:::
+
 ::: {custom-style="MDPI_2.2_heading2"}
@@
 Green points denote the high, mean, and low-confidence limits reported from the t-test.
+:::
+
+::: {custom-style="MDPI_3.1_text"}
+It is vitally important however, to normalize the intensities correctly.
+Figure S3 of the Supplemental Materials shows that with the incorrecte normalization, the corrected and raw p-values become much closer to each other.
 :::
@@
 Not knowing which sample run was most appropriate to use, we removed all of the duplicated samples.
-Using information-content-informed Kendall-tau correlation [@flightInformationContentInformedKendalltauCorrelation2022], we compared the median correlations of each sample to all others in the same disease class for outliers, and removed `r n_out_vector["cancer"]` cancer and `r n_out_vector["non-cancer"]` samples.
+Using information-content-informed Kendall-tau correlation (ICI-Kt) [@flightInformationContentInformedKendalltauCorrelation2022], we compared the median correlations of each sample to all others in the same disease class for outliers, and removed `r n_out_vector["cancer"]` cancer and `r n_out_vector["non-cancer"]` samples.
+The median ICI-Kt correlations are shown in Figure S2 of the Supplemental Materials.
 
@@
 Outlier lipid samples were determined using visualizationQualityControl v 0.4.7 [@flightVisualizationQualityControlDevelopmentVisualization2021].
-Specific data manipulation facilities were provided by dplyr v 1.0.8 [@wickhamDplyrGrammarData2022], tidyr v 1.2.0 [@wickhamTidyrTidyMessy2022], furrr v 0.2.3 [@vaughanFurrrApplyMapping2021a], 
+Specific data manipulation facilities were provided by dplyr v 1.0.8 [@wickhamDplyrGrammarData2022], tidyr v 1.2.0 [@wickhamTidyrTidyMessy2022], furrr v 0.2.3 [@vaughanFurrrApplyMapping2021a].
 This manuscript was generated from rmarkdown v 2.11 [@allaireRmarkdownDynamicDocuments2021; @xieMarkdownDefinitiveGuide2018; @xieMarkdownCookbook2020].
@@
 The code used in the reporting of results is available on GitHub at https://github.com/MoseleyBioinformaticsLab/manuscript.peakCharacterization, and archived on Zenodo [@flightMoseleyBioinformaticsLabManuscriptPeakCharacterization2022].
-The scan-centric peak-characterization is packaged as an R package at https://github.com/MoseleyBioinformaticsLab/FTMS.peakCharacterization, and the specific version used in this work is archived on Zenodo [@flightMoseleyBioinformaticsLabFTMSPeakCharacterization2022].
-:::
-
-::: {custom-style="MDPI_2.1_heading1"}
-## Conclusions
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-
+The scan-centric peak-characterization is available as an R package at https://github.com/MoseleyBioinformaticsLab/FTMS.peakCharacterization, and the specific version used in this work is archived on Zenodo [@flightMoseleyBioinformaticsLabFTMSPeakCharacterization2022].
 :::

### 2022-04-13 doc/peakcharacterization_supplemental.Rmd
supplemental stuff and referring to it
closes #22 and #15

@@
 
+```{r get_setup, include = FALSE}
+source(here::here("packages.R"))
+lapply(list.files(here::here("./R"), full.names = TRUE), source)
+
+figure_count = dn_counter$new("Figure ", prefix2 = "S")
+table_count = dn_counter$new("Table ", prefix2 = "S")
+equation_count = dn_counter$new("")
+```
+
+```{r setup, include = FALSE}
+knitr::opts_chunk$set(echo = FALSE, 
+                      warning = FALSE, 
+                      message = FALSE, 
+                      fig.width = 8, 
+                      fig.height = 6, 
+                      fig.process = dn_modify_path,
+                      dpi = 600,
+                      dev.args = list(png = list(type = "cairo")))
+
+format_equation = function(equation_text, label){
+  number_out = equation_count$label_text(label)
+  paste0('$$\n\\begin{align}\n', equation_text, '&& \\text{(', number_out, ')}', '\n\\end{align}$$')
+}
+```
+
 ## Bruker SolariX ICR Frequency Conversion
@@
 $$frequency = a + x * \frac{1}{mz} + y * \frac{1}{\sqrt{mz}}$$
+
+## Height - NAP Differences of Corrected vs Raw Scan-Centric
+
+```{r increment_nap_intensity}
+figure_count$increment("nap_intensity")
+```
+
+Here in `r figure_count$label_text("nap_intensity")` we show the differences in Height - NAP peaks using the corrected and raw scan-centric intensities so they can be compared with the differences we observed between XCalibur and scan-centric peaks.
+
+```{r nap_intensity, dn_id = figure_count, fig.width = 16, fig.height = 8}
+nap_fontsize = 12
+tar_load(aa_diffs_filtersd_1ecf)
+tar_load(aa_diffs_filtersd_2ecf)
+
+all_diffs = rbind(aa_diffs_filtersd_1ecf$all_diffs %>%
+                    dplyr::mutate(sample = "1ecf"),
+                  aa_diffs_filtersd_2ecf$all_diffs %>%
+                    dplyr::mutate(sample = "2ecf"))
+
+aa_diffplot = all_diffs %>%
+  dplyr::filter(source %in% "char_raw_corrected") %>%
+  dplyr::mutate(diff = -1 * diff) %>%
+  ggplot(aes(x = n_missing, y = diff)) +
+  geom_point() +
+  labs(x = "Number of Missing Scans",
+         y = "Corrected - Raw Scan-Centric NAP - Height Differences") +
+  theme(axis.title.y = element_text(size = nap_fontsize))
+
+aa_diffplot
+```
+
+`r figure_count$label_text("nap_intensity")`.
+The peak-peak NAP - intensity log differences between corrected and raw scan-centric peak heights as a function of the number of scans the peaks were not found in across all of the amino acid assignments in EMFs with more than a single peak in both ECF samples.
+
+## ICI-Kt Median Correlation Outliers
+
+```{r increment_icikt}
+figure_count$increment("icikt_outlier")
+```
+
+For the NSCLC dataset, we used information-content-informed Kendall-tau (ICI-Kt) correlation within each of non-cancer and cancer sample groups to determine outliers that should be removed prior to statistical testing.
+In `r figure_count$label_text("icikt_outlier")` we show the distribution of median ICI-Kt values and the samples determined as outliers.
+
+```{r icikt_outlier, dn_id = figure_count}
+tar_load(qcqa)
+outliers = qcqa$outlier_data
+
+outliers %>%
+  ggplot(aes(x = sample_class, y = med_cor, color = outlier, group = sample_class)) +
+  geom_sina() +
+  theme(legend.position = c(0.1, 0.9)) +
+  labs(x = "Sample Class", y = "Median ICI-Kt")
+```
+
+`r figure_count$label_text("icikt_outlier")`.
+Sina plot of median ICI-Kt correlation for samples within each of the cancer and non-cancer sample groups with outlier status indicated for each sample.
+
+## Incorrect Normalization in P-Value Differences
+
+```{r increment_p_value}
+figure_count$increment("p_value")
+```
+
+During the analysis of the p-values between the two classes of NSCLC samples, we didn't originally have the median intensities for the corrected peaks.
+Therefore, we went ahead and normalized the corrected intensities in each sample with the raw median intensities in each sample, and compared the p-values from each method.
+`r figure_count$label_text("p_value")` shows how this incorrect median normalization affects the difference in p-values between corrected and raw intensity.
+
+```{r p_value, dn_id = figure_count}
+lung_compare = tar_read(lung_compare_wrong)
+lung_compare = lung_compare %>%
+  dplyr::mutate(compare_id = dplyr::case_when(
+    id.test %in% "corrected" ~ "Corrected",
+    id.test %in% "xcalibur" ~ "XCalibur",
+    id.test %in% "msnbase" ~ "MSnbase"
+  ))
+
+lung_compare_corrected = lung_compare %>%
+  dplyr::filter(compare_id %in% "Corrected")
+
+lung_compare_ttest = tar_read(lung_wrong_ttest)
+
+long_ttest = lung_compare_ttest %>%
+  dplyr::select(p.adjust, comparison, estimate, conf.low, conf.high) %>%
+  tidyr::pivot_longer(cols = !comparison, names_to = "measure", values_to = "value") %>%
+  dplyr::mutate(compare_id = dplyr::case_when(
+    comparison %in% "corrected" ~ "Corrected",
+    comparison %in% "xcalibur" ~ "XCalibur",
+    comparison %in% "msnbase" ~ "MSnbase"
+  )) %>%
+  dplyr::filter(compare_id %in% "Corrected")
+
+long_pvalue = long_ttest %>%
+  dplyr::filter(measure %in% "p.adjust") %>%
+  dplyr::filter(compare_id %in% "Corrected")
+
+long_measure = long_ttest %>%
+  dplyr::filter(!(measure %in% "p.adjust")) %>%
+  dplyr::filter(compare_id %in% "Corrected")
+
+compare_sina = lung_compare_corrected %>%
+  ggplot(aes(x = compare_id, y = p_diff)) +
+  geom_hline(yintercept = 0, color = "red") +
+  geom_sina(alpha = 0.5) +
+  geom_point(data = long_measure, aes(x = compare_id, y = value), color = "green") +
+  geom_text(data = long_pvalue, aes(x = compare_id, y = -1, label = format(value, digits = 2))) +
+  labs(x = "", y = "Log-P-Value Difference")
+
+compare_pair = lung_compare_corrected %>%
+  ggplot(aes(x = log_p.ref, y = log_p.test)) +
+  geom_abline(slope = 1, color = "red") +
+  geom_point(alpha = 0.5) +
+  labs(x = "Log-P-Value Raw",
+         y = "Log-P-Value Corrected")
+compare_pair | compare_sina + plot_annotation(tag_levels = "A")
+```
+
+`r figure_count$label_text("p_value")`.
+**A**: Log-p-values generated by comparing non-cancer and cancer sample IMFs using peak intensities from raw and corrected peak intensities, where the corrected intensities were normalized using the sample raw median intensities.
+Red line denotes perfect agreement.
+**D**: Sina plot of differences in the log-p-values generated by corrected intensity normalized by the raw median intensities.
+Also shown are the Bonferroni adjusted p-values from a t-test of the log-p-value differences for each method.
+Green points denote the high, mean, and low-confidence limits reported from the t-test.

### 2022-04-13 _targets.R

@@ -400,6 +400,10 @@ lungcancer_tar = list(
                     xcalibur = lung_xcalibur),
                reference = "raw")
              ),
+  tar_target(lung_wrong_ttest,
+             ttest_compare_diffs(
+               lung_compare_wrong
+             )),
   tar_target(lung_binomial,
              binomial_compare(lung_compare)),
   tar_target(lung_corrected,

### 2022-04-13 doc/peakcharacterization_manuscript.Rmd
@hunter-moseley edits, and changes to the difference histograms
closes # 25

@@
 Fourier transform mass spectrometry (FT-MS) provides high performance in terms of sensitivity, resolution and mass accuracy all in one analytical instrumentation.
-The combination of these capabilities provide several analytical and interpretive improvements including: (i) the ability to resolve distinct isotopologues with identical unit masses but different accurate masses [@higashiStableIsotopeLabeledTracers2014]; (ii)  enable multi-element isotopic natural abundance correction for at least the lower portion of the detected mass range  [@moseleyCorrectingEffectsNatural2010; @carreerComputationalFrameworkHighThroughput2013; @wangAccuCor2IsotopeNatural2021]; (iii) improved assignment accuracy [@kindMetabolomicDatabaseAnnotations2006; @mitchellSmallMoleculeIsotope2019]; and (iv) the detection of metabolites in the sub-femtomolar range, when combined with chromatographic separation  [@eylesMethodsStudyProtein2004; @dettmerMassSpectrometrybasedMetabolomics2007].
+The combination of these capabilities provide several analytical and interpretive improvements: (i) the ability to resolve distinct isotopologues with identical unit masses but different accurate masses [@higashiStableIsotopeLabeledTracers2014]; (ii)  enabling multi-element isotopic natural abundance correction for at least the lower portion of the detected mass range  [@moseleyCorrectingEffectsNatural2010; @carreerComputationalFrameworkHighThroughput2013; @wangAccuCor2IsotopeNatural2021]; (iii) improved assignment accuracy [@kindMetabolomicDatabaseAnnotations2006; @mitchellSmallMoleculeIsotope2019]; and (iv) the detection of metabolites in the sub-femtomolar range, when combined with chromatographic separation  [@eylesMethodsStudyProtein2004; @dettmerMassSpectrometrybasedMetabolomics2007].
 In the metabolomics field, these improvements permit more complicated, but more informative experimental designs such as the use of single and multiple isotope-labeled precursors in stable isotope-resolved metabolomics (SIRM) experiments [@yangChloroformateDerivatizationTracing2017].
@@
 In addition, due to the tendency of increasing point spacing with increasing m/z in Orbitrap type instruments, we derive a method for transforming the m/z data to axial frequency, which has the desirable property of being equally spaced across the full spectrum.
-Finally, we show that in addition to better preserving the expected intensity - NAP relationships, our scan-centric method also results in improved relative standard deviations, automatic identification of high-peak-density artifact peaks, and separation of samples in a lipidomics data analysis.
+Finally, we show that in addition to better preserving the expected intensity to- NAP relationships, our scan-centric method also results in improved relative standard deviations, automatic identification of high-peak-density artifact peaks, and better separation of samples in a lipidomics data analysis.
 :::
@@
 Whereas the number of peaks from MSnbase and Xcalibur are three-fold and 40-fold higher in the 1ecf sample compared to the 97lipid sample.
+
+`r figure_count$label_text("assigned_mz_diff")` shows the distribution of differences between the observed and expected m/z of assigned peaks in the 1ecf and 97lipid spectra.
+For the 1ecf-specific histograms, it is clear that these differences have a narrower unimodal distribution from the scan-centric peak characterization, especially in comparison to MSnbase.
+Also, the MSnbase distribution has far fewer peaks that match a scan-level assigned peak.
+Given the large number of MSnbase-characterized peaks present in the spectrum, many peaks may be outside the matching tolerance of 2pmm. 
+This strongly implies that the peak center error is far higher in the MSnbase-characterized peaks than what the histogram directly shows.
+For the 97lipid-specific histograms, again the difference distribution for the scan-centric peak characterization is narrower, but the improvement is not as pronounced, likely due to the distribution being bimodal.
+Also, far fewer Xcalibur and MSnbase characterized peaks matched assigned scan-centric peaks.
+Again, this strongly implies that their peak centers have far higher error.
 :::
@@
 `r figure_count$label_text("unassigned_upset")`.
-UpSet plot with the counts of common and specific peaks for each of scancentric, MSnbase and XCalibur generated peaks for the 1ecf (**A**) and 97lipid (**B**) samples.
+UpSet plot with the counts of common and specific peaks without consideration of assignments for each of scancentric, MSnbase and XCalibur generated peaks for the 1ecf (**A**) and 97lipid (**B**) samples.
 :::
@@
 :::
+
 ::: {custom-style="MDPI_4.2_table_body"}
@@
 ```{r assigned_mz_diff, dn_id = figure_count}
+get_base_limits = function(in_peaks, bins = 30){
+  in_base = in_peaks %>%
+  dplyr::filter(source2 %in% "MSnbase") %>%
+  ggplot(aes(x = ppm_diff)) + 
+  geom_histogram(bins = bins) 
+  in_limit = ggplot_build(in_base)$data[[1]]
+  in_limit
+}
+
+plot_mz_diffs = function(in_peaks, limits, ylimit = NULL, inset = FALSE){
+  purrr::map(c("Scan-Centric", "XCalibur", "MSnbase"), function(in_method){
+    tmp_peaks = in_peaks %>%
+      dplyr::filter(source2 %in% in_method)
+    peak_count = purrr::map_dfr(seq(1, nrow(limits)), function(in_row){
+      data.frame(x = limits[in_row, "x"],
+                 y = sum(between(tmp_peaks$ppm_diff, limits[in_row, "xmin"], limits[in_row, "xmax"])))
+    })
+    base_plot = peak_count %>%
+      ggplot(aes(x = x, y = y)) +
+      geom_col() +
+      labs(x = "", y = "count", subtitle = in_method)
+
+    if (in_method %in% "MSnbase") {
+      base_plot = base_plot +
+        labs(x = "abs(ppm) Difference")
+    }
+
+    if (!is.null(ylimit)) {
+      base_plot = base_plot +
+        coord_cartesian(ylim = c(NA, ylimit))
+    }
+    if (inset && (in_method %in% c("XCalibur", "MSnbase"))) {
+      inset_plot = peak_count %>%
+        ggplot(aes(x = x, y = y)) +
+        geom_col() +
+        labs(x = "", y = "") +
+        theme(axis.text.x = element_blank(),
+              axis.text.y = element_text(size = 6))
+      out_plot = base_plot + inset_element(inset_plot, 0.2, 0.1, 1, 1)
+    } else {
+      out_plot = base_plot
+    }
+    out_plot
+  })
+}
+
 diffs_1ecf = ass_peaks_1ecf$peaks %>%
-  dplyr::mutate(source = factor(source, levels = c("scanlevel", "xcalibur", "msnbase"), ordered = TRUE))
+  dplyr::mutate(source2 = dplyr::case_when(
+    source %in% "scanlevel" ~ "Scan-Centric",
+    source %in% "xcalibur" ~ "XCalibur",
+    source %in% "msnbase" ~ "MSnbase"
+  ),
+    ppm_diff = abs(mz_diff / mz) * 1e6)
+
+ecf_limits = get_base_limits(diffs_1ecf)
+
+ecf_plots = plot_mz_diffs(diffs_1ecf, ecf_limits, ylimit = 40)
+
 diffs_97lipid = ass_peaks_97lipid$peaks %>%
-  dplyr::mutate(source = factor(source, levels = c("scanlevel", "xcalibur", "msnbase"), ordered = TRUE))
-
-diffs_plot = (diffs_1ecf %>%
-                ggplot(aes(x = mz_diff)) +
-  geom_histogram(bins = 30) +
-  facet_wrap(~ source, ncol = 1) +
-  labs(x = "m/z Difference")) |
-  (diffs_97lipid %>%
-     ggplot(aes(x = mz_diff)) +
-     geom_histogram(bins = 30) +
-     facet_wrap(~ source, ncol = 1) +
-     labs(x = "m/z Difference")) +
-  plot_annotation(tag_levels = "A")
-diffs_plot
+  dplyr::mutate(source2 = dplyr::case_when(
+    source %in% "scanlevel" ~ "Scan-Centric",
+    source %in% "xcalibur" ~ "XCalibur",
+    source %in% "msnbase" ~ "MSnbase"
+  ),
+    ppm_diff = abs(mz_diff / mz) * 1e6)
+
+lipid_limits = get_base_limits(diffs_97lipid)
+
+lipid_plots = plot_mz_diffs(diffs_97lipid, lipid_limits, ylimit = 800, inset = TRUE)
+
+wrap_plots(ecf_plots, ncol = 1) | wrap_plots(lipid_plots, ncol = 1) 
 ```
@@
 `r figure_count$label_text("assigned_mz_diff")`.
-Histograms of the m/z differences to the assigned and matched peak for 1ecf (**A**) and 97lipid (**B**).
+Histograms of the differences between observed and expected m/z values for assigned peaks in the 1ecf (**left**) and 97lipid (**right**) spectra, measured in parts-per-million (ppm) respectively.
+Inset plots are scaled to the maximum counts observed for the particular method.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-[**Is it just me, or does this feel like the conclusions? If so, do we need a Discussion? I think for MDPI they are optional ...**]
 FT-MS direct-injection data provides spectral scans; however, unlike nuclear magnetic resonance, these spectral scans are not identical, due to several instrumentation issues.

### 2022-04-14 doc/peakcharacterization_manuscript.Rmd
crapload of edits and bits of new information

@@
 equation_count = dn_counter$new("")
+
+supp_figure_count = readRDS("doc/supplemental_figure_count.rds")
+supp_table_count = readRDS("doc/supplemental_table_count.rds")
 ```
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("motivation")`.
+**`r figure_count$label_text("motivation")`.**
 Xcalibur intensities (black) and theoretical intensities based on relative NAP (red) for ECF derived threonine.
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("mz_frequency_conversion")`.** **A**: Intensity vs m/z for a single peak from a single scan.
+**`r figure_count$label_text("mz_frequency_conversion")`.** 
+**A**: Intensity vs m/z for a single peak from a single scan.
 Red lines denote the differences between each point, and red dots the average between the pair of points.
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("intensity_scan")`.** **A**: An example of a peak whose height across scans is correlated with scan number.
+**`r figure_count$label_text("intensity_scan")`.** 
+**A**: An example of a peak whose height across scans is correlated with scan number.
 **B**: The peak differences to the same peaks in a reference scan are plotted against peak height.
-Black: Peaks with a height \< 0.7 of the maximum.
-Red: Peaks with a height \>= 0.7 of the maximum.
+Black: Peaks with a height < 0.7 of the maximum.
+Red: Peaks with a height ≥ 0.7 of the maximum.
 :::
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("normalization_factors")`.** **A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity \>= 0.7 of the maximum intensity (singlenorm_int), or the two pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
+**`r figure_count$label_text("normalization_factors")`.** **A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity ≥ 0.7 of the maximum intensity (singlenorm_int), or the two pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
 **B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("hpd_compare_fsd")`.
+**`r figure_count$label_text("hpd_compare_fsd")`.**
 Comparison of HPD and high FSD sites in the 2ecf sample.
@@
 `r figure_count$label_text("rsd_method")` illustrates the peak height RSD distributions for four different samples.
+Up to two modes are reported for each distribution, where a mode location is only reported if it is ≥ 0.2X the most intense mode.
 For samples 1ecf and 2ecf, there is a general shift in the RSD to the left, going from the bottom processing mehtods (msnbase only) to the top processing method (filtersd), representing a visually clear improvement.
@@
 tar_load(rsd_plot)
-rsd_plot
-```
-:::
-
-::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("rsd_method")`.
+wrap_plots(rsd_plot, ncol = 2)
+```
+:::
+
+::: {custom-style="MDPI_5.1_figure_caption"}
+**`r figure_count$label_text("rsd_method")`.**
 Density plots of relative standard deviations (RSD) of peak heights across scans for each of the processing methods.
@@
 `r table_count$label_text("rsd_method")`.
-Relative standard deviation means, medians, modes, and maximum observed values for each sample with different overall processing.
+RSD means, medians, modes, and maximum observed values for each sample with different overall processing.
 :::
@@
 ```
+:::
+
+```{r increment_rsd_intensity}
+figure_count$increment("rsd_intensity")
+```
+
+::: {custom-style="MDPI_3.1_text"}
+The 97lipid filtersd RSD distribution (as well as others) is bi-modal.
+`r figure_count$label_text("rsd_intensity")` plots each peaks RSD as a function of the mean intensity across scans, for every peak present in at least three scans (top), or the peaks present in at least 80% of the scans (bottom).
+At least part of the bi-modality in the RSD distribution appears to be related to this dependence of RSD on intensity, as well as truncation as peaks appear in fewer scans.
+This is supported by part of the distribution disappearing when we require that peaks are in at least 80% of the available scans (difference between `r figure_count$label_text("rsd_intensity")` top and bottom).
+
+We can remove a majority of the bi-modality by only examining those peaks with a Log10(mean) intensity ≥ 5.
+`r supp_figure_count$label_text("rsd_method")` and `r supp_table_count$label_text("rsd_method")` show how the RSD distributions change for each sample when only the peaks with Log10(mean) intensity ≥ 5 are used.
+The distributions are all shifted to lower RSD, however, the overall trends in RSD are the same.
+:::
+
+::: {custom-style="MDPI_5.2_figure"}
+```{r rsd_intensity, dn_id = figure_count, fig.height = 9}
+plot_alpha = 0.4
+x_lim = c(1.8, 7.4)
+y_lim = c(NA, 1)
+tar_load(rsd_combine)
+rsd_all = rsd_combine %>%
+  dplyr::filter(sample %in% "97lipid", processed %in% "filtersd") %>%
+  ggplot(aes(x = log10(mean), y = rsd)) +
+  geom_point(alpha = plot_alpha) + 
+  coord_cartesian(xlim = x_lim, ylim = y_lim) +
+  labs(subtitle = "N-Scan \u2265 3", x = "Log10(Mean)", y = "RSD")
+
+rsd_many = rsd_combine %>%
+  dplyr::filter(sample %in% "97lipid", 
+                processed %in% "filtersd",
+                n_perc >= 0.8) %>%
+  ggplot(aes(x = log10(mean), y = rsd)) +
+  geom_point(alpha = plot_alpha) +
+  coord_cartesian(xlim = x_lim, ylim = y_lim) +
+  labs(x = "Log10(Mean)", y = "RSD", subtitle = "N-Scan \u2265 80%")
+
+tar_load(corrected_rsd_combine)
+rsd_corrected = corrected_rsd_combine %>%
+  dplyr::filter(sample %in% "97lipid",
+                processed %in% "filtersd") %>%
+  ggplot(aes(x = log10(mean), y = rsd)) +
+  geom_point(alpha = plot_alpha) +
+  coord_cartesian(xlim = x_lim, ylim = y_lim) +
+  labs(x = "Log10(Mean)", y = "RSD", subtitle = "N-Scan \u2265 3, Corrected Height")
+(rsd_all / rsd_many / rsd_corrected) + plot_annotation(tag_levels = "A")
+```
+:::
+
+::: {custom-style="MDPI_5.1_figure_caption"}
+**`r figure_count$label_text("rsd_intensity")`.**
+RSD as a function of the log-mean peak intensity for the filtersd method peaks from the 97lipid sample.
+**A** shows all peaks in at least three scans using raw height; **B** shows peaks that were present in at least 80% of the scans; and **C** shows all peaks in at least three scans using the corrected height.
+:::
+
+::: {custom-style="MDPI_3.1_text"}
+`r figure_count$label_text("rsd_intensity")` also provides information about the sources of variance or error in the FT-MS measurements.
+These trends or RSD with intensity (and SD with intensity, see `r supp_figure_count$label_text("peak_variance")`) imply both a constant, baseline additive error component that is independent of intensity, and a proportional error component where the variance increases with intensity.
 :::
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("nap_intensity")`.
+**`r figure_count$label_text("nap_intensity")`.**
 **A**.
@@
 ::: {custom-style="MDPI_3.1_text"}
-Although we expected that the corrected scan-centric intensities would behave similarly or better compared to the raw intensities, Figure S1 shows that while they have smaller differences than XCalibur, there is still a trend of increasing difference with the raw scan-centric intensity - NAP ratios as the number of missing scans increases.
+Although we expected that the corrected scan-centric intensities would behave similarly or better compared to the raw intensities, `r supp_figure_count$label_text("nap_intensity")` shows that while they have smaller differences than XCalibur, there is still a trend of increasing difference with the raw scan-centric intensity - NAP ratios as the number of missing scans increases.
 :::
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("unassigned_upset")`.
+**`r figure_count$label_text("unassigned_upset")`.**
 UpSet plot with the counts of common and specific peaks without consideration of assignments for each of scancentric, MSnbase and XCalibur generated peaks for the 1ecf (**A**) and 97lipid (**B**) samples.
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("assigned_mz_diff")`.
+**`r figure_count$label_text("assigned_mz_diff")`.**
 Histograms of the differences between observed and expected m/z values for assigned peaks in the 1ecf (**left**) and 97lipid (**right**) spectra, measured in parts-per-million (ppm) respectively.
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("p_value_comparison")`.
+**`r figure_count$label_text("p_value_comparison")`.**
 **A - C**: Log-p-values generated by comparing non-cancer and cancer sample IMFs using peak intensities from different methods.
@@
 It is vitally important however, to normalize the intensities correctly.
-Figure S3 of the Supplemental Materials shows that with the incorrecte normalization, the corrected and raw p-values become much closer to each other.
+`r supp_figure_count$label_text("p_value")` of the Supplemental Materials shows that with the incorrect normalization, the corrected and raw p-values become much closer to each other.
 :::
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("coefficients_qcqa")`.
+**`r figure_count$label_text("coefficients_qcqa")`.**
 **A**: Number of scan outliers for each sample in the NSCLC dataset.
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-`r figure_count$label_text("resolution_qcqa")`.
+**`r figure_count$label_text("resolution_qcqa")`.**
 Scan TIC plot noting the square-root coefficient for each scan, where the coefficient is reflective of the resolution used for each scan.
@@
 Fortunately, this distribution has been previously studied [@sampfordTruncatedNegativeBinomial1955a] and we plan to explore this possibility for a future improvement to our methods.
+Access to the scan level peak intensity information also allows the inference of sources and types of error present in the measurements, which we believe have been previously under-described.
 :::
@@
 For each of the two ECF and two lipid samples, the raw data file was converted to profile mzML format.
-Only the MS1 scans were used, and the scan-scan time difference had to be >= 4 seconds, and for the lipid samples only scans acquired before 450 seconds (7.5 minutes) were kept.
+Only the MS1 scans were used, and the scan-scan time difference had to be ≥ 4 seconds, and for the lipid samples only scans acquired before 450 seconds (7.5 minutes) were kept.
 The scan-centric data was then processed in the following eight ways:
@@
 ::: {custom-style="MDPI_2.2_heading2"}
+### Calculation of Relative Standard Deviation
+:::
+
+::: {custom-style="MDPI_3.1_text"}
+For consistency across both the various scan-centric centroiding and MSnbase, the scan level Log10 heights were converted to raw heights, and then averages and standard deviations calculated across scans, removing missing values.
+Relative standard deviation (RSD) was calculated as standard deviation divided by the mean for each peak.
+The number of scans a peak was present in was also noted, as well as the percentage of total scans.
+Only peaks present in at least three scans were kept for analysis of the RSD.
+
+Modes of the RSD are calculated starting from a density approximation of the distribution using default settings in the density function of the R stats package.
+Peaks from the density approximation are determined with the findpeaks function in the pracma package.
+Up to two modes are reported from low to high RSD along the distribution: i) the most intense peak; ii) the next most intense peak if it is ≥ 0.2X the most intense peak.
+:::
+
+::: {custom-style="MDPI_2.2_heading2"}
 ### Scan-Centric Peak Assignment
@@
 Using information-content-informed Kendall-tau correlation (ICI-Kt) [@flightInformationContentInformedKendalltauCorrelation2022], we compared the median correlations of each sample to all others in the same disease class for outliers, and removed `r n_out_vector["cancer"]` cancer and `r n_out_vector["non-cancer"]` samples.
-The median ICI-Kt correlations are shown in Figure S2 of the Supplemental Materials.
+The median ICI-Kt correlations are shown in `r supp_figure_count$label_text("icikt_outlier")`.
 
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Availability
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-The code used in the reporting of results is available on GitHub at https://github.com/MoseleyBioinformaticsLab/manuscript.peakCharacterization, and archived on Zenodo [@flightMoseleyBioinformaticsLabManuscriptPeakCharacterization2022].
+###
+:::
+
+::: {custom-style="MDPI_6.2_BackMatter"}
+**Author Contributions**: All authors have read and agreed to the published version of the manuscript. 
+Conceptualization, R.M.F, and H.N.B.M; Methodology, R.M.F and H.N.B.M.; Software, R.M.F.; Validation, R.M.F, J.M.M.
+and H.N.B.M.; Formal Analysis: R.M.F.
+and H.N.B.M.; Investigation: J.M.M., R.M.F, and H.N.B.M; Resources, H.N.B.M.; Data Curation, R.M.F; Writing – Original Draft Preparation, R.M.F, J.M.M.
+and H.N.B.M; Writing – Review & Editing, R.M.F, J.M.M.
+and H.N.B.M; Visualization, R.M.F.
+and H.N.B.M.; Supervision, H.N.B.M; Project Administration, H.N.B.M.; Funding Acquisition, H.N.B.M.
+
+**Funding**: The work was supported in part by grants NSF 1419282 (PI Moseley), NSF 2020026 (PI Moseley), NIH 1P01CA163223-01A1 (PD Lane), 1U24DK097215-01A1 (PD Higashi), and P30 CA177558 (B.M.
+Evers, PI) via the Markey Cancer Center Biostatistics and Bioinformatics Shared Resource Facility (MCC BB-SRF).
+
+**Data Availability Statement**: The code used in the reporting of results is available on GitHub at https://github.com/MoseleyBioinformaticsLab/manuscript.peakCharacterization, and archived on Zenodo [@flightMoseleyBioinformaticsLabManuscriptPeakCharacterization2022].
 The scan-centric peak-characterization is available as an R package at https://github.com/MoseleyBioinformaticsLab/FTMS.peakCharacterization, and the specific version used in this work is archived on Zenodo [@flightMoseleyBioinformaticsLabFTMSPeakCharacterization2022].
+
+**Acknowledgments**: We thank Timothy Fahrenholz and Qing Jun Wang for their efforts in mass spectrometry data collection. 
+We thank Andrew Lane, Teresa Fan, and Rick Higashi for their efforts in human tissue sample collection and sample preparation.
+
+**Conflicts of Interest**: The authors declare no conflict of interest. 
+The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, or in the decision to publish the results.
 :::

### 2022-04-14 _targets.R
way more stuff, including more rsd plots

@@ -52,7 +52,8 @@ method_tar = tar_map(
   tar_target(zip, write_peaks_for_assignment(method)),
   tar_target(assign, assign_files(zip)),
   tar_target(n_peak, n_final_peak(method)),
-  tar_target(emf, get_assignments(assign))
+  tar_target(emf, get_assignments(assign)),
+  tar_target(corrected_rsd, single_rsd(method, "CorrectedLog10Height"))
 )
 
 
@@ -98,13 +99,22 @@ combine_noise_tar = tar_combine(
 #   rsd_data,
 #   values = ends_with("97lipid")
 # )
-
+min_perc = 0.8
+min_int = 5
 figures_tar = tar_plan(
   tar_target(frequency_conversion, plot_frequency_conversion(data_97lipid)),
   tar_target(peak_ordering, plot_peak_ordering(method_filtersd_97lipid)),
   tar_target(sliding_regions, plot_sliding_window_density(data_97lipid)),
   tar_target(peak_fit_plot, plot_peak_fitting(method_filtersd_97lipid)),
   tar_target(rsd_plot, plot_rsd_differences(rsd_combine)),
+  tar_target(rsd_plot_alt, plot_rsd_differences(rsd_combine %>%
+                                                  dplyr::filter(n_perc >= min_perc))),
+  tar_target(rsd_plot_int, plot_rsd_differences(rsd_combine %>%
+                                                  dplyr::filter(log10(mean) >= min_int),
+                                                limits = list(`1ecf` = c(0, 1),
+                                                              `2ecf` = c(0, 0.6),
+                                                              `49lipid` = c(0, 0.3),
+                                                              `97lipid` = c(0, 0.3)))),
 
   tar_target(find_sub_region, split_regions(method_perc99_nonorm_97lipid)
   ),
@@ -136,6 +146,11 @@ figures_tar = tar_plan(
 
 tables_tar = tar_plan(
   tar_target(rsd_values, summarize_rsd(rsd_combine)),
+  tar_target(rsd_values_alt, summarize_rsd(rsd_combine %>%
+                                             dplyr::filter(n_perc >= min_perc))),
+  tar_target(rsd_values_int, summarize_rsd(rsd_combine %>%
+                                             dplyr::filter(log10(mean) >= min_int))),
[…]

### 2022-04-14 doc/peakcharacterization_supplemental.Rmd
and rsd plots to supplemental

@@
 $$frequency = a + x * \frac{1}{mz} + y * \frac{1}{\sqrt{mz}}$$
+
+## RSD of Methods for Higher Intensity Peaks
+
+```{r increment_rsd_method}
+figure_count$increment("rsd_method")
+table_count$increment("rsd_method")
+```
+
+`r figure_count$label_text("rsd_method")` and `r table_count$label_text("rsd_method")` we plot and summarize the RSD distributions by sample and method for those peaks with a Log10(mean) intensity ≥ 5.
+The distributions are shifted to the left compared to using all peaks, but the overall patterns are the same.
+
+```{r rsd_method, dn_id = figure_count}
+tar_load(rsd_plot_int)
+wrap_plots(rsd_plot_int, ncol = 2)
+```
+
+`r figure_count$label_text("rsd_method")`.
+Density plots of relative standard deviations (RSD) of peak heights across scans for each of the processing methods.
+A peak had to have a Log10(mean) intensity ≥ 5 for the RSD value to be reported.
+
+`r table_count$label_text("rsd_method")`.
+RSD means, medians, modes, and maximum observed values for each sample with different overall processing.
+Only peaks with a Log10(mean) intensity ≥ 5 are included.
+
+```{r rsd_method_table}
+tar_load(rsd_best_int)
+rsd_best_int %>%
+  autofit()
+```
+
+## Peak Variance
+
+```{r increment_peak_variance}
+tar_load(rsd_combine)
+figure_count$increment("peak_variance")
+```
+
+```{r peak_variance, dn_id = figure_count, fig.width = 8, fig.height=9}
+filtersd_only = rsd_combine %>%
+  dplyr::filter(processed %in% "filtersd", n_perc >= 0.9)
+
+filtersd_by_sample = split(filtersd_only, filtersd_only$sample)
+
+sd_by_sample = purrr::map(filtersd_by_sample, function(in_sample){
+  mean_sd = in_sample %>%
+    ggplot(aes(x = log10(mean), y = sd)) +
+    geom_point(alpha = 0.5) +
+    labs(subtitle = in_sample$sample[1])
+  mean_rsd = in_sample %>%
+    ggplot(aes(x = log10(mean), y = rsd)) +
+    geom_point(alpha = 0.5)
+  mean_sd | mean_rsd
+})
+
+wrap_plots(sd_by_sample, ncol = 1)
+```
+
+`r figure_count$label_text("peak_variance")`.
+Standard deviation (SD) and relative standard deviation (RSD) as a function of Log10(mean) intensity for peaks present in ≥ 90% of scans for each sample.
 
@@
 Red line denotes perfect agreement.
-**D**: Sina plot of differences in the log-p-values generated by corrected intensity normalized by the raw median intensities.
+**B**: Sina plot of differences in the log-p-values generated by corrected intensity normalized by the raw median intensities.
 Also shown are the Bonferroni adjusted p-values from a t-test of the log-p-value differences for each method.
 Green points denote the high, mean, and low-confidence limits reported from the t-test.
+
+```{r save_counts}
+saveRDS(figure_count, file = "doc/supplemental_figure_count.rds")
+saveRDS(table_count, file = "doc/supplemental_table_count.rds")
+```

### 2022-04-14 doc/peakcharacterization_manuscript.Rmd
most of @hunter-moseley last edits in

@@
 ::: {custom-style="MDPI_3.1_text"}
-Fourier transform mass spectrometry (FT-MS) provides high performance in terms of sensitivity, resolution and mass accuracy all in one analytical instrumentation.
+Fourier transform mass spectrometry (FT-MS) provides high performance in terms of sensitivity, resolution, and mass accuracy all in one analytical instrumentation.
 The combination of these capabilities provide several analytical and interpretive improvements: (i) the ability to resolve distinct isotopologues with identical unit masses but different accurate masses [@higashiStableIsotopeLabeledTracers2014]; (ii)  enabling multi-element isotopic natural abundance correction for at least the lower portion of the detected mass range  [@moseleyCorrectingEffectsNatural2010; @carreerComputationalFrameworkHighThroughput2013; @wangAccuCor2IsotopeNatural2021]; (iii) improved assignment accuracy [@kindMetabolomicDatabaseAnnotations2006; @mitchellSmallMoleculeIsotope2019]; and (iv) the detection of metabolites in the sub-femtomolar range, when combined with chromatographic separation  [@eylesMethodsStudyProtein2004; @dettmerMassSpectrometrybasedMetabolomics2007].
@@
 In addition, due to the tendency of increasing point spacing with increasing m/z in Orbitrap type instruments, we derive a method for transforming the m/z data to axial frequency, which has the desirable property of being equally spaced across the full spectrum.
-Finally, we show that in addition to better preserving the expected intensity to- NAP relationships, our scan-centric method also results in improved relative standard deviations, automatic identification of high-peak-density artifact peaks, and better separation of samples in a lipidomics data analysis.
+Finally, we show that in addition to better preserving the expected intensity to NAP relationships, our scan-centric method also results in improved relative standard deviations, automatic identification of high-peak-density artifact peaks, and better separation of samples in a lipidomics data analysis.
 :::
@@
 scan acquisition order); and (ii) some peak differences between scans are correlated with intensity.
-The solution to **i** is to do a two pass normalization.
+The solution to **i** is to do a two-pass normalization.
 After the first pass, the peaks whose intensity is correlated with scan order are detected.
 In the second pass, the correlated peaks are removed, and normalization is carried out again.
-`r figure_count$label_text("intensity_scan")`A contains an example peak whose intensity across scans is correlated with scan number.
+`r figure_count$label_text("intensity_scan")`A shows an example peak whose intensity across scans is correlated with scan number.
 The solution to **ii** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("normalization_factors")`.** **A**: Histogram of scan normalization factors using either a single pass normalization using *all* peaks (singlenorm), single pass normalization using peaks with an intensity ≥ 0.7 of the maximum intensity (singlenorm_int), or the two pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
+**`r figure_count$label_text("normalization_factors")`.** **A**: Histogram of scan normalization factors using either a single-pass normalization using *all* peaks (singlenorm), single-pass normalization using peaks with an intensity ≥ 0.7 of the maximum intensity (singlenorm_int), or the two-pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
 **B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.
@@
 `r figure_count$label_text("hpd_compare_fsd")` shows a single HPD site detected in the 2ecf sample, with the peaks from Xcalibur, as well as various scan-centric processing and the peaks from centroiding using MSnbase.
-From this figure, we can see that the point density based noise filtering removes a large number of the peaks in the HPD site, while the FSD outliers removes further peaks that may be suspect.
+From this figure, we can see that the point-density-based noise filtering removes a large number of the peaks in the HPD site, while the FSD outliers removes further peaks that may be suspect.
 After removing the high FSD peaks, the number of peaks left in the HPD site are the same or less than those from MSnbase, and at least in this example, look more likely to be real peaks compared to those from MSnbase.
@@
 **A**: *xcalibur* peaks exported from Thermo-Fisher Xcalibur after averaging scans.
-**B**: *scancentric_00*: Scan-centric peak characterization without any density based filtering (see noperc_nonorm in Methods).
+**B**: *scancentric_00*: Scan-centric peak characterization without any density-based filtering (see noperc_nonorm in Methods).
 **C**: *scancentric_99*: Scan-centric peak characterization using the default point density filtering (see filtersd in Methods).
@@
 A final check we implemented in the conversion to frequency was verifying that the range of frequency point-point differences allowed for creating the m/z to frequency model are the same across scans.
-The initial set of scans we allowed to use included **all** of the MS1 scans, including the precursor scans for the MS2 scans. 
+The initial set of scans we started with included **all** of the MS1 scans, including the precursor scans for the MS2 scans. 
 However, the samples were acquired as a mix of MS1 only scans for the first 7.5 minutes, and then a mix of MS1 precursor scans followed by MS2 scans for the second 7.5 minutes.
 In 20 samples, the MS1 precursor scans have a different resolution than the MS1 main scans, as shown in `r figure_count$label_text("resolution_qcqa")`.
-This is actually easy to detect by differences in the square-root coefficients, but due to using a single frequency model for converting all scans actually **manifested** as different modes in the point-point frequency differences.
+This is easy to detect by differences in the square-root coefficients, but due to using a single frequency model for converting all scans actually **manifested** as different modes in the point-point frequency differences.
 Oddly enough, the check for outlier scans discussed previously, does not flag any of the scans as outliers in this case.
 Ultimately, we were able to resolve this anomaly by only using those scans with an injection time before 7.5 minutes.
-The package will be updated to include checks that the resolution used is the same across the scans, and produce an informative error for the end user before proceeding.
+In the future, the package will be updated check that the resolution is the same across the scans, and produce an informative error for the end user before quitting.
 :::
@@
 FT-MS direct-injection data provides spectral scans; however, unlike nuclear magnetic resonance, these spectral scans are not identical, due to several instrumentation issues.
-Therefore, simple scan summation to an aggregate spectrum followed by peak characterization does not provide optimal data.
-Also, this approach prevents the calculation of useful peak-specific statistics, especially variance.
-In the results presented, we demonstrate superior performance by scan-centric peak characterization in terms of: i) improved QC/QA and noise reduction; ii) improved peak height relative standard deviations; iii) improved peak intensity ratios that better match natural abundance probabilities; and iv) better separation between biological groups.
-These improvements are necessary for downstream derivation of molecular formula based on peak ratio matching to NAP in SMIRFE.
+Therefore, simple scan summation or averaging to an aggregate spectrum followed by peak characterization does not provide optimal data.
+Also, this approach prevents the calculation of useful peak-specific statistics, especially variances, both in m/z and intensity.
+In the results presented, we demonstrate superior performance by scan-centric peak characterization in terms of: i) improved QC/QA that removes low quality scans; ii) intensity independent noise reduction that pragmatically eliminates HPD sites; iii) improved peak height relative standard deviations; iv) improved peak intensity ratios that better match natural abundance probabilities; and v) better separation between biological groups.
+These improvements are necessary for downstream derivation of molecular formula by SMIRFE using spectrum-derived tolerances, an efficient m/z search of a large isotope-resolved cache, and filtering by peak ratio matching to NAP.
 However, scan-centric peak characterization requires a sophisticated pipeline of QC/QA and processing steps to derive a superior characterization with informative statistics and QC/QA metrics.
@@
 But once implemented, large datasets can be processed in a straight-forward manner that supports full computational reproducibility.
+
 Not all of the methods presented here provided an improvement.
 We included our negative results with correcting mean and standard deviations for the intensity of peaks not detected in all scans, so that others do not waste effort pursuing this.
-We initially expected the data to follow a truncated normal distribution; however, upon further contemplation of the results presented here, we believe that the differences between scans still present after normalization imparts a dispersive variance component.
-Thus, we hypothesize that a truncated negative binomial distribution better represents the intensity of peaks not detected across all scans.
+We initially expected the data to follow a truncated normal distribution, since `r figure_count$label_text("rsd_intensity")` clearly demonstrates truncation effects; however, upon further contemplation of the results presented here, we believe that the differences between scans still present after normalization imparts a dispersive variance component.
+Moreover, `r figure_count$label_text("rsd_intensity")` (and `r supp_figure_count$label_text("peak_variance"`) demonstrates the presence of a proportional error component.
+Thus, we hypothesize that a truncated negative binomial distribution may better represent the intensity of peaks not detected across all scans.
 Fortunately, this distribution has been previously studied [@sampfordTruncatedNegativeBinomial1955a] and we plan to explore this possibility for a future improvement to our methods.
-Access to the scan level peak intensity information also allows the inference of sources and types of error present in the measurements, which we believe have been previously under-described.
+However, correction may not be straight-forward.
+As illustrated in `r figure_count$label_text("rsd_intensity")`, there are low intensity peaks that are observed across the majority of the scans.
+This likely indicates that the loss of peak detection involves other factors that are increased at low intensity.
+We theorize that any factor that increases signal decoherence like ion-cloud repulsion would increase the peak flooring performed by the instrument.
+Therefore, access to the scan level peak intensity information as provided by the methods presented here facilitates the inference of sources and types of error present in the measurements, which we believe have been previously under-described.
 :::
@@
 ::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, single pass normalization with all peaks (singlenorm)
+* Noise removal, single-pass normalization with all peaks (singlenorm)
 :::
 ::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, single pass normalization with high ratio peaks (singlenorm_int)
+* Noise removal, single-pass normalization with high ratio peaks (singlenorm_int)
 :::
 ::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, two pass normalization (doublenorm)
+* Noise removal, two-pass normalization (doublenorm)
 :::
 ::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, two pass normalization (filtersd)
+* Noise removal, two-pass normalization (filtersd)
 :::
@@
 The presence of zero intensity points in these Thermo-Fisher Orbitrap spectra are primarily due to flooring implemented by the spectrometer when local spectral quality falls below a certain threshold.
-When this flooring is unstable and incomplete, a partial ringing phenomenom is observed [@mitchellNewMethodsIdentify2018].
+When this flooring is unstable and incomplete, a partial ringing phenomenon is observed [@mitchellNewMethodsIdentify2018].
 
@@
 However, some peaks are not observed in some scans due to the number of ions falling below the detection threshold, or being excluded after filling the trap.
-This results in a left-censored normal distribution of peak intensities across scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
+This should result in a left-censored normal distribution of peak intensities across equivalent scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
 To correct these, either a correction based on a model of the truncated normal distribution can be used, or the differences can be simulated by sampling from data that is present in most of the scans.
@@
 
-We attempted to create peaks that could be assigned by SMIRFE from the MSnbase and Xcalibur peak lists, but SMIRFE would not assign them given its requirements for scan-centric peak characterization data.
+We attempted to create peaks that could be assigned by SMIRFE from the MSnbase and Xcalibur peak lists, but SMIRFE would not assign them given its requirements for scan-centric peak characterization data that includes variance.
 :::

### 2022-04-14 doc/peakcharacterization_manuscript.Rmd
revisions in

@@
 These trends or RSD with intensity (and SD with intensity, see `r supp_figure_count$label_text("peak_variance")`) imply both a constant, baseline additive error component that is independent of intensity, and a proportional error component where the variance increases with intensity.
+**When the intensity is log-transformed, this additive component becomes a dispersive variance component.**[this is true, but I'm not sure it describes these figures**]
+In addition, there is an unusual boot-shaped curve at the lower log mean intensities, representing left-censorship (truncation) effects due to detection limits.
 :::
@@
 ```
+:::
+
+::: {custom-style="MDPI_2.2_heading2"}
+###
 :::
@@
 We initially expected the data to follow a truncated normal distribution, since `r figure_count$label_text("rsd_intensity")` clearly demonstrates truncation effects; however, upon further contemplation of the results presented here, we believe that the differences between scans still present after normalization imparts a dispersive variance component.
-Moreover, `r figure_count$label_text("rsd_intensity")` (and `r supp_figure_count$label_text("peak_variance"`) demonstrates the presence of a proportional error component.
+Moreover, `r figure_count$label_text("rsd_intensity")` (and `r supp_figure_count$label_text("peak_variance")`) demonstrates the presence of a proportional error component.
 Thus, we hypothesize that a truncated negative binomial distribution may better represent the intensity of peaks not detected across all scans.
@@
 The ratio standard deviation across peaks can then be fitted to a cubic model of the fraction used, and a correction factor predicted for those peaks that are present in fewer scans.
+Our correction uses log-transformed peak heights, and differences instead of raw heights and ratios directly to make some of the calculations easier.
 The corrected standard deviations can then be used to correct the mean height assuming that it is the result of a left-censored normal distribution [@TruncatedNormalDistribution2022; @burkardtTruncatedNormalDistribution].

### 2022-04-15 doc/peakcharacterization_manuscript.Rmd
jmm edits in

--- 2022-04-14 doc/peakcharacterization_manuscript.Rmd
@@
       email: jmmi243@uky.edu
-      institute: [markey, biochem, rcsirm, ibi]
+      institute: [markey, biochem, rcsirm]
   - Hunter NB Moseley:
@@
   - tox: Department of Toxicology and Cancer Biology, University of Kentucky, Lexington, KY 40536, USA
-date: "`r Sys.time()`"
 output: 
@@
 **Abstract:** We introduce a novel, scan-centric method for characterizing peaks from direct injection multi-scan Fourier transform mass spectra of complex samples that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
-Our peak characterization method utilizes intensity independent noise removal and normalization of scan level data to provide a much better fit of relative intensity to natural abundance probabilities for low abundance isotopologues that are not present in all of the acquired scans.
+Our peak characterization method utilizes intensity independent noise removal and normalization of scan-level data to provide a much better fit of relative intensity to natural abundance probabilities for low abundance isotopologues that are not present in all of the acquired scans.
 Moreover, our method calculates both peak- and scan-specific statistics incorporated within a series of quality control steps that are designed to robustly derive peak centers, intensities and intensity ratios with their scan-level variances.
@@
 ::: {custom-style="MDPI_1.8_keywords"}
-**Keywords**: Fourier-transform mass-spectrometry
+**Keywords**: Fourier-transform mass spectrometry
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-Fourier transform mass spectrometry (FT-MS) provides high performance in terms of sensitivity, resolution, and mass accuracy all in one analytical instrumentation.
+Fourier-transform mass spectrometry (FT-MS) provides high performance in terms of sensitivity, resolution, and mass accuracy all in one analytical instrumentation.
 The combination of these capabilities provide several analytical and interpretive improvements: (i) the ability to resolve distinct isotopologues with identical unit masses but different accurate masses [@higashiStableIsotopeLabeledTracers2014]; (ii)  enabling multi-element isotopic natural abundance correction for at least the lower portion of the detected mass range  [@moseleyCorrectingEffectsNatural2010; @carreerComputationalFrameworkHighThroughput2013; @wangAccuCor2IsotopeNatural2021]; (iii) improved assignment accuracy [@kindMetabolomicDatabaseAnnotations2006; @mitchellSmallMoleculeIsotope2019]; and (iv) the detection of metabolites in the sub-femtomolar range, when combined with chromatographic separation  [@eylesMethodsStudyProtein2004; @dettmerMassSpectrometrybasedMetabolomics2007].
 In the metabolomics field, these improvements permit more complicated, but more informative experimental designs such as the use of single and multiple isotope-labeled precursors in stable isotope-resolved metabolomics (SIRM) experiments [@yangChloroformateDerivatizationTracing2017].
-These stable isotop tracing experiments provide a wealth of isotope flux data that is interpretable in terms of metabolite flux information that is specific to a metabolic model, pathway, and subcellular location [@higashiStableIsotopeLabeledTracers2014; @creekStableIsotopeAssistedMetabolomics2012; @hillerNontargetedElucidationMetabolic2010; @fanStableIsotoperesolvedMetabolomics2012; @moseleyNovelDeconvolutionMethod2011; @sellersPyruvateCarboxylaseCritical2015; @verdegemMAIMSSoftwareTool2017; @jinMoietyModelingFramework2019; @jinRobustMoietyModel2020].
+These stable isotope tracing experiments provide a wealth of isotope flux data that is interpretable in terms of metabolite flux information that is specific to a metabolic model, pathway, and subcellular location [@higashiStableIsotopeLabeledTracers2014; @creekStableIsotopeAssistedMetabolomics2012; @hillerNontargetedElucidationMetabolic2010; @fanStableIsotoperesolvedMetabolomics2012; @moseleyNovelDeconvolutionMethod2011; @sellersPyruvateCarboxylaseCritical2015; @verdegemMAIMSSoftwareTool2017; @jinMoietyModelingFramework2019; @jinRobustMoietyModel2020].
 
 While these advantages of FT-MS are significant, when deployed in a high-throughput environment, the volume of data produced requires automated tools for data reduction, quality control, feature assignment, and downstream analyses.
-Furthermore within the context of direct infusion, assignment of FT-MS spectral features lack orthogonal sources of information such as chromatographic retention times, reducing the reliability of assignment with most MS assignment software tools.
-
-Within this combined high-throughput with direct injection context, we previously developed methods to remove high-peak-density artifacts [@mitchellNewMethodsIdentify2018], annotate peaks with assignments using SMIRFE [@mitchellSmallMoleculeIsotope2019], and generate lipid classifications for spectral peak assignments [@mitchellDerivingLipidClassification2020].
-We applied all of these methods together in detecting and assigning differential lipids in non-small-cell lung cancer (NSCLC) [@mitchellUntargetedLipidomicsNonSmall2021].
-
-The input peak-data in these studies are derived from multi-scan direct-injection FT-MS data.
+Furthermore within the context of direct infusion, assignment of FT-MS spectral features lacks orthogonal sources of information such as chromatographic retention times, reducing the reliability of assignment with most MS assignment software tools.
+
+Within this combined high-throughput with direct injection context, we previously developed methods to remove high-peak-density artefacts [@mitchellNewMethodsIdentify2018], annotate peaks with assignments using SMIRFE [@mitchellSmallMoleculeIsotope2019], and generate lipid classifications for spectral peak assignments [@mitchellDerivingLipidClassification2020].
+We applied all of these methods together in detecting and assigning differential lipids in non-small-cell lung carcinoma (NSCLC) [@mitchellUntargetedLipidomicsNonSmall2021].
+
+The input peak data in these studies are derived from multi-scan direct-injection FT-MS data.
 SMIRFE uses expected relationships of relative peak height or intensity to natural abundance probability (NAP) across multiple peak pairs to determine the likelihood of an assignment for the peaks in question.
-In the Thermo-Fisher Orbitrap FT-MS instruments, data is acquired in micro-scans and scans, where each scan is an aggregate of multiple micro-scans.
+In the Thermo-Fisher Orbitrap FT-MS instruments, data is acquired in microscans and scans, where each scan is an aggregate of multiple microscans.
 While these micro-scan and subsequent scans are expected to be analytical replicates, spray instability and temporal delay in automatic gain control pragmatically break this expectation.
@@
 This is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points.
-The m/z point-to-point differences are not constant, but can be approximated by a loess linear model [@Cleveland1992loess]; however, it is very difficult to create a loess model with an intercept of 0.
-In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the loess model.
-Frequency based points suffer none of these drawbacks, and the conversion from m/z can be derived from the raw profile level data itself, which is incredibly useful.
+The m/z point-to-point differences are not constant, but can be approximated by a LOESS linear model [@Cleveland1992loess]; however, it is very difficult to create a LOESS model with an intercept of 0.
+In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the LOESS model.
+Frequency-based points suffer none of these drawbacks, and the conversion from m/z can be derived from the raw profile level data itself, which is incredibly useful.
 :::
@@
 The differences for all peaks across all scans vs m/z are shown in **D**, with those differences that lie within 0.49 - 0.51 shown in red.
-**E** shows the plot of derived frequency *vs* m/z, with fitted values from the linear model in red.
+**E** shows the plot of derived frequency vs m/z, with fitted values from the linear model in red.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-Although many other types of mass-spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo-Fisher Fusion has a baseline of 0 due to manipulations in the Thermo-Fisher firmware, making the determination of the centroided values considerably easier.
+Although many other types of mass spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo-Fisher Fusion has a baseline of 0 due to manipulations in the Thermo-Fisher firmware, making the determination of the centroided values considerably easier.
 For each region initially created, the peaks in each scan within that region can be characterized (i.e. centroiding).
@@
 The solution to **i** is to do a two-pass normalization.
-After the first pass, the peaks whose intensity is correlated with scan order are detected.
+After the first pass, the peaks whose intensity is correlated with scan order are detected (absolute Pearson correlation > 0.5).
 In the second pass, the correlated peaks are removed, and normalization is carried out again.
@@
 Up to two modes are reported for each distribution, where a mode location is only reported if it is ≥ 0.2X the most intense mode.
-For samples 1ecf and 2ecf, there is a general shift in the RSD to the left, going from the bottom processing mehtods (msnbase only) to the top processing method (filtersd), representing a visually clear improvement.
+For samples 1ecf and 2ecf, there is a general shift in the RSD to the left, going from the bottom processing methods (msnbase_only) to the top processing method (filtersd), representing a visually clear improvement.
 For samples 49lipid and 97lipid, the improvements are visually more subtle, having a slight shift to the left as well as a narrowing and smoothing of the RSD distribution.
-Part of this sublety is likely due to the bimodality of these two distributions.
+Part of this sublety is likely due to the bi-modality of these two distributions.
 This is not surprising, since these two samples are non-polar extractions from tissue and are biochemically more complex.
@@
 These trends or RSD with intensity (and SD with intensity, see `r supp_figure_count$label_text("peak_variance")`) imply both a constant, baseline additive error component that is independent of intensity, and a proportional error component where the variance increases with intensity.
+When the intensity is log-transformed, this additive component becomes a dispersive variance component.
+In addition, there is an unusual boot-shaped curve at the lower log mean intensities, representing left-censorship (truncation) effects due to detection limits.
 :::
@@
 **A**.
-The peak-peak NAP - intensity log differences from scan-centric peak heights (red) and Xcalibur peak heights (blue) from the threonine amino acid assignments with Na adduct, with point size reflecting how many peaks were missing across scans.
+The peak-peak NAP - intensity log differences from scan-centric peak heights (red) and Xcalibur peak heights (blue) from the ECF derivatized threonine amino acid assignments with Na adduct, with point size reflecting how many peaks were missing across scans.
 **B**. The difference of Xcalibur to scan-centric ratios plotted directly as a function of the number of scans the peak was not found in.
@@
 Also, the MSnbase distribution has far fewer peaks that match a scan-level assigned peak.
-Given the large number of MSnbase-characterized peaks present in the spectrum, many peaks may be outside the matching tolerance of 2pmm. 
+Given the large number of MSnbase-characterized peaks present in the spectrum, many peaks may be outside the matching tolerance of 2ppm. 
 This strongly implies that the peak center error is far higher in the MSnbase-characterized peaks than what the histogram directly shows.
-For the 97lipid-specific histograms, again the difference distribution for the scan-centric peak characterization is narrower, but the improvement is not as pronounced, likely due to the distribution being bimodal.
+For the 97lipid-specific histograms, again the difference distribution for the scan-centric peak characterization is narrower, but the improvement is not as pronounced, likely due to the distribution being bi-modal.
 Also, far fewer Xcalibur and MSnbase characterized peaks matched assigned scan-centric peaks.
@@
 This echoes the patterns of low overlapping assigned peaks observed in both ECF samples, where many of the scan-centric amino-acid peaks only found one match from the MSnbase peaks.
-Surprisingly, the truncated normal distribution corrected intensities generated very different p-values compared to the raw intensities, and XCalibur showed the most agreement with the raw p-values.
+Surprisingly, the truncated log-normal distribution corrected intensities generated very different p-values compared to the raw intensities, and XCalibur showed the most agreement with the raw p-values.
 We tested the statistical significance of these differences using a t-test of the actual difference in log-p-values (shown in `r figure_count$label_text("p_value_comparison")`D).
@@
 Ultimately, we were able to resolve this anomaly by only using those scans with an injection time before 7.5 minutes.
-In the future, the package will be updated check that the resolution is the same across the scans, and produce an informative error for the end user before quitting.
+In the future, the package will be updated to check that the resolution is the same across the scans, and produce an informative error for the end user before quitting.
 :::
@@
 **`r figure_count$label_text("resolution_qcqa")`.**
-Scan TIC plot noting the square-root coefficient for each scan, where the coefficient is reflective of the resolution used for each scan.
+Scan-level total ion chromatogram (TIC) plot noting the square-root coefficient for each scan, where the coefficient is reflective of the resolution used for each scan.
 The y-axis is cropped to 2x10^7^ so that the MS1 precursor scans are visible, as their TIC's are much smaller than the others.
@@
 We included our negative results with correcting mean and standard deviations for the intensity of peaks not detected in all scans, so that others do not waste effort pursuing this.
-We initially expected the data to follow a truncated normal distribution, since `r figure_count$label_text("rsd_intensity")` clearly demonstrates truncation effects; however, upon further contemplation of the results presented here, we believe that the differences between scans still present after normalization imparts a dispersive variance component.
-Moreover, `r figure_count$label_text("rsd_intensity")` (and `r supp_figure_count$label_text("peak_variance"`) demonstrates the presence of a proportional error component.
+We initially expected the data to follow a truncated log-normal distribution, since `r figure_count$label_text("rsd_intensity")` clearly demonstrates truncation effects; however, upon further contemplation of the results presented here, we believe that the differences between scans still present after normalization imparts a dispersive variance component.
+Moreover, `r figure_count$label_text("rsd_intensity")` (and `r supp_figure_count$label_text("peak_variance")`) demonstrates the presence of an additive error component that will become dispersive after log-transformation.
 Thus, we hypothesize that a truncated negative binomial distribution may better represent the intensity of peaks not detected across all scans.
 Fortunately, this distribution has been previously studied [@sampfordTruncatedNegativeBinomial1955a] and we plan to explore this possibility for a future improvement to our methods.
-However, correction may not be straight-forward.
+However, correction may not be straight-forward and it is unclear if it is better to apply the correction to the raw or log-transformed peak height data.
+Either way, a dispersive variance component is present and must be accounted for.
 As illustrated in `r figure_count$label_text("rsd_intensity")`, there are low intensity peaks that are observed across the majority of the scans.
@@
 We theorize that any factor that increases signal decoherence like ion-cloud repulsion would increase the peak flooring performed by the instrument.
-Therefore, access to the scan level peak intensity information as provided by the methods presented here facilitates the inference of sources and types of error present in the measurements, which we believe have been previously under-described.
+Therefore, access to the scan level peak intensity information as provided by the methods presented here facilitates the inference of sources and types of error present in the measurements, which we believe have been previously under-described and is required for future data processing improvements.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-Two different sets of samples were used to evaluate the various methods, ethylchloroformate (ECF) amino acid samples and non-small-cell-lung-cancer (NSCLC) lipid-extracted tissue samples.
+Two different sets of samples were used to evaluate the various methods, ethylchloroformate (ECF) derived amino acid samples and non-small-cell-lung-carcinoma (NSCLC) lipid-extracted tissue samples.
 
 The method for generating the amino-acid samples was adapted from a previously published method for performing ECF amino acid derivatization [@yangChloroformateDerivatizationTracing2017]. 
-Two replicate samples were prepared and spectra were obtained for both samples using a Tribrid Fusion Orbitrap at 500k resolution and 150 to 1000m/z.
+Two replicate samples were prepared and spectra were obtained for both samples using a Tribrid Fusion Orbitrap at 500k resolution and a mass range of 150 to 1000m/z.
 
@@
 ::: {custom-style="MDPI_3.9_equation"}
-`r format_equation('frequency = a + y * \\frac{1}{\\sqrt{mz}} + z * \\frac{1}{\\sqrt[3]{mz}}', 'mzfrequency')`
+`r format_equation('frequency = a + y \\times \\frac{1}{\\sqrt{mz}} + z \\times \\frac{1}{\\sqrt[3]{mz}}', 'mzfrequency')`
 :::
@@
 ::: {custom-style="MDPI_3.9_equation"}
-`r format_equation('mz = a + x * \\frac{1}{frequency} + y * \\frac{1}{frequency^2} + z *\\frac{1}{frequency^3}', 'frequencymz')`
+`r format_equation('mz = a + x \\times \\frac{1}{frequency} + y \\times \\frac{1}{frequency^2} + z *\\frac{1}{frequency^3}', 'frequencymz')`
 :::
@@
 ::: {custom-style="MDPI_3.9_equation"}
-`r format_equation('\\ln{intensity} = a + x * position + y * position^2', 'modelintensity')`
+`r format_equation('\\ln(intensity) = a + x \\times position + y \\times position^2', 'modelintensity')`
 :::
@@
 However, some peaks are not observed in some scans due to the number of ions falling below the detection threshold, or being excluded after filling the trap.
-This should result in a left-censored normal distribution of peak intensities across equivalent scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
-To correct these, either a correction based on a model of the truncated normal distribution can be used, or the differences can be simulated by sampling from data that is present in most of the scans.
+This should result in a left-censored log-normal distribution of peak intensities across equivalent scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
+To correct these, either a correction based on a model of the truncated normal distribution can be used on log-tansformed data, or the differences can be simulated by sampling from data that is present in most of the scans.
 To simulate the effect of peaks missing from some scans on the standard deviations, the peaks present in all scans are used.
@@
 ::: {custom-style="MDPI_3.1_text"}
-High-peak-density (HPD) artifacts [@mitchellNewMethodsIdentify2018] present as groups of singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
+High-peak-density (HPD) artefacts [@mitchellNewMethodsIdentify2018] present as groups of singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
 Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and peaks with FSDs greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are marked.
@@
 For the lipid samples, an initial EMF database was generated using an m/z limit of 1605 m/z, and maximum numbers for each element were set to C: 130, N: 7, O: 28, P: 3, H: 230. 
-Assigned formulas were allowed to have K, Na, H, and NH4 adducts (only positive mode samples were assigned). 
+Assigned formulas were allowed to have K^+^, Na^+^, H^+^, and NH4^+^ adducts (only positive mode samples were assigned). 
 Assigned molecular formulas were then classified into one or more lipid categories using our lipid classifier tool [@mitchellDerivingLipidClassification2020].
-Lipid classes that were previously observed to be overclassified by our models (neutral and acidic glycosphingolipids) were excluded.
 For the ECF derivatized amino-acid samples, the EMF database was generated using an m/z limit of 1005 m/z, and maximum numbers for each element were set to C: 100, N: 7, O: 40, H: 230, P: 3, S: 3.
+^15^N labeling was included in the database generation.
+Assigned formulas were allowed to have H^+^ and Na^+^ adducts.
 
@@
 
+**Institutional Review Board Statement**: The study was conducted according to the guidelines of the Declaration of Helsinki and approved by the Institutional Review Boards of the University of Louisville (#05.0523, approval date: 10/19/2005) and the University of Kentucky (#14-0288-F6A, approval date: 05/07/2014).
+
+**Informed Consent Statement**: Informed consent was obtained from all subjects involved in the study.
+
 **Data Availability Statement**: The code used in the reporting of results is available on GitHub at https://github.com/MoseleyBioinformaticsLab/manuscript.peakCharacterization, and archived on Zenodo [@flightMoseleyBioinformaticsLabManuscriptPeakCharacterization2022].

### 2022-04-15 doc/peakcharacterization_manuscript.Rmd
making sure figures get labeled

@@
 ::: {custom-style="MDPI_5.2_figure"}
-```{r coefficients_qcqa}
+```{r coefficients_qcqa, dn_id = figure_count}
 coef_df1 = data.frame(N_out = coefficients_qcqa$n_coef_out)
@@
 ::: {custom-style="MDPI_5.2_figure"}
-```{r resolution_qcqa}
+```{r resolution_qcqa, dn_id = figure_count}
 tar_load(resolution_qcqa)

### 2022-04-15 doc/peakcharacterization_manuscript.Rmd
getting all to artifacts

@@
 
-Within this combined high-throughput with direct injection context, we previously developed methods to remove high-peak-density artefacts [@mitchellNewMethodsIdentify2018], annotate peaks with assignments using SMIRFE [@mitchellSmallMoleculeIsotope2019], and generate lipid classifications for spectral peak assignments [@mitchellDerivingLipidClassification2020].
+Within this combined high-throughput with direct injection context, we previously developed methods to remove high-peak-density artifacts [@mitchellNewMethodsIdentify2018], annotate peaks with assignments using SMIRFE [@mitchellSmallMoleculeIsotope2019], and generate lipid classifications for spectral peak assignments [@mitchellDerivingLipidClassification2020].
 We applied all of these methods together in detecting and assigning differential lipids in non-small-cell lung carcinoma (NSCLC) [@mitchellUntargetedLipidomicsNonSmall2021].
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Mitigation of High Peak Density Artefacts
+### Mitigation of High Peak Density Artifacts
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-We have previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [@mitchellNewMethodsIdentify2018].
+We have previously described the presence of high peak density (HPD) artifacts in FT-MS spectra [@mitchellNewMethodsIdentify2018].
 Ideally, the peak characterization procedure should reduce and mitigate their presence and effect in the resultant reported peaks.
@@
 Having a scan-centric workflow for generating the peak centroids means that we have opportunities to evaluate the scans as a whole, as well as the peaks across scans.
-For example, we can mark peaks with unusually high frequency standard deviations (FSD) and mitigate the presence of high-peak-density artefacts (see **Mitigation of High Peak Density Artefacts**).
+For example, we can mark peaks with unusually high frequency standard deviations (FSD) and mitigate the presence of high-peak-density artifacts (see **Mitigation of High Peak Density Artifacts**).
 We also sometimes find that from initial peak region that has been marked as a single region, still results in two peaks being characterized in a single scan.
@@
 ::: {custom-style="MDPI_3.1_text"}
-High-peak-density (HPD) artefacts [@mitchellNewMethodsIdentify2018] present as groups of singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
+High-peak-density (HPD) artifacts [@mitchellNewMethodsIdentify2018] present as groups of singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations.
 Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and peaks with FSDs greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are marked.

### 2022-04-15 doc/peakcharacterization_manuscript.Rmd
removing things we don't actually need anymore

@@
 ::: {custom-style="MDPI_3.1_text"}
-Two different sets of samples were used to evaluate the various methods, ethylchloroformate (ECF) derived amino acid samples and non-small-cell-lung-carcinoma (NSCLC) lipid-extracted tissue samples.
+Two different sets of samples were used to evaluate the various methods, ethylchloroformate (ECF) derived amino acid samples described in [@mitchellSmallMoleculeIsotope2019] and non-small-cell-lung-carcinoma (NSCLC) lipid-extracted tissue samples described in [@mitchellUntargetedLipidomicsNonSmall2021].
 
@@
 Written informed consent was collected from all subjects prior to inclusion and all samples were collected under a University of Louisville or University of Kentucky IRB protocol. 
-Lipid extracts were prepared using a modified Folch extraction and reconstituted for direct infusion ultra-high resolution mass spectrometry on a pair of Thermo-Fisher Tribrid Fusion Orbitrap instruments (FSN10115 and FSN10352, referred to as Fusion 1 and Fusion 2 respectively) coupled to an Advion nanoelectrospray system.
-53 patients (102 spectra total) were acquired from Fusion 1 (serial # FSN10115) and 40 patients (77 spectra total) were acquired from Fusion 2 (serial # FSN10352).
-Fusion 1 samples were exclusively from the University of Louisville while Fusion 2 were a mix.
-Three of the patients had only a cancer or non-cancer sample acquired.
+Lipid extracts were prepared using a modified Folch extraction and reconstituted for direct infusion ultra-high resolution mass spectrometry on a pair of Thermo-Fisher Tribrid Fusion Orbitrap instruments coupled to an Advion nanoelectrospray system.
 Two of the cancer lipid samples (49Cpos and 97Cpos) were used for the majority of examples in this manuscript.
@@
 
-**Acknowledgments**: We thank Timothy Fahrenholz and Qing Jun Wang for their efforts in mass spectrometry data collection. 
-We thank Andrew Lane, Teresa Fan, and Rick Higashi for their efforts in human tissue sample collection and sample preparation.
-
 **Conflicts of Interest**: The authors declare no conflict of interest. 

### 2022-04-15 doc/peakcharacterization_manuscript.Rmd
a few more edits to get to v1.4.1

@@
 ::: {custom-style="MDPI_3.9_equation"}
-`r format_equation('mz = a + x \\times \\frac{1}{frequency} + y \\times \\frac{1}{frequency^2} + z *\\frac{1}{frequency^3}', 'frequencymz')`
+`r format_equation('mz = a + x \\times \\frac{1}{frequency} + y \\times \\frac{1}{frequency^2} + z \\times \\frac{1}{frequency^3}', 'frequencymz')`
 :::

### 2022-04-15 doc/peakcharacterization_supplemental.Rmd
a few more edits to get to v1.4.1

@@
 
-$$frequency = a + x * \frac{1}{mz} + y * \frac{1}{\sqrt{mz}}$$
+$$frequency = a + x \times \frac{1}{mz} + y \times \frac{1}{\sqrt{mz}}$$
 

### 2022-04-22 doc/peakcharacterization_manuscript.Rmd
removing IRB statement b/c its already in the NSCLC paper and everything here is reuse of data and deidentified

@@
 
-**Institutional Review Board Statement**: The study was conducted according to the guidelines of the Declaration of Helsinki and approved by the Institutional Review Boards of the University of Louisville (#05.0523, approval date: 10/19/2005) and the University of Kentucky (#14-0288-F6A, approval date: 05/07/2014).
-
-**Informed Consent Statement**: Informed consent was obtained from all subjects involved in the study.
-
 **Data Availability Statement**: The code used in the reporting of results is available on GitHub at https://github.com/MoseleyBioinformaticsLab/manuscript.peakCharacterization, and archived on Zenodo [@flightMoseleyBioinformaticsLabManuscriptPeakCharacterization2022].

### 2022-05-11 doc/response_to_reviewers.Rmd
first draft of response document

---
title: "Scan-Centric, Frequency-Based Method for Characterizing Peaks from Direct Injection Fourier transform Mass Spectrometry Experiments: Response to Reviewers"
output: 
  word_document:
    reference_docx: response_to_reviewers_template.docx
    keep_md: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

::: {custom-style="Response"}
This is how our comments in return will show up.
:::

::: {custom-style="Modified"}
And this is how any inserted text we want to highlight as to how we actually modified the text will show up.
:::

## Reviewer 1

Mass spectrometry-based metabolite identification and characterization are rapidly developing. Ultra-sensitive detection, high throughput analysis, and intelligent data processing strategies are enabling the identification of very low abundant metabolites and other biomolecules. Downstream identification of metabolites is based on the FT-MS signal in mass spectra. The well-characterized peaks lead to confident identification. In the current manuscript, the authors reported the methods to characterize the peaks to improve the identification. The authors are quickly describing the different methods without much of its discussion. However, the manuscript is interesting and could be further improved.

### Major Points:

1. In Results section 2.1. the authors are trying to demonstrate that the Xcalibur calculated intensities and theoretical peak intensities differ with the help of Figure 1. The observed and theoretical intensities often differ and it is well-known fact. What new things that authors are trying to show are hard to find. Figure 1 is hard to understand and there is not enough description of that.

::: {custom-style="Response"}
We need to add description of the importance that the deviations increase as the peak intensity gets lower.
:::

2. “The Thermo-Fisher Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 450K or 500K depending on the sample, has a mode of 0.5, as shown in Figure 2C and D” is not clear as if the frequency difference range already within 0.5 or should be within that range? Also, the peak ordering calculations from m/z to frequency models need more description.

::: {custom-style="Response"}
I wonder if there is something we didn't explain very well here, like how M/Z does not have a constant difference?

Or that we weren't sure if frequency would be constant? I'm also guessing it's not clear that Figure 2D that a whole lot of points are outside 0.5 +- 0.05 (i.e. the points in black).
:::

3. In the section “Sliding Window Density to Remove Noise” the authors describe the method to remove the noise. The authors need to give a real example demonstrating the removal of noise from the actual signal scans. Not all no-zero points are noise. This needs to be better explained as it is not clear how the authors reached a cutoff of 7.5.

::: {custom-style="Response"}
I'm not surprised, it's not the easiest thing to follow. I'm wondering if some kind of schematic might help here. Will have to think carefully on this one.
:::

4. Breaking up the peaks when the co-isolated ions are overlapped may be helpful. The authors may want to clearly demonstrate the advantages of splitting the peaks of non-zero values.

::: {custom-style="Response"}
I think this relates to Figure 6, splitting initial regions.

I also think the figure caption needs more work, as we need to make sure we are indicating that each trace is a peak in a scan.
:::

5. Normalization of Scans, Mitigation of High Peak Density Artifacts, and Changes in Relative Standard Deviation (RSD) need more discussion. The singlenorm_int and the two-pass normalization are expected to be similar due to the most intense peak selection.

::: {custom-style="Response"}
They are correct. I think the RSD comes back to visiting #28 (rsd changes at peak level) so we can actually talk about how the RSD changes for each individual peak across the incremental changes to SCPC.
:::

6. The authors indicated that not all methods reported in the manuscript are improved, and they also reported negative results. This is highly valuable in science. However, the authors need to include which methods are really improved and which are not. This would be very helpful to draw conclusions.

::: {custom-style="Response"}
I guess we need to be more convincing about which of the bits result in real improvements. Probably gets back to #36 honestly. And pushing more on the peak overlap amongst XCalibur and MSnbase??
:::

### Minor points:

1. Abstract needs revision.

2. Figure 1 could be improved. From where does the 18O isotopologue appear?

::: {custom-style="Response"}
I'm not 100% sure about this comment. the 18O comes from natural abundance, no? Again, does this come down to bad caption, because we don't emphasize enough that we use the LOG transform of the intensities? Or do we need an inset where we show the raw intensities ....
:::

3. Figure 13 is hard to understand and could be improved.

::: {custom-style="Response"}
UpSet plots are definitely hard to parse the first time they are encountered. We need to explain this better so that they see how it matches to a Venn diagram.
:::

## Reviewer 2

### OVERVIEW AND GENERAL RECOMMENDATION

The manuscript ‘Scan-Centric, Frequency-Based Method for Characterizing Peaks from Direct Injection Fourier transform Mass Spectrometry Experiments’ is on a topic of analytical relevance as complete analyses/characterization of complex samples makes complimentary of instrumentation inevitable. I must admit that I found the paper to be scientifically/analytically sound and well
described. Please consider the following comments to polish the draft to a publishable standard.

### SPECIFIC COMMENTS

#### ABSTRACT

* L13: introduce >> report.

> Abstract: We introduce a novel, scan-centric method for characterizing peaks from direct injection

::: {custom-style="Response"}
I'm not sure I agree with this.
:::

* L21-22: This line needs to be rephrased; if it is suitable as stated, the authors should have applied the same method and reported the results in this submission.

> These cross-scan characterized peaks are suitable for use in our previously published peak assignment methodology, Small Molecule Isotope Resolved Formula Enumeration (SMIRFE).

::: {custom-style="Response"}
I guess we need to be clear that we've been developing this for a while ....
:::

#### KEYWORDS

A single keyword may reduce the article’s visibility in search engines. I suggest adding scan-centric peak characterization, molecular fingerprint, multiplex or Fellgett’s advantage, orbitrap and interferograms as author-suggested indexing keywords.

::: {custom-style="Response"}
We will definitely add some of these, and some others we can think of
:::

#### INTRODUCTION

The section is well written but I am putting forth a suggestion as introductions are meant to start from a bigger picture before narrowing to the study context. Consider including a brief introduction on analytical techniques/instruments and the fact that a single technique may not achieve full characterization or analyses of (complex) samples. Thus, complementarity of techniques is becoming a powerful approach in the analyses of samples.

::: {custom-style="Response"}
I think we didn't go "big picture" enough here? And I think they would like us to mention DI, LC, GC, NMR??
:::

#### RESULTS

* L75: I do not find ‘‘Naïvely’’ scientific enough to be used in this line.

::: {custom-style="Response"}
For reference, this refers to the topical heading of 2.1

> 2.1. Naïvely Averaged Data Have Bad Relative Intensities

I guess we look at synonyms ...
:::

* L190, 195: better to use (a) and (b) instead of i and ii, respectively.

::: {custom-style="Response"}
This should be simple to change, and part of the issue probably was that I didn't use brackets. We'll just switch to (a) and (b).
:::

* In Figure 10, rsd should not be used interchangeably with RSD, as only the latter was defined before.

::: {custom-style="Response"}
Yep, definitely. This was inconsistent on my part.
:::

* For consistency, supplementary figures (Figure S1 to S5) should be cited consecutively. In the present draft, Figures S3 and S5 were cited before Figures S2 and S4.

::: {custom-style="Response"}
Definitely. Thought we had fixed this, but I guess we didn't get it all right.
:::

## Reviewer 3

The text suggests an important statistical approach to improve the data. More chemometric analyzes would be needed to improve data processing. good job

### 2022-05-12 doc/peakcharacterization_manuscript.Rmd
switching to (a) and (b) in description of issues with normalization. close #45

@@
 Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
-However, there are two issues with these peak intensities across scans: (i) some peak intensities correlate with the scan number (i.e.
-scan acquisition order); and (ii) some peak differences between scans are correlated with intensity.
-The solution to **i** is to do a two-pass normalization.
+However, there are two issues with these peak intensities across scans: (a) some peak intensities correlate with the scan number (i.e.
+scan acquisition order); and (b) some peak differences between scans are correlated with intensity.
+The solution to (a) is to do a two-pass normalization.
 After the first pass, the peaks whose intensity is correlated with scan order are detected (absolute Pearson correlation > 0.5).
@@
 `r figure_count$label_text("intensity_scan")`A shows an example peak whose intensity across scans is correlated with scan number.
-The solution to **ii** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
+The solution to (b) (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
 The highlighted peaks in `r figure_count$label_text("intensity_scan")`B are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.

### 2022-05-13 doc/peakcharacterization_manuscript.Rmd
added section numbers so we match what is in the MDPI version

@@
 ::: {custom-style="MDPI_2.1_heading1"}
-## Introduction
+## 1. Introduction
 :::
@@
 ::: {custom-style="MDPI_2.1_heading1"}
-## Results
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Naïvely Averaged Data Have Bad Relative Intensities
+## 2. Results
+:::
+
+::: {custom-style="MDPI_2.2_heading2"}
+### 2.1 Naïvely Averaged Data Have Bad Relative Intensities
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### m/z to Frequency
+### 2.2 m/z to Frequency
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Sliding Window Density to Remove Noise
+### 2.3 Sliding Window Density to Remove Noise
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Peak Characterization Using Quadratic Fit
+### 2.4 Peak Characterization Using Quadratic Fit
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Breaking Up Initial Regions
+### 2.5 Breaking Up Initial Regions
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Normalization of Scans
+### 2.6 Normalization of Scans
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Mitigation of High Peak Density Artifacts
+### 2.7 Mitigation of High Peak Density Artifacts
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Changes in Relative Standard Deviation (RSD)
+### 2.8 Changes in Relative Standard Deviation (RSD)
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Difference to Relative Natural Abundance
+### 2.9 Difference to Relative Natural Abundance
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Method Specific Peaks
+### 2.10 Method Specific Peaks
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Changes in P-Values On a Large Dataset
+### 2.11 Changes in P-Values On a Large Dataset
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Quality Control and Quality Analysis
+### 2.12 Quality Control and Quality Analysis
 :::
@@
 ::: {custom-style="MDPI_2.1_heading1"}
-## Discussion
+## 3. Discussion
 :::
@@
 ::: {custom-style="MDPI_2.1_heading1"}
-## Materials and Methods
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### Samples and Overall Processing
+## 4. Materials and Methods
+:::
+
+::: {custom-style="MDPI_2.2_heading2"}
+### 4.1 Samples and Overall Processing
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Matching Peaks
+### 4.2 Matching Peaks
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Conversion of m/z to Frequency
+### 4.3 Conversion of m/z to Frequency
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Frequency Intervals
+### 4.4 Frequency Intervals
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Interval Range Based Data
+### 4.5 Interval Range Based Data
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Peak Containing Intervals
+### 4.6 Peak Containing Intervals
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Peak Detection and Centroided Values
+### 4.7 Peak Detection and Centroided Values
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Scan to Scan Normalization
+### 4.8 Scan to Scan Normalization
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Full Scan-Centric Characterization
+### 4.9 Full Scan-Centric Characterization
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Correction of Height and Standard Deviation
+### 4.10 Correction of Height and Standard Deviation
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Marking High Frequency Standard Deviation Peaks
+### 4.11 Marking High Frequency Standard Deviation Peaks
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Calculation of Relative Standard Deviation
+### 4.12 Calculation of Relative Standard Deviation
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Scan-Centric Peak Assignment
+### 4.13 Scan-Centric Peak Assignment
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Consistently Assigned Lipid Spectral Feature (Corresponded Peak) Generation and Peak Intensity Normalization
+### 4.14 Consistently Assigned Lipid Spectral Feature (Corresponded Peak) Generation and Peak Intensity Normalization
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Peak - Peak NAP Height Ratios
+### 4.15 Peak - Peak NAP Height Ratios
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Differential Analysis of Large Dataset
+### 4.16 Differential Analysis of Large Dataset
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### Software Used
+### 4.17 Software Used
 :::

### 2022-05-13 lungcancer_all/lungcancer_smirfe_assignmentprocessing.R
updating path of where to save the JSON output so it moves around properly

@@ -98,4 +98,4 @@ json_data = purrr::map(all_json_files, function(in_json){
   tmp_json = jsonlite::fromJSON(in_json)
 })
 names(json_data) = gsub(".json$", "", basename(all_json_files))
-saveRDS(json_data, file = "data/data_output/lung_json_data_2022-04-02.rds")
+saveRDS(json_data, file = "data/data_output/lung_data/lung_json_data_2022-04-02.rds")

### 2022-05-13 doc/peakcharacterization_supplemental.Rmd
included frequency models
close #31

@@
 $$frequency = a + x \times \frac{1}{mz} + y \times \frac{1}{\sqrt{mz}}$$
+
+## Alternative Frequency Models
+
+We can define several models, and test them on the four samples.
+
+```{r increment_models}
+figure_count$increment("frequency_models")
+table_count$increment("frequency_models")
+```
+
+$$frequency = a + \frac{y}{\sqrt{mz}} (1)$$
+
+$$frequency = a + \frac{x}{mz} + \frac{y}{\sqrt{mz}} (2)$$
+
+$$frequency = a + \frac{x}{mz} + \frac{y}{\sqrt{mz}} + \frac{z}{\sqrt[3]{mz}} (3)$$
+
+$$frequency = a + \frac{y}{\sqrt{mz}} + \frac{z}{\sqrt[3]{mz}} (4)$$
+
+`r figure_count$label_text("frequency_models")` shows the median absolute deviation (MAD) and medians of the residuals between the calculated and predicted frequency values after fitting of frequency to m/z in each scan using each model, and then across each sample.
+It is clear that model 3 that includes all three of the m/z terms performs the best in terms of overall fit, although there may be some disagreement between the ECF and lipid samples based on median of the residual differences.
+`r table_count$label_text("frequency_models")` shows the coefficients for each term of the models for the 1ecf sample.
+The coefficients of the **y** term in each of the models is orders of magnitude larger than the coefficients for any of the other terms in the other models.
+That term also appears relatively constant, with some small variations to account for the presence and absence of the other terms of the model.
+
+```{r load_model_results}
+tar_load(frequency_1ecf)
+tar_load(frequency_2ecf)
+tar_load(frequency_49lipid)
+tar_load(frequency_97lipid)
+
+all_models = list(
+  `1ecf` = frequency_1ecf,
+  `2ecf` = frequency_2ecf,
+  `49lipid` = frequency_49lipid,
+  `97lipid` = frequency_97lipid
+)
+
+```
+
+```{r frequency_models, dn_id = figure_count, fig.width = 8, fig.height = 9}
+create_plot = function(in_sample){
+  mad_plot = in_sample$mad_data %>%
+    ggplot(aes(x = model, y = mad)) +
+    geom_sina() +
+    labs(x = "Frequency Model", y = "MAD(residuals)",
+         subtitle = in_sample$sample)
+  median_plot = in_sample$mad_data %>%
+    ggplot(aes(x = model, y = median)) +
+    geom_sina() +
+    labs(x = "Frequency Model", y = "Median(residuals)")
+  mad_plot | median_plot
+}
+sample_plots = purrr::map(all_models, create_plot)
+
+wrapped_plots = wrap_plots(sample_plots, ncol = 1)
+wrapped_plots
+```
+
+`r figure_count$label_text("frequency_models")`.
+Sina plots of the median absolute deviation (MAD) and median of the residuals between the calculated frequency using the m/z spacing and the predicted frequency after fitting a frequency model, for each of the four samples and for each of the models (1 - 4) defined above.
+
+`r table_count$label_text("frequency_models")`.
+Coefficients for each term in the models after fitting of the calculated frequency to m/z.
+
+```{r coefficient_table}
+ecf_coef_table = all_models[[1]]$model_data %>%
+  flextable() %>%
+  colformat_double(digits = 0) %>%
+  autofit()
+ecf_coef_table
+```
 

### 2022-05-14 doc/peakcharacterization_manuscript.Rmd
all the frequency crap

@@
 To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur as well as our scan-centric peak characterization, and found matching Xcalibur peaks to our assignments.
-As an example, in `r figure_count$label_text("motivation")`, we show the Xcalibur calculated intensities and theoretical peak intensities based on NAP for four peaks matching threonine.
-As the NAP gets lower, the deviations of the Xcalibur calculated intensities from the NAP become larger.
+As an example, in `r figure_count$label_text("motivation")`, we show the Xcalibur calculated intensities and theoretical peak intensities based on NAP for four peaks assigned to threonine.
+As the NAP decreases, the deviations of the Xcalibur calculated intensities from the NAP become larger.
 :::
@@
 **`r figure_count$label_text("motivation")`.**
-Xcalibur intensities (black) and theoretical intensities based on relative NAP (red) for ECF derived threonine.
+**A** Xcalibur intensities (black) and theoretical intensities based on relative NAP (red) for ECF derived threonine.
 The 18O isotopologue is shifted by 0.01 m/z and the NAP peaks by 0.05 m/z to aid visualization.
+**B** Theoretical NAP-based-Log10-intensities and Log10 XCalibur intensities.
+The red line indicates perfect agreement.
 :::

### 2022-05-14 R/frequency_models.R

@@ -1,0 +1,110 @@
+model_rtime_filter = function(scan_times, min_time_difference = 4, rtime_limit = 7.5*60){
+
+  low_rtime = scan_times$rtime <= rtime_limit
+
+  high_lag <- scan_times$rtime_lag >= min_time_difference
+  high_lag[is.na(high_lag)] <- TRUE
+  high_lead <- scan_times$rtime_lead >= min_time_difference
+  high_lead[is.na(high_lead)] <- TRUE
+
+  na_lead_high_lag <- is.na(scan_times$rtime_lead) & high_lag
+  na_lag_high_lead <- is.na(scan_times$rtime_lag) & high_lead
+
+  keep_scans <- low_rtime & (na_lead_high_lag | high_lag) & (na_lag_high_lead | high_lead)
+  keep_scans
+}
+
+model_time_filter = function(scan_times, min_time_difference = 4){
+
+  high_lag <- scan_times$rtime_lag >= min_time_difference
+  high_lag[is.na(high_lag)] <- TRUE
+  high_lead <- scan_times$rtime_lead >= min_time_difference
+  high_lead[is.na(high_lead)] <- TRUE
+
+  na_lead_high_lag <- is.na(scan_times$rtime_lead) & high_lag
+  na_lag_high_lead <- is.na(scan_times$rtime_lag) & high_lead
+
+  keep_scans <- (na_lead_high_lag | high_lag) & (na_lag_high_lead | high_lead)
+  keep_scans
+}
+
+
+run_frequency_models = function(mzml_file){
+  mod_1 = c("a.freq" = 0, "y.freq" = -1/2)
+  mod_2 = c("a.freq" = 0, "x.freq" = -1, "y.freq" = -1/2)
+  mod_3 = c("a.freq" = 0, "x.freq" = -1, "y.freq" = -1/2, "z.freq" = -1/3)
+  mod_4 = c("a.freq" = 0, "y.freq" = -1/2, "z.freq" = -1/3)
+
+  list_models = list(mod1 = mod_1,
+                     mod2 = mod_2,
+                     mod3 = mod_3,
+                     mod4 = mod_4)
+
+  if (grepl("pos|lipid", mzml_file)) {
+    use_filter = model_rtime_filter
+  } else {
+    use_filter = model_time_filter
+  }
+
+
+  sc_raw = ScanCentricPeakCharacterization::SCRaw$new(mzml_file)
+  sc_raw$extract_raw_data()
+
+  sc_raw$frequency_fit_description = mod_3
[…]

### 2022-05-16 doc/peakcharacterization_manuscript.Rmd
added histogram to frequency conversion
close #33

@@
 The subsequent differences in frequency should be relatively constant with respect to m/z, in contrast to the differences of adjacent m/z points, as shown in `r figure_count$label_text("mz_frequency_conversion")`B and C.
-The Thermo-Fisher Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 450K or 500K depending on the sample, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C and D.
-Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`E.
+The Thermo-Fisher Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 450K or 500K depending on the sample, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C, D and E.
+It is important to note that although the frequency difference has been consistent regardless of instrument resolution, we do not know a priori what it should be, and the method makes no assumptions about the frequency difference, but rather calculates the mode of the frequency differences from those observed in the data itself.
+Restricting to those points that fall into a narrow range around the mode of frequency differences (0.49 - 0.51 in this work), a model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`F.
 This model seems to fit the known relationship between frequency and m/z, where the frequency is related to $1/\sqrt{mz}$.

### 2022-05-16 R/graph_generation.R

@@ -49,6 +49,7 @@ plot_nap_peaks = function(raw_data, processed_data, assignments, interesting_reg
 }
 
 plot_frequency_conversion = function(char_obj){
+  # char_obj = tar_read(data_97lipid)
   raw_points = as.data.frame(char_obj$zip_ms$peak_finder$peak_regions$frequency_point_regions$frequency[[1]]@elementMetadata)
   is_convertable = dplyr::between(raw_points$mean_freq_diff, 0.49, 0.51)
   is_convertable[is.na(is_convertable)] = FALSE
@@ -95,8 +96,17 @@ plot_frequency_conversion = function(char_obj){
   point_difference_plot = ggplot(example_data, aes(x = pair_frequency, y = pair_frequency_offset)) + geom_point() +
     labs(x = "Frequency", y = "Frequency Differences")
 
-  all_frequency_difference_plot = ggplot(raw_points, aes(x = mz, y = log10(abs(mean_freq_diff)))) + geom_point() +
-    geom_point(data = dplyr::filter(raw_points, org_convertable), color = "red") + labs(x = "M/Z", y = "Log10(Frequency Differences)")
+  all_frequency_difference_hist = ggplot(raw_points, aes(x = log10(abs(mean_freq_diff)))) +
+    geom_histogram(bins = 100) +
+    geom_vline(xintercept = log10(0.5), color = "red") +
+    scale_y_continuous(expand = expansion(mult = c(0, .01))) +
+    labs(x = "Log10(Frequency Differences)",
+         y = "Number of Points")
+
+  y_title_size = 12
+  all_frequency_difference_plot = ggplot(raw_points, aes(x = mz, y = log10(abs(mean_freq_diff)))) + geom_point(alpha = 0.5) +
+    geom_point(data = dplyr::filter(raw_points, org_convertable), color = "red") + labs(x = "M/Z", y = "Log10(Frequency Differences)") +
+    theme(axis.title.y = element_text(size = y_title_size))
 
   convertable_raw2 = dplyr::filter(raw_points, org_convertable)
   raw2_model = FTMS.peakCharacterization:::fit_exponentials(convertable_raw2$mean_mz, convertable_raw2$mean_frequency, c(0, -1/2, -1/3))
@@ -123,7 +133,7 @@ plot_frequency_conversion = function(char_obj){
     dplyr::mutate(residuals = (predicted - frequency) / frequency)
 
[…]

### 2022-05-16 _targets.R
lung results with new frequency model

@@ -334,17 +334,17 @@ lungcancer_tar = list(
                patient_file
              )),
   tar_target(scancentric_json,
-             "data/data_output/lung_data/lung_json_data_2022-04-02.rds",
+             "data/data_output/lung_data/lung_json_data_2022-05-11.rds",
              format = "file"),
   tar_target(scancentric_med_file_corrected,
-             "data/data_output/lung_data/corrected_medians_2022-04-02.rds",
+             "data/data_output/lung_data/corrected_medians_2022-05-11.rds",
              format = "file"),
   tar_target(scancentric_medians_corrected,
              readRDS(scancentric_med_file_corrected)),
   tar_target(scancentric_medians,
              get_scancentric_medians(scancentric_json)),
   tar_target(emf_file,
-             "data/data_output/lung_data/lung_voted_all_2022-04-02.rds",
+             "data/data_output/lung_data/lung_voted_all_2022-05-11.rds",
              format = "file"),
   tar_target(scancentric_imfs_raw,
              extract_scancentric_imfs(

### 2022-05-16 lungcancer_all/lungcancer_smirfe_assignmentprocessing.R

@@ -1,6 +1,6 @@
 # cd smirfe_related/smirfe_code/SMIRFE_assigner
 # python3 -m pipenv shell
-# python3 -m AssignMany.py 12 "../../smirfe_dbs/none_nosulfur_1600.db FILE '_assigned.json' '[]' '[\"H\", \"Na\", \"K\", \"NH4\"]' '[1]'" /mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-04-02/*.zip
+# python3 -m AssignMany.py 4 "../../smirfe_dbs/none_nosulfur_1600.db FILE '_assigned.json' '[]' '[\"H\", \"Na\", \"K\", \"NH4\"]' '[1]'" /mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-05-11/*.zip
 # Database limits: C: 130, N: 7, O: 28, S: 0, P: 3, H: 230, max M/Z 1605, adducts: K, Na, NH4, NAP cutoff: 0.4
 library(smirfeTools)
 library(ggplot2)
@@ -14,7 +14,7 @@ plan(multicore)
 set_internal_map(furrr::future_map)
 options(future.globals.maxSize = 800 * 1024 ^ 2)
 
-assigned_files <- dir("/mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-04-02",
+assigned_files <- dir("/mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-05-11",
                       full.names = TRUE, pattern = "assigned.json")
 pb <- progress_estimated(length(assigned_files))
 
@@ -29,11 +29,11 @@ names(assigned_data) = purrr::map_chr(assigned_data, ~ .x$sample)
 
 all_emfs = unique(unlist(purrr::map(assigned_data, ~ .x$assignments$isotopologue_EMF)))
 
-saveRDS(assigned_data, "data/data_output/lung_data/lung_matched_tissue_raw_smirfe_assignments_2022-04-02.rds")
+saveRDS(assigned_data, "data/data_output/lung_data/lung_matched_tissue_raw_smirfe_assignments_2022-05-11.rds")
 
-cat(all_emfs, sep = "\n", file = "data/data_output/lung_data/all_emfs_2022-04-02.txt")
+cat(all_emfs, sep = "\n", file = "data/data_output/lung_data/all_emfs_2022-05-11.txt")
 
[…]

### 2022-05-16 lungcancer_all/lungcancer_smirfe_correctedmedians.R

@@ -5,7 +5,7 @@
 library(FTMS.peakCharacterization)
 library(knitrProgressBar)
 
-zip_files <- dir("/mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-04-02",
+zip_files <- dir("/mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-05-11",
                       full.names = TRUE, pattern = ".zip$")
 pb <- progress_estimated(length(zip_files))
 
@@ -20,4 +20,4 @@ corrected_medians <- purrr::map_df(zip_files, function(in_file){
   out_df
 })
 
-saveRDS(corrected_medians, file = "data/data_output/lung_data/corrected_medians_2022-04-02.rds")
+saveRDS(corrected_medians, file = "data/data_output/lung_data/corrected_medians_2022-05-11.rds")

### 2022-05-16 doc/peakcharacterization_manuscript.Rmd
incorporating all the changes related to frequency
and we've copied over the changes to our master document too

@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.1 Naïvely Averaged Data Have Bad Relative Intensities
+### 2.1 Simplistically Averaged Data Have Bad Relative Intensities
 :::
@@
 To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur as well as our scan-centric peak characterization, and found matching Xcalibur peaks to our assignments.
-As an example, in `r figure_count$label_text("motivation")`, we show the Xcalibur calculated intensities and theoretical peak intensities based on NAP for four peaks assigned to threonine.
-As the NAP decreases, the deviations of the Xcalibur calculated intensities from the NAP become larger.
+As an example, in `r figure_count$label_text("motivation")`, we show the Xcalibur calculated intensities from the simplistically averaged spectrum and theoretical peak intensities based on NAP for four peaks assigned to threonine.
+As the peak intensity decreases, the deviation between the Xcalibur calculated intensity and its corresponding NAP-based theoretical intensity become larger.
+This becomes more apparent when plotting the two sets of intensities directly against each other, as shown in `r figure_count$label_text("motivation")`B.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-FT-MS data from the Thermo-Fisher Orbitrap instruments used to acquire the data does not provide any information about the raw frequency data.
+FT-MS data from the Thermo-Fisher Orbitrap instruments used to acquire the data does not provide any information about the raw spectral frequency data.
 Outside of the meta-data, it merely contains the m/z and intensity values for profile spectra acquired across multiple scans.
-However, the frequency can be calculated by dividing the average m/z of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points representing the average m/z of two adjacent points, red lines representing the difference between the two adjacent points).
-The subsequent differences in frequency should be relatively constant with respect to m/z, in contrast to the differences of adjacent m/z points, as shown in `r figure_count$label_text("mz_frequency_conversion")`B and C.
-The Thermo-Fisher Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 450K or 500K depending on the sample, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C, D and E.
+However, the spectral frequency can be calculated by dividing the midpoint m/z of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points representing the midpoint m/z of two adjacent points, length of red lines representing the difference between the two adjacent points).
+The subsequent adjacent point differences in frequency are expected to be relatively constant with respect to m/z, as shown in `r figure_count$label_text("mz_frequency_conversion")`C, D and E, in contrast to the adjacent point differences in m/z, which are not constant with respect to m/z.
+The Thermo-Fisher Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 450K or 500K depending on the sample, has an adjacent frequency point difference mode of 0.5, as shown in  `r figure_count$label_text("mz_frequency_conversion")`D.
 It is important to note that although the frequency difference has been consistent regardless of instrument resolution, we do not know a priori what it should be, and the method makes no assumptions about the frequency difference, but rather calculates the mode of the frequency differences from those observed in the data itself.
-Restricting to those points that fall into a narrow range around the mode of frequency differences (0.49 - 0.51 in this work), a model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`F.
-This model seems to fit the known relationship between frequency and m/z, where the frequency is related to $1/\sqrt{mz}$.
-This is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points.
+Restricting to those points that fall into a narrow range around the mode of frequency differences (0.49 - 0.51 in this work), a regression model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`F.
+This regression model seems to fit the known relationship between frequency and m/z, where the frequency is related to $1/\sqrt{mz}$.
+The actual regression model used in this work includes terms of the m/z, $\sqrt{m/z}$, and $\sqrt[3]{m/z}$ (see **Conversion of m/z to Frequency** in Methods for further description of the regression model).
+We did investigate a variety of frequency regression models to see which one seemed to be the best, as shown in Supplemental `r supp_figure_count$label_text("frequency_models")` and `r supp_table_count$label_text("frequency_models")`.
+
+The constant point difference of frequency space is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points.
 The m/z point-to-point differences are not constant, but can be approximated by a LOESS linear model [@Cleveland1992loess]; however, it is very difficult to create a LOESS model with an intercept of 0.
@@
 **A**: Intensity vs m/z for a single peak from a single scan.
-Red lines denote the differences between each point, and red dots the average between the pair of points.
-The difference over the point is used to derive the frequency values in **B**.
+Red lines denote the difference between each adjacent pair of points, and red dots the m/z midpoint between the pair of points.
+The difference divided by the midpoint is used to derive the frequency values in **B**.
 **B** plots the intensity vs the converted frequency points derived from **A**.
-The red lines denote pairwise differences, which are shown in **C** for this single peak.
-The differences for all peaks across all scans vs m/z are shown in **D**, with those differences that lie within 0.49 - 0.51 shown in red.
-**E** shows the plot of derived frequency vs m/z, with fitted values from the linear model in red.
+The red lines denote the differenc between adjacent points, which are shown in the y-axis of **C** for this single peak.
+A histogram of adjacent point differences for all points is shown in **D**.
+The differences for all points in a single scan vs m/z are shown in **E**, with those differences that lie within 0.49 - 0.51 shown in red.
+**F** shows the plot of derived spectral frequency vs m/z, with fitted values from the linear regression model in red.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-The m/z to frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
+The m/z to frequency regression models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
 While scan specific models **could** be used in the conversion of m/z to frequency, doing so results in changes to the relative peak ordering compared to m/z space, as shown in `r figure_count$label_text("peak_ordering")`.
@@
 **`r figure_count$label_text("peak_ordering")`.** Peak ordering in m/z compared with ordering in frequency space when a single m/z to frequency model is used or scan specific m/z to frequency models are used.
-For a single peak, the scan level peak m/z's were extracted, and then frequency values for those m/z generated using a single common model of m/z to frequency (*single_order*), or models derived from each scan (*scan_order*).
-A subset of the peaks end up out of order using scan specific models, implying that a single model should be used across all the scan level data.
+For a single peak, the scan level peak m/z's were extracted, and then frequency values for those m/z generated using a single common model of m/z to frequency (*Single Model*), or models derived from each scan (*Individual Models*).
+A subset of the peaks ends up out of order using scan specific models, implying that a single model should be used across all the scan level data.
 :::
@@
 ::: {custom-style="MDPI_3.9_equation"}
-`r format_equation('frequency = a + y \\times \\frac{1}{\\sqrt{mz}} + z \\times \\frac{1}{\\sqrt[3]{mz}}', 'mzfrequency')`
+`r format_equation('frequency = a + \\frac{x}{mz} + \\frac{y}{\\sqrt{mz}} + \\frac{z}{\\sqrt[3]{mz}}', 'mzfrequency')`
 :::
@@
 From the known physical properties of the Orbitrap, only the square root term should be necessary [@ledfordSpaceChargeEffects1984].
-Practically, we have found the combination of square and cube-roots to provide a better fit when processing data from the Thermo-Fisher Fusion instrument, likely due to issues with slight imperfections in the orbitrap geometry, contributions from space charge effects and magnetronic motion, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
+Practically, we have found the combination of no root, square and cube-roots to provide a better fit when processing data from the Thermo-Fisher Fusion instrument, likely due to issues with slight imperfections in the orbitrap geometry, contributions from space charge effects and magnetronic motion, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
 We do note that when working with Bruker SolariX ICR data, the equation is slightly different, and does not require the cubic square root term (unpublished results, see Supplemental Materials).

### 2022-05-18 doc/peakcharacterization_manuscript.Rmd
some more changes
close #50
close #42
close #41
close #40
close #37
close #36
close #34

@@
 ::: {custom-style="MDPI_1.7_abstract"}
-**Abstract:** We introduce a novel, scan-centric method for characterizing peaks from direct injection multi-scan Fourier transform mass spectra of complex samples that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
+**Abstract:** We present a novel, scan-centric method for characterizing peaks from direct injection multi-scan Fourier transform mass spectra of complex samples that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
 Our peak characterization method utilizes intensity independent noise removal and normalization of scan-level data to provide a much better fit of relative intensity to natural abundance probabilities for low abundance isotopologues that are not present in all of the acquired scans.
@@
 While these micro-scan and subsequent scans are expected to be analytical replicates, spray instability and temporal delay in automatic gain control pragmatically break this expectation.
-Therefore, it is advantageous to keep the data at the scan level, so that any given poor quality (i.e.
-"bad") scan can be removed prior to aggregating the data across scans and generating a centroided peak m/z and intensity.
+Therefore, it is advantageous to keep the data at the scan level, so that any given poor quality (i.e., "bad") scan can be removed prior to aggregating the data across scans and generating a centroided peak m/z and intensity.
 If the scan level data is not processed and aggregated correctly, the final peak intensities in arbitrary units and centers in m/z will have high uncertainty (total variance) [@moseleyErrorAnalysisPropagation2013].
@@
 The 18O isotopologue is shifted by 0.01 m/z and the NAP peaks by 0.05 m/z to aid visualization.
-**B** Theoretical NAP-based-Log10-intensities and Log10 XCalibur intensities.
+**B** Theoretical NAP-based-Log10-intensities and Log10 Xcalibur intensities.
 The red line indicates perfect agreement.
@@
 The constant point difference of frequency space is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points.
-The m/z point-to-point differences are not constant, but can be approximated by a LOESS linear model [@Cleveland1992loess]; however, it is very difficult to create a LOESS model with an intercept of 0.
+The m/z point-to-point differences are not constant, but can be approximated by a LOESS linear model [@Cleveland1992loess]; however, it is exceedingly difficult to create a LOESS model with an intercept of 0.
 In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the LOESS model.
@@
 Red lines denote the difference between each adjacent pair of points, and red dots the m/z midpoint between the pair of points.
-The difference divided by the midpoint is used to derive the frequency values in **B**.
+The difference divided by the midpoint is used to derive the spectral frequency values in **B**.
 **B** plots the intensity vs the converted frequency points derived from **A**.
-The red lines denote the differenc between adjacent points, which are shown in the y-axis of **C** for this single peak.
+The red lines denote the difference between adjacent points, which are shown in the y-axis of **C** for this single peak.
 A histogram of adjacent point differences for all points is shown in **D**.
@@
 figure_count$increment("slidingwindow_count")
+figure_count$increment("percentile_variation")
 tar_load(mn_ratios)
 tar_load(sliding_regions)
+tar_load(example_sliding_region)
 tar_load(noise_plot)
@@
 ::: {custom-style="MDPI_3.1_text"}
-In a dataset of this nature, we expect that much of the data is really just noise, and doesn't contribute that much to the analysis.
+In a dataset of this nature, we expect that much of the data is really just noise, and does not contribute meaningfully to the analysis.
 Furthermore, it is expected that noise is randomly distributed across the scans.
 Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
-Subsequently, we divide the counts into tiled regions of a set size (2000 in this work), and examine the 99th percentile of the sliding window counts within the tiled region.
-A histogram of sliding window counts in a single tiled region is shown in `r figure_count$label_text("slidingwindow_count")`A.
-The distribution of 99th percentiles across all of the tiled regions for a single sample is shown in `r figure_count$label_text("slidingwindow_count")`B.
-We can use the median (50th percentile) value of all the 99th percentiles from all of the tiled regions to determine a cutoff value above which any non-zero points are potentially **signal** and not just noise.
-In the example below, the median of the tiled windows is 5, resulting in a cutoff of 7.5.
-The number of peaks detected without applying a cutoff explodes anywhere from `r mn_ratios[["97lipid"]]$ratio`-fold (from `r format(mn_ratios[["97lipid"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["97lipid"]]$count_0, big.mark = ",")` in the 97lipid sample), to `r mn_ratios[["2ecf"]]$ratio`-fold (from `r format(mn_ratios[["2ecf"]]$count_99, big.mark = ",")` to `r format(mn_ratios[["2ecf"]]$count_0, big.mark = ",")` in the 2ecf sample) with a vast majority of the peaks being present in a single scan.
-By discarding those windows with a non-zero point density below 1.5x the median of 99th percentiles, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
-:::
-
-::: {custom-style="MDPI_5.2_figure"}
-
-```{r slidingwindow_count, dn_id = figure_count, fig.width = 8, fig.height = 8}
-(sliding_regions / noise_plot) + plot_annotation(tag_levels = "A")
+Subsequently, we divide the counts into tiled regions of a set size (1000 frequency in this work), and examine the 99th percentile of the sliding window counts within the tiled region.
+In `r figure_count$label_text("slidingwindow_count")`A we plotted the non-zero point intensities from all scans for one tiled region 2000 points wide.
+The number of non-zero points in the sliding windows across this region are shown in `r figure_count$label_text("slidingwindow_count")`B, and zoomed in in C.
+These provide a measure of the non-zero point density across this region.
+The blue line indicates the cumulative 99th percentile for this one region, and is high, given that it looks like there are some consistent peaks across the scans in this region.
+However, examining the cumulative 99th percentiles across all the regions of the spectrum, as shown in the histogram of `r figure_count$label_text("slidingwindow_count")`F, a rather large fraction of the regions have very low cumulative 99th percentile values, with a median value of 5, which results in a cutoff value of 8 (ceiling of 5 x 1.5).
+Keeping only those sliding regions with a non-zero point count greater than 8 (above the red line in `r figure_count$label_text("slidingwindow_count")`B and C) results in those sliding regions shown in D, and then the point intensities in E.
+A non-zero point density based method to removing potential noise is desirable because the peak intensities across scans are not identical, as shown by the need for normalization (see **Normalization of Scans**), and the distribution of RSD values before and after normalization (see **Changes in Relative Standard Deviation**).
+:::
+
+::: {custom-style="MDPI_5.2_figure"}
+
+```{r slidingwindow_count, dn_id = figure_count, fig.width = 12, fig.height = 8}
+(example_sliding_region | sliding_regions) + plot_annotation(tag_levels = "A")
 ```
@@
 **`r figure_count$label_text("slidingwindow_count")`.**
-**A**: Histogram of the number of non-zero points from sliding windows ten frequency points wide and a stride of one point, from a single tiled window 2000 points wide.
-Blue line is the 99th percentile.
-**B**: Histogram of the 99th percentile cutoffs from all of the tiled windows.
-Red line is the median x 1.5.
-**C** and **D** show the number of initial regions as a function of the percentile cutoff used for either the AA ECF (C) or lipid (D) samples.
+**A - E**: A single tiled region 2000 data points wide.
+**A**: Non-zero point intensities in frequency space across all scans.
+**B**: Number of non-zero points in sliding regions. Blue line indicates 99th percentile of non-zero counts in this one region.
+Red line is the rounded value of median x 1.5 of 99th percentiles from **F**.
+**C**: Zoomed y-axis of **B** to show the non-zero counts that are below the median cutoff.
+**D**: Sliding region non-zero counts that are greater than the cutoff.
+**E**: Non-zero point intensities across scans in the regions from D.
+**F**: Histogram of the 99th percentile cutoffs from all of the tiled windows. Red line is the median x 1.5.
+:::
+
+::: {custom-style="MDPI_3.1_text"}
+Although it appears that only a handful of potential signal regions are removed between `r figure_count$label_text("slidingwindow_count")`A and E with a cutoff of 8 non-zero points in a sliding region, the number of initial signal regions rapidly explodes as the cumulative percentile cutoff is lowered, going from thousands to tens of thousands of potential peak regions to find and fit peaks within, as shown in `r figure_count$label_text("percentile_variation")`A and B for both the amino-acid ECF and lipid samples, respectively.
+:::
+
+::: {custom-style="MDPI_5.2_figure"}
+
+```{r percentile_variation, dn_id = figure_count, fig.width = 10, fig.height = 6}
+noise_plot + plot_annotation(tag_levels = "A")
+```
+:::
+
+::: {custom-style="MDPI_5.1_figure_caption"}
+**`r figure_count$label_text("percentile_variation")`.**
+The number of initial regions as a function of the percentile cutoff used for either the AA ECF (**A**) or lipid (**B**) samples.
 :::
@@
 Although many other types of mass spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo-Fisher Fusion has a baseline of 0 due to manipulations in the Thermo-Fisher firmware, making the determination of the centroided values considerably easier.
-For each region initially created, the peaks in each scan within that region can be characterized (i.e. centroiding).
+For each region initially created, the peaks in each scan within that region can be characterized (i.e., centroiding).
 For centroiding, we use a simple weighted quadratic model of log(intensity) to m/z. 
@@
 **A**: The full set of raw frequency and intensity data across all scans for the region are shown.
+Each horizontal trace are the point intensities in a single scan.
 Clearly the region has two separate peaks within it.
@@
 Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
-However, there are two issues with these peak intensities across scans: (a) some peak intensities correlate with the scan number (i.e.
-scan acquisition order); and (b) some peak differences between scans are correlated with intensity.
+However, there are two issues with these peak intensities across scans: (a) some peak intensities correlate with the scan number (i.e., scan acquisition order); and (b) some peak differences between scans are correlated with intensity.
 The solution to (a) is to do a two-pass normalization.
@@
 If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and B.
-Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B and C), removing the scan-correlated peaks does change some of the normalization factors.
+Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B and C), which is expected as they both use the most intense peaks.
+But removing the scan-correlated peaks does change the majority of the normalization factors.
 :::
@@
 These peaks can be detected by simply examining the distribution of FSDs and marking outlier peaks.
-To verify the mitigation of HPD regions, we converted centroided m/z's from XCalibur to frequency values using the previously calculated model for that sample, and measured peak density to detect HPD regions, and compare them with the scan-centric peaks and FSD outliers.
-XCalibur peak density was measured using a sliding window ten points wide and a stride of one point.
+To verify the mitigation of HPD regions, we converted centroided m/z's from Xcalibur to frequency values using the previously calculated model for that sample, and measured peak density to detect HPD regions, and compare them with the scan-centric peaks and FSD outliers.
+Xcalibur peak density was measured using a sliding window ten points wide and a stride of one point.
 
@@
 For samples 49lipid and 97lipid, the improvements are visually more subtle, having a slight shift to the left as well as a narrowing and smoothing of the RSD distribution.
-Part of this sublety is likely due to the bi-modality of these two distributions.
+Part of this subtlety is likely due to the bi-modality of these two distributions.
 This is not surprising, since these two samples are non-polar extractions from tissue and are biochemically more complex.
@@
 At least part of the bi-modality in the RSD distribution appears to be related to this dependence of RSD on intensity, as well as truncation as peaks appear in fewer scans.
-This is supported by part of the distribution disappearing when we require that peaks are in at least 80% of the available scans (difference between `r figure_count$label_text("rsd_intensity")` top and bottom).
+This is supported by part of the distribution disappearing when we require that peaks be in at least 80% of the available scans (difference between `r figure_count$label_text("rsd_intensity")` top and bottom).
 
@@
 ::: {custom-style="MDPI_3.1_text"}
-Although we expected that the corrected scan-centric intensities would behave similarly or better compared to the raw intensities, `r supp_figure_count$label_text("nap_intensity")` shows that while they have smaller differences than XCalibur, there is still a trend of increasing difference with the raw scan-centric intensity - NAP ratios as the number of missing scans increases.
+Although we expected that the corrected scan-centric intensities would behave similarly or better compared to the raw intensities, `r supp_figure_count$label_text("nap_intensity")` shows that while they have smaller differences than Xcalibur, there is still a trend of increasing difference with the raw scan-centric intensity - NAP ratios as the number of missing scans increases.
 :::
@@
 **`r figure_count$label_text("unassigned_upset")`.**
-UpSet plot with the counts of common and specific peaks without consideration of assignments for each of scancentric, MSnbase and XCalibur generated peaks for the 1ecf (**A**) and 97lipid (**B**) samples.
+UpSet plot with the counts of common and specific peaks without consideration of assignments for each of scancentric, MSnbase and Xcalibur generated peaks for the 1ecf (**A**) and 97lipid (**B**) samples.
+The black points with connected vertical lines identify which set intersections are represented in the bar at the top.
+A single black point identifies what was specific to the set and not in any other sets.
 :::
@@
 plot_mz_diffs = function(in_peaks, limits, ylimit = NULL, inset = FALSE){
-  purrr::map(c("Scan-Centric", "XCalibur", "MSnbase"), function(in_method){
+  purrr::map(c("Scan-Centric", "Xcalibur", "MSnbase"), function(in_method){
     tmp_peaks = in_peaks %>%
@@
     }
-    if (inset && (in_method %in% c("XCalibur", "MSnbase"))) {
+    if (inset && (in_method %in% c("Xcalibur", "MSnbase"))) {
       inset_plot = peak_count %>%
@@
     source %in% "scanlevel" ~ "Scan-Centric",
-    source %in% "xcalibur" ~ "XCalibur",
+    source %in% "xcalibur" ~ "Xcalibur",
     source %in% "msnbase" ~ "MSnbase"
@@
     id.test %in% "corrected" ~ "Corrected",
-    id.test %in% "xcalibur" ~ "XCalibur",
+    id.test %in% "xcalibur" ~ "Xcalibur",
     id.test %in% "msnbase" ~ "MSnbase"
@@
 One way to test this is to compare p-values generated from the different intensities.
-`r figure_count$label_text("p_value_comparison")` compares the -1 * Log10(p-values) from `r nrow(compare_split$Corrected)` (corrected and XCalibur) or `r nrow(compare_split$MSnbase)` (MSnbase) isotope resolved molecular formulas (IMFs) assigned by SMIRFE for the scan-centric peaks and then peaks matched and intensities extracted from the other methods.
+`r figure_count$label_text("p_value_comparison")` compares the -1 * Log10(p-values) from `r nrow(compare_split$Corrected)` (corrected and Xcalibur) or `r nrow(compare_split$MSnbase)` (MSnbase) isotope resolved molecular formulas (IMFs) assigned by SMIRFE for the scan-centric peaks and then peaks matched and intensities extracted from the other methods.
 Although all of the p-vaues are somewhat (and statistically) different from those reported by the raw intensities, there are some interesting patterns of differences.
@@
 This echoes the patterns of low overlapping assigned peaks observed in both ECF samples, where many of the scan-centric amino-acid peaks only found one match from the MSnbase peaks.
-Surprisingly, the truncated log-normal distribution corrected intensities generated very different p-values compared to the raw intensities, and XCalibur showed the most agreement with the raw p-values.
+Surprisingly, the truncated log-normal distribution corrected intensities generated very different p-values compared to the raw intensities, and Xcalibur showed the most agreement with the raw p-values.
 We tested the statistical significance of these differences using a t-test of the actual difference in log-p-values (shown in `r figure_count$label_text("p_value_comparison")`D).
@@
     comparison %in% "corrected" ~ "Corrected",
-    comparison %in% "xcalibur" ~ "XCalibur",
+    comparison %in% "xcalibur" ~ "Xcalibur",
     comparison %in% "msnbase" ~ "MSnbase"
@@
 
-lung_plot = (compare_plots$Corrected | compare_plots$MSnbase) / (compare_plots$XCalibur | compare_sina) +
+lung_plot = (compare_plots$Corrected | compare_plots$MSnbase) / (compare_plots$Xcalibur | compare_sina) +
   plot_annotation(tag_levels = "A")
@@
 Also, this approach prevents the calculation of useful peak-specific statistics, especially variances, both in m/z and intensity.
-In the results presented, we demonstrate superior performance by scan-centric peak characterization in terms of: i) improved QC/QA that removes low quality scans; ii) intensity independent noise reduction that pragmatically eliminates HPD sites; iii) improved peak height relative standard deviations; iv) improved peak intensity ratios that better match natural abundance probabilities; and v) better separation between biological groups.
+In the results presented, we demonstrate superior performance by scan-centric peak characterization in terms of: (a) improved QC/QA that removes low quality scans; (b) intensity independent noise reduction that pragmatically eliminates HPD sites; (c) improved peak height relative standard deviations; (d) improved peak intensity ratios that better match natural abundance probabilities; and (e) better separation between biological groups.
 These improvements are necessary for downstream derivation of molecular formula by SMIRFE using spectrum-derived tolerances, an efficient m/z search of a large isotope-resolved cache, and filtering by peak ratio matching to NAP.
@@
 But once implemented, large datasets can be processed in a straight-forward manner that supports full computational reproducibility.
+We also kept all steps that provided a demonstrable improvement, no matter how small that improvement was, but the following steps provided the most improvement: (a) noise removal by non-zero point density, (b) scan normalization using median peak intensity differences, and (c) removal of outlier scans.
+However, both steps a and c require the m/z to frequency regression model. 
+In the end, it is the combination of steps that synergistically provide the presented superior peak characterization results. 
+We note that the non-zero point density based noise removal is particularly useful, as the peak intensities from scan to scan are not identical, which would possibly hamper any kind of intensity based noise removal.
 
@@
 ::: {custom-style="MDPI_3.8_bullet"}
-* Scans merged and peak-list exported by XCalibur
+* Scans merged and peak-list exported by Xcalibur
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-To associate the XCalibur and MSnbase peaks with the scan-centric peaks, a 4 ppm (2ppm low and high) window is calculated for the scan-centric centroid, and if any peaks are found within the window, the one with the smallest m/z difference is kept as the matching peak to the scan-centric one under consideration.
+To associate the Xcalibur and MSnbase peaks with the scan-centric peaks, a 4 ppm (2ppm low and high) window is calculated for the scan-centric centroid, and if any peaks are found within the window, the one with the smallest m/z difference is kept as the matching peak to the scan-centric one under consideration.
 :::
@@
 To find intervals that contain points that represent actual signal and not just random noise, the number of non-zero intensity points in each sliding window are counted.
-Subsequently, we break these counts into fixed width tiles (default width of 2000) and calculate the 99th percentile of non-zero points for each tile.
-The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions containing potential signal.
+Subsequently, we break these counts into fixed width tiles (default width of 1000 in frequency space) and calculate the 99th percentile of non-zero points for each tile.
+The rounded up median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions containing potential signal.
 Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are kept and overlapping sliding windows are merged to create the initial peak regions.
@@
 This should result in a left-censored log-normal distribution of peak intensities across equivalent scans, which is expected for analytical measurements with detection limits [@flightInformationContentInformedKendalltauCorrelation2022].
-To correct these, either a correction based on a model of the truncated normal distribution can be used on log-tansformed data, or the differences can be simulated by sampling from data that is present in most of the scans.
+To correct these, either a correction based on a model of the truncated normal distribution can be used on log-transformed data, or the differences can be simulated by sampling from data that is present in most of the scans.
 To simulate the effect of peaks missing from some scans on the standard deviations, the peaks present in all scans are used.
@@
 ::: {custom-style="MDPI_3.1_text"}
-Thermo-Fisher XCalibur was used to export peak lists from all samples used.
+Thermo-Fisher Xcalibur was used to export peak lists from all samples used.
 SMIRFE v 1.0 [@mitchellSmallMoleculeIsotope2019] running under Python 3.8 [@vanrossumPython3] was used for assignments.

### 2022-05-18 R/compare_noise_cutoff.R

@@ -40,6 +40,115 @@ compare_noise_cutoffs <- function(in_char) {
   n_region
 }
 
+plot_example_sliding_counts = function(peak_data){
+  #source("./packages.R")
+  #peak_data = tar_read(data_97lipid)
+  sliding_regions = peak_data$zip_ms$peak_finder$peak_regions$sliding_regions
+  point_region_list = peak_data$zip_ms$peak_finder$peak_regions$frequency_point_regions$frequency
+  region_multiplier = peak_data$zip_ms$peak_finder$peak_regions$frequency_multiplier
+  multiplier = peak_data$zip_ms$peak_finder$quantile_multiplier
+  n_point_region = peak_data$zip_ms$peak_finder$n_point_region
+
+  nz_counts = FTMS.peakCharacterization:::count_overlaps(sliding_regions, point_region_list[[1]])
+  n_region = seq(2, length(point_region_list))
+  for (iregion in n_region) {
+    nz_counts_iregion = FTMS.peakCharacterization:::count_overlaps(sliding_regions, point_region_list[[iregion]])
+    nz_counts = nz_counts + nz_counts_iregion
+  }
+
+  mcols(sliding_regions) = DataFrame(nz_counts = nz_counts, frequency = start(sliding_regions) / region_multiplier)
+
+  chunk_indices = seq(1, length(nz_counts), by = n_point_region)
+
+  chunk_perc = purrr::map_dbl(chunk_indices, function(in_index){
+    use_counts = nz_counts[seq(in_index, min(in_index + (n_point_region - 1), length(nz_counts)))]
+    if (max(use_counts) > 0) {
+      return(stats::quantile(use_counts, 0.99))
+    } else {
+      return(0)
+    }
+  })
+
+  chunk_cutoff = ceiling(median(chunk_perc) * multiplier)
+
+  freq_800 = FTMS.peakCharacterization:::predict_exponentials(800, peak_data$peak_finder$peak_regions$frequency_point_regions$metadata$frequency_coefficients, peak_data$peak_finder$peak_regions$frequency_point_regions$metadata$frequency_fit_description)[1]
+  iranges_800 = IRanges::IRanges(start = (freq_800 - 500) * 400, end = (freq_800 + 500) * 400)
+
+  point_list_nz = purrr::map(point_region_list, function(.x){
+    tmp_df = as.data.frame(mcols(.x))
[…]

### 2022-05-18 packages.R

@@ -26,6 +26,7 @@ library(documentNumbering)
 library(flextable)
 library(ggforce)
 library(visualizationQualityControl)
+library(IRanges)
 suppressPackageStartupMessages(library(MSnbase))
 suppressPackageStartupMessages(library(ComplexHeatmap))
 conflict_prefer("combine", "dplyr")

### 2022-05-18 doc/peakcharacterization_manuscript.Rmd
a few more edits in

@@
 ::: {custom-style="MDPI_1.8_keywords"}
-**Keywords**: Fourier-transform mass spectrometry
+**Keywords**: Fourier-transform mass spectrometry, orbitrap, scan-centric peak characterization, Fellgett's advantage, interferograms, frequency spectrum
 :::
@@
 To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur as well as our scan-centric peak characterization, and found matching Xcalibur peaks to our assignments.
-As an example, in `r figure_count$label_text("motivation")`, we show the Xcalibur calculated intensities from the simplistically averaged spectrum and theoretical peak intensities based on NAP for four peaks assigned to threonine.
-As the peak intensity decreases, the deviation between the Xcalibur calculated intensity and its corresponding NAP-based theoretical intensity become larger.
+As an example, in `r figure_count$label_text("motivation")`, we show the Xcalibur calculated intensities from the simplistically averaged spectrum and theoretical peak intensities based on NAP for four peaks matching threonine.
+As the peak intensity decreases, the deviation between the Xcalibur calculated intensity and its corresponding NAP-based theoretical intensity becomes larger.
 This becomes more apparent when plotting the two sets of intensities directly against each other, as shown in `r figure_count$label_text("motivation")`B.
@@
 ```{r centroided_peaks, dn_id = figure_count}
-# figure showing fit of quadratic model to actual data points
+# showing fit of quadratic model to actual data points
 tar_load(peak_fit_plot)
@@
 ```{r breaking_regions, dn_id = figure_count}
-# figure showing multiple "peaks" in what is initially a single region
+# showing multiple "peaks" in what is initially a single region
 tar_load(split_region_plot)
@@
 ```{r intensity_scan, dn_id = figure_count}
-# figure showing peak intensity correlation with m/z, intensity varying with m/z
+# showing peak intensity correlation with m/z, intensity varying with m/z
 tar_load(intensity_scan_cor_plot)
@@
 These peaks can be detected by simply examining the distribution of FSDs and marking outlier peaks.
-To verify the mitigation of HPD regions, we converted centroided m/z's from Xcalibur to frequency values using the previously calculated model for that sample, and measured peak density to detect HPD regions, and compare them with the scan-centric peaks and FSD outliers.
+To verify the mitigation of HPD regions, we converted centroided m/z's from Xcalibur to frequency values using the previously calculated regression model for that sample, measured peak density to detect HPD regions, and compared them with the scan-centric peaks and FSD outliers.
 Xcalibur peak density was measured using a sliding window ten points wide and a stride of one point.
@@
 
-To convert m/z back into frequency, we can use a similar model without the roots, as well as an extra simple linear term that doesn't have an equivalent in the above frequency model.
+To convert m/z back into frequency, we can use a similar model without the roots, as well as an extra simple linear term that does not have an equivalent in the above frequency model.
 :::

### 2022-05-18 doc/peakcharacterization_supplemental.Rmd
getting supplemental figures in the correct order
close #47

@@
 
-Here in `r figure_count$label_text("nap_intensity")` we show the differences in Height - NAP peaks using the corrected and raw scan-centric intensities so they can be compared with the differences we observed between XCalibur and scan-centric peaks.
+Here in `r figure_count$label_text("nap_intensity")` we show the differences in Height - NAP peaks using the corrected and raw scan-centric intensities so they can be compared with the differences we observed between Xcalibur and scan-centric peaks.
 
@@
 
-## ICI-Kt Median Correlation Outliers
-
-```{r increment_icikt}
-figure_count$increment("icikt_outlier")
-```
-
-For the NSCLC dataset, we used information-content-informed Kendall-tau (ICI-Kt) correlation within each of non-cancer and cancer sample groups to determine outliers that should be removed prior to statistical testing.
-In `r figure_count$label_text("icikt_outlier")` we show the distribution of median ICI-Kt values and the samples determined as outliers.
-
-```{r icikt_outlier, dn_id = figure_count}
-tar_load(qcqa)
-outliers = qcqa$outlier_data
-
-outliers %>%
-  ggplot(aes(x = sample_class, y = med_cor, color = outlier, group = sample_class)) +
-  geom_sina() +
-  theme(legend.position = c(0.1, 0.9)) +
-  labs(x = "Sample Class", y = "Median ICI-Kt")
-```
-
-`r figure_count$label_text("icikt_outlier")`.
-Sina plot of median ICI-Kt correlation for samples within each of the cancer and non-cancer sample groups with outlier status indicated for each sample.
-
 ## Incorrect Normalization in P-Value Differences
@@
 
+## ICI-Kt Median Correlation Outliers
+
+```{r increment_icikt}
+figure_count$increment("icikt_outlier")
+```
+
+For the NSCLC dataset, we used information-content-informed Kendall-tau (ICI-Kt) correlation within each of non-cancer and cancer sample groups to determine outliers that should be removed prior to statistical testing.
+In `r figure_count$label_text("icikt_outlier")` we show the distribution of median ICI-Kt values and the samples determined as outliers.
+
+```{r icikt_outlier, dn_id = figure_count}
+tar_load(qcqa)
+outliers = qcqa$outlier_data
+
+outliers %>%
+  ggplot(aes(x = sample_class, y = med_cor, color = outlier, group = sample_class)) +
+  geom_sina() +
+  theme(legend.position = c(0.1, 0.9)) +
+  labs(x = "Sample Class", y = "Median ICI-Kt")
+```
+
+`r figure_count$label_text("icikt_outlier")`.
+Sina plot of median ICI-Kt correlation for samples within each of the cancer and non-cancer sample groups with outlier status indicated for each sample.
+
 ```{r save_counts}

### 2022-05-19 doc/peakcharacterization_manuscript.Rmd
regenerate qcqa stuff

@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.1 Simplistically Averaged Data Have Bad Relative Intensities
+### 2.1. Simplistically Averaged Data Have Bad Relative Intensities
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.2 m/z to Frequency
+### 2.2. m/z to Frequency
 :::
@@
 Restricting to those points that fall into a narrow range around the mode of frequency differences (0.49 - 0.51 in this work), a regression model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`F.
-This regression model seems to fit the known relationship between frequency and m/z, where the frequency is related to $1/\sqrt{mz}$.
+This regression model seems to fit the known relationship between frequency and m/z, where the frequency is related to $1/\sqrt{m/z}$.
 The actual regression model used in this work includes terms of the m/z, $\sqrt{m/z}$, and $\sqrt[3]{m/z}$ (see **Conversion of m/z to Frequency** in Methods for further description of the regression model).
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.3 Sliding Window Density to Remove Noise
+### 2.3. Sliding Window Density to Remove Noise
 :::
@@
 **`r figure_count$label_text("slidingwindow_count")`.**
-**A - E**: A single tiled region 2000 data points wide.
+**A - E**: A single tiled region 1000 frequency wide.
 **A**: Non-zero point intensities in frequency space across all scans.
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.4 Peak Characterization Using Quadratic Fit
+### 2.4. Peak Characterization Using Quadratic Fit
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.5 Breaking Up Initial Regions
+### 2.5. Breaking Up Initial Regions
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.6 Normalization of Scans
+### 2.6. Normalization of Scans
 :::
@@
 However, there are two issues with these peak intensities across scans: (a) some peak intensities correlate with the scan number (i.e., scan acquisition order); and (b) some peak differences between scans are correlated with intensity.
-The solution to (a) is to do a two-pass normalization.
+The solution to **a** is to do a two-pass normalization.
 After the first pass, the peaks whose intensity is correlated with scan order are detected (absolute Pearson correlation > 0.5).
@@
 `r figure_count$label_text("intensity_scan")`A shows an example peak whose intensity across scans is correlated with scan number.
-The solution to (b) (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
+The solution to **b** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
 The highlighted peaks in `r figure_count$label_text("intensity_scan")`B are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
 If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and B.
-Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B and C), which is expected as they both use the most intense peaks.
+Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B), which is expected as they both use the most intense peaks.
 But removing the scan-correlated peaks does change the majority of the normalization factors.
@@
 **B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.
-**C**: **B**, but with a zoomed y-axis to illustrate the differences in the normalization factors obtained with the different methods.
-:::
-
-::: {custom-style="MDPI_2.2_heading2"}
-### 2.7 Mitigation of High Peak Density Artifacts
+:::
+
+::: {custom-style="MDPI_2.2_heading2"}
+### 2.7. Mitigation of High Peak Density Artifacts
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.8 Changes in Relative Standard Deviation (RSD)
+### 2.8. Changes in Relative Standard Deviation (RSD)
 :::
@@
 For samples 49lipid and 97lipid, the improvements are visually more subtle, having a slight shift to the left as well as a narrowing and smoothing of the RSD distribution.
-Part of this subtlety is likely due to the bi-modality of these two distributions.
+The narrowing of the distribution is detected in the change of the standard deviations (SD) of the RSDs.
+Part of this visual subtlety is likely due to the bi-modality of these two distributions.
 This is not surprising, since these two samples are non-polar extractions from tissue and are biochemically more complex.
@@
 `r table_count$label_text("rsd_method")`.
-RSD means, medians, modes, and maximum observed values for each sample with different overall processing.
+RSD means, standard deviations (sd), medians, modes, and maximum observed values for each sample with different overall processing.
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.9 Difference to Relative Natural Abundance
+### 2.9. Difference to Relative Natural Abundance
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.10 Method Specific Peaks
+### 2.10. Method Specific Peaks
 :::
@@
 Again, this strongly implies that their peak centers have far higher error.
+When we matched assigned peaks across the NSCLC samples for examination of changes in p-values (see **Changes in P-Values On a Large Dataset**), the standard deviation of the peak locations in frequency space was 0.5 (results not shown).
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.11 Changes in P-Values On a Large Dataset
+### 2.11. Changes in P-Values On a Large Dataset
 :::
@@
   geom_point(data = long_measure, aes(x = compare_id, y = value), color = "green") +
-  geom_text(data = long_pvalue, aes(x = compare_id, y = -5, label = format(value, digits = 2))) +
+  geom_text(data = long_pvalue, aes(x = compare_id, y = 2, label = format(value, digits = 2))) +
   labs(x = "", y = "Log-P-Value Difference")
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.12 Quality Control and Quality Analysis
+### 2.12. Quality Control and Quality Analysis
 :::
@@
 The conversion of m/z to frequency also provides opportunities to evaluate each scan and whether it should be kept for further processing.
-Checks that have helped us in the development of the scan-centric peak characterization have included: (i) verifying that the ordering of data points remains identical in m/z and frequency space after conversion to frequency; and (ii) examining the R^2^ fit of the predicted frequency points to the original frequency points across scans for outliers.
+Checks that have helped us in the development of the scan-centric peak characterization have included: (a) verifying that the ordering of data points remains identical in m/z and frequency space after conversion to frequency; and (b) examining the R^2^ fit of the predicted frequency points to the original frequency points across scans for outliers.
 A third quality control step is to examine the coefficients of the square-root term for each scan (see term **y** in Equation `r equation_count$label_text("mzfrequency")`) and look for any outliers, and remove them before continuing the sample processing.
@@
 ::: {custom-style="MDPI_3.1_text"}
-Finally, in the scan-scan normalization step, we require that there are at least 25 peaks in each of the scans that can be used to calculate normalization factors between the scans and the reference scan.
+Finally, in the scan-scan normalization step, we require that there be at least 25 peaks in each of the scans that can be used to calculate normalization factors between the scans and the reference scan.
 If there are no scans with at least 25 peaks, then the processing of that sample will error and stop.
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.1 Samples and Overall Processing
+### 4.1. Samples and Overall Processing
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.2 Matching Peaks
+### 4.2. Matching Peaks
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.3 Conversion of m/z to Frequency
+### 4.3. Conversion of m/z to Frequency
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.4 Frequency Intervals
+### 4.4. Frequency Intervals
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.5 Interval Range Based Data
+### 4.5. Interval Range Based Data
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.6 Peak Containing Intervals
+### 4.6. Peak Containing Intervals
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.7 Peak Detection and Centroided Values
+### 4.7. Peak Detection and Centroided Values
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.8 Scan to Scan Normalization
+### 4.8. Scan to Scan Normalization
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.9 Full Scan-Centric Characterization
+### 4.9. Full Scan-Centric Characterization
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.10 Correction of Height and Standard Deviation
+### 4.10. Correction of Height and Standard Deviation
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.11 Marking High Frequency Standard Deviation Peaks
+### 4.11. Marking High Frequency Standard Deviation Peaks
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.12 Calculation of Relative Standard Deviation
+### 4.12. Calculation of Relative Standard Deviation
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.13 Scan-Centric Peak Assignment
+### 4.13. Scan-Centric Peak Assignment
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.14 Consistently Assigned Lipid Spectral Feature (Corresponded Peak) Generation and Peak Intensity Normalization
+### 4.14. Consistently Assigned Lipid Spectral Feature (Corresponded Peak) Generation and Peak Intensity Normalization
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.15 Peak - Peak NAP Height Ratios
+### 4.15. Peak - Peak NAP Height Ratios
 :::
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.16 Differential Analysis of Large Dataset
+### 4.16. Differential Analysis of Large Dataset
 :::
@@
 
-Of the starting 181 samples, five did not finish during scan-centric characterization due to not having any scans left during normalization.
+Of the starting 181 samples, nine did not finish during scan-centric characterization for various reasons.
 In addition, some of the samples were acquired multiple times, creating duplicate samples.
 Not knowing which sample run was most appropriate to use, we removed all of the duplicated samples.
-Using information-content-informed Kendall-tau correlation (ICI-Kt) [@flightInformationContentInformedKendalltauCorrelation2022], we compared the median correlations of each sample to all others in the same disease class for outliers, and removed `r n_out_vector["cancer"]` cancer and `r n_out_vector["non-cancer"]` samples.
+Using information-content-informed Kendall-tau correlation (ICI-Kt) [@flightInformationContentInformedKendalltauCorrelation2022], we compared the median correlations of each sample to all others in the same disease class for outliers, and removed `r n_out_vector["cancer"]` cancer and `r n_out_vector["non-cancer"]` non-cancer samples.
 The median ICI-Kt correlations are shown in `r supp_figure_count$label_text("icikt_outlier")`.
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 4.17 Software Used
+### 4.17. Software Used
 :::

### 2022-05-19 lungcancer_all/qcqa_related.R

@@ -1,6 +1,6 @@
 # this is where we are running some stuff to check where we are getting scan
 # level QC/QA, besides in the actual peaks.
-all_zip = dir("/mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-04-02", pattern = "zip$", full.names = TRUE)
+all_zip = dir("/mlab/scratch/cesb_data/zip_files/lung_matched_tissue-2022-05-11", pattern = "zip$", full.names = TRUE)
 library(FTMS.peakCharacterization)
 
 scan_time_rtime_filter = function(scan_times, min_time_difference = 4, rtime_limit = 7.5*60){
@@ -133,4 +133,4 @@ check_scans_removed = function(in_zip){
 
 checked_zip = purrr::map(all_zip, check_scans_removed)
 
-saveRDS(checked_zip, file = "data/data_output/lung_data/lung_scandiffs_2022-04-02.rds")
+saveRDS(checked_zip, file = "data/data_output/lung_data/lung_scandiffs_2022-05-11.rds")

### 2022-05-20 lungcancer_all/check_frequency_models.R
after installing new version of SCPC, running again

@@ -188,7 +188,7 @@ run_mzml_list_SC = function(mzml_files, json_files = NULL, progress = TRUE, save
   list(time = tmp, zip_files = zip_files)
 }
 ScanCentricPeakCharacterization::set_internal_map(furrr::future_map)
-zip_sc_dir = file.path("/mlab/scratch/cesb_data/zip_files/check_frequency_models", "SC_new_model")
+zip_sc_dir = file.path("/mlab/scratch/cesb_data/zip_files/check_frequency_models", "SC_new_model_2")
 dir.create(zip_sc_dir)
 zip_sc = run_mzml_list_SC(use_files, json_files = json_files, save_loc = zip_sc_dir)
 saveRDS(zip_sc, file = file.path(zip_sc_dir, "zip_data.rds"))

### 2022-05-20 R/graph_generation.R
making the graph use log10 so can see all the counts

@@ -99,7 +99,7 @@ plot_frequency_conversion = function(char_obj){
   all_frequency_difference_hist = ggplot(raw_points, aes(x = log10(abs(mean_freq_diff)))) +
     geom_histogram(bins = 100) +
     geom_vline(xintercept = log10(0.5), color = "red") +
-    scale_y_continuous(expand = expansion(mult = c(0, .01))) +
+    scale_y_log10(expand = expansion(mult = c(0, .01))) +
     labs(x = "Log10(Frequency Differences)",
          y = "Number of Points")
 

### 2022-05-20 R/qcqa_analysis.R
have less things with the new scans

@@ -13,7 +13,7 @@ coefficients_analysis <- function(coefficients_data) {
     length(.x$frequency$bad)
   })
 
-  example_freq = coefficients_data[[(which(n_freq == 7)[1])]]
+  example_freq = coefficients_data[[(which(n_freq == 5)[1])]]
 
   df_freq = data.frame(coefficient = example_freq$frequency$coefficients,
                        sample = gsub(".zip$", "", basename(example_freq$zip)))

### 2022-05-20 _targets.R
using the new version of lung characterized data

@@ -105,6 +105,7 @@ figures_tar = tar_plan(
   tar_target(frequency_conversion, plot_frequency_conversion(data_97lipid)),
   tar_target(peak_ordering, plot_peak_ordering(method_filtersd_97lipid)),
   tar_target(sliding_regions, plot_sliding_window_density(data_97lipid)),
+  tar_target(example_sliding_region, plot_example_sliding_counts(data_97lipid)),
   tar_target(peak_fit_plot, plot_peak_fitting(method_filtersd_97lipid)),
   tar_target(rsd_plot, plot_rsd_differences(rsd_combine)),
   tar_target(rsd_plot_alt, plot_rsd_differences(rsd_combine %>%
@@ -452,7 +453,7 @@ lungcancer_tar = list(
 
 qcqa_tar = list(
   tar_target(coefficients_file,
-             "data/data_output/lung_data/lung_scandiffs_2022-04-02.rds",
+             "data/data_output/lung_data/lung_scandiffs_2022-05-11.rds",
              format = "file"),
   tar_target(coefficients_data,
              readRDS(coefficients_file)),

### 2022-05-20 doc/peakcharacterization_manuscript.Rmd
edits and more edits

@@
 In the Thermo-Fisher Orbitrap FT-MS instruments, data is acquired in microscans and scans, where each scan is an aggregate of multiple microscans.
-While these micro-scan and subsequent scans are expected to be analytical replicates, spray instability and temporal delay in automatic gain control pragmatically break this expectation.
+While these microscan and subsequent scans are expected to be analytical replicates, spray instability and temporal delay in automatic gain control pragmatically break this expectation.
 Therefore, it is advantageous to keep the data at the scan level, so that any given poor quality (i.e., "bad") scan can be removed prior to aggregating the data across scans and generating a centroided peak m/z and intensity.
@@
 To motivate our solution, we generated peak lists using the peak exporting functionality in Xcalibur as well as our scan-centric peak characterization, and found matching Xcalibur peaks to our assignments.
-As an example, in `r figure_count$label_text("motivation")`, we show the Xcalibur calculated intensities from the simplistically averaged spectrum and theoretical peak intensities based on NAP for four peaks matching threonine.
+As an example, in `r figure_count$label_text("motivation")`A, we show the Xcalibur calculated intensities from the simplistically averaged spectrum and theoretical peak intensities based on NAP for four peaks matching threonine.
 As the peak intensity decreases, the deviation between the Xcalibur calculated intensity and its corresponding NAP-based theoretical intensity becomes larger.
@@
 **`r figure_count$label_text("motivation")`.**
-**A** Xcalibur intensities (black) and theoretical intensities based on relative NAP (red) for ECF derived threonine.
+**A**: Xcalibur intensities (black) and theoretical intensities based on relative NAP (red) for ECF derived threonine.
 The 18O isotopologue is shifted by 0.01 m/z and the NAP peaks by 0.05 m/z to aid visualization.
-**B** Theoretical NAP-based-Log10-intensities and Log10 Xcalibur intensities.
+**B**: Theoretical NAP-based-Log10-intensities and Log10 Xcalibur intensities.
 The red line indicates perfect agreement.
@@
 However, the spectral frequency can be calculated by dividing the midpoint m/z of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points representing the midpoint m/z of two adjacent points, length of red lines representing the difference between the two adjacent points).
-The subsequent adjacent point differences in frequency are expected to be relatively constant with respect to m/z, as shown in `r figure_count$label_text("mz_frequency_conversion")`C, D and E, in contrast to the adjacent point differences in m/z, which are not constant with respect to m/z.
-The Thermo-Fisher Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 450K or 500K depending on the sample, has an adjacent frequency point difference mode of 0.5, as shown in  `r figure_count$label_text("mz_frequency_conversion")`D.
+The subsequent adjacent point differences in frequency are expected to be relatively constant with respect to m/z, as shown in `r figure_count$label_text("mz_frequency_conversion")`C, `r figure_count$just_count("mz_frequency_conversion")`D and `r figure_count$just_count("mz_frequency_conversion")`E, in contrast to the adjacent point differences in m/z, which are not constant with respect to m/z.
+The Thermo-Fisher Fusion instrument from which most of our collaborators' data has been acquired, at a resolution of 450K or 500K depending on the sample, has an adjacent frequency point difference mode of 0.5, as shown in  `r figure_count$label_text("mz_frequency_conversion")`D.
 It is important to note that although the frequency difference has been consistent regardless of instrument resolution, we do not know a priori what it should be, and the method makes no assumptions about the frequency difference, but rather calculates the mode of the frequency differences from those observed in the data itself.
@@
 This regression model seems to fit the known relationship between frequency and m/z, where the frequency is related to $1/\sqrt{m/z}$.
-The actual regression model used in this work includes terms of the m/z, $\sqrt{m/z}$, and $\sqrt[3]{m/z}$ (see **Conversion of m/z to Frequency** in Methods for further description of the regression model).
+The actual regression model used in this work includes terms for m/z, $\sqrt{m/z}$, and $\sqrt[3]{m/z}$ (see **Conversion of m/z to Frequency** in Methods for further description of the regression model).
 We did investigate a variety of frequency regression models to see which one seemed to be the best, as shown in Supplemental `r supp_figure_count$label_text("frequency_models")` and `r supp_table_count$label_text("frequency_models")`.
@@
 The red lines denote the difference between adjacent points, which are shown in the y-axis of **C** for this single peak.
-A histogram of adjacent point differences for all points is shown in **D**.
+**D** shows a histogram of adjacent point differences for all points. Note that the y-axis is Log10 scaled.
 The differences for all points in a single scan vs m/z are shown in **E**, with those differences that lie within 0.49 - 0.51 shown in red.
@@
 Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
-Subsequently, we divide the counts into tiled regions of a set size (1000 frequency in this work), and examine the 99th percentile of the sliding window counts within the tiled region.
-In `r figure_count$label_text("slidingwindow_count")`A we plotted the non-zero point intensities from all scans for one tiled region 2000 points wide.
-The number of non-zero points in the sliding windows across this region are shown in `r figure_count$label_text("slidingwindow_count")`B, and zoomed in in C.
+Subsequently, we divide the counts into tiled regions of a set size (1000 frequency in this work), and examine the cumulative 99^th^ percentile of the sliding window counts within the tiled region.
+In `r figure_count$label_text("slidingwindow_count")`A, we plotted the non-zero point intensities from all scans for one tiled region 1000 frequency wide.
+The number of non-zero points in the sliding windows across this region are shown in `r figure_count$label_text("slidingwindow_count")`B, and zoomed closer in `r figure_count$just_count("slidingwindow_count")`C.
 These provide a measure of the non-zero point density across this region.
-The blue line indicates the cumulative 99th percentile for this one region, and is high, given that it looks like there are some consistent peaks across the scans in this region.
-However, examining the cumulative 99th percentiles across all the regions of the spectrum, as shown in the histogram of `r figure_count$label_text("slidingwindow_count")`F, a rather large fraction of the regions have very low cumulative 99th percentile values, with a median value of 5, which results in a cutoff value of 8 (ceiling of 5 x 1.5).
-Keeping only those sliding regions with a non-zero point count greater than 8 (above the red line in `r figure_count$label_text("slidingwindow_count")`B and C) results in those sliding regions shown in D, and then the point intensities in E.
-A non-zero point density based method to removing potential noise is desirable because the peak intensities across scans are not identical, as shown by the need for normalization (see **Normalization of Scans**), and the distribution of RSD values before and after normalization (see **Changes in Relative Standard Deviation**).
+The blue line indicates the cumulative 99^th^ percentile for this one region, and is high, given that it looks like there are some consistent peaks across the scans in this region.
+However, examining the cumulative 99^th^ percentiles across all the regions of the spectrum, as shown in the histogram of `r figure_count$label_text("slidingwindow_count")`F, a rather large fraction of the regions have very low cumulative 99^th^ percentile values, with a median value of 5, which results in a cutoff value of 8 (ceiling of 5 x 1.5).
+Keeping only those sliding regions with a non-zero point count greater than 8 (above the red line in `r figure_count$label_text("slidingwindow_count")`B and `r figure_count$just_count("slidingwindow_count")`C) results in those sliding regions shown in `r figure_count$just_count("slidingwindow_count")`D, and then the point intensities in `r figure_count$just_count("slidingwindow_count")`E.
+A non-zero point density based method for removing potential noise is desirable because the peak intensities across scans are not identical, as shown by the need for normalization (see **Normalization of Scans**), and the distribution of RSD values before and after normalization (see **Changes in Relative Standard Deviation**).
 :::
@@
 **`r figure_count$label_text("slidingwindow_count")`.**
-**A - E**: A single tiled region 1000 frequency wide.
+**A – E**: A single tiled region 1000 frequency wide.
 **A**: Non-zero point intensities in frequency space across all scans.
-**B**: Number of non-zero points in sliding regions. Blue line indicates 99th percentile of non-zero counts in this one region.
-Red line is the rounded value of median x 1.5 of 99th percentiles from **F**.
+**B**: Number of non-zero points in sliding regions. The blue line indicates 99^th^ percentile of non-zero counts in this one region.
+The red line is the rounded value of median x 1.5 of 99^th^ percentiles from **F**.
 **C**: Zoomed y-axis of **B** to show the non-zero counts that are below the median cutoff.
@@
 **E**: Non-zero point intensities across scans in the regions from D.
-**F**: Histogram of the 99th percentile cutoffs from all of the tiled windows. Red line is the median x 1.5.
-:::
-
-::: {custom-style="MDPI_3.1_text"}
-Although it appears that only a handful of potential signal regions are removed between `r figure_count$label_text("slidingwindow_count")`A and E with a cutoff of 8 non-zero points in a sliding region, the number of initial signal regions rapidly explodes as the cumulative percentile cutoff is lowered, going from thousands to tens of thousands of potential peak regions to find and fit peaks within, as shown in `r figure_count$label_text("percentile_variation")`A and B for both the amino-acid ECF and lipid samples, respectively.
+**F**: Histogram of the 99^th^ percentile cutoffs from all of the tiled windows. The red line is the median x 1.5.
+:::
+
+::: {custom-style="MDPI_3.1_text"}
+Although it appears that only a handful of potential signal regions are removed between `r figure_count$label_text("slidingwindow_count")`A and `r figure_count$just_count("slidingwindow_count")`E with a cutoff of 8 non-zero points in a sliding region, the number of initial signal regions rapidly explodes as the cumulative percentile cutoff is lowered, going from thousands to tens of thousands of potential peak regions to find and fit peaks within, as shown in `r figure_count$label_text("percentile_variation")`A and `r figure_count$just_count("percentile_variation")`B for both the amino-acid ECF and lipid samples, respectively.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-Due to differences in how many ions are captured in the trap and the limited dynamic range of the detector, the observed peak intensities for the same analyte vary between scans (see `r figure_count$label_text("breaking_regions")`A and B for example).
+Due to differences in how many ions are captured in the trap and the limited dynamic range of the detector, the observed peak intensities for the same analyte vary between scans (see `r figure_count$label_text("breaking_regions")`A and `r figure_count$just_count("breaking_regions")`B for example).
 Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
@@
 The highlighted peaks in `r figure_count$label_text("intensity_scan")`B are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
-If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and B.
+If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and `r figure_count$just_count("normalization_factors")`B.
 Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B), which is expected as they both use the most intense peaks.
@@
 To verify the mitigation of HPD regions, we converted centroided m/z's from Xcalibur to frequency values using the previously calculated regression model for that sample, measured peak density to detect HPD regions, and compared them with the scan-centric peaks and FSD outliers.
-Xcalibur peak density was measured using a sliding window ten points wide and a stride of one point.
+The Xcalibur peak density was measured using a sliding window ten points wide and a stride of one point.
 
@@
 For samples 49lipid and 97lipid, the improvements are visually more subtle, having a slight shift to the left as well as a narrowing and smoothing of the RSD distribution.
-The narrowing of the distribution is detected in the change of the standard deviations (SD) of the RSDs.
+The narrowing of the distribution is detected in the change of the standard deviations (SD) of the RSDs, however.
 Part of this visual subtlety is likely due to the bi-modality of these two distributions.
@@
 To find intervals that contain points that represent actual signal and not just random noise, the number of non-zero intensity points in each sliding window are counted.
-Subsequently, we break these counts into fixed width tiles (default width of 1000 in frequency space) and calculate the 99th percentile of non-zero points for each tile.
-The rounded up median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions containing potential signal.
+Subsequently, we break these counts into fixed width tiles (default width of 1000 in frequency space) and calculate the cumulative 99^th^ percentile of non-zero points for each tile.
+The rounded up median value x 1.5 of these 99^th^ percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions containing potential signal.
 Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are kept and overlapping sliding windows are merged to create the initial peak regions.
@@
 ::: {custom-style="MDPI_3.1_text"}
-From this weighted parabolic fit, the center, intensity, integrated area and sum-of-square residuals can be extracted for the peak.
+From this weighted parabolic fit, the center, intensity, integrated area, and sum-of-square residuals can be extracted for the peak.
 The center and intensity from this model are equivalent to the centroided peak center and intensity.
@@
 Modes of the RSD are calculated starting from a density approximation of the distribution using default settings in the density function of the R stats package.
-Peaks from the density approximation are determined with the findpeaks function in the pracma package.
+Peaks from the density approximation are determined with the find_peaks function in the pracma package.
 Up to two modes are reported from low to high RSD along the distribution: i) the most intense peak; ii) the next most intense peak if it is ≥ 0.2X the most intense peak.
@@
 
-Of the starting 181 samples, nine did not finish during scan-centric characterization for various reasons.
+Of the starting 181 samples, 12 did not finish during scan-centric peak characterization for various reasons, leaving 169.
 In addition, some of the samples were acquired multiple times, creating duplicate samples.

### 2022-05-20 doc/peakcharacterization_manuscript.Rmd
updating for html output too

@@
       - '--lua-filter=author-info-blocks.lua'
+  html_document:
+    pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
 bibliography: '`r here::here("doc/peakcharacterization.json")`'

### 2022-05-20 doc/peakcharacterization_supplemental.Rmd
updating for html output too

@@
     keep_md: true
+    pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
+  html_document:
+    pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
+  pdf_document:
     pandoc_args:

### 2022-05-20 R/strip_mdpi.R

@@ -1,4 +1,4 @@
-strip_mdpi_render = function(in_rmd = "doc/peakcharacterization_manuscript.Rmd", out_rmd = "doc/peakcharacterization_nostyle.Rmd"){
+strip_mdpi_render = function(in_rmd = "doc/peakcharacterization_manuscript.Rmd", out_rmd = "doc/peakcharacterization_nostyle.Rmd", out_file = here::here("doc/peakcharacterization_nostyle.docx"), output_format = "word_document"){
   rmd_doc = readLines(in_rmd)
 
   template_line = grepl("metabolites-template", rmd_doc)
@@ -8,7 +8,11 @@ strip_mdpi_render = function(in_rmd = "doc/peakcharacterization_manuscript.Rmd",
   rmd_nostyle = rmd_notemplate[!has_style]
 
   cat(rmd_nostyle, file = out_rmd, sep = "\n")
-  rmarkdown::render(out_rmd, knit_root_dir = getwd(), quiet = TRUE)
+  rmarkdown::render(out_rmd,
+                    output_format = output_format,
+                    output_file = out_file,
+                    knit_root_dir = getwd(),
+                    quiet = TRUE)
   beepr::beep(4)
   out_rmd
 }

### 2022-05-20 _targets.R

@@ -139,8 +139,17 @@ figures_tar = tar_plan(
              "doc/peakcharacterization_manuscript.Rmd"),
   tar_render(supplemental_materials,
              "doc/peakcharacterization_supplemental.Rmd"),
+  tar_render(supplemental_html,
+             "doc/peakcharacterization_supplemental.Rmd",
+             output_file = here::here("docs/scancentricpeakcharacterization_supplemental.html"),
+             output_format = "html_document"),
+             # output_format = "pdf_document"),
   tar_target(manuscript_nostyle,
              strip_mdpi_render(base_manuscript)),
+  tar_target(manuscript_html,
+             strip_mdpi_render(base_manuscript,
+                               out_file = here::here("docs/scancentricpeakcharacterization.html"),
+                               output_format = "html_document")),
   tar_target(manuscript_mdpi,
              strip_headers_render(base_manuscript))
 )

### 2022-06-01 doc/peakcharacterization_manuscript.Rmd
some minor edits

@@
 ::: {custom-style="MDPI_1.7_abstract"}
-**Abstract:** We present a novel, scan-centric method for characterizing peaks from direct injection multi-scan Fourier transform mass spectra of complex samples that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
+**Abstract:** We present a novel, scan-centric method for characterizing peaks from direct injection multi-scan Fourier transform mass spectra of complex samples that utilizes frequency values derived directly from the spacing of raw *m/z* points in spectral scans.
 Our peak characterization method utilizes intensity independent noise removal and normalization of scan-level data to provide a much better fit of relative intensity to natural abundance probabilities for low abundance isotopologues that are not present in all of the acquired scans.
@@
 ::: {custom-style="MDPI_1.8_keywords"}
-**Keywords**: Fourier-transform mass spectrometry, orbitrap, scan-centric peak characterization, Fellgett's advantage, interferograms, frequency spectrum
+**Keywords**: Fourier-transform mass spectrometry; orbitrap; scan-centric peak characterization; Fellgett's advantage; interferograms; frequency spectrum
 :::
@@
 Fourier-transform mass spectrometry (FT-MS) provides high performance in terms of sensitivity, resolution, and mass accuracy all in one analytical instrumentation.
-The combination of these capabilities provide several analytical and interpretive improvements: (i) the ability to resolve distinct isotopologues with identical unit masses but different accurate masses [@higashiStableIsotopeLabeledTracers2014]; (ii)  enabling multi-element isotopic natural abundance correction for at least the lower portion of the detected mass range  [@moseleyCorrectingEffectsNatural2010; @carreerComputationalFrameworkHighThroughput2013; @wangAccuCor2IsotopeNatural2021]; (iii) improved assignment accuracy [@kindMetabolomicDatabaseAnnotations2006; @mitchellSmallMoleculeIsotope2019]; and (iv) the detection of metabolites in the sub-femtomolar range, when combined with chromatographic separation  [@eylesMethodsStudyProtein2004; @dettmerMassSpectrometrybasedMetabolomics2007].
+The combination of these capabilities provides several analytical and interpretive improvements: (i) the ability to resolve distinct isotopologues with identical unit masses but different accurate masses [@higashiStableIsotopeLabeledTracers2014]; (ii)  enabling multi-element isotopic natural abundance correction for at least the lower portion of the detected mass range  [@moseleyCorrectingEffectsNatural2010; @carreerComputationalFrameworkHighThroughput2013; @wangAccuCor2IsotopeNatural2021]; (iii) improved assignment accuracy [@kindMetabolomicDatabaseAnnotations2006; @mitchellSmallMoleculeIsotope2019]; and (iv) the detection of metabolites in the sub-femtomolar range, when combined with chromatographic separation  [@eylesMethodsStudyProtein2004; @dettmerMassSpectrometrybasedMetabolomics2007].
 In the metabolomics field, these improvements permit more complicated, but more informative experimental designs such as the use of single and multiple isotope-labeled precursors in stable isotope-resolved metabolomics (SIRM) experiments [@yangChloroformateDerivatizationTracing2017].
@@
 SMIRFE uses expected relationships of relative peak height or intensity to natural abundance probability (NAP) across multiple peak pairs to determine the likelihood of an assignment for the peaks in question.
-In the Thermo-Fisher Orbitrap FT-MS instruments, data is acquired in microscans and scans, where each scan is an aggregate of multiple microscans.
+In the Thermo-Fisher Orbitrap FT-MS instruments, data are acquired in microscans and scans, where each scan is an aggregate of multiple microscans.
 While these microscan and subsequent scans are expected to be analytical replicates, spray instability and temporal delay in automatic gain control pragmatically break this expectation.
-Therefore, it is advantageous to keep the data at the scan level, so that any given poor quality (i.e., "bad") scan can be removed prior to aggregating the data across scans and generating a centroided peak m/z and intensity.
-If the scan level data is not processed and aggregated correctly, the final peak intensities in arbitrary units and centers in m/z will have high uncertainty (total variance) [@moseleyErrorAnalysisPropagation2013].
+Therefore, it is advantageous to keep the data at the scan level, so that any given poor quality (i.e., "bad") scan can be removed prior to aggregating the data across scans and generating a centroided peak *m/z* and intensity.
+If the scan level data are not processed and aggregated correctly, the final peak intensities in arbitrary units and centers in *m/z* will have high uncertainty (total variance) [@moseleyErrorAnalysisPropagation2013].
 
@@
 Here we describe novel scan-centric FT-MS metabolomics data processing methods that better preserve the expected relative intensity to NAP.
-In addition, due to the tendency of increasing point spacing with increasing m/z in Orbitrap type instruments, we derive a method for transforming the m/z data to axial frequency, which has the desirable property of being equally spaced across the full spectrum.
+In addition, due to the tendency of increasing point spacing with increasing *m/z* in Orbitrap type instruments, we derive a method for transforming the *m/z* data to axial frequency, which has the desirable property of being equally spaced across the full spectrum.
 Finally, we show that in addition to better preserving the expected intensity to NAP relationships, our scan-centric method also results in improved relative standard deviations, automatic identification of high-peak-density artifact peaks, and better separation of samples in a lipidomics data analysis.
@@
 **A**: Xcalibur intensities (black) and theoretical intensities based on relative NAP (red) for ECF derived threonine.
-The 18O isotopologue is shifted by 0.01 m/z and the NAP peaks by 0.05 m/z to aid visualization.
+The 18O isotopologue is shifted by 0.01 *m/z* and the NAP peaks by 0.05 *m/z* to aid visualization.
 **B**: Theoretical NAP-based-Log10-intensities and Log10 Xcalibur intensities.
@@
 ::: {custom-style="MDPI_2.2_heading2"}
-### 2.2. m/z to Frequency
+### 2.2. *m/z* to Frequency
 :::
@@
 FT-MS data from the Thermo-Fisher Orbitrap instruments used to acquire the data does not provide any information about the raw spectral frequency data.
-Outside of the meta-data, it merely contains the m/z and intensity values for profile spectra acquired across multiple scans.
-However, the spectral frequency can be calculated by dividing the midpoint m/z of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points representing the midpoint m/z of two adjacent points, length of red lines representing the difference between the two adjacent points).
-The subsequent adjacent point differences in frequency are expected to be relatively constant with respect to m/z, as shown in `r figure_count$label_text("mz_frequency_conversion")`C, `r figure_count$just_count("mz_frequency_conversion")`D and `r figure_count$just_count("mz_frequency_conversion")`E, in contrast to the adjacent point differences in m/z, which are not constant with respect to m/z.
-The Thermo-Fisher Fusion instrument from which most of our collaborators' data has been acquired, at a resolution of 450K or 500K depending on the sample, has an adjacent frequency point difference mode of 0.5, as shown in  `r figure_count$label_text("mz_frequency_conversion")`D.
+Outside of the meta-data, it merely contains the *m/z* and intensity values for profile spectra acquired across multiple scans.
+However, the spectral frequency can be calculated by dividing the midpoint *m/z* of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points representing the midpoint *m/z* of two adjacent points, length of red lines representing the difference between the two adjacent points).
+The subsequent adjacent point differences in frequency are expected to be relatively constant with respect to *m/z*, as shown in `r figure_count$label_text("mz_frequency_conversion")`C, `r figure_count$just_count("mz_frequency_conversion")`D and `r figure_count$just_count("mz_frequency_conversion")`E, in contrast to the adjacent point differences in *m/z*, which are not constant with respect to *m/z*.
+The Thermo-Fisher Fusion instrument from which most of our collaborators' data has been acquired, at a resolution of 450 K or 500 K depending on the sample, has an adjacent frequency point difference mode of 0.5, as shown in  `r figure_count$label_text("mz_frequency_conversion")`D.
 It is important to note that although the frequency difference has been consistent regardless of instrument resolution, we do not know a priori what it should be, and the method makes no assumptions about the frequency difference, but rather calculates the mode of the frequency differences from those observed in the data itself.
-Restricting to those points that fall into a narrow range around the mode of frequency differences (0.49 - 0.51 in this work), a regression model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`F.
-This regression model seems to fit the known relationship between frequency and m/z, where the frequency is related to $1/\sqrt{m/z}$.
+Restricting to those points that fall into a narrow range around the mode of frequency differences (0.49 - 0.51 in this work), a regression model of frequency to *m/z* can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`F.
+This regression model seems to fit the known relationship between frequency and *m/z*, where the frequency is related to $1/\sqrt{m/z}$.
 The actual regression model used in this work includes terms for m/z, $\sqrt{m/z}$, and $\sqrt[3]{m/z}$ (see **Conversion of m/z to Frequency** in Methods for further description of the regression model).
 We did investigate a variety of frequency regression models to see which one seemed to be the best, as shown in Supplemental `r supp_figure_count$label_text("frequency_models")` and `r supp_table_count$label_text("frequency_models")`.
-
-The constant point difference of frequency space is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points.
-The m/z point-to-point differences are not constant, but can be approximated by a LOESS linear model [@Cleveland1992loess]; however, it is exceedingly difficult to create a LOESS model with an intercept of 0.
-In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the LOESS model.
-Frequency-based points suffer none of these drawbacks, and the conversion from m/z can be derived from the raw profile level data itself, which is incredibly useful.
 :::
@@
 **`r figure_count$label_text("mz_frequency_conversion")`.** 
-**A**: Intensity vs m/z for a single peak from a single scan.
-Red lines denote the difference between each adjacent pair of points, and red dots the m/z midpoint between the pair of points.
+**A**: Intensity vs *m/z* for a single peak from a single scan.
+Red lines denote the difference between each adjacent pair of points, and red dots the *m/z* midpoint between the pair of points.
 The difference divided by the midpoint is used to derive the spectral frequency values in **B**.
@@
 **D** shows a histogram of adjacent point differences for all points. Note that the y-axis is Log10 scaled.
-The differences for all points in a single scan vs m/z are shown in **E**, with those differences that lie within 0.49 - 0.51 shown in red.
-**F** shows the plot of derived spectral frequency vs m/z, with fitted values from the linear regression model in red.
+The differences for all points in a single scan vs *m/z* are shown in **E**, with those differences that lie within 0.49 - 0.51 shown in red.
+**F** shows the plot of derived spectral frequency vs *m/z*, with fitted values from the linear regression model in red.
+:::
+
+::: {custom-style="MDPI_3.1_text"}
+The constant point difference in frequency space is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points.
+The *m/z* point-to-point differences are not constant, but can be approximated by a LOESS linear model [@Cleveland1992loess]; however, it is exceedingly difficult to create a LOESS model with an intercept of 0.
+In addition, we would also need to vary the width of sliding windows according to the *m/z* difference at a particular *m/z* based on the LOESS model.
+Frequency-based points suffer from none of these drawbacks, and the conversion from *m/z* can be derived from the raw profile level data itself, which is incredibly useful.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-The m/z to frequency regression models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
-While scan specific models **could** be used in the conversion of m/z to frequency, doing so results in changes to the relative peak ordering compared to m/z space, as shown in `r figure_count$label_text("peak_ordering")`.
+The *m/z* to frequency regression models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
+While scan specific models **could** be used in the conversion of *m/z* to frequency, doing so results in changes to the relative peak ordering compared to *m/z* space, as shown in `r figure_count$label_text("peak_ordering")`.
 Therefore, a single model for all scans based on the scan with the slope closest to the median of slopes across scans is used for converting *all* remaining scan level data.
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("peak_ordering")`.** Peak ordering in m/z compared with ordering in frequency space when a single m/z to frequency model is used or scan specific m/z to frequency models are used.
-For a single peak, the scan level peak m/z's were extracted, and then frequency values for those m/z generated using a single common model of m/z to frequency (*Single Model*), or models derived from each scan (*Individual Models*).
+**`r figure_count$label_text("peak_ordering")`.** Peak ordering in *m/z* compared with ordering in frequency space when a single *m/z* to frequency model is used or scan specific *m/z* to frequency models are used.
+For a single peak, the scan level peak *m/z*'s were extracted, and then frequency values for those *m/z* generated using a single common model of *m/z* to frequency (*Single Model*), or models derived from each scan (*Individual Models*).
 A subset of the peaks ends up out of order using scan specific models, implying that a single model should be used across all the scan level data.
@@
 ::: {custom-style="MDPI_3.1_text"}
-Although the original model is created from only those points that had frequency point-to-point differences within a narrow range, **all** m/z points are converted to frequency for subsequent steps in the workflow.
+Although the original model is created from only those points that had frequency point-to-point differences within a narrow range, **all** *m/z* points are converted to frequency for subsequent steps in the workflow.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-In a dataset of this nature, we expect that much of the data is really just noise, and does not contribute meaningfully to the analysis.
+In a dataset of this nature, we expect that much of the data are really just noise, and do not contribute meaningfully to the analysis.
 Furthermore, it is expected that noise is randomly distributed across the scans.
@@
 ::: {custom-style="MDPI_3.1_text"}
-Although many other types of mass spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo-Fisher Fusion has a baseline of 0 due to manipulations in the Thermo-Fisher firmware, making the determination of the centroided values considerably easier.
+Although many other types of mass spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo-Fisher Fusion have a baseline of 0 due to manipulations in the Thermo-Fisher firmware, making the determination of the centroided values considerably easier.
 For each region initially created, the peaks in each scan within that region can be characterized (i.e., centroiding).
-For centroiding, we use a simple weighted quadratic model of log(intensity) to m/z. 
-We add a small constant to enable using the zero values, and weight the values by their ratio to the most intense value, which is normally the value closest to the center of peak, helping to ensure that the **true** centroid is determined.
+For centroiding, we use a simple weighted quadratic model of log(intensity) to *m/z*. 
+We add a small constant to enable using the zero values, and weight the values by their ratio to the most intense value, which is normally the value closest to the center of the peak, helping to ensure that the **true** centroid is determined.
 From the fitted model, we can derive the centroided center and the intensity of the peak, as shown in `r figure_count$label_text("centroided_peaks")`.
@@
 **`r figure_count$label_text("centroided_peaks")`.**
-Log-intensity (A) and Intensity (B) of points for a single peak against frequency. Black points are the original data points, and the blue point represents the calculated centroid.
+Log-intensity (A) and intensity (B) of points for a single peak against frequency. Black points are the original data points, and the blue point represents the calculated centroid.
 :::
@@
 Clearly the region has two separate peaks within it.
-**B**: The peak centroids (frequency and intensity) for each peak in black.
+**B**: The peak centroids (frequency and intensity) for each peak are in black.
 The tiled regions (red) are used to quantify the number of peaks.
-**C**: The number of peaks within each tiled region shown as a histogram.
+**C**: The number of peaks within each tiled region are shown as a histogram.
 Each group of non-zero adjacent regions will be merged to form a new peak region.
@@
 If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and `r figure_count$just_count("normalization_factors")`B.
-Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B), which is expected as they both use the most intense peaks.
-But removing the scan-correlated peaks does change the majority of the normalization factors.
+Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B), this is expected as they both use the most intense peaks.
+However, removing the scan-correlated peaks does change the majority of the normalization factors.
 :::
@@
 ```{r intensity_scan, dn_id = figure_count}
-# showing peak intensity correlation with m/z, intensity varying with m/z
+# showing peak intensity correlation with *m/z*, intensity varying with *m/z*
 tar_load(intensity_scan_cor_plot)
@@
 ::: {custom-style="MDPI_5.1_figure_caption"}
-**`r figure_count$label_text("normalization_factors")`.** **A**: Histogram of scan normalization factors using either a single-pass normalization using *all* peaks (singlenorm), single-pass normalization using peaks with an intensity ≥ 0.7 of the maximum intensity (singlenorm_int), or the two-pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
-**B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.
+**`r figure_count$label_text("normalization_factors")`.** **A**: Histogram of scan normalization factors using either a single-pass normalization using *all* peaks (*singlenorm*), single-pass normalization using peaks with an intensity ≥ 0.7 of the maximum intensity (*singlenorm_int*), or the two-pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (*doublenorm*).
+**B**: The difference of the normalization factors obtained from either *doublenorm* or *singlenorm_int* compared to *singlenorm*.
 :::
@@
 These peaks can be detected by simply examining the distribution of FSDs and marking outlier peaks.
-To verify the mitigation of HPD regions, we converted centroided m/z's from Xcalibur to frequency values using the previously calculated regression model for that sample, measured peak density to detect HPD regions, and compared them with the scan-centric peaks and FSD outliers.
+To verify the mitigation of HPD regions, we converted centroided *m/z*'s from Xcalibur to frequency values using the previously calculated regression model for that sample, measured peak density to detect HPD regions, and compared them with the scan-centric peaks and FSD outliers.
 The Xcalibur peak density was measured using a sliding window ten points wide and a stride of one point.
@@
 **`r figure_count$label_text("rsd_method")`.**
-Density plots of relative standard deviations (RSD) of peak heights across scans for each of the processing methods.
-A peak had to be present in at least three scans for the RSD value to be reported.
+Density plots for relative standard deviations (RSD) of peak heights across scans for each of the processing methods.
+A peak was required to be present in at least three scans for the RSD value to be reported.
 :::
@@
 The 97lipid filtersd RSD distribution (as well as others) is bi-modal.
-`r figure_count$label_text("rsd_intensity")` plots each peaks RSD as a function of the mean intensity across scans, for every peak present in at least three scans (top), or the peaks present in at least 80% of the scans (bottom).
+`r figure_count$label_text("rsd_intensity")` plots each peak RSD as a function of the mean intensity across scans, for every peak present in at least three scans (top), or the peaks present in at least 80% of the scans (bottom).
 At least part of the bi-modality in the RSD distribution appears to be related to this dependence of RSD on intensity, as well as truncation as peaks appear in fewer scans.
 This is supported by part of the distribution disappearing when we require that peaks be in at least 80% of the available scans (difference between `r figure_count$label_text("rsd_intensity")` top and bottom).
-
-We can remove a majority of the bi-modality by only examining those peaks with a Log10(mean) intensity ≥ 5.
-`r supp_figure_count$label_text("rsd_method")` and `r supp_table_count$label_text("rsd_method")` show how the RSD distributions change for each sample when only the peaks with Log10(mean) intensity ≥ 5 are used.
-The distributions are all shifted to lower RSD, however, the overall trends in RSD are the same.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
+We can remove a majority of the bi-modality by only examining those peaks with a Log10(mean) intensity ≥ 5.
+`r supp_figure_count$label_text("rsd_method")` and `r supp_table_count$label_text("rsd_method")` show how the RSD distributions change for each sample when only the peaks with Log10(mean) intensity ≥ 5 are used.
+The distributions are all shifted to lower RSD; however, the overall trends in RSD are the same.
+
 `r figure_count$label_text("rsd_intensity")` also provides information about the sources of variance or error in the FT-MS measurements.
@@
 ::: {custom-style="MDPI_3.1_text"}
-As an alternative to RSD, we can also compare the fit of relative intensities after assignment using SMIRFE [@mitchellSmallMoleculeIsotope2019] to the theoretical relative natural abundances (relNAP) of the assigned isotopic molecular formula's (IMFs) within the assigned elemental molecular formula's (EMFs).
+As an alternative to RSD, we can also compare the fit of relative intensities after assignment using SMIRFE [@mitchellSmallMoleculeIsotope2019] to the theoretical relative natural abundances (relNAP) of the assigned isotopic molecular formulas (IMFs) within the assigned elemental molecular formulas (EMFs).
 Theoretically, we expect lower quality data to have both lower numbers of assignments, and for those things that are assigned, the fit between relative intensity and relNAP to be worse.
@@
 
-`r figure_count$label_text("nap_intensity")` compares the peak-peak isotopic natural abundance probability and height log-ratio differences (Equation `r equation_count$label_text("napintensity")` in Methods) generated using heights from Xcalibur and from our scan-centric peak characterization.
+`r figure_count$label_text("nap_intensity")` compares the peak-to-peak isotopic natural abundance probability and height log-ratio differences (Equation `r equation_count$label_text("napintensity")` in Methods) generated using heights from Xcalibur and from our scan-centric peak characterization.
 From this figure, it is clear that inconsistency in peak presence across scans leads to larger deviations between measured peak heights from aggregate spectra and expected relNAP.
@@
 **A**.
-The peak-peak NAP - intensity log differences from scan-centric peak heights (red) and Xcalibur peak heights (blue) from the ECF derivatized threonine amino acid assignments with Na adduct, with point size reflecting how many peaks were missing across scans.
-**B**. The difference of Xcalibur to scan-centric ratios plotted directly as a function of the number of scans the peak was not found in.
-**C**. The differences of Xcalibur to scan-centric ratios for all of the amino acid assignments in EMFs with more than a single peak in both ECF samples.
+The peak-to-peak NAP - intensity log differences from scan-centric peak heights (red) and Xcalibur peak heights (blue) from the ECF derivatized threonine amino acid assignments with Na adduct, with point size reflecting how many peaks were missing across scans.
+**B**. The difference in Xcalibur to scan-centric ratios plotted directly as a function of the number of scans the peak was not found in.
+**C**. The differences in Xcalibur to scan-centric ratios for all of the amino acid assignments in EMFs with more than a single peak in both ECF samples.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-Each set of peaks generated may or may not be specific to the particular method used to generate centroided peak m/z and intensities.
+Each set of peaks generated may or may not be specific to the particular method used to generate centroided peak *m/z* and intensities.
 This is true for the different scan-centric combinations, as well as the peaks from Xcalibur and MSnbase.
@@
 The 1ecf sample has all of the scan-centric peaks shared with either Xcalibur or MSnbase peaks, whereas the 97lipid sample has 2/3 of the peaks specific to scan-centric characterization and not matched to either of the other methods.
-Notably, for both samples, the scan-centric characterization produces similar numbers of peaks, even though the upper mass limit in 1ecf is 1000 m/z compared to 1600 m/z for the 97lipid sample.
-Whereas the number of peaks from MSnbase and Xcalibur are three-fold and 40-fold higher in the 1ecf sample compared to the 97lipid sample.
-
-`r figure_count$label_text("assigned_mz_diff")` shows the distribution of differences between the observed and expected m/z of assigned peaks in the 1ecf and 97lipid spectra.
+Notably, for both samples, the scan-centric characterization produces similar numbers of peaks, even though the upper mass limit in 1ecf is 1000 *m/z* compared to 1600 *m/z* for the 97lipid sample, whereas the number of peaks from MSnbase and Xcalibur are three-fold and 40-fold higher in the 1ecf sample compared to the 97lipid sample.
+
+`r figure_count$label_text("assigned_mz_diff")` shows the distribution of differences between the observed and expected *m/z* of assigned peaks in the 1ecf and 97lipid spectra.
 For the 1ecf-specific histograms, it is clear that these differences have a narrower unimodal distribution from the scan-centric peak characterization, especially in comparison to MSnbase.
-Also, the MSnbase distribution has far fewer peaks that match a scan-level assigned peak.
-Given the large number of MSnbase-characterized peaks present in the spectrum, many peaks may be outside the matching tolerance of 2ppm. 
+In addition, the MSnbase distribution has far fewer peaks that match a scan-level assigned peak.
+Given the large number of MSnbase-characterized peaks present in the spectrum, many peaks may be outside the matching tolerance of 2 ppm. 
 This strongly implies that the peak center error is far higher in the MSnbase-characterized peaks than what the histogram directly shows.
 For the 97lipid-specific histograms, again the difference distribution for the scan-centric peak characterization is narrower, but the improvement is not as pronounced, likely due to the distribution being bi-modal.
-Also, far fewer Xcalibur and MSnbase characterized peaks matched assigned scan-centric peaks.
+Moreover, far fewer Xcalibur and MSnbase characterized peaks matched assigned scan-centric peaks.
 Again, this strongly implies that their peak centers have far higher error.
@@
 set_order = c("scancentric", "msnbase", "xcalibur")
+
 u1 = grid::grid.grabExpr(draw(UpSet(um_peaks_1ecf$comb_matrix, 
@@
 **`r figure_count$label_text("assigned_mz_diff")`.**
-Histograms of the differences between observed and expected m/z values for assigned peaks in the 1ecf (**left**) and 97lipid (**right**) spectra, measured in parts-per-million (ppm) respectively.
+Histograms of the differences between observed and expected *m/z* values for assigned peaks in the 1ecf (**left**) and 97lipid (**right**) spectra, measured in parts-per-million (ppm) respectively.
 Inset plots are scaled to the maximum counts observed for the particular method.
@@
 `r figure_count$label_text("p_value_comparison")` compares the -1 * Log10(p-values) from `r nrow(compare_split$Corrected)` (corrected and Xcalibur) or `r nrow(compare_split$MSnbase)` (MSnbase) isotope resolved molecular formulas (IMFs) assigned by SMIRFE for the scan-centric peaks and then peaks matched and intensities extracted from the other methods.
-Although all of the p-vaues are somewhat (and statistically) different from those reported by the raw intensities, there are some interesting patterns of differences.
+Although all of the p-values are somewhat (and statistically) different from those reported by the raw intensities, there are some interesting patterns of differences.
 MSnbase generated p-values show the widest distribution of differences, as well as the smallest number of peaks that are present in 50% of both the cancer and non-cancer samples.
@@
     coord_equal() +
-    labs(x = "Log-P-Value Raw",
-         y = paste0("Log-P-Value ", id))
+    labs(x = expression(paste("Log-", italic("P"), "-Value Raw", sep = "")),
+         y = bquote(paste("Log-", italic("P"), "-Value ", .(id), sep = "")))
   out_plot
@@
 
+scientific_10 = function(x) {
+  gsub("e\\+*", " %*% 10^", scales::scientific_format()(x))
+}
+
+scientific_10_label = function(x) {
+  parse(text = gsub("e\\+*", " %*% 10^", scales::scientific_format()(x)))
+}
+
 long_pvalue = long_ttest %>%
-  dplyr::filter(measure %in% "p.adjust")
+  dplyr::filter(measure %in% "p.adjust") %>%
+  dplyr::mutate(text_value = dplyr::case_when(
+    value < 1 ~ scientific_10(value),
+    TRUE ~ "1"
+  ))
 
@@
   geom_point(data = long_measure, aes(x = compare_id, y = value), color = "green") +
-  geom_text(data = long_pvalue, aes(x = compare_id, y = 2, label = format(value, digits = 2))) +
-  labs(x = "", y = "Log-P-Value Difference")
+  geom_text(data = long_pvalue, aes(x = compare_id, y = 2, label = text_value), parse = TRUE) +
+  labs(x = "", y = expression(paste("Log-", italic("P"), "-Value Difference", sep = "")))
 
@@
 **D**: Sina plot of differences in the log-p-values generated by different methods compared to the raw scan-centric log-p-values.
-Also shown are the Bonferroni adjusted p-values from a t-test of the log-p-value differences for each method.
+The Bonferroni adjusted p-values from a t-test of the log-p-value differences for each method are also shown.
 Green points denote the high, mean, and low-confidence limits reported from the t-test.
@@
 For example, we can mark peaks with unusually high frequency standard deviations (FSD) and mitigate the presence of high-peak-density artifacts (see **Mitigation of High Peak Density Artifacts**).
-We also sometimes find that from initial peak region that has been marked as a single region, still results in two peaks being characterized in a single scan.
+We also sometimes find that from the initial peak region that has been marked as a single region, it still results in two peaks being characterized in a single scan.
 This results in that peak being removed from that scan entirely.
 
-The conversion of m/z to frequency also provides opportunities to evaluate each scan and whether it should be kept for further processing.
-Checks that have helped us in the development of the scan-centric peak characterization have included: (a) verifying that the ordering of data points remains identical in m/z and frequency space after conversion to frequency; and (b) examining the R^2^ fit of the predicted frequency points to the original frequency points across scans for outliers.
+The conversion of *m/z* to frequency also provides opportunities to evaluate each scan and whether it should be kept for further processing.
+Checks that have helped us in the development of the scan-centric peak characterization have included: (a) verifying that the ordering of data points remains identical in *m/z* and frequency space after conversion to frequency; and (b) examining the R^2^ fit of the predicted frequency points to the original frequency points across scans for outliers.
 A third quality control step is to examine the coefficients of the square-root term for each scan (see term **y** in Equation `r equation_count$label_text("mzfrequency")`) and look for any outliers, and remove them before continuing the sample processing.
@@
 ::: {custom-style="MDPI_3.1_text"}
-A final check we implemented in the conversion to frequency was verifying that the range of frequency point-point differences allowed for creating the m/z to frequency model are the same across scans.
+A final check we implemented in the conversion to frequency was verifying that the range of frequency point-point differences allowed for creating the *m/z* to frequency model are the same across scans.
 The initial set of scans we started with included **all** of the MS1 scans, including the precursor scans for the MS2 scans. 
-However, the samples were acquired as a mix of MS1 only scans for the first 7.5 minutes, and then a mix of MS1 precursor scans followed by MS2 scans for the second 7.5 minutes.
+However, the samples were acquired as a mix of MS1 only scans for the first 7.5 min, and then a mix of MS1 precursor scans followed by MS2 scans for the second 7.5 min.
 In 20 samples, the MS1 precursor scans have a different resolution than the MS1 main scans, as shown in `r figure_count$label_text("resolution_qcqa")`.
@@
 Oddly enough, the check for outlier scans discussed previously, does not flag any of the scans as outliers in this case.
-Ultimately, we were able to resolve this anomaly by only using those scans with an injection time before 7.5 minutes.
+Ultimately, we were able to resolve this anomaly by only using those scans with an injection time before 7.5 min.
 In the future, the package will be updated to check that the resolution is the same across the scans, and produce an informative error for the end user before quitting.
@@
 tar_load(resolution_qcqa)
+
+resolution_qcqa$sqrt = scientific_10(as.numeric(resolution_qcqa$sqrt))
 
@@
   coord_cartesian(ylim = c(NA, 2e7)) +
-  scale_y_continuous(expand = expansion(mult = c(0, .1))) +
+  scale_y_continuous(expand = expansion(mult = c(0, .1)), labels = scientific_10_label) +
+  scale_color_discrete(labels = scales::parse_format()) +
   scale_x_continuous(expand = expansion(mult = c(0, .1)))
@@
 ::: {custom-style="MDPI_3.1_text"}
-FT-MS direct-injection data provides spectral scans; however, unlike nuclear magnetic resonance, these spectral scans are not identical, due to several instrumentation issues.
+FT-MS direct-injection data provide spectral scans; however, unlike nuclear magnetic resonance, these spectral scans are not identical, due to several instrumentation issues.
 Therefore, simple scan summation or averaging to an aggregate spectrum followed by peak characterization does not provide optimal data.
-Also, this approach prevents the calculation of useful peak-specific statistics, especially variances, both in m/z and intensity.
+Additionally, this approach prevents the calculation of useful peak-specific statistics, especially variances, both in *m/z* and intensity.
 In the results presented, we demonstrate superior performance by scan-centric peak characterization in terms of: (a) improved QC/QA that removes low quality scans; (b) intensity independent noise reduction that pragmatically eliminates HPD sites; (c) improved peak height relative standard deviations; (d) improved peak intensity ratios that better match natural abundance probabilities; and (e) better separation between biological groups.
-These improvements are necessary for downstream derivation of molecular formula by SMIRFE using spectrum-derived tolerances, an efficient m/z search of a large isotope-resolved cache, and filtering by peak ratio matching to NAP.
+These improvements are necessary for downstream derivation of molecular formula by SMIRFE using spectrum-derived tolerances, an efficient *m/z* search of a large isotope-resolved cache, and filtering by peak ratio matching to NAP.
 However, scan-centric peak characterization requires a sophisticated pipeline of QC/QA and processing steps to derive a superior characterization with informative statistics and QC/QA metrics.
 Years of methods development, testing, and optimization have gone into the methods presented here.
-But once implemented, large datasets can be processed in a straight-forward manner that supports full computational reproducibility.
+Nevertheless, once implemented, large datasets can be processed in a straightforward manner that supports full computational reproducibility.
 We also kept all steps that provided a demonstrable improvement, no matter how small that improvement was, but the following steps provided the most improvement: (a) noise removal by non-zero point density, (b) scan normalization using median peak intensity differences, and (c) removal of outlier scans.
-However, both steps a and c require the m/z to frequency regression model. 
+However, both steps a and c require the *m/z* to frequency regression model. 
 In the end, it is the combination of steps that synergistically provide the presented superior peak characterization results. 
-We note that the non-zero point density based noise removal is particularly useful, as the peak intensities from scan to scan are not identical, which would possibly hamper any kind of intensity based noise removal.
+We note that the non-zero point density based noise removal is particularly useful, as the peak intensities from scan to scan are not identical, which would possibly hamper any kind of intensity-based noise removal.
 
@@
 Fortunately, this distribution has been previously studied [@sampfordTruncatedNegativeBinomial1955a] and we plan to explore this possibility for a future improvement to our methods.
-However, correction may not be straight-forward and it is unclear if it is better to apply the correction to the raw or log-transformed peak height data.
+However, correction may not be straightforward and it is unclear if it is better to apply the correction to the raw or log-transformed peak height data.
 Either way, a dispersive variance component is present and must be accounted for.
@@
 This likely indicates that the loss of peak detection involves other factors that are increased at low intensity.
-We theorize that any factor that increases signal decoherence like ion-cloud repulsion would increase the peak flooring performed by the instrument.
+We theorize that any factor that increases signal decoherence, such as ion-cloud repulsion would increase the peak flooring performed by the instrument.
 Therefore, access to the scan level peak intensity information as provided by the methods presented here facilitates the inference of sources and types of error present in the measurements, which we believe have been previously under-described and is required for future data processing improvements.
@@
 The method for generating the amino-acid samples was adapted from a previously published method for performing ECF amino acid derivatization [@yangChloroformateDerivatizationTracing2017]. 
-Two replicate samples were prepared and spectra were obtained for both samples using a Tribrid Fusion Orbitrap at 500k resolution and a mass range of 150 to 1000m/z.
+Two replicate samples were prepared and spectra were obtained for both samples using a Tribrid Fusion Orbitrap at 500k resolution and a mass range of 150 to 1000*m/z*.
 
@@
 For each of the two ECF and two lipid samples, the raw data file was converted to profile mzML format.
-Only the MS1 scans were used, and the scan-scan time difference had to be ≥ 4 seconds, and for the lipid samples only scans acquired before 450 seconds (7.5 minutes) were kept.
-The scan-centric data was then processed in the following eight ways:
+Only the MS1 scans were used, and the scan-scan time difference had to be ≥ 4 s, and for the lipid samples only scans acquired before 450 s (7.5 min) were kept.
+The scan-centric data were then processed in the following eight ways:
 :::
@@
 ::: {custom-style="MDPI_3.8_bullet"}
-* No noise removal, no normalization (noperc_nonorm)
+* No noise removal, no normalization (noperc_nonorm);
 :::
 ::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, no normalization (perc99_nonorm)
+* Noise removal, no normalization (perc99_nonorm);
 :::
 ::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, single-pass normalization with all peaks (singlenorm)
+* Noise removal, single-pass normalization with all peaks (singlenorm);
 :::
 ::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, single-pass normalization with high ratio peaks (singlenorm_int)
+* Noise removal, single-pass normalization with high ratio peaks (singlenorm_int);
 :::
 ::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, two-pass normalization (doublenorm)
+* Noise removal, two-pass normalization (doublenorm);
 :::
 ::: {custom-style="MDPI_3.8_bullet"}
-* Noise removal, two-pass normalization (filtersd)
+* Noise removal, two-pass normalization (filtersd);
 :::
@@
 ::: {custom-style="MDPI_3.8_bullet"}
-* Scans merged and then centroids generated by MSnbase (using `combineSpectra` and `pickPeaks`)
+* Scans merged and then centroids generated by MSnbase (using `combineSpectra` and `pickPeaks`);
 :::
@@
 ::: {custom-style="MDPI_3.8_bullet"}
-* Scans merged and peak-list exported by Xcalibur
+* Scans merged and peak-list exported by Xcalibur.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-To associate the Xcalibur and MSnbase peaks with the scan-centric peaks, a 4 ppm (2ppm low and high) window is calculated for the scan-centric centroid, and if any peaks are found within the window, the one with the smallest m/z difference is kept as the matching peak to the scan-centric one under consideration.
+To associate the Xcalibur and MSnbase peaks with the scan-centric peaks, a 4 ppm (2 ppm low and high) window is calculated for the scan-centric centroid, and if any peaks are found within the window, the one with the smallest *m/z* difference is kept as the matching peak to the scan-centric one under consideration.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-The input data consists of profile mode m/z spectra from multiple scans encoded as m/z and intensity values for each scan.
+The input data consist of profile mode *m/z* spectra from multiple scans encoded as *m/z* and intensity values for each scan.
 No information about the original observed frequency values is available in either the `raw` files or the `mzML` files.
-However, proxy frequency values can be generated by averaging the m/z of adjacent points and dividing them by the m/z difference.
+However, proxy frequency values can be generated by averaging the m/z of adjacent points and dividing them by the *m/z* difference.
 Ideally, the difference between subsequent points in this proxy frequency space is constant, in practice there is a range of differences in frequency space.
-The *actual, constant* difference can be obtained by examining the median of the calculated frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to m/z) to be within 2% of the mode value.
-These *useful* points can be used to construct a linear model relating m/z to frequency for individual scans based on the formula:
+The *actual, constant* difference can be obtained by examining the median of the calculated frequency differences, and then constraining *useful* points (those that can be used for generating a model of frequency to *m/z*) to be within 2% of the mode value.
+These *useful* points can be used to construct a linear model relating *m/z* to frequency for individual scans based on the formula:
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-From the known physical properties of the Orbitrap, only the square root term should be necessary [@ledfordSpaceChargeEffects1984].
-Practically, we have found the combination of no root, square and cube-roots to provide a better fit when processing data from the Thermo-Fisher Fusion instrument, likely due to issues with slight imperfections in the orbitrap geometry, contributions from space charge effects and magnetronic motion, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
+From the known physical properties of the Orbitrap, only the square-root term should be necessary [@ledfordSpaceChargeEffects1984].
+Practically, we found the combination of no root, square and cube-roots to provide a better fit when processing data from the Thermo-Fisher Fusion instrument, likely due to issues with slight imperfections in the orbitrap geometry, contributions from space charge effects and magnetronic motion, control of the magnetic fields, and the Fourier-like transform conversion used by the spectrometer.
 We do note that when working with Bruker SolariX ICR data, the equation is slightly different, and does not require the cubic square root term (unpublished results, see Supplemental Materials).
-A frequency model is generated for each scan, and then a single model using the scan with the square-root term closest to the median of the square-root terms from all scans.
-We observed that this single model better preserved the relative ordering of the peaks in both m/z and frequency-space compared to the scan specific models (see Results).
-
-To convert m/z back into frequency, we can use a similar model without the roots, as well as an extra simple linear term that does not have an equivalent in the above frequency model.
+A frequency model was generated for each scan, followed by a single model using the scan with the square-root term closest to the median of the square-root terms from all scans.
+We observed that this single model better preserved the relative ordering of the peaks in both *m/z* and frequency-space compared to the scan specific models (see Results).
+
+To convert *m/z* back into frequency, we can use a similar model without the roots, as well as an extra simple linear term that does not have an equivalent in the above frequency model.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-Two types of frequency intervals are used, sliding and tiled windows.
+Two types of frequency intervals were used; sliding and tiled windows.
 In this work, the sliding windows are the equivalent of 10 frequency points wide with a stride of one point.
@@
 ::: {custom-style="MDPI_3.1_text"}
-To find intervals that contain points that represent actual signal and not just random noise, the number of non-zero intensity points in each sliding window are counted.
-Subsequently, we break these counts into fixed width tiles (default width of 1000 in frequency space) and calculate the cumulative 99^th^ percentile of non-zero points for each tile.
-The rounded up median value x 1.5 of these 99^th^ percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions containing potential signal.
+To find intervals that contain points that represent actual signal and not just random noise, the number of non-zero intensity points in each sliding window were counted.
+Subsequently, we broke these counts into fixed width tiles (default width of 1000 in frequency space) and calculated the cumulative 99^th^ percentile of non-zero points for each tile.
+The rounded up median value x 1.5 of these 99^th^ percentile values from the fixed width tiles were used as the cutoff value to determine which of the initial sliding regions should be kept as regions containing potential signal.
 Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are kept and overlapping sliding windows are merged to create the initial peak regions.
@@
 On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package [@borchersPracmaPracticalNumerical2021].
-These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either m/z or frequency), and the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.
+These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either *m/z* or frequency), and the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.
 :::
@@
 ::: {custom-style="MDPI_3.1_text"}
-Scans are normalized to a single *reference* scan based on the log-intensity differences of a subset of peaks present in at least the same number of scans as the 95th percentile of scan counts for the peaks.
+Scans are normalized to a single *reference* scan based on the log-intensity differences in a subset of peaks present in at least the same number of scans as the 95th percentile of scan counts for the peaks.
 In addition, only those peaks with an intensity greater than 0.7 times the highest intensity peak in the scan are used.
-Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each scan to all other scans.
+Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference in each scan to all other scans.
 The scan with the lowest summed overall distance is chosen as the *reference* scan ($scan_{ref}$), and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the $scan_{ref}$.
-This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan order is checked, and those peaks with correlation of greater than 0.5 with scan order are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.
+This normalization was carried out twice, once using all possible peaks, after which the correlation of peak intensity with scan order was checked, and those peaks with correlation of greater than 0.5 with scan order were removed, and the normalization factors were calculated again, and then applied to both the centroided peak height and the raw point intensities.
 Peaks correlated to scan order represent an artifact that we speculate results from a gradient in the sample well and are marked in the final output.
@@
 
-In addition to these values, the frequency point-to-point median difference is calculated across all of the raw data points for those points that could be used for modeling frequency to m/z, and this difference of a single point from the peak center is calculated in frequency space, and converted to m/z space to provide an "offset" value that is potentially useful to define the search space around the peak for any assignment algorithm.
+In addition to these values, the frequency point-to-point median difference was calculated across all of the raw data points for those points that could be used for modeling frequency to *m/z*, and this difference of a single point from the peak center is calculated in frequency space, and converted to *m/z* space to provide an "offset" value that is potentially useful to define the search space around the peak for any assignment algorithm.
 :::
@@
 To correct these, either a correction based on a model of the truncated normal distribution can be used on log-transformed data, or the differences can be simulated by sampling from data that is present in most of the scans.
-To simulate the effect of peaks missing from some scans on the standard deviations, the peaks present in all scans are used.
-For each peak, a sample of the heights across scans are taken (ranging from 5% to 95% of scans), and a new standard deviation calculated for that fraction, and a ratio of the fractional standard deviation to the "true" standard deviation calculated.
+To simulate the effect of peaks missing from some scans on the standard deviations, the peaks present in all scans were used.
+For each peak, a sample of the heights across scans was obtained (ranging from 5% to 95% of scans), and a new standard deviation was calculated for that fraction, and a ratio of the fractional standard deviation to the "true" standard deviation was calculated.
 The ratio standard deviation across peaks can then be fitted to a cubic model of the fraction used, and a correction factor predicted for those peaks that are present in fewer scans.
@@
 Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and peaks with FSDs greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are marked.
-The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
-For HPD detection, the peaks in excel output from Xcalibur were used after converting the m/z peak centers into frequency space.
+The HPD detection algorithm from Mitchell et al. [@mitchellNewMethodsIdentify2018] was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
+For HPD detection, the peaks in excel output from Xcalibur were used after converting the *m/z* peak centers into frequency space.
 Sliding windows that are 1000 frequency points wide with a stride of 100 points were used for the density calculations.
@@
 
-Modes of the RSD are calculated starting from a density approximation of the distribution using default settings in the density function of the R stats package.
-Peaks from the density approximation are determined with the find_peaks function in the pracma package.
-Up to two modes are reported from low to high RSD along the distribution: i) the most intense peak; ii) the next most intense peak if it is ≥ 0.2X the most intense peak.
+Modes of the RSD were calculated starting from a density approximation of the distribution using default settings in the density function of the R stats package.
+Peaks from the density approximation were determined with the find_peaks function in the pracma package.
+Up to two modes are reported from low to high RSD along the distribution: (i) the most intense peak; (ii) the next most intense peak if it is ≥ 0.2X the most intense peak.
 :::
@@
 Our previously described SMIRFE algorithm [@mitchellSmallMoleculeIsotope2019] was used to assign molecular formulas to scan-centric characterized peaks in an untargeted manner. 
-For the lipid samples, an initial EMF database was generated using an m/z limit of 1605 m/z, and maximum numbers for each element were set to C: 130, N: 7, O: 28, P: 3, H: 230. 
+For the lipid samples, an initial EMF database was generated using an *m/z* limit of 1605 *m/z*, and maximum numbers for each element were set to C: 130, N: 7, O: 28, P: 3, H: 230. 
 Assigned formulas were allowed to have K^+^, Na^+^, H^+^, and NH4^+^ adducts (only positive mode samples were assigned). 
 Assigned molecular formulas were then classified into one or more lipid categories using our lipid classifier tool [@mitchellDerivingLipidClassification2020].
-For the ECF derivatized amino-acid samples, the EMF database was generated using an m/z limit of 1005 m/z, and maximum numbers for each element were set to C: 100, N: 7, O: 40, H: 230, P: 3, S: 3.
+For the ECF derivatized amino-acid samples, the EMF database was generated using an *m/z* limit of 1005 *m/z*, and maximum numbers for each element were set to C: 100, N: 7, O: 40, H: 230, P: 3, S: 3.
 ^15^N labeling was included in the database generation.
@@
 Each IMF represents an isotopologue of a given elemental molecular formula (EMF) (e.g., 13C112C51H1216O6 is an IMF representing the m+13C1 isotopologue of EMF C6H12O6).
-Consistently assigned (i.e., corresponded peaks) are identified using an in-house method we have named EMF voting.
+Consistently assigned formulas (i.e., corresponded peaks) were identified using an in-house method we have named EMF voting.
 
 Elemental molecular formula (EMF) voting was used to match peaks across samples and determine the most likely assignments from SMIRFE. 
-First, for each sample, the assignments are extracted, filtered, and scored.
-Any assignments that contain only peaks that were marked as being questionable (high FSD or correlated with scan number) removed.
+First, for each sample, the assignments were extracted, filtered, and scored.
+Any assignments that contained only peaks that were marked as being questionable (high FSD or correlated with scan number) were removed.
 Scores are calculated as 1 – E-value, and for the lipids, EMFs that were classified as lipids had their score multiplied by 2.
 
-For each sample, peaks were grouped to a sample specific EMF by determining the list of shared EMFs across a group of peaks (grouped_EMF). 
-Scores for each formula in the grouped_EMF in the sample were taken as the best score for that formula from available scores in the group. 
+For each sample, peaks were grouped to a sample-specific EMF by determining the list of shared EMFs across a group of peaks (grouped_EMF). 
+Scores for each formula in the grouped_EMF in the sample were obtained as the best score for that formula from available scores in the group. 
 After all sample grouped_EMFs are generated, additional scores for a formula are considered in actual voting by looking for the same unadducted base EMF with different adducts and adding these scores into the final total score.
@@
 The pseudo_EMFs are collections of grouped_EMFs across samples. 
-They are generated across samples by iteratively merging grouped_EMFs with shared formulas, creating a new list of formulas in an EMF, and merging any pseudo_EMFs that have shared formulas again.
+They were generated across samples by iteratively merging grouped_EMFs with shared formulas, creating a new list of formulas in an EMF, and merging any pseudo_EMFs that have shared formulas again.
 
@@
 
-After voting, all of the peaks are checked to make sure that the peak locations are within previously defined m/z specific cutoffs. 
-Any peak outside of its specific cutoff is removed.
+After voting, all of the peaks were checked to make sure that the peak locations are within previously defined *m/z* specific cutoffs. 
+Any peak outside of its specific cutoff were removed.
 
@@
 
-Peak locations from our custom data processing pipeline are reported both in m/z and frequency, which are derived from the m/z values. 
+Peak locations from our custom data processing pipeline are reported both in *m/z* and frequency, which are derived from the *m/z* values. 
 The frequency values are more reliable (i.e., more consistent with less variance), but differ between instruments. 
-Therefore, for each set of samples from a particular instrument, we use high confidence assignments (e-value <= 0.1, m/z <= 600) to derive a frequency cutoff using the mode of the distribution of frequency standard deviations across both groups of samples.
+Therefore, for each set of samples from a particular instrument, we use high confidence assignments (e-value <= 0.1, *m/z* <= 600) to derive a frequency cutoff using the mode of the distribution of frequency standard deviations across both groups of samples.
 This frequency cutoff is used to do EMF voting within an instrument. 
-The m/z standard deviations of the voted peaks are then fit to m/z using a generalized additive model, and the standard deviation of the predicted m/z values are multiplied by 2 to derive an m/z cutoff so that voting can be performed across the instruments.
+The *m/z* standard deviations of the voted peaks are then fit to *m/z* using a generalized additive model, and the standard deviation of the predicted *m/z* values are multiplied by 2 to derive an *m/z* cutoff so that voting can be performed across the instruments.
 
@@
 ::: {custom-style="MDPI_3.1_text"}
-SMIRFE assignments include the NAP for each of the isotopologue molecular formulas (IMFs) in the particular labelled and adducted EMF.
-For each peak pair in the peak group for a particular EMF, we calculate the log-ratio for the NAPs and the log-ratio for the intensities, and their absolute differences.
+SMIRFE assignments include the NAP for each of the isotopologue molecular formulas (IMFs) in the particular labeled and adducted EMF.
+For each peak pair in the peak group for a particular EMF, we calculated the log-ratio for the NAPs and the log-ratio for the intensities, and their absolute differences.
 This calculation is done for the final characterized intensity, as well as the intensities at the scan level, and for both the *raw* intensity and *corrected* intensity, and for matched Xcalibur peak intensities, and matched MSnbase peak intensities if there were two or more matching peaks in the labeled, adducted EMF.
@@
 
-Each samples intensities were normalized by dividing by the median intensities from that sample and intensity method.
+Each sample intensity was normalized by dividing by the median intensities from that sample and intensity method.
 Only those peaks present in 50% of both the non-cancer and cancer samples were kept for differential analysis.
@@
 Lipid classifications were generated by LipidClassifier v 1.0 [@mitchellDerivingLipidClassification2020].
-All other calculations were performed in R v 4.1.0 [@rcoreteam_rlanguage_2020].
+All other calculations were performed in R v 4.1.0 [@rcoreteamLanguageEnvironmentStatistical2021a].
 The targets package v 0.10.0 was used to control processing and aggregation of results [@landauTargetsPackageDynamic2021], and renv v 0.15.3 [@usheyRenvProjectEnvironments2022] to create a reproducible R package environment.
@@
 
-**Data Availability Statement**: The code used in the reporting of results is available on GitHub at https://github.com/MoseleyBioinformaticsLab/manuscript.peakCharacterization, and archived on Zenodo [@flightMoseleyBioinformaticsLabManuscriptPeakCharacterization2022].
+**Data Availability Statement**: The code used in the reporting of results is available on GitHub at https://github.com/MoseleyBioinformaticsLab/manuscript.peakCharacterization (accessed 2022-06-01), and archived on Zenodo [@flightMoseleyBioinformaticsLabManuscriptPeakCharacterization2022].
 The scan-centric peak-characterization is available as an R package at https://github.com/MoseleyBioinformaticsLab/FTMS.peakCharacterization, and the specific version used in this work is archived on Zenodo [@flightMoseleyBioinformaticsLabFTMSPeakCharacterization2022].

### 2022-06-02 doc/peakcharacterization_supplemental.Rmd
last set of changes

@@
   flextable() %>%
-  colformat_double(digits = 0) %>%
+  colformat_double(digits = 0, big.mark = ",") %>%
   autofit()
@@
 ```{r peak_variance, dn_id = figure_count, fig.width = 8, fig.height=9}
+scientific_10 = function(x) {
+  gsub("e\\+*", " %*% 10^", scales::scientific_format()(x))
+}
+
+scientific_10_label = function(x) {
+  parse(text = gsub("e\\+*", " %*% 10^", scales::scientific_format()(x)))
+}
+
 filtersd_only = rsd_combine %>%
@@
     geom_point(alpha = 0.5) +
-    labs(subtitle = in_sample$sample[1])
+    labs(subtitle = in_sample$sample[1]) +
+    scale_y_continuous(labels = scientific_10_label)
   mean_rsd = in_sample %>%
@@
   geom_point(data = long_measure, aes(x = compare_id, y = value), color = "green") +
-  geom_text(data = long_pvalue, aes(x = compare_id, y = -1, label = format(value, digits = 2))) +
-  labs(x = "", y = "Log-P-Value Difference")
+  geom_text(data = long_pvalue, aes(x = compare_id, y = -1, label = scientific_10(value)), parse = TRUE) +
+  labs(x = "", y = expression(paste("Log-", italic("P"), "-Value Difference", sep = "")))
 
@@
   geom_point(alpha = 0.5) +
-  labs(x = "Log-P-Value Raw",
-         y = "Log-P-Value Corrected")
-compare_pair | compare_sina + plot_annotation(tag_levels = "A")
+  labs(x = expression(paste("Log-", italic("P"), "-Value Raw", sep = "")),
+         y = expression(paste("Log-", italic("P"), "-Value Corrected", sep = "")))
+(compare_pair | compare_sina) + plot_annotation(tag_levels = "A")
 ```

### 2022-06-02 R/best_rsd.R

@@ -23,8 +23,8 @@ find_best_rsd = function(rsd_values){
   rsd_ft = rsd_values2 %>%
     flextable() %>%
     hline(i = filtersd_rsd[2:(length(filtersd_rsd))] - 1) %>%
-    colformat_double(digits = 2) %>%
-    bold(i = c(other_loc, filtersd_rsd), j = seq(3, 6), part = "body")
+    colformat_double(digits = 2) #%>%
+    #bold(i = c(other_loc, filtersd_rsd), j = seq(3, 6), part = "body")
   rsd_ft
 
 }

### 2022-06-02 R/compare_noise_cutoff.R

@@ -141,6 +141,7 @@ plot_example_sliding_counts = function(peak_data){
     geom_point(alpha = 0.5) +
     coord_cartesian(xlim = xlim, ylim = c(2, NA)) +
     labs(y = "Log10(Intensity)", x = "Frequency") +
+    scale_x_continuous(labels = scales::comma) +
     theme(axis.title.y = element_text(size = y_title_size))
 
 
@@ -195,13 +196,13 @@ plot_sliding_window_density = function(peak_data){
   single_plot = ggplot(single_chunk, aes(x = use_counts)) +
     geom_histogram(bins = 100) +
     geom_vline(xintercept = single_99, color = "blue") +
-    scale_y_log10(expand = c(0, 0), limits = c(1, NA)) +
+    scale_y_log10(expand = c(0, 0), limits = c(1, NA), labels = scales::comma) +
     labs(x = "Number of Non-Zero Points",
          y = "Log10(Counts)")
   all_plot = ggplot(all_99, aes(x = perc_99)) +
     geom_histogram(bins = 100) +
     geom_vline(xintercept = all_cut, color = "red") +
-    scale_y_continuous(expand = c(0, 0)) +
+    scale_y_continuous(expand = c(0, 0), labels = scales::comma) +
     labs(y = "# Tiled Regions",
          x = "99th Percentiles of Tiled Regions")
   all_plot
@@ -215,6 +216,7 @@ create_noise_plot = function(noise_combine){
     geom_point() +
     geom_line() +
     theme(legend.position = c(0.1, 0.5)) +
+    scale_y_continuous(labels = scales::comma) +
     labs(y = "# of Initial Peak Regions", x = "Percentile Cutoff")
 
   lipid_plot = noise_combine %>%
@@ -223,6 +225,7 @@ create_noise_plot = function(noise_combine){
     geom_point() +
     geom_line() +
     theme(legend.position = c(0.1, 0.5)) +
+    scale_y_continuous(labels = scales::comma) +
     labs(y = "# of Initial Peak Regions", x = "Percentile Cutoff")
 
   (ecf_plot | lipid_plot)

### 2022-06-02 R/examine_hpd.R

@@ -212,6 +212,8 @@ plot_hpds = function(hpd_res){
       geom_segment() +
       coord_cartesian(xlim = plot_range,
                       ylim = c(NA, mode_height)) +
+      scale_x_continuous(labels = scales::comma) +
+      scale_y_continuous(labels = scales::comma) +
       labs(y = "Intensity", subtitle = in_type$source[1])
     base_plot
   })

### 2022-06-02 R/get_assignments.R

@@ -531,7 +531,7 @@ motivating_plot = function(nap_height_1ecf, aa_id = "Threonine", formula_id = "C
              size = font_size) +
     annotate("text", x = 243.75, y = 4.2, label = peak_data$complete_IMF[4],
              size = font_size) +
-    labs(x = "m/z", y = "Log10(Intensity)") +
+    labs(x = expression(~italic("m/z")), y = "Log10(Intensity)") +
     scale_y_continuous(expand = expansion(mult = c(0, 0.05)))
   nap_int_plot = peak_data %>%
     ggplot(aes(x = log10intNAP, y = log10intensity)) +

### 2022-06-02 R/graph_generation.R

@@ -88,24 +88,26 @@ plot_frequency_conversion = function(char_obj){
   example_data[1, "pair_mz"] = NA
   mz_point_plot = ggplot(example_data, aes(x = mz, y = log10(intensity+1))) + geom_point() +
     geom_segment(aes(x = mz_group_xstart, xend = mz_group_xend, y = level1_intensity, yend = level1_intensity), color = "red") +
-    geom_point(aes(x = pair_mz, y = level1_intensity), color = "red") + labs(y = "Log10(Intensity)", x = "M/Z")
+    geom_point(aes(x = pair_mz, y = level1_intensity), color = "red") + labs(y = "Log10(Intensity)", x = expression(~italic("m/z")))
 
   frequency_point_plot = ggplot(example_data, aes(x = pair_frequency, y = log10(intensity+1))) + geom_point() +
+    scale_x_continuous(labels = scales::comma) +
     labs(y = "Log10(Intensity)", x = "Frequency") + geom_segment(aes(x = frequency_group_xstart, xend = frequency_group_xend, y = level1_intensity, yend = level1_intensity), color = "red")
 
   point_difference_plot = ggplot(example_data, aes(x = pair_frequency, y = pair_frequency_offset)) + geom_point() +
-    labs(x = "Frequency", y = "Frequency Differences")
+    labs(x = "Frequency", y = "Frequency Differences") +
+    scale_x_continuous(labels = scales::comma)
 
   all_frequency_difference_hist = ggplot(raw_points, aes(x = log10(abs(mean_freq_diff)))) +
     geom_histogram(bins = 100) +
     geom_vline(xintercept = log10(0.5), color = "red") +
-    scale_y_log10(expand = expansion(mult = c(0, .01))) +
+    scale_y_log10(expand = expansion(mult = c(0, .01)), labels = scales::comma) +
     labs(x = "Log10(Frequency Differences)",
          y = "Number of Points")
 
   y_title_size = 12
   all_frequency_difference_plot = ggplot(raw_points, aes(x = mz, y = log10(abs(mean_freq_diff)))) + geom_point(alpha = 0.5) +
-    geom_point(data = dplyr::filter(raw_points, org_convertable), color = "red") + labs(x = "M/Z", y = "Log10(Frequency Differences)") +
[…]
